### 2022-11-29 Draft1.Rmd
Draft 1

---
title: "mathml: Translate R expressions to MathML"
date: "2022-11-09"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan-Rochon
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.britannica.com/animal/quokka
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.britannica.com/animal/bilby
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(ggplot2)
library(palmerpenguins)
```

# Introduction

The R extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
However, so far, R expressions such as `pbinom(k, N, p)` are printed as-is; 
prettier mathematical expressions such as $P_\mathrm{Bi}(X \le k; N, p)$ require
Latex commands like `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for
minimalistic examples, these commands are tedious to type in and hard to read in
the original code. Since recently [@R, version 4.2], manual pages include
support for mathematical expressions [@Sarkar2022], which already is a big
improvement. Rules for the translation of R's built-in language elements to
their mathematical representation are lacking.

In this R package, such rules are defined for the _automatic_ translation of R
expressions mathematical output. The translation is done by an embedded
Prolog interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively, for typesetting of R Markdown documents in
different formats or ShinyApp webpages. User-defined hooks enable to extend the
set of rules and/or replace specific R elements by special mathematical signs.

The paper is organized as follow. Section 2 presents the background of the mathml package, namely Prolog and the R package `rolog`. Section 3 illustrates the features of the mathml package, providing some examples. Section 4 embeds the Discussion and the presentation of a Case Study. Section 5 is devoted to Conclusions and Further Research. The reader may also find some additional characteristics of the package and its implementation in the Appendix.

# Background

The automatic translation of R expressions to mathematical output is achieved through a Prolog interpreter. Prolog is a classical logic programming language with many applications in expert systems, computer linguistics and traditional, that is, symbolic artificial intelligence. The main strength of Prolog is its concise representation of facts and rules for the representation of knowledge and grammar, as well as its efficient built-in search engine for closed world domains. R, as it is well-known, is a statistical programming language for data analysis and statistical modeling which is widely used in academia and industry. Besides the core library, a lot of packages have been developed for all kinds of statistical problems, including statistics-based artificial intelligence tools such as neural networks for machine learning and deep learning. Whereas Prolog is weak in statistical computation, but strong in symbolic manipulation, the converse may be said for the R language. The `rolog` package then bridges this gap by providing an interface to a SWI-Prolog distribution in R. The communication between the two systems is mainly in the form of queries from R to Prolog, but two predicates allow Prolog to ring back and evaluate terms in R. 
Here is reported a first trivial example. At the Prolog end, there is a handler for `pbinom/3` that translates the term into a pretty MathML syntax like P_bi(X <= k; N, pi).

```{r, echo=TRUE}
library(rolog)
library(mathml)
term = quote(pbinom(k, N, p))

# Pretty print
mathml(term)
```

_I encounter a problem, when I explicit the 'asis' in the results='', the execution of the Rmd is halted. Yet, only when I use mathml()._

Beside the graphical rendition of the term, an important feature is that the package allows also for its calculation, for example:

```{r, results = 'asis', echo = TRUE}
k = 10
N = 22
p = 0.4
eval(term)
```

Here we present a second less trivial example where Prolog needs to find out the name of
the integration variable for `sin`. For that purpose, rolog provides a predicate `r_eval/2` that calls R from Prolog. Here, the predicate is used for the  R\ function `formalArgs(args(sin))`, which returns the name of the function argument of `sin`, that is, `x`.

```{r, echo = TRUE}
term = quote(integrate(sin, 0L, 2L*pi))
mathml(term)
eval(term)
```

Note that the Prolog end, the handler for `integrate/3` is rather rigid; it
accepts only these three arguments in that particular order, and without names,
that is, `integrate(sin, lower=0L, upper=2L * pi)` would not print the desired
result.

Therefore, `pl/mathml.pl` includes two handlers that accept terms with named
arguments, `integrate(f=Fn, lower=Lower, upper=Upper)`, as well as terms of the
form `$(integrate(Fn, Lower, Upper), value)` that are needed for the evaluation
below.

```{r, echo = TRUE}
library(rolog)
# Apply match.call to all components of a term
canonical <- function(term)
{
  if(is.call(term))
  {
    f <- match.fun(term[[1]])
    if(!is.primitive(f))
      term <- match.call(f, term)

    # Recurse into arguments
    term[-1] <- lapply(term[-1], canonical)
  }
  return(term)
}
# A custom function
g <- function(u)
{
  sin(u)
}
# # Mixture of (partially) named and positional arguments in unusual order
term <- quote(2L * integrate(low=-Inf, up=Inf, g)$value)
# mathml(canonical(term))
# # It is a bit of a mystery that R knows the result of this integral.
# eval(term)
```

The extra R function `canonical()` applies `match.call()` to non-primitive R
calls, basically cleaning up the arguments and bringing them into the correct
order.

<!-- Some packages on interactive graphics include \CRANpkg{plotly} [@plotly] that interfaces with Javascript for web-based interactive graphics, \CRANpkg{crosstalk} [@crosstalk] that specializes cross-linking elements across individual graphics.  The recent R Journal paper \CRANpkg{tsibbletalk} [@RJ-2021-050] provides a good example of including interactive graphics into an article for the journal. It has both a set of linked plots, and also an animated gif example, illustrating linking between time series plots and feature summaries. -->

# Customizing tooltip design with \pkg{ToOoOlTiPs}

\pkg{ToOoOlTiPs} is a packages for customizing tooltips in interactive graphics, it features these possibilities.

# A gallery of tooltips examples

The \CRANpkg{palmerpenguins} data [@palmerpenguins]  features three penguin species which has a lovely illustration by Alison Horst in Figure \@ref(fig:penguins-alison).

```{r penguins-alison, out.width = "100%", out.height = "30%", fig.cap = "Artwork by \\@allison\\_horst", fig.alt="A picture of three different penguins with their species: Chinstrap, Gentoo, and Adelie. "}
knitr::include_graphics("penguins.png")
```

Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:penguins-tab-interactive)', '\\@ref(tab:penguins-tab-static)'))` prints at the first few rows of the `penguins` data:

```{r penguins-tab-interactive, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(head(penguins), format = "html", caption = "A basic table")
```

```{r penguins-tab-static, eval = knitr::is_latex_output()}
knitr::kable(head(penguins), format = "latex", caption = "A basic table") %>% 
  kableExtra::kable_styling(font_size = 7)
```

Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:penguins-plotly)', '\\@ref(fig:penguins-ggplot)'))` shows an `r ifelse(knitr::is_html_output(), 'interactive', '')` plot of the penguins data, made using the `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\CRANpkg{plotly}', '\\CRANpkg{ggplot2}'))` package.

```{r penguins-plotly, echo = TRUE, fig.height = 5, fig.cap="A basic interactive plot made with the plotly package on palmer penguin data. Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis. When hovering on a point, a tooltip will show the exact value of the bill depth and length for that point, along with the species name.", include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.alt = "A scatterplot of bill length against bill depth, both measured in millimetre. The three species are shown in different colours and loosely forms three clusters. Adelie has small bill length and large bill depth, Gentoo has small bill depth but large bill length, and Chinstrap has relatively large bill depth and bill length."}
p <- penguins %>% 
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
             color = species)) + 
  geom_point()
ggplotly(p)
```

```{r penguins-ggplot, echo = TRUE, fig.height = 5, fig.cap="A basic non-interactive plot made with the ggplot2 package on palmer penguin data. Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis. Visit the online article to access the interactive version made with the plotly package.", include=knitr::is_latex_output(), eval=knitr::is_latex_output()}
penguins %>% 
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
             color = species)) + 
  geom_point()
```

# Summary

We have displayed various tooltips that are available in the package \pkg{ToOoOlTiPs}.

### 2022-11-29 Draft1.Rmd
added an abstract

@@
 ---
-title: "mathml: Translate R expressions to MathML"
+title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
 date: "2022-11-09"
 abstract: >
-  An abstract of less than 150 words.
+  This R package translates R objects to suitable entities in MathML or LaTeX,
+  thereby enabling pretty mathematical representation of R functions or
+  algorithms in data analysis, scientific reports and interactive web content.
+  R code text descriptions and mathematical content already coexist,
+  side-by-side, in the RMarkdown document rendering language. The present
+  package allows to use the same R\ objects for both data analysis and
+  typesetting in documents or web content. By tightening the link between the
+  statistical analysis and its verbal or symbolic description, this new feature
+  is a futher step towards reprocible science. The package can be extended by
+  user-defined hooks that allow to replace variable names and new R functions by
+  a suitable symbolic representation. We provide two working examples that
+  illustrate the use of the new package in a scientific article and in an
+  interactive web page.
 draft: true
@@
   # see ?rjournal_article for more information
-  - name: Matthias Gondan-Rochon
+  - name: Matthias Gondan
     affiliation: University of Innsbruck
@@
 "}
-knitr::include_graphics("penguins.png")
+# knitr::include_graphics("penguins.png")
 ```

### 2022-11-30 Draft1.Rmd
update 30/11

@@
   statistical analysis and its verbal or symbolic description, this new feature
-  is a futher step towards reprocible science. The package can be extended by
+  is a futher step towards reproducible science. The package can be extended by
   user-defined hooks that allow to replace variable names and new R functions by
@@
 library(palmerpenguins)
+
+catmath <- function(term)
+{
+  s <- substitute(term)
+  c <- canonical(s)
+  mathml(c)
+}
+
+catmathjax <- function(term)
+{
+  s <- substitute(term)
+  c <- canonical(s)
+  mathjax(c)
+}
+
 ```
@@
 The paper is organized as follow.
-Section 2 presents the background of the mathml package, namely Prolog and the R package `rolog`.
-Section 3 illustrates the features of the mathml package, providing some examples.
+Section 2 presents the background of the mathml package, namely Prolog and the R package \pkg{rolog}.
+Section 3 illustrates the features of the \pkg{mathml} package, providing some examples.
 Section 4 embeds the Discussion and the presentation of a Case Study.
@@
 Whereas Prolog is weak in statistical computation, but strong in symbolic manipulation, the converse may be said for the R language.
-The `rolog` package then bridges this gap by providing an interface to a SWI-Prolog distribution in R.
+The \pkg{rolog} package then bridges this gap by providing an interface to a SWI-Prolog distribution in R.
 The communication between the two systems is mainly in the form of queries from R to Prolog, but two predicates allow Prolog to ring back and evaluate terms in R.
@@
 
-```{r, echo=TRUE}
+```{r, echo=TRUE, results='asis', eval=knitr::is_html_output()}
 library(rolog)
@@
 
-_I encounter a problem, when I explicit the 'asis' in the results='', the execution of the Rmd is halted. Yet, only when I use mathml()._
+```{r, echo=TRUE, results='asis', eval=knitr::is_latex_output()}
+library(rolog)
+library(mathml)
+term = quote(pbinom(k, N, p))
+
+# Pretty print
+mathjax(term)
+```
 
@@
 
-```{r, echo = TRUE}
+```{r,  results = 'asis', echo = TRUE, eval=knitr::is_html_output()}
 term = quote(integrate(sin, 0L, 2L*pi))
 mathml(term)
+```
+
+```{r, echo = TRUE}
 eval(term)
@@
 # # It is a bit of a mystery that R knows the result of this integral.
-# eval(term)
+eval(term)
 ```
@@
 
-# Customizing tooltip design with \pkg{ToOoOlTiPs}
+# Package capabilities
 
-\pkg{ToOoOlTiPs} is a packages for customizing tooltips in interactive graphics, it features these possibilities.
+\pkg{mathml} is a package for pretty mathematical representation of R functions or algorithms in data analysis, scientific reports and interactive web content.
+The features currently supported are listed below, following the order proposed by [@murrell2000].
 
-# A gallery of tooltips examples
+## Arithmetic operators
+The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division.
+Additionally, we have provided two possible ways to represent multiplication: With sign and by juxtaposition and two different kinds of division.
 
-The \CRANpkg{palmerpenguins} data [@palmerpenguins]  features three penguin species which has a lovely illustration by Alison Horst in Figure \@ref(fig:penguins-alison).
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+catmath(x + y)
+catmathjax(x + y)
 
-```{r penguins-alison, out.width = "100%", out.height = "30%", fig.cap = "Artwork by \\@allison\\_horst", fig.alt="A picture of three different penguins with their species: Chinstrap, Gentoo, and Adelie.
-"}
-# knitr::include_graphics("penguins.png")
+catmath(x - y)
+catmathjax(x - y)
+
+# Plus Minus
+catmath(x %+-% y)
+catmathjax(x %+-% y)
+
+catmath(x*y)
+catmathjax(x*y)
+
+# %*%
+catmath(x %*% y)
+catmathjax(x %*% y)
+
+# %.%
+catmath(x %.% y)
+catmathjax(x %.% y)
+
+catmath(X %x% Y)
+catmathjax(X %x% Y)
+
+catmath(x/y)
+catmathjax(x/y)
+
+# %/%: integer division
+catmath(x %/% y)
+catmathjax(x %/% y)
 ```
 
-Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:penguins-tab-interactive)', '\\@ref(tab:penguins-tab-static)'))` prints at the first few rows of the `penguins` data:
+## Fractions
+As in [@murrell2000] the `frac` and `over` functions aim to represent fractions. `choose` and `lchoose`, as in R\ \pkg{base}, display the binomial coefficient.
 
-```{r penguins-tab-interactive, eval = knitr::is_html_output(), layout = "l-body-outset"}
-knitr::kable(head(penguins), format = "html", caption = "A basic table")
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+catmath(frac(x, y))
+catmathjax(frac(x, y))
+
+catmath(over(x, y))
+catmathjax(over(x, y))
+
+catmath(choose(n, k))
+catmathjax(choose(n, k))
+
+catmath(choose(k=K, n=N))
+catmathjax(choose(k=K, n=N))
+
+catmath(lchoose(n, k))
+catmathjax(lchoose(n, k))
 ```
 
-```{r penguins-tab-static, eval = knitr::is_latex_output()}
-knitr::kable(head(penguins), format = "latex", caption = "A basic table") %>% 
-  kableExtra::kable_styling(font_size = 7)
+## Relations
+
+Among the possible relational operators, following [@murrell2000], we included:
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+catmath(x==y)
+catmathjax(x==y)
+
+catmath(x < y)
+catmathjax(x < y)
+
+catmath(x <= y)
+catmathjax(x <= y)
+
+catmath(x >= y)
+catmathjax(x >= y)
+
+catmath(!x)
+catmathjax(!x)
+
+catmath(x != y)
+catmathjax(x != y)
+
+catmath(X ~ Y)
+catmathjax(X ~ Y)
+
+catmath(x %~~% y)
+catmathjax(x %~~% y)
+
+catmath(x %==% y)
+catmathjax(x %==% y)
+
+catmath(x %=~% y)
+catmathjax(x %=~% y)
+
+catmath(x %prop% y)
+catmathjax(x %prop% y)
+
 ```
 
-Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:penguins-plotly)', '\\@ref(fig:penguins-ggplot)'))` shows an `r ifelse(knitr::is_html_output(), 'interactive', '')` plot of the penguins data, made using the `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\CRANpkg{plotly}', '\\CRANpkg{ggplot2}'))` package.
+## Set relations
 
-```{r penguins-plotly, echo = TRUE, fig.height = 5, fig.cap="A basic interactive plot made with the plotly package on palmer penguin data.
-Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis.
-When hovering on a point, a tooltip will show the exact value of the bill depth and length for that point, along with the species name.", include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.alt = "A scatterplot of bill length against bill depth, both measured in millimetre.
-The three species are shown in different colours and loosely forms three clusters.
-Adelie has small bill length and large bill depth, Gentoo has small bill depth but large bill length, and Chinstrap has relatively large bill depth and bill length."}
-p <- penguins %>% 
-  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
-             color = species)) + 
-  geom_point()
-ggplotly(p)
+Here is provided a series of useful set relations.
+
+_Still to be implemented_
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+catmath(a %in% A)
+catmathjax(a %in% A)
+
+catmath(intersect(A, B))
+catmathjax(intersect(A, B))
+
+catmath(union(A, B))
+catmathjax(union(A, B))
+
+catmath(is.null(intersect(A, B)))
+catmathjax(is.null(intersect(A, B)))
 ```
 
-```{r penguins-ggplot, echo = TRUE, fig.height = 5, fig.cap="A basic non-interactive plot made with the ggplot2 package on palmer penguin data.
-Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis.
-Visit the online article to access the interactive version made with the plotly package.", include=knitr::is_latex_output(), eval=knitr::is_latex_output()}
-penguins %>% 
-  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
-             color = species)) + 
-  geom_point()
+## Arrows
+
+Following once again [@murrell2000], we provide here a list of useful arrows
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+# x %<->% y
+catmath(x %<->% y)
+catmathjax(x %<->% y)
+
+# x %->% y
+catmath(x %->% y)
+catmathjax(x %->% y)
+
+# x %<-% y
+catmath(x %<-% y)
+catmathjax(x %<-% y)
+
+# x %up% y:
+catmath(x %up% y)
+catmathjax(x %up% y)
+
+# x %down% y:
+catmath(x %down% y)
+catmathjax(x %down% y)
+
+# x %<=>% y: solved it using iff
+catmath(x %<=>% y)
+catmathjax(x %<=>% y)
+
+# x %=>% y: solved using rArr
+catmath(x %=>% y)
+catmathjax(x %=>% y)
+
+# x %<=% y: solved using lArr
+catmath(x %<=% y)
+catmathjax(x %<=% y)
+
+# x %dblup% y:
+catmath(x %dblup% y)
+catmathjax(x %dblup% y)
+
+# x %dbldown% y:
+catmath(x %dbldown% y)
+catmathjax(x %dbldown% y)
+
 ```
 
-# Summary
+# A case study
 
-We have displayed various tooltips that are available in the package \pkg{ToOoOlTiPs}.
+For a more clear representation of what the \pkg{mathml} package is supposed to do, we provide here a case study, based on the work by Schwarz [-@schwarz1994], in which the features of \pkg{mathml} are implemented.
+The reader
+may recognize that Schwarz [-@schwarz1994]'s study can be traced back to the strand of
+decision-making models, where decision-making is assumed to be a process of
+noisy accumulation of information over time [e.g., @ratcliff2016]. The aim of
+Schwarz' [-@schwarz1994] study is to present a new explanation of redundancy gains when
+observers make speeded responses to stimuli of different sources, and the same
+information is presented on two channels. Schwarz [-@schwarz1994] describes a model that assumes
+the superposition of channel-specific diffusion processes that eventually reach
+an absorbing barrier to elicit the response. For a detailed description the
+reader may refer to the original article.
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+# Improve display
+hook(quote(mu_A), quote(mu["A"]))
+hook(quote(mu_B), quote(mu["B"]))
+hook(quote(sigma_A), quote(sigma["A"]))
+hook(quote(sigma_B), quote(sigma["B"]))
+hook(quote(mu_M), quote(mu["M"]))
+hook(quote(zn), quote(z[tau]))
+hook(quote(M), quote(overline(X)))
+```
+
+### The model
+
+Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
+alone or in combination (redundant stimuli), with the redundant stimuli being
+presented either simultaneously or with onset asynchrony \(\tau\). The channel
+activation is described as a two-dimensional Wiener process with
+drifts \(\mu_\mathrm A\), \(\mu_\mathrm B\),
+variances \(\sigma^2_\mathrm A\), \(\sigma^2_\mathrm B\),
+correlation \(\rho_\mathrm{AB}\) and initial
+conditions \(X_i(t=0)=0, i=\mathrm{A, B}\). The response is elicited when the
+process reaches the absorbing barrier \(c > 0\) for the first time. In combined
+stimuli ("redundant" stimuli), the overall diffusion process
+is \(X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)\) , which is again a
+Wiener process with drift \(\mu_\mathrm A + \mu_\mathrm B\) and
+variance \(\sigma^2_\mathrm A + \sigma^2_\mathrm B
+  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B\).
+
+In single-target trials, the first passages at \(c\) are expected at
+\(E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}\), whereas in synchronous
+redundant-target trials, we have
+\(E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}\).
+
+For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
+first-passage time \(E[D_{\mathrm A(\tau)\mathrm B}]\) as a function of the
+stimulus onset asyncrony \(\tau > 0\) (Eq.\ 10),
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
+    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
+      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
+        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
+}
+
+mathml(f)
+```
+
+The observable response time is assumed to be the sum of \(D\), the time
+employed to reach the threshold for the decision, and \(M\), denoting the of the
+other processes.
+
+\(E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M\)
+
+### Methods
+
+Schwarz [-@schwarz1994] applied the model to Miller’s [-@miller1986] data from
+a redundant signals task with 13 onset
+asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
+where 0 refers to synchronous AB, \(\infty\) refers to the single-target
+presentation, and negative \(\tau\) denote those conditions in which B is
+presented before A. Each condition was replicated 400 times. The 13 mean RTs and
+standard deviations are given in Table 2 below.
+
+Then, from Schwarz' [-@schwarz1994] Equations [2] [3] and [10], we calculated the expected first time passage for the respective \(\tau\) and the parameters \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)
+
+````{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+mu_A = 0.53
+mu_B = 1.32
+sigma_A = 4.4
+sigma_B = 18.3
+tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
+c = 100
+mu_M = 161
+````
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+# Conditions
+
+g <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+{
+    if( tau == Inf) return(dfrac(c, mu_A));
+    if( tau == -Inf) return(dfrac(c, mu_B));
+    if( tau > 0L)  return(f(tau, c, mu_A, sigma_A, mu_B, sigma_B));
+    if( tau == 0L) return(dfrac(c, mu_A + mu_B));
+    if( tau < 0L) return(f(-tau, c,  mu_B, sigma_B, mu_A, sigma_A))
+}
+
+mathml(g)
+
+```
+
+For instance, we can see that with \(\tau = -167\) the expected first passage time is:
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+mu_A = 0.53
+mu_B = 1.32
+sigma_A = 4.4
+sigma_B = 18.3
+tau = -167
+c = 100
+mu_M = 161
+
+g(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+
+```
+
+To which we have to add \(\mu_M = 161\) (which represents the amount of time that the movement of providing the answer takes) and we obtain:
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+g(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
+```
+
+Similarly, we obtain the first-time passage for every \(\tau\):
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+G <- Vectorize(g, vectorize.args = 'tau')
+
+h <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+{
+  G(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
+}
+mathml(h)
+```
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+mu_M = 161
+mu_A = 0.53
+mu_B = 1.32
+sigma_A = 4.4
+sigma_B = 18.3
+tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
+c = 100
+
+#h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+
+```
+
+### Model fitting and Parameters Estimation
+
+Then we proceeded with the model fitting, using the data provided by Miller [-@miller1986] study in Schwarz [-@schwarz1994]. First, we calculated the predicted means RT as \(E[RT]=E[T]+\mu_M\).
+
+We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed [@ratcliff2008] with at the first passage of \(c\) \(X(t) \sim N(\mu t , \sigma^2 t)\).
+Where \(T\) is expressed as \(T= \arg \min_t X(t)\geq c\) with \(T \sim IG (c, \mu, \sigma^2)\) with mean \(E(T)=\displaystyle \frac{c}{\mu}\) and variance \(Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}\).
+Furthermore, for the Central Limit Theorem \(\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})\)
+
+For these reasons, in order to obtain the Goodness of Fit measure we proceeded with a z-standardisation.
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+zf <-  function(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+{
+  E <- h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+  dfrac(M-E, s/sqrt(N))
+}
+
+mathml(zf)
+```
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+# The vector X contains the observed values, the vector s their observed standard deviation and N = 400 are the number of trials.
+
+M = c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
+s = c(56, 58, 40, 40, 32, 28, 28, 28, 26, 30, 32, 34, 92)
+N = c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400)
+
+zn <- zf(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+# zn
+
+```
+
+Then, we calculated the Goodness of Fit as follows:
+
+\(X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)\).
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+par <- c(mu_A = 0.53, sigma_A = 4.4, mu_B = 1.32, sigma_B = 18.3, mu_M = 161)
+
+# Goodness of Fit
+
+gof <- function(par, tau, M, s, N)
+{
+  zn <- zf(M, s, N, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])
+  sum(zn^2L)
+}
+
+mathml(gof)
+
+n <- gof(par, tau, M, s, N)
+#n
+
+#pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)
+
+```
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+lsq = function(tau, M, s, N)
+{
+  start = c(mu_A = 0.3, sigma_A = 2, mu_B = 1 , sigma_B = 5, mu_M = 200)
+  optim(start, fn = gof,
+    tau = tau, M = M, s = s, N = N, hessian = TRUE, control = list(fnscale = 1))
+}
+
+# mathml(lsq)
+
+# Parameters' estimates
+fit <-  lsq(tau, M, s, N)
+
+est = fit$par
+se = sqrt(diag(solve(fit$hessian)))
+cilo = qnorm(0.025, est, se)
+ciup = qnorm(0.975, est, se)
+
+# Lower bound
+#cilo
+
+# Upper bound
+#ciup
+
+```
+
+### Results
+
+The calculated Goodness of Fit value is \(X^2= 30.54\), \(p= 0.0002\). Additionally, we proceeded with the parameters' estimation with the Least Squares Method.
+
+The values of the estimated parameters reported in the table below are consistent with those reported by Schwarz [-@schwarz1994].
+
+: Table 1: Estimated Model's Parameters
+
+| Parameters | Estimate | 95% Confidence Interval
+| :---: | :---: | :---:|
+| \( \mu_ \mathrm A \)| 0.53 | 0.51, 0.56
+| \( \sigma_ \mathrm A \)| 4.4 | 3.8, 5.1
+| \( \mu_ \mathrm B \)| 1.32 | 1.2, 1.4
+| \( \sigma_ \mathrm B \)| 18.3 | 11.6, 24.9
+| \( \mu_ \mathrm M \)| 161 | 157, 165
+
+Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i.
+\(\tau= 100\)) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model.
+Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.
+
+: Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's [-@miller1986] study in Schwarz [-@schwarz1994]
+
+| SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
+|:--------:|:-----------------:|:----------------:|:-----------:|
+\(-\infty\)| 237 | 231 | 56 |
+\(-167\)   | 231 | 234 | 58 |
+\(-133\)   | 230 | 230 | 40 |
+\(-100\)   | 228 | 227 | 40 |
+\(-67\)    | 226 | 228 | 32 |
+\(-33\)    | 222 | 221 | 28 |
+\(0\)      | 215 | 217 | 28 |
+\(33\)     | 239 | 238 | 28 |
+\(67\)     | 262 | 263 | 26 |
+\(100\)    | 283 | 277 | 30 |
+\(133\)    | 300 | 298 | 32 |
+\(167\)    | 313 | 316 | 34 |
+\(\infty\) | 350 | 348 | 92 |
+
+ _Note_.  Parameters: \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)
+
+# Conclusions
+
+* General introduction on R and its features
+* What is mathml 
+* issues the user must be aware of
+* points of strength
+* future directions
+* (inspirational ending)
+* ways to install rolog and mathml
+
+# Acknowledgments

### 2022-11-30 inst/pl/mathml.pl

@@ -932,6 +932,10 @@ math(Flags, $(integrate(Fn, Lower, Upper), value), New, M)
  => Flags = New,
     M = integrate(Fn, Lower, Upper).
 
+math(Flags, integrate(f=Fn, lower=Lower, upper=Upper), New, M)
+ => Flags = New,
+    M = integrate(Fn, Lower, Upper).
+
 % No argument names
 math(Flags, integrate(Fn, Lower, Upper), New, M)
  => Flags = New,

### 2022-12-01 Draft1.Rmd
operators

@@
 
+## Operators
+A list of possible operators is available here.
+
+```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
+
+catmath(sum(x[i], i==1L, n))
+catmathjax(sum(x[i], i==1L, n))
+
+# fixme: braces not needed when right to *
+catmath(a * sum(b))
+catmathjax(a * sum(b))
+
+# Multiplication (capital Pi) is missing
+
+# Integral is missing
+
+catmath(A & B)
+catmathjax(A & B)
+
+catmath(A | B)
+catmathjax(A | B)
+
+catmath(!A)
+catmathjax(!A)
+
+catmath(xor(A, B))
+catmathjax(xor(A, B))
+
+catmath(exp(2L*pi*i))
+catmathjax(exp(2L*pi*i))
+
+catmath(expm1(2L*pi*i))
+catmathjax(expm1(2L*pi*i))
+
+catmath(log(x))
+catmathjax(log(x))
+
+catmath(log10(x))
+catmathjax(log10(x))
+
+catmath(log2(x))
+catmathjax(log2(x))
+
+catmath(logb(x, e))
+catmathjax(logb(x, e))
+
+catmath(log1p(x))
+catmathjax(log1p(x))
+
+catmath(ceiling(pi))
+catmathjax(ceiling(pi))
+
+catmath(floor(pi))
+catmathjax(floor(pi))
+
+catmath(sin(pi/2L))
+catmathjax(sin(pi/2L))
+
+catmath(cos(pi/2L))
+catmathjax(cos(pi/2L))
+
+catmath(tan(pi/2L))
+catmathjax(tan(pi/2L))
+
+```
+
+## Absolute value
+
+Is it also possible to calculate and represent the absolute value, as:
+
+```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
+catmath(abs(x))
+catmathjax(abs(alpha))
+
+```
+
 # A case study
@@
 
+## Mistakes
+
+The case study presented above displays several possible ways of how \pkg{mathml} can be a useful tool in data representation and calculation.
+Furthermore, the careful reader might have also noticed that \pkg{mathml} allows also for a more smoothly detection of possible mistakes in writing.
+
+An interesting use case of the package is in this sense the readability of self-written
+functions in R code. R code is notoriously hard to read, as are most programming
+languages [@green1977]. The poor legibility of the language is one of the main
+sources of mistakes. For illustration, we consider Equation 10 in
+Schwarz [-@schwarz1994], namely the same equation reported above. The first version has a wrong parenthesis, which is barely visible in the R code.
+
+````{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+f <- function(tau)
+{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
+    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
+      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
+        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
+}
+
+attributes(f)$name = quote(f)
+mathml(f)
+````
+
+But, if the function is rendered in MathML, the wrong parenthesis appears as a curly
+brace, and we immediately see the mistake. The correct version is shown below
+for comparison.
+
+````{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+f <- function(tau)
+{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
+    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
+      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
+        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
+}
+
+attributes(f)$name = quote(f)
+mathml(f)
+````
+
+This example may seem trivial.
+However, as the reader well knows, a missed parenthesis is often the cause of untraceable errors and contradictory results.
+Thus, rendering the expression in MathMl improves the readability of the text, allowing better identification of the sources of possible errors.
+
 # Conclusions
 
-* General introduction on R and its features
-* What is mathml 
-* issues the user must be aware of
-* points of strength
-* future directions
-* (inspirational ending)
-* ways to install rolog and mathml
+This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time.
+This makes it easier for the user to notice writing errors and correct them more smoothly.
+Building on current features of R\ and existing packages for displaying formulas in R\ (e.g.
+Murrell and Ihaka [-@murrell2000], Allaire et al.
+[-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility.
+To do so, it uses the Prolog's grammar rules, via the \pkg{rolog}, and translates the R\ expressions to MathML.
+Then, two possible solutions for typesetting are available: LaTeX and MathJax to export R\ Markdown documents in different formats or ShinyApp webpages.
+The article and Appendix show examples of how the package can be used and what operations are currently possible.
+Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.
+In its current version \pkg{mathml} presents some limitations, as, for instance, (...). However, several solutions to this issue are possible (...).
+
+\pkg{mathml} is available for R\ version (…) and later, and can be easily installed using the usual `install.packages("mathml")`.
+The source code of the package is found at (…) including installation instructions for Unix, Windows and macOS.
 

### 2022-12-01 Draft1.Rmd
%.%

@@
 
-# Integral is missing
+catmath(integrate(x))
+catmathjax(integrate(x))
 
@@
 For a more clear representation of what the \pkg{mathml} package is supposed to do, we provide here a case study, based on the work by Schwarz [-@schwarz1994], in which the features of \pkg{mathml} are implemented.
-The reader
-may recognize that Schwarz [-@schwarz1994]'s study can be traced back to the strand of
-decision-making models, where decision-making is assumed to be a process of
+
+The reader may recognize that Schwarz' [-@schwarz1994] study can be traced back to the strand of decision-making models, where decision-making is assumed to be a process of
 noisy accumulation of information over time [e.g., @ratcliff2016]. The aim of
@@
 Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.
-In its current version \pkg{mathml} presents some limitations, as, for instance, (...). However, several solutions to this issue are possible (...).
+In its current version \pkg{mathml} has some limitations.
+For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation.
+However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations.
+Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications.
 

### 2022-12-01 Draft1.Rmd
removed duplicate canonical

@@
 library(rolog)
-# Apply match.call to all components of a term
-canonical <- function(term)
-{
-  if(is.call(term))
-  {
-    f <- match.fun(term[[1]])
-    if(!is.primitive(f))
-      term <- match.call(f, term)
 
-    # Recurse into arguments
-    term[-1] <- lapply(term[-1], canonical)
-  }
-  return(term)
-}
 # A custom function

### 2022-12-02 Draft1.Rmd
to do

@@
 
+#### To do
+
+* paste(x, y, z)
+* Spacing
+* Comma separated lists
+* Ellipsis
+* Radicals
+* Subsets
+* Greek letters in MathJax
+* Typeface
+* Stylechanges
+* Phantoms
+
 # Introduction
@@
 support for mathematical expressions [@Sarkar2022], which already is a big
-improvement. Rules for the translation of R's built-in language elements to
-their mathematical representation are lacking.
+improvement. However, rules for the translation of R's built-in language elements to
+their mathematical representation are still lacking.
 
@@
 # Background
+
+## Prolog: the \pkg{rolog} package
 
@@
 the integration variable for `sin`.
-For that purpose, rolog provides a predicate `r_eval/2` that calls R from Prolog.
+For that purpose, \pkg{rolog} provides a predicate `r_eval/2` that calls R from Prolog.
 Here, the predicate is used for the  R\ function `formalArgs(args(sin))`, which returns the name of the function argument of `sin`, that is, `x`.
@@
 
-Note that the Prolog end, the handler for `integrate/3` is rather rigid; it
+Note that, on the Prolog end, the handler for `integrate/3` is rather rigid; it
 accepts only these three arguments in that particular order, and without names,
-that is, `integrate(sin, lower=0L, upper=2L * pi)` would not print the desired
+that is, `integrate(sin, lower=0L, upper=2L * pi)`, would not print the desired
 result.
@@
 
-# Plus Minus
 catmath(x %+-% y)
@@
 
-# %*%
 catmath(x %*% y)
@@
 
-# %.%
 catmath(x %.% y)
@@
 
-# %/%: integer division
 catmath(x %/% y)
 catmathjax(x %/% y)
+```
+
+## Subscripts and Superscripts
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+catmath(a["x"])
+catmathjax(a["x"])
+
+catmath(a^x)
+catmathjax(a^x)
+
+catmath(x["b"]^"a")
+catmathjax(x["b"]^"a")
+
 ```
@@
 
-catmath(choose(k=K, n=N))
-catmathjax(choose(k=K, n=N))
-
 catmath(lchoose(n, k))
 catmathjax(lchoose(n, k))
+```
+
+## Radicals
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+
+catmath(sqrt(x))
+catmathjax(sqrt(x))
+
 ```
@@
 
+## Accents
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+catmath(hat(x))
+catmathjax(hat(x))
+
+catmath(hat(xy))
+catmathjax(hat(xy))
+
+catmath(mean(x))
+catmathjax(mean(x))
+
+catmath(mean(xy))
+catmathjax(mean(xy))
+
+```
+
+## Symbolic names
+
+```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+
+catmath(Alpha-Omega)
+catmathjax(Alpha-Omega)
+
+catmath(alpha-omega)
+catmathjax(alpha-omega)
+
+catmath(Inf)
+catmathjax(Inf)
+
+```
+
 ## Operators
@@
 
-catmath(integrate(x))
-catmathjax(integrate(x))
+catmath(integrate(sin, 0L, 2L*pi))
+catmathjax(integrate(sin, 0L, 2L*pi))
 
@@
 
+## Grouping
+
+```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
+catmath(x^y + z)
+catmathjax(x^y + z)
+
+catmath(x^(y + z))
+catmathjax(x^(y + z))
+
+# without parenthesis
+mathml(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
+mathjax(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
+
+# shortcut
+catmath(x^{y + z})
+catmathjax(x^{y + z})
+
+```
+
 ## Absolute value
@@
 catmathjax(abs(alpha))
+
+```
+
+## Matrix and Vectors
+
+Since also Matrices and Vectors are important elements in R\, we decided to implement them and render them graphically in the following ways.
+
+```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
+
+vec <- 1:3
+mathml(canonical(vec))
+mathml(vec)
+mathjax(canonical(vec))
+mathjax(vec)
+
+mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
+mathml(canonical(mat))
+mathjax(canonical(mat))
 
@@
 This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time.
-This makes it easier for the user to notice writing errors and correct them more smoothly.
+This makes it easier for the user to notice writing errors and correct them smoothly.
 Building on current features of R\ and existing packages for displaying formulas in R\ (e.g.
@@
 [-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility.
-To do so, it uses the Prolog's grammar rules, via the \pkg{rolog}, and translates the R\ expressions to MathML.
-Then, two possible solutions for typesetting are available: LaTeX and MathJax to export R\ Markdown documents in different formats or ShinyApp webpages.
+To do so, it uses the Prolog's grammar rules and translates the R\ expressions to MathML.
+Then, two possible solutions for typesetting are available to export R\ Markdown documents in different formats or ShinyApp webpages: MathML and MathJax.
+
 The article and Appendix show examples of how the package can be used and what operations are currently possible.
 Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.
+
 In its current version \pkg{mathml} has some limitations.

### 2022-12-02 inst/pl/mathml.pl

@@ -293,9 +293,10 @@ mathjax :-
 math(R, M),
     atom(R),
     memberchk(R, [alpha, beta, gamma, delta, epsilon, varepsilon, zeta, eta,
-        theta, vartheta, iota, kappa, lambda, mu, nu, xi, pi, rho, sigma, varsigma, tau,
-        upsilon, phi, varphi, chi, psi, omega, 'Gamma', 'Delta', 'Theta',
-        'Lambda', 'Xi', 'Pi', 'Sigma', 'Upsilon', 'Phi', 'Psi', 'Omega'])
+        theta, vartheta, iota, kappa, lambda, mu, nu, xi, omikron, pi, rho, sigma, varsigma, tau,
+        upsilon, phi, varphi, chi, psi, omega, 'Alpha', 'Beta', 'Gamma', 'Delta', 'Epsilon',
+        'Zeta', 'Eta','Theta', 'Iota', 'Kappa', 'Lambda', 'Mi', 'Ni', 'Xi', 'Omikron', 'Pi',
+        'Rho','Sigma', 'Tau','Upsilon', 'Phi', 'Chi','Psi', 'Omega'])
  => M = greek(R).
 
 ml(_Flags, greek(R), M)

### 2022-12-05 Draft1.Rmd
errors
1 sentence

@@
 ## Subscripts and Superscripts
+Secondly, \pkg{mathml} allows also for the display of subscripts, superscripts and their combination.
 
@@
 
+## Errors
+
+An additional feature that this package offers is related to its application in educational scenarios.
+The functions here reported are intended to identify and highlight possible errors that students make, as well as provide the correct answer for the task.
+
+```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+
+```
+
 # Conclusions

### 2022-12-05 Draft1.Rmd
www

@@
     - Innsbruck, Austria
-    url: https://www.britannica.com/animal/quokka
+    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/
     orcid: 0000-0001-9974-0057

### 2022-12-05 Draft2.Rmd
so many changes

--- 2022-12-05 Draft1.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+# knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
 library(plotly)
@@
 
+mathout <- function(term)
+{
+  if(knitr::is_html_output())
+    return(mathml(term))
+
+  if(knitr::is_latex_output())
+    return(mathjax(term))
+
+  warning("no output specified")  
+  mathjax(term)
+}
+
 catmath <- function(term)
@@
   s <- substitute(term)
-  c <- canonical(s)
-  mathml(c)
+  mathout(canonical(s))
 }
-
-catmathjax <- function(term)
-{
-  s <- substitute(term)
-  c <- canonical(s)
-  mathjax(c)
-}
-
 ```
@@
 
-```{r, echo=TRUE, results='asis', eval=knitr::is_html_output()}
+```{r, echo=TRUE, results='asis'}
 library(rolog)
@@
 # Pretty print
-mathml(term)
-```
-
-```{r, echo=TRUE, results='asis', eval=knitr::is_latex_output()}
-library(rolog)
-library(mathml)
-term = quote(pbinom(k, N, p))
-
-# Pretty print
-mathjax(term)
+mathout(term)
 ```
@@
 
-```{r,  results = 'asis', echo = TRUE, eval=knitr::is_html_output()}
+```{r,  results = 'asis', echo = TRUE}
 term = quote(integrate(sin, 0L, 2L*pi))
-mathml(term)
+mathout(term)
 ```
@@
 
-```{r, echo = TRUE}
-library(rolog)
-
+```{r, echo = TRUE, results='asis'}
 # A custom function
@@
 }
-# # Mixture of (partially) named and positional arguments in unusual order
-term <- quote(2L * integrate(low=-Inf, up=Inf, g)$value)
-# mathml(canonical(term))
-# # It is a bit of a mystery that R knows the result of this integral.
-eval(term)
+
+# Mixture of (partially) named and positional arguments in unusual order
+term <- call("integrate", low=-Inf, up=Inf, g)
+# mathout(canonical(term))
+
+# term <- quote(2L * integrate(low=-Inf, up=Inf, g)$value)
+# mathout(canonical(term))
+
+# It is a bit of a mystery that R knows the result of this integral.
+# eval(term)
 ```
@@
 order.
-
-<!-- Some packages on interactive graphics include \CRANpkg{plotly} [@plotly] that interfaces with Javascript for web-based interactive graphics, \CRANpkg{crosstalk} [@crosstalk] that specializes cross-linking elements across individual graphics.
-The recent R Journal paper \CRANpkg{tsibbletalk} [@RJ-2021-050] provides a good example of including interactive graphics into an article for the journal.
-It has both a set of linked plots, and also an animated gif example, illustrating linking between time series plots and feature summaries.
--->
 
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 catmath(x + y)
-catmathjax(x + y)
 
 catmath(x - y)
-catmathjax(x - y)
 
-catmath(x %+-% y)
-catmathjax(x %+-% y)
+# Note wrong precedence of custom operator
+catmath(mean(m) %+-% 1.96 * s/sqrt(N))
+
+# This is what we actually want
+catmath(mean(m) %+-% (1.96 * s/sqrt(N)))
+
+# Without parentheses
+catmath(mean(m) %+-% {1.96 * s/sqrt(N)})
 
 catmath(x*y)
-catmathjax(x*y)
 
 catmath(x %*% y)
-catmathjax(x %*% y)
 
 catmath(x %.% y)
-catmathjax(x %.% y)
 
 catmath(X %x% Y)
-catmathjax(X %x% Y)
 
 catmath(x/y)
-catmathjax(x/y)
 
 catmath(x %/% y)
-catmathjax(x %/% y)
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 catmath(a["x"])
-catmathjax(a["x"])
 
 catmath(a^x)
-catmathjax(a^x)
 
 catmath(x["b"]^"a")
-catmathjax(x["b"]^"a")
-
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 catmath(frac(x, y))
-catmathjax(frac(x, y))
 
 catmath(over(x, y))
-catmathjax(over(x, y))
 
 catmath(choose(n, k))
-catmathjax(choose(n, k))
 
 catmath(lchoose(n, k))
-catmathjax(lchoose(n, k))
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
-
+```{r, results="asis", echo=TRUE}
 catmath(sqrt(x))
-catmathjax(sqrt(x))
-
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 catmath(x==y)
-catmathjax(x==y)
 
 catmath(x < y)
-catmathjax(x < y)
 
 catmath(x <= y)
-catmathjax(x <= y)
 
 catmath(x >= y)
-catmathjax(x >= y)
 
 catmath(!x)
-catmathjax(!x)
 
 catmath(x != y)
-catmathjax(x != y)
 
 catmath(X ~ Y)
-catmathjax(X ~ Y)
 
 catmath(x %~~% y)
-catmathjax(x %~~% y)
 
 catmath(x %==% y)
-catmathjax(x %==% y)
 
 catmath(x %=~% y)
-catmathjax(x %=~% y)
 
 catmath(x %prop% y)
-catmathjax(x %prop% y)
-
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 catmath(a %in% A)
-catmathjax(a %in% A)
 
 catmath(intersect(A, B))
-catmathjax(intersect(A, B))
 
 catmath(union(A, B))
-catmathjax(union(A, B))
 
 catmath(is.null(intersect(A, B)))
-catmathjax(is.null(intersect(A, B)))
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 # x %<->% y
 catmath(x %<->% y)
-catmathjax(x %<->% y)
 
@@
 catmath(x %->% y)
-catmathjax(x %->% y)
 
@@
 catmath(x %<-% y)
-catmathjax(x %<-% y)
 
@@
 catmath(x %up% y)
-catmathjax(x %up% y)
 
@@
 catmath(x %down% y)
-catmathjax(x %down% y)
 
@@
 catmath(x %<=>% y)
-catmathjax(x %<=>% y)
 
@@
 catmath(x %=>% y)
-catmathjax(x %=>% y)
 
@@
 catmath(x %<=% y)
-catmathjax(x %<=% y)
 
@@
 catmath(x %dblup% y)
-catmathjax(x %dblup% y)
 
@@
 catmath(x %dbldown% y)
-catmathjax(x %dbldown% y)
-
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 catmath(hat(x))
-catmathjax(hat(x))
 
 catmath(hat(xy))
-catmathjax(hat(xy))
 
 catmath(mean(x))
-catmathjax(mean(x))
 
 catmath(mean(xy))
-catmathjax(mean(xy))
-
 ```
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
-
+```{r, results="asis", echo=TRUE}
 catmath(Alpha-Omega)
-catmathjax(Alpha-Omega)
 
 catmath(alpha-omega)
-catmathjax(alpha-omega)
+
+catmath(pi-Pi)
 
 catmath(Inf)
-catmathjax(Inf)
-
 ```
@@
 ## Operators
+
 A list of possible operators is available here.
 
-```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
-
+```{r, results='asis', echo=TRUE}
 catmath(sum(x[i], i==1L, n))
-catmathjax(sum(x[i], i==1L, n))
 
@@
 catmath(a * sum(b))
-catmathjax(a * sum(b))
 
 # Multiplication (capital Pi) is missing
+catmath(a * prod(b))
 
 catmath(integrate(sin, 0L, 2L*pi))
-catmathjax(integrate(sin, 0L, 2L*pi))
 
 catmath(A & B)
-catmathjax(A & B)
 
 catmath(A | B)
-catmathjax(A | B)
 
 catmath(!A)
-catmathjax(!A)
 
 catmath(xor(A, B))
-catmathjax(xor(A, B))
 
 catmath(exp(2L*pi*i))
-catmathjax(exp(2L*pi*i))
 
 catmath(expm1(2L*pi*i))
-catmathjax(expm1(2L*pi*i))
 
 catmath(log(x))
-catmathjax(log(x))
 
 catmath(log10(x))
-catmathjax(log10(x))
 
 catmath(log2(x))
-catmathjax(log2(x))
 
 catmath(logb(x, e))
-catmathjax(logb(x, e))
 
 catmath(log1p(x))
-catmathjax(log1p(x))
 
 catmath(ceiling(pi))
-catmathjax(ceiling(pi))
 
 catmath(floor(pi))
-catmathjax(floor(pi))
 
 catmath(sin(pi/2L))
-catmathjax(sin(pi/2L))
 
 catmath(cos(pi/2L))
-catmathjax(cos(pi/2L))
 
 catmath(tan(pi/2L))
-catmathjax(tan(pi/2L))
-
 ```
@@
 
-```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results='asis', echo=TRUE}
 catmath(x^y + z)
-catmathjax(x^y + z)
 
 catmath(x^(y + z))
-catmathjax(x^(y + z))
 
 # without parenthesis
-mathml(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
-mathjax(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
+mathout(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
 
@@
 catmath(x^{y + z})
-catmathjax(x^{y + z})
-
 ```
@@
 
-```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results='asis', echo=TRUE}
 catmath(abs(x))
-catmathjax(abs(alpha))
-
 ```
@@
 
-```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
-
+```{r, results='asis', echo=TRUE}
 vec <- 1:3
-mathml(canonical(vec))
-mathml(vec)
-mathjax(canonical(vec))
-mathjax(vec)
+mathout(canonical(vec))
+mathout(vec)
 
 mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
-mathml(canonical(mat))
-mathjax(canonical(mat))
-
+mathout(canonical(mat))
 ```
@@
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
-
+```{r, echo=FALSE}
 # Improve display
@@
 
-```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=TRUE}
 f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
@@
 
-mathml(f)
+mathout(f)
 ```
@@
 
-````{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+````{r, echo=FALSE}
 mu_A = 0.53
@@
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
-
+```{r, results="asis", echo=FALSE}
 # Conditions
@@
 
-mathml(g)
-
+mathout(g)
 ```
@@
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+```{r, echo=FALSE}
 mu_A = 0.53
@@
 g(tau, c, mu_A, sigma_A, mu_B, sigma_B)
-
 ```
@@
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+```{r, echo=FALSE}
 g(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
@@
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=FALSE}
 
@@
 }
-mathml(h)
+
+mathout(h)
 ```
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+```{r, echo=FALSE}
 
@@
 
-#h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
-
+h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
 ```
@@
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
-
+```{r, results="asis", echo=FALSE}
 zf <-  function(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
@@
 
-mathml(zf)
+mathout(zf)
 ```
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
-
+```{r, echo=FALSE}
 # The vector X contains the observed values, the vector s their observed standard deviation and N = 400 are the number of trials.
@@
 zn <- zf(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
-# zn
-
+zn
 ```
@@
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
-
+```{r, results="asis", echo=FALSE}
 par <- c(mu_A = 0.53, sigma_A = 4.4, mu_B = 1.32, sigma_B = 18.3, mu_M = 161)
@@
 
-mathml(gof)
-
-n <- gof(par, tau, M, s, N)
-#n
-
-#pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)
-
+mathout(gof)
 ```
 
-```{r, results="asis", echo=FALSE,  eval=knitr::is_html_output()}
+```{r, results="asis", echo=FALSE}
+n <- gof(par, tau, M, s, N)
+n
+pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)
+```
 
+```{r, results="asis", echo=FALSE}
 lsq = function(tau, M, s, N)
@@
 
-# mathml(lsq)
+mathout(lsq)
 
@@
 
-````{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
+````{r, results="asis", echo=TRUE}
 f <- function(tau)
@@
 attributes(f)$name = quote(f)
-mathml(f)
+mathout(f)
 ````

### 2022-12-05 R/mathml.R

@@ -92,7 +92,8 @@ mathml_preproc <- function(query=quote(2 != 2))
 #'
 mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
 {
-  t = rolog::once(call("r2mathjax", term, expression(X)),
+  flags = attributes(term)
+  t = rolog::once(call("r2mathjax", flags, term, expression(X)),
     options=list(preproc=mathml_preproc))
   cat(paste(t$X, collapse=""))
 }

### 2022-12-05 Draft2.Rmd
text

@@
   - name: Irene Alfarone
-    url: https://www.britannica.com/animal/bilby
+    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/
     email: Irene.Alfarone@uibk.ac.at
@@
 
+## Hooks
+Before start listing all the possible operations with \pkg{mathml},  we mention here that \pkg{mathml} also allows the user to create hooks for specific functions. As an example:
+
+```{r, echo=TRUE}
+# Improve display
+hook(quote(mu_A), quote(mu["A"]))
+hook(quote(mu_B), quote(mu["B"]))
+hook(quote(sigma_A), quote(sigma["A"]))
+hook(quote(sigma_B), quote(sigma["B"]))
+hook(quote(mu_M), quote(mu["M"]))
+hook(quote(zn), quote(z[tau]))
+hook(quote(M), quote(overline(X)))
+```
+
+The hooks reported here will be the one used in Section @ref[] to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance
+
 ## Arithmetic operators
-The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division.
-Additionally, we have provided two possible ways to represent multiplication: With sign and by juxtaposition and two different kinds of division.
+The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division. For instance:
 
 ```{r, results="asis", echo=TRUE}
-catmath(x + y)
+catmath(x + y - z)
+```
 
-catmath(x - y)
+In addition, with \pkg{mathml} it is also possible to combine `minus` and `plus` and get a new operator.
+However, in R\ the operator \(\pm\) is not present and this leads to some interesting questions.
+If we let R\ present the expression, this is what we get:
 
+```{r, results='asis', echo=TRUE}
 # Note wrong precedence of custom operator
 catmath(mean(m) %+-% 1.96 * s/sqrt(N))
+```
 
+As can be seen, the new custom operator takes precedence over multiplication, as R\ automatically puts `mean(m) %+-% 1.96` in parentheses.
+Unfortunately, this is not consistent with the rules of mathematics.
+As the reader well knows, it is not possible in R\ to change the precedence of custom operators.
+This leaves two possible solutions.
+The first involves a mathematically unnecessary parenthesis around `(1.96 * s/sqrt(N))`.
+The second is more elegant and mathematically consistent: here `1.96 * s/sqrt(N)` is put in curly brackets and mathematical formality is respected.
+In both solutions R\ is able to produce correct output once the order of operations is made explicit.
+
+```{r, results='asis', echo=TRUE}
 # This is what we actually want
@@
 catmath(mean(m) %+-% {1.96 * s/sqrt(N)})
+```
+
+_The following can be inserted in a table_
+
+```{r, results='asis', echo=TRUE}
 
@@
 
+At first glance, converting operators may seem trivial enough.
+However, when you want to deal with, for example, exponentiation, things can get more complicated.
+We see here that if we want `x` to the power of `y + z`, following R\ notation we should add a parenthesis around the two terms:
+
 ```{r, results='asis', echo=TRUE}
@@
 catmath(x^(y + z))
+```
 
+As we can see, R\ automatically put `y + z` in parentheses even if, from a mathematical point of view, the brackets are not useful.
+To solve this issue two solutions are possible, the first one is cumbersome since it requires many specifications.
+The second, the shortcut, uses a syntax similar to LateX and puts the power in curly brackets.
+In both cases the output produced is consistent with the mathematical notation.
+
+```{r, results='asis', echo=TRUE}
 # without parenthesis
@@
 
+blabla `quote(a * (b + c))[[3]][[1]]`
 ## Absolute value
@@
 reader may refer to the original article.
-
-```{r, echo=FALSE}
-# Improve display
-hook(quote(mu_A), quote(mu["A"]))
-hook(quote(mu_B), quote(mu["B"]))
-hook(quote(sigma_A), quote(sigma["A"]))
-hook(quote(sigma_B), quote(sigma["B"]))
-hook(quote(mu_M), quote(mu["M"]))
-hook(quote(zn), quote(z[tau]))
-hook(quote(M), quote(overline(X)))
-```
 

### 2022-12-05 Draft2.Rmd
integral of custom R functions

@@
     - Innsbruck, Austria
-    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/
+    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
     orcid: 0000-0001-9974-0057
@@
   - name: Irene Alfarone
-    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/
+    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
     email: Irene.Alfarone@uibk.ac.at
@@
 
-```{r, echo = TRUE, results='asis'}
+```{r, echo=TRUE, results='asis'}
 # A custom function
@@
 # Mixture of (partially) named and positional arguments in unusual order
-term <- call("integrate", low=-Inf, up=Inf, g)
-# mathout(canonical(term))
+term = call("integrate", lower=-Inf, upper=Inf, g)
+attributes(term)$name = quote(g)
+mathout(canonical(term))
 
-# term <- quote(2L * integrate(low=-Inf, up=Inf, g)$value)
-# mathout(canonical(term))
+# Even more complex
+term <- call("*", 2L, call("$", call("integrate", low=-Inf, up=Inf, g), "value"))
+attributes(term)$name = quote(g)
+mathout(canonical(term))
 
 # It is a bit of a mystery that R knows the result of this integral.
-# eval(term)
+eval(term)
 ```

### 2022-12-05 inst/pl/mathml.pl

@@ -929,7 +929,7 @@ math(Flags, cal(A), New, M)
 % Integrate over range
 %
 % extract value
-math(Flags, $(integrate(Fn, Lower, Upper), value), New, M)
+math(Flags, $(integrate(Fn, Lower, Upper), "value"), New, M)
  => Flags = New,
     M = integrate(Fn, Lower, Upper).
 
@@ -938,6 +938,20 @@ math(Flags, integrate(f=Fn, lower=Lower, upper=Upper), New, M)
     M = integrate(Fn, Lower, Upper).
 
 % No argument names
+math(Flags, integrate(Fn, Lower, Upper), New, M),
+    Fn = (Head :- _Body),
+    member(name-Name, Flags)
+ => Flags = New,
+    compound_name_arguments(Head, _, [DX | _]),
+    M = integrate(fn(Name, [DX]), Lower, Upper, DX).
+
+math(Flags, integrate(Fn, Lower, Upper), New, M),
+    r_eval('is.null'('$'('.GlobalEnv', Fn)), false)
+ => Flags = New,
+%    r_eval('['(formalArgs(args('$'('.GlobalEnv', Fn))), 1), Arg1),
+    atom_string(DX, "Arg1"),
+    M = integrate(fn(Fn, [DX]), Lower, Upper, DX).
+
 math(Flags, integrate(Fn, Lower, Upper), New, M)
  => Flags = New,
     r_eval('['(formalArgs(args(Fn)), 1), Arg1),
@@ -953,7 +967,6 @@ ml(Flags, integrate(Fn, From, To, DX), M)
     ml(Flags, space, Space),
     M = mrow([munderover([mo(&(int)), XFrom, XTo]), XFn, Space, mi(d), XDX]).
 
-% Internal
 jax(Flags, integrate(Fn, From, To, DX), M)
  => jax(Flags, Fn, XFn),
     jax(Flags, From, XFrom),

### 2022-12-06 Draft2.Rmd
denote commented

@@
 
+## Abbreviations
+
+\pkg{mathml} allows also for the use of abbreviations in formulas. Let us consider here the t-statistic for independent samples with equal variance assumed. 
+The procedure consists of three steps.
+The first one is to define the hooks to nicely print the variables.
+Second, we write the t-statistic specifying in the formula how to calculate the pooled variance with the function  `denote(abbr, expr, info)`.
+Finally, since \pkg{mathml} also allows computation, we can directly compute the value of the t statistic with `eval()`.
+
+```{r, results='asis', echo=TRUE}
+
+hook(quote(x_1), quote(mean(x["1"])))
+hook(quote(x_2), quote(mean(x["2"])))
+hook(quote(n_1), quote(n["1"]))
+hook(quote(n_2), quote(n["2"]))
+hook(quote(s_1), quote(s["1"]))
+hook(quote(s_2), quote(s["2"]))
+hook(quote(s_p), quote(s["pool"]))
+
+t <- quote(dfrac(x_1 - x_2, sqrt(denote(s_p^2L, frac((n_1 - 1L)*s_1^2L + (n_2 - 1L)*s_2^2L, n_1+n_2-2L), "the pooled variance.") * (dfrac(1L, n_1)+dfrac(1L,n_2)))))
+
+mathout(t)
+
+# An example
+
+x_1 = 20.5
+x_2 = 35.7
+n_1 = 150
+n_2 = 160
+s_1 = 3.4
+s_2 = 4.2
+
+eval(t)
+```
+
 # A case study

### 2022-12-06 R/mathml.R

@@ -493,5 +493,23 @@ e1*e2
 '%dbldown%' <- function(x, y)
   x=y
 
+#' denote
+#' This is a function that allows the user to insert abbreviations in the formula,
+#' explain them and make the needed computations
+#'
+#' @param abbr
+#' Abbreviation used in the text to refer to the calculation, for example 's_p' for the pooled
+#' variance.
+#'
+#' @param expr
+#' Expression: calculations to be made in order to obtain the value to which the abbreviation
+#' refers to.
+#'
+#' @param info
+#' Information: Explanation of the formula used to provide the value of the abbreviation.
+#' e.g. 'the pooled variance'
+#'
+#' @return expr
+#'
 denote <- function(abbr, expr, info)
   return(expr)

### 2022-12-07 Draft2.Rmd
add mistakes

@@
 
-mathout <- function(term)
+mathout <- function(...)
 {
   if(knitr::is_html_output())
-    return(mathml(term))
+    return(mathml(...))
 
   if(knitr::is_latex_output())
-    return(mathjax(term))
+    return(mathjax(...))
 
   warning("no output specified")  
-  mathjax(term)
+  mathjax(...)
 }
 
-catmath <- function(term)
+catmath <- function(term, ...)
 {
   s <- substitute(term)
-  mathout(canonical(s))
+  mathout(canonical(s), ...)
 }
@@
 ```{r, results='asis', echo=TRUE}
+first <- quote(a)
+second <- quote(b + c)
+mathout(call("*", first, second))
+
 catmath(x^y + z)
+mathout(call("+", call("^", as.symbol("x"), as.symbol("y")), as.symbol("z")))
 
@@
 
-```{r, results="asis", echo=FALSE}
+```{r, results="asis"}
+t <- quote(omit_left(a + b))
+mathout(t)
+```
 
+```{r, results="asis"}
+t <- quote(omit_right(a + b))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- list(quote(a), quote(b), quote(omit(c)))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- quote(add_left(a + b))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- quote(add_right(a + b))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- list(quote(a), quote(b), quote(add(c)), quote(d))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- quote(instead(a, b) + c)
+mathout(t)
+```
+
+also mention the use of these flags for context-dependent rendering
+
+```{r, results="asis"}
+t <- quote(omit_left(a + b))
+mathout(t, flags=quote(error(highlight))) # default
+```
+
+```{r, results="asis"}
+mathout(t, flags=quote(error(ignore)))
+```
+
+```{r, results="asis"}
+mathout(t, flags=quote(error(fix)))
 ```

### 2022-12-07 R/mathml.R

@@ -12,6 +12,9 @@
 #' an R call or symbol/number. This function translates _term_ into a
 #' MathML string.
 #'
+#' @param flags (default NULL)
+#' list of flags that control the translation
+#'
 #' @return
 #' A string with the MathML representation or _term_.
 #'
@@ -22,9 +25,9 @@
 #' @examples
 #' mathml(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
 #'
-mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
+mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
 {
-  flags = attributes(term)
+  flags = c(attributes(term), flags)
   t = rolog::once(call("r2mathml", flags, term, expression(X)),
     options=list(preproc=mathml_preproc))
   cat(paste(t$X, collapse=""))
@@ -80,6 +83,9 @@ mathml_preproc <- function(query=quote(2 != 2))
 #' an R call or symbol/number. This function translates _term_ into a
 #' LaTeX/MathJax string.
 #'
+#' @param flags (default NULL)
+#' list of flags that control the translation
+#'
 #' @return
 #' A string with the MathJax representation or _term_.
 #'
@@ -90,9 +96,9 @@ mathml_preproc <- function(query=quote(2 != 2))
 #' @examples
 #' mathjax(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
 #'
-mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
+mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
 {
-  flags = attributes(term)
+  flags = c(attributes(term), flags)
   t = rolog::once(call("r2mathjax", flags, term, expression(X)),
     options=list(preproc=mathml_preproc))
   cat(paste(t$X, collapse=""))

### 2022-12-07 inst/pl/mathml.pl

@@ -2122,6 +2122,87 @@ math(Flags, omit_right(Expr), New, M),
  => Flags = New,
     M = list(space, [L, cancel(list(space, [op(Op), R]))]).
 
+math(Flags, omit(Expr), New, M),
+    option(error(ignore), Flags, highlight)
+ => Flags = New,
+    M = Expr.
+
+math(Flags, omit(Expr), New, M),
+    option(error(fix), Flags, highlight)
+ => Flags = New,
+    M = box(Expr).
+
+math(Flags, omit(Expr), New, M),
+    option(error(highlight), Flags, highlight)
+ => Flags = New,
+    M = cancel(Expr).
+
+math(Flags, add_left(Expr), New, M),
+    option(error(ignore), Flags, highlight),
+    Expr =.. [_Op, _L, R]
+ => Flags = New,
+    M = R.
+
+math(Flags, add_left(Expr), New, M),
+    option(error(fix), Flags, highlight),
+    Expr =.. [Op, L, R]
+ => Flags = New,
+    M = list(space, [cancel(list(space, [L, op(Op)])), R]).
+
+math(Flags, add_left(Expr), New, M),
+    option(error(highlight), Flags, highlight),
+    Expr =.. [Op, L, R]
+ => Flags = New,
+    M = list(space, [box(list(space, [L, op(Op)])), R]).
+
+math(Flags, add_right(Expr), New, M),
+    option(error(ignore), Flags, highlight),
+    Expr =.. [_Op, L, _R]
+ => Flags = New,
+    M = L.
+
+math(Flags, add_right(Expr), New, M),
+    option(error(fix), Flags, highlight),
+    Expr =.. [Op, L, R]
+ => Flags = New,
+    M = list(space, [L, cancel(list(space, [op(Op), R]))]).
+
+math(Flags, add_right(Expr), New, M),
+    option(error(highlight), Flags, highlight),
+    Expr =.. [Op, L, R]
+ => Flags = New,
+    M = list(space, [L, box(list(space, [op(Op), R]))]).
+
+math(Flags, omit(Expr), New, M),
+    option(error(ignore), Flags, highlight)
+ => Flags = New,
+    M = Expr.
+
+math(Flags, omit(Expr), New, M),
+    option(error(fix), Flags, highlight)
+ => Flags = New,
+    M = box(Expr).
+
+math(Flags, omit(Expr), New, M),
+    option(error(highlight), Flags, highlight)
+ => Flags = New,
+    M = cancel(Expr).
+
[…]

### 2022-12-07 Draft2.Rmd
add mistakes
errors

@@
 
-The hooks reported here will be the one used in Section @ref[] to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance
+The hooks reported here will be the one used in Section _ref_ to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance
 
@@
 ```{r, results="asis", echo=TRUE}
-catmath(Alpha-Omega)
+catmath(A-Omega)
 
@@
 
-At first glance, converting operators may seem trivial enough.
-However, when you want to deal with, for example, exponentiation, things can get more complicated.
-We see here that if we want `x` to the power of `y + z`, following R\ notation we should add a parenthesis around the two terms:
+At first glance, grouping expressions may seem trivial enough.
+However, when you want to deal with, for instance, exponentiation or multiplication, things can get more complicated.
+Let us consider a trivial example.
+If we want to calculate and evaluate the expression `a * (b + c)`, we find that R\ automatically gives precedence to the addition even if we do not specify it in the text, as in the following.
 
@@
 
+# precedence
+quote(a * (b + c))[[3]][[1]]
+```
+
+With exponentiation the situation can be less trivial.
+If we consider `x` to the power of `y + z`, following R\ notation we should add a parenthesis around the two terms.
+Otherwise R\ would consider only `y` as power of `x`.
+
+```{r, results='asis', echo=TRUE}
 catmath(x^y + z)
-mathout(call("+", call("^", as.symbol("x"), as.symbol("y")), as.symbol("z")))
+
+# mathout(call("+", call("^", as.symbol("x"), as.symbol("y")), as.symbol("z")))
 
@@
 
-As we can see, R\ automatically put `y + z` in parentheses even if, from a mathematical point of view, the brackets are not useful.
+Unfortunately, R\ automatically put `y + z` in parentheses even if, from a mathematical point of view, the brackets are not needed.
 To solve this issue two solutions are possible, the first one is cumbersome since it requires many specifications.
-The second, the shortcut, uses a syntax similar to LateX and puts the power in curly brackets.
+The second, that can be considered a shortcut, uses a syntax similar to LateX and puts the power in curly brackets.
 In both cases the output produced is consistent with the mathematical notation.
@@
 
-blabla `quote(a * (b + c))[[3]][[1]]`
 ## Absolute value
@@
 mathout(t)
+
+catmath(omit_left(a + b))
 ```
@@
 mathout(t)
+
+catmath(omit_right(a + b))
 ```
@@
 mathout(t)
+
+catmath(add_left(a + b))
 ```
@@
 mathout(t)
+
+catmath(add_right(a + b))
+```
+
+```{r, results="asis"}
+t <- list(quote(a), quote(b), quote(add(c)), quote(d))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- quote(add_right(a + b))
+mathout(t)
+
+catmath(add_right(a + b))
 ```
@@
 mathout(t)
+
+catmath(instead(a, b) + c)
 ```
 
-also mention the use of these flags for context-dependent rendering
+It is also possible to use flags for context-dependent rendering.
+The highlight flag is the default one.
+It shows where the error is and what is its solution.
+The ignore flag allows the program to ignore the fact that there is an error and print the expression as if it were correct.
+Finally, the third flag, fix, tells you what is the term that you are supposed to fix without giving the solution.
+Below the code with its comment is provided.
 
 ```{r, results="asis"}
+# It highlights what has been omitted
 t <- quote(omit_left(a + b))
@@
 ```{r, results="asis"}
+# It prints the correct version
 mathout(t, flags=quote(error(ignore)))
@@
 ```{r, results="asis"}
+# It highlights the correct version by framing it
+mathout(t, flags=quote(error(fix)))
+```
+
+```{r, results="asis"}
+# It frames the redundant term
+t <- quote(add_right(a + b))
+mathout(t, flags=quote(error(highlight))) # default
+```
+
+```{r, results="asis"}
+# It prints the correct expression without the redundant term
+mathout(t, flags=quote(error(ignore)))
+```
+
+```{r, results="asis"}
+# It strikes the redundant term 
+mathout(t, flags=quote(error(fix)))
+```
+
+```{r, results="asis"}
+# It highlights that a is wrong and b is correct
+t <- quote(instead(a, b) + c)
+mathout(t, flags=quote(error(highlight))) # default
+```
+
+```{r, results="asis"}
+# It provides the correct solution
+mathout(t, flags=quote(error(ignore)))
+```
+
+```{r, results="asis"}
+# It provides the correct solution framing the correct term
 mathout(t, flags=quote(error(fix)))

### 2022-12-07 R/mathml.R

@@ -519,3 +519,84 @@ e1*e2
 #'
 denote <- function(abbr, expr, info)
   return(expr)
+
+#' omit_left
+#'
+#' @param expr
+#'
+#' @eval  (substitute(expr)[[3]])
+#'
+omit_left <- function(expr)
+{
+# use third element of [-, A, B]
+  eval(substitute(expr)[[3]])
+}
+
+#' omit_right
+#'
+#' @param expr
+#'
+#' @eval  (substitute(expr)[[3]])
+#'
+omit_right <-  function(expr)
+{
+  eval(substitute(expr)[[2]])
+}
+
+#' omit
+#'
+#' @param expr
+#'
+#' @eval  (substitute(expr))
+#'
+omit<-  function(expr)
+{
+  eval(substitute(expr))
+}
+
+
+#' add_left
+#'
+#' @param expr
+#'
+#' @return expr
+#'
+add_left <- function(expr)
+{
+  return(expr)
+}
+
+#' add_right
+#'
+#' @param expr
+#'
+#' @return expr
+#'
+add_right <- function(expr)
+{
+  return(expr)
+}
+
+#' add
+#'
+#' @param expr
+#'
+#' @return expr
+#'
+add <- function(expr)
+{
+  return(expr)
+}
+
+#' instead
+#'
+#' @param inst
+#'
+#' @param of
+#'
+#' @return inst
+#'
+instead <- function(inst, of)
+{
+  return(inst)
+}

### 2022-12-09 Draft2.Rmd
intro

@@
 title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
-date: "2022-11-09"
+date: "2022-12-09"
 abstract: >
-  This R package translates R objects to suitable entities in MathML or LaTeX,
-  thereby enabling pretty mathematical representation of R functions or
+  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
+  thereby enabling pretty mathematical representation of R\ functions or
   algorithms in data analysis, scientific reports and interactive web content.
-  R code text descriptions and mathematical content already coexist,
+  R\ code text descriptions and mathematical content already coexist,
   side-by-side, in the RMarkdown document rendering language. The present
@@
   typesetting in documents or web content. By tightening the link between the
-  statistical analysis and its verbal or symbolic description, this new feature
-  is a futher step towards reproducible science. The package can be extended by
-  user-defined hooks that allow to replace variable names and new R functions by
-  a suitable symbolic representation. We provide two working examples that
-  illustrate the use of the new package in a scientific article and in an
+  statistical analysis and its verbal or symbolic representation, this new
+  feature is a futher step towards reproducible science. User-defined hooks
+  allow to extend the package by mapping new variable names and functions to
+  new MathML and LaTeX entities. We provide a few working examples that
+  illustrate the use of the package in a little scientific report and in an
   interactive web page.
@@
 bibliography: RJreferences.bib
-
 ---
@@
 ```{r setup, include=FALSE}
-# knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+# To do
+#
+# paste(x, y, z)
+# Spacing
+# Comma separated lists
+# Ellipsis
+# Radicals
+# Subsets
+# Greek letters in MathJax
+# Typeface
+# Stylechanges
+# Phantoms
+
 library(plotly)
 library(ggplot2)
-library(palmerpenguins)
 
@@
 
-#### To do
-
-* paste(x, y, z)
-* Spacing
-* Comma separated lists
-* Ellipsis
-* Radicals
-* Subsets
-* Greek letters in MathJax
-* Typeface
-* Stylechanges
-* Phantoms
-
 # Introduction
@@
 statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
-However, so far, R expressions such as `pbinom(k, N, p)` are printed as-is; 
-prettier mathematical expressions such as $P_\mathrm{Bi}(X \le k; N, p)$ require
-Latex commands like `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for
-minimalistic examples, these commands are tedious to type in and hard to read in
-the original code. Since recently [@R, version 4.2], manual pages include
-support for mathematical expressions [@Sarkar2022], which already is a big
-improvement. However, rules for the translation of R's built-in language elements to
-their mathematical representation are still lacking.
+Since recently [@R, version 4.2], R's manual pages include support for
+mathematical expressions [@Sarkar2022], which already is a big improvement.
+However, rules for the translation of R's built-in language elements to their
+mathematical representation are still lacking. So far, R expressions such
+as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
+like \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
+is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
+these commands are tedious to type in and hard to read in their source code.
 
-In this R package, such rules are defined for the _automatic_ translation of R
-expressions mathematical output. The translation is done by an embedded
-Prolog interpreter that maps nested expressions recursively to MathML and
-LaTeX/MathJax, respectively, for typesetting of R Markdown documents in
-different formats or ShinyApp webpages. User-defined hooks enable to extend the
-set of rules and/or replace specific R elements by special mathematical signs.
+The present R package defines a set of rules for an _automatic_ translation of
+R expressions to mathematical output in RMarkdown documents [@Xie2020] and
+ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
+interpreter that maps nested expressions recursively to MathML and
+LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
+rules, for example, to represent specific R elements by custom mathematical
+signs.
 
-The paper is organized as follow.
-Section 2 presents the background of the mathml package, namely Prolog and the R package \pkg{rolog}.
-Section 3 illustrates the features of the \pkg{mathml} package, providing some examples.
-Section 4 embeds the Discussion and the presentation of a Case Study.
-Section 5 is devoted to Conclusions and Further Research.
-The reader may also find some additional characteristics of the package and its implementation in the Appendix.
+*todo*: mention that the _same_ R expression is used for mathematical
+calculations and printing.
+
+The paper is organized as follows. Section\ 2 presents the background of the
+package, including Prolog and the R package \pkg{rolog} and the two main classes
+of rules for translating R\ objects to mathematical expressions. Section\ 3
+illustrates some important features of the \pkg{mathml} package, potential
+issues and workarounds using examples from the day-to-day perspective of the
+user. Sections\ 4 and\ 5 present two case studies for the use of the package in
+a scientific report and a ShinyApp web page. Section\ 6 concludes with a
+discussion and ideas for further development. A complete list of the R elements
+covered by the package is found in the Appendix.
 

### 2022-12-09 Draft2.Rmd
few comments

--- 2022-12-07 Draft2.Rmd
@@
 
+In this first example the function highlights that the student forgot to add the left-hand side of the expression and it does that by striking the missing part.
+In this case the letter A and the operator `+`.
+Similarly the function `omit_right` does the same for the omissions in the right-hand side of the expression.
+Finally the function `omit()` can be used, for instance, to highlight the omission in the elements of a list.
+
 ```{r, results="asis"}
@@
 mathout(t)
-
-catmath(omit_left(a + b))
 ```
@@
 mathout(t)
-
-catmath(omit_right(a + b))
 ```
@@
 
+The functions `add_left()`, `add_right()`, `add()` reported below follow the same way of reasoning, but they account for the mistakes made in adding incorrect elements in the expression.
+They highlight the redundant element placing it in a box.
+
 ```{r, results="asis"}
@@
 mathout(t)
-
-catmath(add_left(a + b))
 ```
@@
 mathout(t)
-
-catmath(add_right(a + b))
 ```
@@
 
-```{r, results="asis"}
-t <- quote(add_right(a + b))
-mathout(t)
-
-catmath(add_right(a + b))
-```
-
-```{r, results="asis"}
-t <- list(quote(a), quote(b), quote(add(c)), quote(d))
-mathout(t)
-```
+Finally the function `instead( , )` place a curly bracket under the incorrect element  and provides its correct version under the curly bracket.
+In order to provide the correction with this function one must fill it with its wrong and correct terms in this order: `instead(wrong, correct)`.
+An example of its use its provided below.
 
@@
 mathout(t)
-
-catmath(instead(a, b) + c)
 ```

### 2022-12-09 Draft2.Rmd

@@
+<<<<<<< HEAD
 ---
@@
 # Acknowledgments
+
+=======
+---
+title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
+date: "2022-12-09"
+abstract: >
+  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
+  thereby enabling pretty mathematical representation of R\ functions or
+  algorithms in data analysis, scientific reports and interactive web content.
+  R\ code text descriptions and mathematical content already coexist,
+  side-by-side, in the RMarkdown document rendering language. The present
+  package allows to use the same R\ objects for both data analysis and
+  typesetting in documents or web content. By tightening the link between the
+  statistical analysis and its verbal or symbolic representation, this new
+  feature is a futher step towards reproducible science. User-defined hooks
+  allow to extend the package by mapping new variable names and functions to
+  new MathML and LaTeX entities. We provide a few working examples that
+  illustrate the use of the package in a little scientific report and in an
+  interactive web page.
+draft: true
+author:  
+  # see ?rjournal_article for more information
+  - name: Matthias Gondan
+    affiliation: University of Innsbruck
+    address:
+    - Department of Psychology
+    - Innsbruck, Austria
+    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
+    orcid: 0000-0001-9974-0057
+    email:  Matthias.Gondan-Rochon@uibk.ac.at
+  - name: Irene Alfarone
+    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
+    email: Irene.Alfarone@uibk.ac.at
+    orcid: 0000-0002-8409-8900
+    affiliation: University of Innsbruck
+    address:
+    - Department of Psychology
+    - Innsbruck, Austria
+type: package
+output: 
+  rjtools::rjournal_web_article:
+    self_contained: yes
+    toc: no
+bibliography: RJreferences.bib
+---
+
+```{r setup, include=FALSE}
+# To do
+#
+# paste(x, y, z)
+# Spacing
+# Comma separated lists
+# Ellipsis
+# Radicals
+# Subsets
+# Greek letters in MathJax
+# Typeface
+# Stylechanges
+# Phantoms
+
+library(plotly)
+library(ggplot2)
+
+mathout <- function(...)
+{
+  if(knitr::is_html_output())
+    return(mathml(...))
+
+  if(knitr::is_latex_output())
+    return(mathjax(...))
+
+  warning("no output specified")  
+  mathjax(...)
+}
+
+catmath <- function(term, ...)
+{
+  s <- substitute(term)
+  mathout(canonical(s), ...)
+}
+```
+
+# Introduction
+
+The R extension of the markdown language [@Xie2020] enables reproducible
+statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
+Since recently [@R, version 4.2], R's manual pages include support for
+mathematical expressions [@Sarkar2022], which already is a big improvement.
+However, rules for the translation of R's built-in language elements to their
+mathematical representation are still lacking. So far, R expressions such
+as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
+like \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
+is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
+these commands are tedious to type in and hard to read in their source code.
+
+The present R package defines a set of rules for an _automatic_ translation of
+R expressions to mathematical output in RMarkdown documents [@Xie2020] and
+ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
+interpreter that maps nested expressions recursively to MathML and
+LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
+rules, for example, to represent specific R elements by custom mathematical
+signs.
+
+*todo*: mention that the _same_ R expression is used for mathematical
+calculations and printing.
+
+The paper is organized as follows. Section\ 2 presents the background of the
+package, including Prolog and the R package \pkg{rolog} and the two main classes
+of rules for translating R\ objects to mathematical expressions. Section\ 3
+illustrates some important features of the \pkg{mathml} package, potential
+issues and workarounds using examples from the day-to-day perspective of the
+user. Sections\ 4 and\ 5 present two case studies for the use of the package in
+a scientific report and a ShinyApp web page. Section\ 6 concludes with a
+discussion and ideas for further development. A complete list of the R elements
+covered by the package is found in the Appendix.
+
+# Background
+
+## Prolog: the \pkg{rolog} package
+
+The automatic translation of R expressions to mathematical output is achieved through a Prolog interpreter.
+Prolog is a classical logic programming language with many applications in expert systems, computer linguistics and traditional, that is, symbolic artificial intelligence.
+The main strength of Prolog is its concise representation of facts and rules for the representation of knowledge and grammar, as well as its efficient built-in search engine for closed world domains.
+R, as it is well-known, is a statistical programming language for data analysis and statistical modeling which is widely used in academia and industry.
+Besides the core library, a lot of packages have been developed for all kinds of statistical problems, including statistics-based artificial intelligence tools such as neural networks for machine learning and deep learning.
+Whereas Prolog is weak in statistical computation, but strong in symbolic manipulation, the converse may be said for the R language.
+The \pkg{rolog} package then bridges this gap by providing an interface to a SWI-Prolog distribution in R.
+The communication between the two systems is mainly in the form of queries from R to Prolog, but two predicates allow Prolog to ring back and evaluate terms in R.
+Here is reported a first trivial example. At the Prolog end, there is a handler for `pbinom/3` that translates the term into a pretty MathML syntax like P_bi(X <= k; N, pi).
+
+```{r, echo=TRUE, results='asis'}
+library(rolog)
+library(mathml)
+term = quote(pbinom(k, N, p))
+
+# Pretty print
+mathout(term)
+```
+
+Beside the graphical rendition of the term, an important feature is that the package allows also for its calculation, for example:
+
+```{r, results = 'asis', echo = TRUE}
+k = 10
+N = 22
+p = 0.4
+eval(term)
+```
+
+Here we present a second less trivial example where Prolog needs to find out the name of
+the integration variable for `sin`.
+For that purpose, \pkg{rolog} provides a predicate `r_eval/2` that calls R from Prolog.
+Here, the predicate is used for the  R\ function `formalArgs(args(sin))`, which returns the name of the function argument of `sin`, that is, `x`.
+
+```{r,  results = 'asis', echo = TRUE}
+term = quote(integrate(sin, 0L, 2L*pi))
+mathout(term)
+```
+
+```{r, echo = TRUE}
+eval(term)
+```
+
+Note that, on the Prolog end, the handler for `integrate/3` is rather rigid; it
+accepts only these three arguments in that particular order, and without names,
+that is, `integrate(sin, lower=0L, upper=2L * pi)`, would not print the desired
+result.
+
+Therefore, `pl/mathml.pl` includes two handlers that accept terms with named
+arguments, `integrate(f=Fn, lower=Lower, upper=Upper)`, as well as terms of the
+form `$(integrate(Fn, Lower, Upper), value)` that are needed for the evaluation
+below.
+
+```{r, echo=TRUE, results='asis'}
+# A custom function
+g <- function(u)
+{
+  sin(u)
+}
+
+# Mixture of (partially) named and positional arguments in unusual order
+term = call("integrate", lower=-Inf, upper=Inf, g)
+attributes(term)$name = quote(g)
+mathout(canonical(term))
+
+# Even more complex
+term <- call("*", 2L, call("$", call("integrate", low=-Inf, up=Inf, g), "value"))
+attributes(term)$name = quote(g)
+mathout(canonical(term))
+
+# It is a bit of a mystery that R knows the result of this integral.
+eval(term)
+```
+
+The extra R function `canonical()` applies `match.call()` to non-primitive R
+calls, basically cleaning up the arguments and bringing them into the correct
+order.
+
+# Package capabilities
+
+\pkg{mathml} is a package for pretty mathematical representation of R functions or algorithms in data analysis, scientific reports and interactive web content.
+The features currently supported are listed below, following the order proposed by [@murrell2000].
+
+## Hooks
+Before start listing all the possible operations with \pkg{mathml},  we mention here that \pkg{mathml} also allows the user to create hooks for specific functions. As an example:
+
+```{r, echo=TRUE}
+# Improve display
+hook(quote(mu_A), quote(mu["A"]))
+hook(quote(mu_B), quote(mu["B"]))
+hook(quote(sigma_A), quote(sigma["A"]))
+hook(quote(sigma_B), quote(sigma["B"]))
+hook(quote(mu_M), quote(mu["M"]))
+hook(quote(zn), quote(z[tau]))
+hook(quote(M), quote(overline(X)))
+```
+
+The hooks reported here will be the one used in Section _ref_ to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance
+
+## Arithmetic operators
+The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division. For instance:
+
+```{r, results="asis", echo=TRUE}
+catmath(x + y - z)
+```
+
+In addition, with \pkg{mathml} it is also possible to combine `minus` and `plus` and get a new operator.
+However, in R\ the operator \(\pm\) is not present and this leads to some interesting questions.
+If we let R\ present the expression, this is what we get:
+
+```{r, results='asis', echo=TRUE}
+# Note wrong precedence of custom operator
+catmath(mean(m) %+-% 1.96 * s/sqrt(N))
+```
+
+As can be seen, the new custom operator takes precedence over multiplication, as R\ automatically puts `mean(m) %+-% 1.96` in parentheses.
+Unfortunately, this is not consistent with the rules of mathematics.
+As the reader well knows, it is not possible in R\ to change the precedence of custom operators.
+This leaves two possible solutions.
+The first involves a mathematically unnecessary parenthesis around `(1.96 * s/sqrt(N))`.
+The second is more elegant and mathematically consistent: here `1.96 * s/sqrt(N)` is put in curly brackets and mathematical formality is respected.
+In both solutions R\ is able to produce correct output once the order of operations is made explicit.
+
+```{r, results='asis', echo=TRUE}
+# This is what we actually want
+catmath(mean(m) %+-% (1.96 * s/sqrt(N)))
+
+# Without parentheses
+catmath(mean(m) %+-% {1.96 * s/sqrt(N)})
+```
+
+_The following can be inserted in a table_
+
+```{r, results='asis', echo=TRUE}
+
+catmath(x*y)
+
+catmath(x %*% y)
+
+catmath(x %.% y)
+
+catmath(X %x% Y)
+
+catmath(x/y)
+
+catmath(x %/% y)
+```
+
+## Subscripts and Superscripts
+Secondly, \pkg{mathml} allows also for the display of subscripts, superscripts and their combination.
+
+```{r, results="asis", echo=TRUE}
+catmath(a["x"])
+
+catmath(a^x)
+
+catmath(x["b"]^"a")
+```
+
+## Fractions
+As in [@murrell2000] the `frac` and `over` functions aim to represent fractions. `choose` and `lchoose`, as in R\ \pkg{base}, display the binomial coefficient.
+
+```{r, results="asis", echo=TRUE}
+catmath(frac(x, y))
+
+catmath(over(x, y))
+
+catmath(choose(n, k))
+
+catmath(lchoose(n, k))
+```
+
+## Radicals
+
+```{r, results="asis", echo=TRUE}
+catmath(sqrt(x))
+```
+
+## Relations
+
+Among the possible relational operators, following [@murrell2000], we included:
+
+```{r, results="asis", echo=TRUE}
+catmath(x==y)
+
+catmath(x < y)
+
+catmath(x <= y)
+
+catmath(x >= y)
+
+catmath(!x)
+
+catmath(x != y)
+
+catmath(X ~ Y)
+
+catmath(x %~~% y)
+
+catmath(x %==% y)
+
+catmath(x %=~% y)
+
+catmath(x %prop% y)
+```
+
+## Set relations
+
+Here is provided a series of useful set relations.
+
+_Still to be implemented_
+
+```{r, results="asis", echo=TRUE}
+catmath(a %in% A)
+
+catmath(intersect(A, B))
+
+catmath(union(A, B))
+
+catmath(is.null(intersect(A, B)))
+```
+
+## Arrows
+
+Following once again [@murrell2000], we provide here a list of useful arrows
+
+```{r, results="asis", echo=TRUE}
+# x %<->% y
+catmath(x %<->% y)
+
+# x %->% y
+catmath(x %->% y)
+
+# x %<-% y
+catmath(x %<-% y)
+
+# x %up% y:
+catmath(x %up% y)
+
+# x %down% y:
+catmath(x %down% y)
+
+# x %<=>% y: solved it using iff
+catmath(x %<=>% y)
+
+# x %=>% y: solved using rArr
+catmath(x %=>% y)
+
+# x %<=% y: solved using lArr
+catmath(x %<=% y)
+
+# x %dblup% y:
+catmath(x %dblup% y)
+
+# x %dbldown% y:
+catmath(x %dbldown% y)
+```
+
+## Accents
+
+```{r, results="asis", echo=TRUE}
+catmath(hat(x))
+
+catmath(hat(xy))
+
+catmath(mean(x))
+
+catmath(mean(xy))
+```
+
+## Symbolic names
+
+```{r, results="asis", echo=TRUE}
+catmath(A-Omega)
+
+catmath(alpha-omega)
+
+catmath(pi-Pi)
+
+catmath(Inf)
+```
+
+## Operators
+
+A list of possible operators is available here.
+
+```{r, results='asis', echo=TRUE}
+catmath(sum(x[i], i==1L, n))
+
+# fixme: braces not needed when right to *
+catmath(a * sum(b))
+
+# Multiplication (capital Pi) is missing
+catmath(a * prod(b))
+
+catmath(integrate(sin, 0L, 2L*pi))
+
+catmath(A & B)
+
+catmath(A | B)
+
+catmath(!A)
+
+catmath(xor(A, B))
+
+catmath(exp(2L*pi*i))
+
+catmath(expm1(2L*pi*i))
+
+catmath(log(x))
+
+catmath(log10(x))
+
+catmath(log2(x))
+
+catmath(logb(x, e))
+
+catmath(log1p(x))
+
+catmath(ceiling(pi))
+
+catmath(floor(pi))
+
+catmath(sin(pi/2L))
+
+catmath(cos(pi/2L))
+
+catmath(tan(pi/2L))
+```
+
+## Grouping
+
+At first glance, grouping expressions may seem trivial enough.
+However, when you want to deal with, for instance, exponentiation or multiplication, things can get more complicated.
+Let us consider a trivial example.
+If we want to calculate and evaluate the expression `a * (b + c)`, we find that R\ automatically gives precedence to the addition even if we do not specify it in the text, as in the following.
+
+```{r, results='asis', echo=TRUE}
+first <- quote(a)
+second <- quote(b + c)
+mathout(call("*", first, second))
+
+# precedence
+quote(a * (b + c))[[3]][[1]]
+```
+
+With exponentiation the situation can be less trivial.
+If we consider `x` to the power of `y + z`, following R\ notation we should add a parenthesis around the two terms.
+Otherwise R\ would consider only `y` as power of `x`.
+
+```{r, results='asis', echo=TRUE}
+catmath(x^y + z)
+
+# mathout(call("+", call("^", as.symbol("x"), as.symbol("y")), as.symbol("z")))
+
+catmath(x^(y + z))
+```
+
+Unfortunately, R\ automatically put `y + z` in parentheses even if, from a mathematical point of view, the brackets are not needed.
+To solve this issue two solutions are possible, the first one is cumbersome since it requires many specifications.
+The second, that can be considered a shortcut, uses a syntax similar to LateX and puts the power in curly brackets.
+In both cases the output produced is consistent with the mathematical notation.
+
+```{r, results='asis', echo=TRUE}
+# without parenthesis
+mathout(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
+
+# shortcut
+catmath(x^{y + z})
+```
+
+## Absolute value
+
+Is it also possible to calculate and represent the absolute value, as:
+
+```{r, results='asis', echo=TRUE}
+catmath(abs(x))
+```
+
+## Matrix and Vectors
+
+Since also Matrices and Vectors are important elements in R\, we decided to implement them and render them graphically in the following ways.
+
+```{r, results='asis', echo=TRUE}
+vec <- 1:3
+mathout(canonical(vec))
+mathout(vec)
+
+mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
+mathout(canonical(mat))
+```
+
+## Abbreviations
+
+\pkg{mathml} allows also for the use of abbreviations in formulas. Let us consider here the t-statistic for independent samples with equal variance assumed. 
+The procedure consists of three steps.
+The first one is to define the hooks to nicely print the variables.
+Second, we write the t-statistic specifying in the formula how to calculate the pooled variance with the function  `denote(abbr, expr, info)`.
+Finally, since \pkg{mathml} also allows computation, we can directly compute the value of the t statistic with `eval()`.
+
+```{r, results='asis', echo=TRUE}
+
+hook(quote(x_1), quote(mean(x["1"])))
+hook(quote(x_2), quote(mean(x["2"])))
+hook(quote(n_1), quote(n["1"]))
+hook(quote(n_2), quote(n["2"]))
+hook(quote(s_1), quote(s["1"]))
+hook(quote(s_2), quote(s["2"]))
+hook(quote(s_p), quote(s["pool"]))
+
+t <- quote(dfrac(x_1 - x_2, sqrt(denote(s_p^2L, frac((n_1 - 1L)*s_1^2L + (n_2 - 1L)*s_2^2L, n_1+n_2-2L), "the pooled variance.") * (dfrac(1L, n_1)+dfrac(1L,n_2)))))
+
+mathout(t)
+
+# An example
+
+x_1 = 20.5
+x_2 = 35.7
+n_1 = 150
+n_2 = 160
+s_1 = 3.4
+s_2 = 4.2
+
+eval(t)
+```
+
+# A case study
+
+For a more clear representation of what the \pkg{mathml} package is supposed to do, we provide here a case study, based on the work by Schwarz [-@schwarz1994], in which the features of \pkg{mathml} are implemented.
+
+The reader may recognize that Schwarz' [-@schwarz1994] study can be traced back to the strand of decision-making models, where decision-making is assumed to be a process of
+noisy accumulation of information over time [e.g., @ratcliff2016]. The aim of
+Schwarz' [-@schwarz1994] study is to present a new explanation of redundancy gains when
+observers make speeded responses to stimuli of different sources, and the same
+information is presented on two channels. Schwarz [-@schwarz1994] describes a model that assumes
+the superposition of channel-specific diffusion processes that eventually reach
+an absorbing barrier to elicit the response. For a detailed description the
+reader may refer to the original article.
+
+### The model
+
+Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
+alone or in combination (redundant stimuli), with the redundant stimuli being
+presented either simultaneously or with onset asynchrony \(\tau\). The channel
+activation is described as a two-dimensional Wiener process with
+drifts \(\mu_\mathrm A\), \(\mu_\mathrm B\),
+variances \(\sigma^2_\mathrm A\), \(\sigma^2_\mathrm B\),
+correlation \(\rho_\mathrm{AB}\) and initial
+conditions \(X_i(t=0)=0, i=\mathrm{A, B}\). The response is elicited when the
+process reaches the absorbing barrier \(c > 0\) for the first time. In combined
+stimuli ("redundant" stimuli), the overall diffusion process
+is \(X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)\) , which is again a
+Wiener process with drift \(\mu_\mathrm A + \mu_\mathrm B\) and
+variance \(\sigma^2_\mathrm A + \sigma^2_\mathrm B
+  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B\).
+
+In single-target trials, the first passages at \(c\) are expected at
+\(E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}\), whereas in synchronous
+redundant-target trials, we have
+\(E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}\).
+
+For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
+first-passage time \(E[D_{\mathrm A(\tau)\mathrm B}]\) as a function of the
+stimulus onset asyncrony \(\tau > 0\) (Eq.\ 10),
+
+```{r, results="asis", echo=TRUE}
+f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
+    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
+      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
+        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
+}
+
+mathout(f)
+```
+
+The observable response time is assumed to be the sum of \(D\), the time
+employed to reach the threshold for the decision, and \(M\), denoting the of the
+other processes.
+
+\(E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M\)
+
+### Methods
+
+Schwarz [-@schwarz1994] applied the model to Miller’s [-@miller1986] data from
+a redundant signals task with 13 onset
+asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
+where 0 refers to synchronous AB, \(\infty\) refers to the single-target
+presentation, and negative \(\tau\) denote those conditions in which B is
+presented before A. Each condition was replicated 400 times. The 13 mean RTs and
+standard deviations are given in Table 2 below.
+
+Then, from Schwarz' [-@schwarz1994] Equations [2] [3] and [10], we calculated the expected first time passage for the respective \(\tau\) and the parameters \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)
+
+````{r, echo=FALSE}
+mu_A = 0.53
+mu_B = 1.32
+sigma_A = 4.4
+sigma_B = 18.3
+tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
+c = 100
+mu_M = 161
+````
+
+```{r, results="asis", echo=FALSE}
+# Conditions
+
+g <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+{
+    if( tau == Inf) return(dfrac(c, mu_A));
+    if( tau == -Inf) return(dfrac(c, mu_B));
+    if( tau > 0L)  return(f(tau, c, mu_A, sigma_A, mu_B, sigma_B));
+    if( tau == 0L) return(dfrac(c, mu_A + mu_B));
+    if( tau < 0L) return(f(-tau, c,  mu_B, sigma_B, mu_A, sigma_A))
+}
+
+mathout(g)
+```
+
+For instance, we can see that with \(\tau = -167\) the expected first passage time is:
+
+```{r, echo=FALSE}
+mu_A = 0.53
+mu_B = 1.32
+sigma_A = 4.4
+sigma_B = 18.3
+tau = -167
+c = 100
+mu_M = 161
+
+g(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+```
+
+To which we have to add \(\mu_M = 161\) (which represents the amount of time that the movement of providing the answer takes) and we obtain:
+
+```{r, echo=FALSE}
+g(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
+```
+
+Similarly, we obtain the first-time passage for every \(\tau\):
+
+```{r, results="asis", echo=FALSE}
+
+G <- Vectorize(g, vectorize.args = 'tau')
+
+h <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+{
+  G(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
+}
+
+mathout(h)
+```
+
+```{r, echo=FALSE}
+
+mu_M = 161
+mu_A = 0.53
+mu_B = 1.32
+sigma_A = 4.4
+sigma_B = 18.3
+tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
+c = 100
+
+h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+```
+
+### Model fitting and Parameters Estimation
+
+Then we proceeded with the model fitting, using the data provided by Miller [-@miller1986] study in Schwarz [-@schwarz1994]. First, we calculated the predicted means RT as \(E[RT]=E[T]+\mu_M\).
+
+We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed [@ratcliff2008] with at the first passage of \(c\) \(X(t) \sim N(\mu t , \sigma^2 t)\).
+Where \(T\) is expressed as \(T= \arg \min_t X(t)\geq c\) with \(T \sim IG (c, \mu, \sigma^2)\) with mean \(E(T)=\displaystyle \frac{c}{\mu}\) and variance \(Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}\).
+Furthermore, for the Central Limit Theorem \(\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})\)
+
+For these reasons, in order to obtain the Goodness of Fit measure we proceeded with a z-standardisation.
+
+```{r, results="asis", echo=FALSE}
+zf <-  function(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+{
+  E <- h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+  dfrac(M-E, s/sqrt(N))
+}
+
+mathout(zf)
+```
+
+```{r, echo=FALSE}
+# The vector X contains the observed values, the vector s their observed standard deviation and N = 400 are the number of trials.
+
+M = c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
+s = c(56, 58, 40, 40, 32, 28, 28, 28, 26, 30, 32, 34, 92)
+N = c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400)
+
+zn <- zf(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+zn
+```
+
+Then, we calculated the Goodness of Fit as follows:
+
+\(X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)\).
+
+```{r, results="asis", echo=FALSE}
+par <- c(mu_A = 0.53, sigma_A = 4.4, mu_B = 1.32, sigma_B = 18.3, mu_M = 161)
+
+# Goodness of Fit
+
+gof <- function(par, tau, M, s, N)
+{
+  zn <- zf(M, s, N, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])
+  sum(zn^2L)
+}
+
+mathout(gof)
+```
+
+```{r, results="asis", echo=FALSE}
+n <- gof(par, tau, M, s, N)
+n
+pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)
+```
+
+```{r, results="asis", echo=FALSE}
+lsq = function(tau, M, s, N)
+{
+  start = c(mu_A = 0.3, sigma_A = 2, mu_B = 1 , sigma_B = 5, mu_M = 200)
+  optim(start, fn = gof,
+    tau = tau, M = M, s = s, N = N, hessian = TRUE, control = list(fnscale = 1))
+}
+
+mathout(lsq)
+
+# Parameters' estimates
+fit <-  lsq(tau, M, s, N)
+
+est = fit$par
+se = sqrt(diag(solve(fit$hessian)))
+cilo = qnorm(0.025, est, se)
+ciup = qnorm(0.975, est, se)
+
+# Lower bound
+#cilo
+
+# Upper bound
+#ciup
+
+```
+
+### Results
+
+The calculated Goodness of Fit value is \(X^2= 30.54\), \(p= 0.0002\). Additionally, we proceeded with the parameters' estimation with the Least Squares Method.
+
+The values of the estimated parameters reported in the table below are consistent with those reported by Schwarz [-@schwarz1994].
+
+: Table 1: Estimated Model's Parameters
+
+| Parameters | Estimate | 95% Confidence Interval
+| :---: | :---: | :---:|
+| \( \mu_ \mathrm A \)| 0.53 | 0.51, 0.56
+| \( \sigma_ \mathrm A \)| 4.4 | 3.8, 5.1
+| \( \mu_ \mathrm B \)| 1.32 | 1.2, 1.4
+| \( \sigma_ \mathrm B \)| 18.3 | 11.6, 24.9
+| \( \mu_ \mathrm M \)| 161 | 157, 165
+
+Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i.
+\(\tau= 100\)) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model.
+Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.
+
+: Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's [-@miller1986] study in Schwarz [-@schwarz1994]
+
+| SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
+|:--------:|:-----------------:|:----------------:|:-----------:|
+\(-\infty\)| 237 | 231 | 56 |
+\(-167\)   | 231 | 234 | 58 |
+\(-133\)   | 230 | 230 | 40 |
+\(-100\)   | 228 | 227 | 40 |
+\(-67\)    | 226 | 228 | 32 |
+\(-33\)    | 222 | 221 | 28 |
+\(0\)      | 215 | 217 | 28 |
+\(33\)     | 239 | 238 | 28 |
+\(67\)     | 262 | 263 | 26 |
+\(100\)    | 283 | 277 | 30 |
+\(133\)    | 300 | 298 | 32 |
+\(167\)    | 313 | 316 | 34 |
+\(\infty\) | 350 | 348 | 92 |
+
+ _Note_.  Parameters: \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)
+
+## Mistakes
+
+The case study presented above displays several possible ways of how \pkg{mathml} can be a useful tool in data representation and calculation.
+Furthermore, the careful reader might have also noticed that \pkg{mathml} allows also for a more smoothly detection of possible mistakes in writing.
+
+An interesting use case of the package is in this sense the readability of self-written
+functions in R code. R code is notoriously hard to read, as are most programming
+languages [@green1977]. The poor legibility of the language is one of the main
+sources of mistakes. For illustration, we consider Equation 10 in
+Schwarz [-@schwarz1994], namely the same equation reported above. The first version has a wrong parenthesis, which is barely visible in the R code.
+
+````{r, results="asis", echo=TRUE}
+f <- function(tau)
+{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
+    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
+      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
+        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
+}
+
+attributes(f)$name = quote(f)
+mathout(f)
+````
+
+But, if the function is rendered in MathML, the wrong parenthesis appears as a curly
+brace, and we immediately see the mistake. The correct version is shown below
+for comparison.
+
+````{r, results="asis", echo=TRUE}
+f <- function(tau)
+{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
+    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
+      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
+        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
+}
+
+attributes(f)$name = quote(f)
+mathout(f)
+````
+
+This example may seem trivial.
+However, as the reader well knows, a missed parenthesis is often the cause of untraceable errors and contradictory results.
+Thus, rendering the expression in MathMl improves the readability of the text, allowing better identification of the sources of possible errors.
+
+## Errors
+
+An additional feature that this package offers is related to its application in educational scenarios.
+The functions here reported are intended to identify and highlight possible errors that students make, as well as provide the correct answer for the task.
+
+```{r, results="asis"}
+t <- quote(omit_left(a + b))
+mathout(t)
+
+catmath(omit_left(a + b))
+```
+
+```{r, results="asis"}
+t <- quote(omit_right(a + b))
+mathout(t)
+
+catmath(omit_right(a + b))
+```
+
+```{r, results="asis"}
+t <- list(quote(a), quote(b), quote(omit(c)))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- quote(add_left(a + b))
+mathout(t)
+
+catmath(add_left(a + b))
+```
+
+```{r, results="asis"}
+t <- quote(add_right(a + b))
+mathout(t)
+
+catmath(add_right(a + b))
+```
+
+```{r, results="asis"}
+t <- list(quote(a), quote(b), quote(add(c)), quote(d))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- quote(add_right(a + b))
+mathout(t)
+
+catmath(add_right(a + b))
+```
+
+```{r, results="asis"}
+t <- list(quote(a), quote(b), quote(add(c)), quote(d))
+mathout(t)
+```
+
+```{r, results="asis"}
+t <- quote(instead(a, b) + c)
+mathout(t)
+
+catmath(instead(a, b) + c)
+```
+
+It is also possible to use flags for context-dependent rendering.
+The highlight flag is the default one.
+It shows where the error is and what is its solution.
+The ignore flag allows the program to ignore the fact that there is an error and print the expression as if it were correct.
+Finally, the third flag, fix, tells you what is the term that you are supposed to fix without giving the solution.
+Below the code with its comment is provided.
+
+```{r, results="asis"}
+# It highlights what has been omitted
+t <- quote(omit_left(a + b))
+mathout(t, flags=quote(error(highlight))) # default
+```
+
+```{r, results="asis"}
+# It prints the correct version
+mathout(t, flags=quote(error(ignore)))
+```
+
+```{r, results="asis"}
+# It highlights the correct version by framing it
+mathout(t, flags=quote(error(fix)))
+```
+
+```{r, results="asis"}
+# It frames the redundant term
+t <- quote(add_right(a + b))
+mathout(t, flags=quote(error(highlight))) # default
+```
+
+```{r, results="asis"}
+# It prints the correct expression without the redundant term
+mathout(t, flags=quote(error(ignore)))
+```
+
+```{r, results="asis"}
+# It strikes the redundant term 
+mathout(t, flags=quote(error(fix)))
+```
+
+```{r, results="asis"}
+# It highlights that a is wrong and b is correct
+t <- quote(instead(a, b) + c)
+mathout(t, flags=quote(error(highlight))) # default
+```
+
+```{r, results="asis"}
+# It provides the correct solution
+mathout(t, flags=quote(error(ignore)))
+```
+
+```{r, results="asis"}
+# It provides the correct solution framing the correct term
+mathout(t, flags=quote(error(fix)))
+```
+
+# Conclusions
+
+This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time.
+This makes it easier for the user to notice writing errors and correct them smoothly.
+Building on current features of R\ and existing packages for displaying formulas in R\ (e.g.
+Murrell and Ihaka [-@murrell2000], Allaire et al.
+[-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility.
+To do so, it uses the Prolog's grammar rules and translates the R\ expressions to MathML.
+Then, two possible solutions for typesetting are available to export R\ Markdown documents in different formats or ShinyApp webpages: MathML and MathJax.
+
+The article and Appendix show examples of how the package can be used and what operations are currently possible.
+Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.
+
+In its current version \pkg{mathml} has some limitations.
+For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation.
+However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations.
+Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications.
+
+\pkg{mathml} is available for R\ version (…) and later, and can be easily installed using the usual `install.packages("mathml")`.
+The source code of the package is found at (…) including installation instructions for Unix, Windows and macOS.
+
+# Acknowledgments
+
+>>>>>>> 9ed4866198fb10109aebbba2714ff06f1e6e9df5

### 2022-12-09 Draft3.Rmd
Draft 3

--- 2022-12-09 Draft2.Rmd
@@
 mathout(t, flags=quote(error(ignore)))
+```
+
+```{r, results="asis"}
+# It prints the incorrect version
+mathout(t, flags=quote(error(asis)))
 ```

### 2022-12-09 Draft3.tex
Draft 3

% !TeX root = RJwrapper.tex
\title{mathml: Translate R expressions to MathML and LaTeX/MathJax}
\author{by Matthias Gondan and Irene Alfarone}

\maketitle

\abstract{%
This R~package translates R~objects to suitable entities in MathML or LaTeX, thereby enabling pretty mathematical representation of R~functions or algorithms in data analysis, scientific reports and interactive web content. R~code text descriptions and mathematical content already coexist, side-by-side, in the RMarkdown document rendering language. The present package allows to use the same R~objects for both data analysis and typesetting in documents or web content. By tightening the link between the statistical analysis and its verbal or symbolic representation, this new feature is a futher step towards reproducible science. User-defined hooks allow to extend the package by mapping new variable names and functions to new MathML and LaTeX entities. We provide a few working examples that illustrate the use of the package in a little scientific report and in an interactive web page.
}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The R extension of the markdown language (Xie, Dervieux, and Riederer 2020) enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently (R Core Team 2021, version 4.2), R's manual pages include support for
mathematical expressions (Sarkar and Hornik 2022), which already is a big improvement.
However, rules for the translation of R's built-in language elements to their
mathematical representation are still lacking. So far, R expressions such
as \texttt{pbinom(k,\ N,\ p)} are printed as they are; pretty mathematical formulae
like \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
is, \texttt{P\_\textbackslash{}mathrm\{Bi\}\textbackslash{}left(X\ \textbackslash{}le\ k;\ N,\ p\textbackslash{}right)}. Except for minimalistic examples,
these commands are tedious to type in and hard to read in their source code.

The present R package defines a set of rules for an \emph{automatic} translation of
R expressions to mathematical output in RMarkdown documents (Xie, Dervieux, and Riederer 2020) and
ShinyApp webpages (Chang et al. 2022). The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R elements by custom mathematical
signs.

\emph{todo}: mention that the \emph{same} R expression is used for mathematical
calculations and printing.

The paper is organized as follows. Section~2 presents the background of the
package, including Prolog and the R package \pkg{rolog} and the two main classes
of rules for translating R~objects to mathematical expressions. Section~3
illustrates some important features of the \pkg{mathml} package, potential
issues and workarounds using examples from the day-to-day perspective of the
user. Sections~4 and~5 present two case studies for the use of the package in
a scientific report and a ShinyApp web page. Section~6 concludes with a
discussion and ideas for further development. A complete list of the R elements
covered by the package is found in the Appendix.

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{prolog-the-package}{%
\subsection{\texorpdfstring{Prolog: the \pkg{rolog} package}{Prolog: the  package}}\label{prolog-the-package}}

The automatic translation of R expressions to mathematical output is achieved through a Prolog interpreter. Prolog is a classical logic programming language with many applications in expert systems, computer linguistics and traditional, that is, symbolic artificial intelligence. The main strength of Prolog is its concise representation of facts and rules for the representation of knowledge and grammar, as well as its efficient built-in search engine for closed world domains. R, as it is well-known, is a statistical programming language for data analysis and statistical modeling which is widely used in academia and industry. Besides the core library, a lot of packages have been developed for all kinds of statistical problems, including statistics-based artificial intelligence tools such as neural networks for machine learning and deep learning. Whereas Prolog is weak in statistical computation, but strong in symbolic manipulation, the converse may be said for the R language. The \pkg{rolog} package then bridges this gap by providing an interface to a SWI-Prolog distribution in R. The communication between the two systems is mainly in the form of queries from R to Prolog, but two predicates allow Prolog to ring back and evaluate terms in R.
Here is reported a first trivial example. At the Prolog end, there is a handler for \texttt{pbinom/3} that translates the term into a pretty MathML syntax like P\_bi(X \textless= k; N, pi).

\begin{verbatim}
library(rolog)
\end{verbatim}

\begin{verbatim}
#> Found SWI-Prolog at SWI_HOME_DIR: /Library/Frameworks/R.framework/Versions/4.2/Resources/library/rswipl/swipl/lib/swipl
\end{verbatim}

\begin{verbatim}
#> Welcome to SWI-Prolog (threaded, 64 bits, version 8.5.20)
#> SWI-Prolog comes with ABSOLUTELY NO WARRANTY. This is free software.
#> Please run ?- license. for legal details.
#> 
#> For online help and background, visit https://www.swi-prolog.org
#> For built-in help, use ?- help(Topic). or ?- apropos(Word).
\end{verbatim}

\begin{verbatim}
library(mathml)
term = quote(pbinom(k, N, p))

# Pretty print
mathout(term)
\end{verbatim}

\({{{P}_\mathrm{Bi}}\left({{{X}\le{k}}{;{{N}{,{p}}}}}\right)}\)

Beside the graphical rendition of the term, an important feature is that the package allows also for its calculation, for example:

\begin{verbatim}
k = 10
N = 22
p = 0.4
eval(term)
\end{verbatim}

{[}1{]} 0.77195

Here we present a second less trivial example where Prolog needs to find out the name of
the integration variable for \texttt{sin}. For that purpose, \pkg{rolog} provides a predicate \texttt{r\_eval/2} that calls R from Prolog. Here, the predicate is used for the R~function \texttt{formalArgs(args(sin))}, which returns the name of the function argument of \texttt{sin}, that is, \texttt{x}.

\begin{verbatim}
term = quote(integrate(sin, 0L, 2L*pi))
mathout(term)
\end{verbatim}

\(\int_{{0}}^{{{2}{}\pi}}{{\sin{x}}}\,{d{x}}\)

\begin{verbatim}
eval(term)
\end{verbatim}

\begin{verbatim}
#> 2.221501e-16 with absolute error < 4.4e-14
\end{verbatim}

Note that, on the Prolog end, the handler for \texttt{integrate/3} is rather rigid; it
accepts only these three arguments in that particular order, and without names,
that is, \texttt{integrate(sin,\ lower=0L,\ upper=2L\ *\ pi)}, would not print the desired
result.

Therefore, \texttt{pl/mathml.pl} includes two handlers that accept terms with named
arguments, \texttt{integrate(f=Fn,\ lower=Lower,\ upper=Upper)}, as well as terms of the
form \texttt{\$(integrate(Fn,\ Lower,\ Upper),\ value)} that are needed for the evaluation
below.

\begin{verbatim}
# A custom function
g <- function(u)
{
  sin(u)
}

# Mixture of (partially) named and positional arguments in unusual order
term = call("integrate", lower=-Inf, upper=Inf, g)
attributes(term)$name = quote(g)
mathout(canonical(term))
\end{verbatim}

\(\int_{{-\infty}}^{\infty}{{{g}\left({u}\right)}}\,{d{u}}\)

\begin{verbatim}
# Even more complex
term <- call("*", 2L, call("$", call("integrate", low=-Inf, up=Inf, g), "value"))
attributes(term)$name = quote(g)
mathout(canonical(term))
\end{verbatim}

\({{2}\cdot\int_{{-\infty}}^{\infty}{{{g}\left({u}\right)}}\,{d{u}}}\)

\begin{verbatim}
# It is a bit of a mystery that R knows the result of this integral.
eval(term)
\end{verbatim}

{[}1{]} 0

The extra R function \texttt{canonical()} applies \texttt{match.call()} to non-primitive R
calls, basically cleaning up the arguments and bringing them into the correct
order.

\hypertarget{package-capabilities}{%
\section{Package capabilities}\label{package-capabilities}}

\pkg{mathml} is a package for pretty mathematical representation of R functions or algorithms in data analysis, scientific reports and interactive web content. The features currently supported are listed below, following the order proposed by (Murrell and Ihaka 2000).

\hypertarget{hooks}{%
\subsection{Hooks}\label{hooks}}

Before start listing all the possible operations with \pkg{mathml}, we mention here that \pkg{mathml} also allows the user to create hooks for specific functions. As an example:

\begin{verbatim}
# Improve display
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
\end{verbatim}

The hooks reported here will be the one used in Section \emph{ref} to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance

\hypertarget{arithmetic-operators}{%
\subsection{Arithmetic operators}\label{arithmetic-operators}}

The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division. For instance:

\begin{verbatim}
catmath(x + y - z)
\end{verbatim}

\({{{x}+{y}}-{z}}\)

In addition, with \pkg{mathml} it is also possible to combine \texttt{minus} and \texttt{plus} and get a new operator. However, in R~the operator \(\pm\) is not present and this leads to some interesting questions. If we let R~present the expression, this is what we get:

\begin{verbatim}
# Note wrong precedence of custom operator
catmath(mean(m) %+-% 1.96 * s/sqrt(N))
\end{verbatim}

\({{\left({\overline{{m}}\pm{1.96}}\right)\cdot{s}}/\sqrt{{N}}}\)

As can be seen, the new custom operator takes precedence over multiplication, as R~automatically puts \texttt{mean(m)\ \%+-\%\ 1.96} in parentheses. Unfortunately, this is not consistent with the rules of mathematics. As the reader well knows, it is not possible in R~to change the precedence of custom operators. This leaves two possible solutions. The first involves a mathematically unnecessary parenthesis around \texttt{(1.96\ *\ s/sqrt(N))}. The second is more elegant and mathematically consistent: here \texttt{1.96\ *\ s/sqrt(N)} is put in curly brackets and mathematical formality is respected. In both solutions R~is able to produce correct output once the order of operations is made explicit.

\begin{verbatim}
# This is what we actually want
catmath(mean(m) %+-% (1.96 * s/sqrt(N)))
\end{verbatim}

\({\overline{{m}}\pm\left({{{1.96}{}{s}}/\sqrt{{N}}}\right)}\)

\begin{verbatim}
# Without parentheses
catmath(mean(m) %+-% {1.96 * s/sqrt(N)})
\end{verbatim}

\({\overline{{m}}\pm{{{1.96}{}{s}}/\sqrt{{N}}}}\)

\emph{The following can be inserted in a table}

\begin{verbatim}
catmath(x*y)
\end{verbatim}

\({{x}{}{y}}\)

\begin{verbatim}
catmath(x %*% y)
\end{verbatim}

\({{x}\times{y}}\)

\begin{verbatim}
catmath(x %.% y)
\end{verbatim}

\({{x}\cdot{y}}\)

\begin{verbatim}
catmath(X %x% Y)
\end{verbatim}

\({{X}\otimes{Y}}\)

\begin{verbatim}
catmath(x/y)
\end{verbatim}

\({{x}/{y}}\)

\begin{verbatim}
catmath(x %/% y)
\end{verbatim}

\({\lfloor{{x}/{y}}\rfloor}\)

\hypertarget{subscripts-and-superscripts}{%
\subsection{Subscripts and Superscripts}\label{subscripts-and-superscripts}}

Secondly, \pkg{mathml} allows also for the display of subscripts, superscripts and their combination.

\begin{verbatim}
catmath(a["x"])
\end{verbatim}

\({{a}_\mathrm{x}}\)

\begin{verbatim}
catmath(a^x)
\end{verbatim}

\({{{a}}^{{x}}}\)

\begin{verbatim}
catmath(x["b"]^"a")
\end{verbatim}

\({{x}_\mathrm{b}^\mathrm{a}}\)

\hypertarget{fractions}{%
\subsection{Fractions}\label{fractions}}

As in (Murrell and Ihaka 2000) the \texttt{frac} and \texttt{over} functions aim to represent fractions. \texttt{choose} and \texttt{lchoose}, as in R~\pkg{base}, display the binomial coefficient.

\begin{verbatim}
catmath(frac(x, y))
\end{verbatim}

\(\frac{{x}}{{y}}\)

\begin{verbatim}
catmath(over(x, y))
\end{verbatim}

\(\frac{{x}}{{y}}\)

\begin{verbatim}
catmath(choose(n, k))
\end{verbatim}

\(\binom{{n}}{{k}}\)

\begin{verbatim}
catmath(lchoose(n, k))
\end{verbatim}

\({\log\binom{{n}}{{k}}}\)

\hypertarget{radicals}{%
\subsection{Radicals}\label{radicals}}

\begin{verbatim}
catmath(sqrt(x))
\end{verbatim}

\(\sqrt{{x}}\)

\hypertarget{relations}{%
\subsection{Relations}\label{relations}}

Among the possible relational operators, following (Murrell and Ihaka 2000), we included:

\begin{verbatim}
catmath(x==y)
\end{verbatim}

\({{x}={y}}\)

\begin{verbatim}
catmath(x < y)
\end{verbatim}

\({{x}<{y}}\)

\begin{verbatim}
catmath(x <= y)
\end{verbatim}

\({{x}\le{y}}\)

\begin{verbatim}
catmath(x >= y)
\end{verbatim}

\({{x}\ge{y}}\)

\begin{verbatim}
catmath(!x)
\end{verbatim}

\({\lnot{x}}\)

\begin{verbatim}
catmath(x != y)
\end{verbatim}

\({{x}\ne{y}}\)

\begin{verbatim}
catmath(X ~ Y)
\end{verbatim}

\({{X}\sim{Y}}\)

\begin{verbatim}
catmath(x %~~% y)
\end{verbatim}

\({{x}\approx{y}}\)

\begin{verbatim}
catmath(x %==% y)
\end{verbatim}

\({{x}\equiv{y}}\)

\begin{verbatim}
catmath(x %=~% y)
\end{verbatim}

\({{x}\cong{y}}\)

\begin{verbatim}
catmath(x %prop% y)
\end{verbatim}

\({{x}\propto{y}}\)

\hypertarget{set-relations}{%
\subsection{Set relations}\label{set-relations}}

Here is provided a series of useful set relations.

\emph{Still to be implemented}

\begin{verbatim}
catmath(a %in% A)
\end{verbatim}

\({{a}\in{A}}\)

\begin{verbatim}
catmath(intersect(A, B))
\end{verbatim}

\({{A}\cap{B}}\)

\begin{verbatim}
catmath(union(A, B))
\end{verbatim}

\({{A}\cup{B}}\)

\begin{verbatim}
catmath(is.null(intersect(A, B)))
\end{verbatim}

\({{{A}\cap{B}}=\emptyset}\)

\hypertarget{arrows}{%
\subsection{Arrows}\label{arrows}}

Following once again (Murrell and Ihaka 2000), we provide here a list of useful arrows

\begin{verbatim}
# x %<->% y
catmath(x %<->% y)
\end{verbatim}

\({{x}\leftrightarrow{y}}\)

\begin{verbatim}
# x %->% y
catmath(x %->% y)
\end{verbatim}

\({{x}\rightarrow{y}}\)

\begin{verbatim}
# x %<-% y
catmath(x %<-% y)
\end{verbatim}

\({{x}\leftarrow{y}}\)

\begin{verbatim}
# x %up% y:
catmath(x %up% y)
\end{verbatim}

\({{x}\uparrow{y}}\)

\begin{verbatim}
# x %down% y:
catmath(x %down% y)
\end{verbatim}

\({{x}\downarrow{y}}\)

\begin{verbatim}
# x %<=>% y: solved it using iff
catmath(x %<=>% y)
\end{verbatim}

\({{x}\iff{y}}\)

\begin{verbatim}
# x %=>% y: solved using rArr
catmath(x %=>% y)
\end{verbatim}

\({{x}\Rightarrow{y}}\)

\begin{verbatim}
# x %<=% y: solved using lArr
catmath(x %<=% y)
\end{verbatim}

\({{x}\Leftarrow{y}}\)

\begin{verbatim}
# x %dblup% y:
catmath(x %dblup% y)
\end{verbatim}

\({{x}\Uparrow{y}}\)

\begin{verbatim}
# x %dbldown% y:
catmath(x %dbldown% y)
\end{verbatim}

\({{x}\Downarrow{y}}\)

\hypertarget{accents}{%
\subsection{Accents}\label{accents}}

\begin{verbatim}
catmath(hat(x))
\end{verbatim}

\(\hat{{x}}\)

\begin{verbatim}
catmath(hat(xy))
\end{verbatim}

\(\hat{{xy}}\)

\begin{verbatim}
catmath(mean(x))
\end{verbatim}

\(\overline{{x}}\)

\begin{verbatim}
catmath(mean(xy))
\end{verbatim}

\(\overline{{xy}}\)

\hypertarget{symbolic-names}{%
\subsection{Symbolic names}\label{symbolic-names}}

\begin{verbatim}
catmath(A-Omega)
\end{verbatim}

\({{A}-\Omega}\)

\begin{verbatim}
catmath(alpha-omega)
\end{verbatim}

\({\alpha-\omega}\)

\begin{verbatim}
catmath(pi-Pi)
\end{verbatim}

\({\pi-\Pi}\)

\begin{verbatim}
catmath(Inf)
\end{verbatim}

\(\infty\)

\hypertarget{operators}{%
\subsection{Operators}\label{operators}}

A list of possible operators is available here.

\begin{verbatim}
catmath(sum(x[i], i==1L, n))
\end{verbatim}

\({{\sum}\left({{{x}_{i}}{,{{i}={1}}{,{n}}}}\right)}\)

\begin{verbatim}
# fixme: braces not needed when right to *
catmath(a * sum(b))
\end{verbatim}

\({{a}\cdot{{\sum}{b}}}\)

\begin{verbatim}
# Multiplication (capital Pi) is missing
catmath(a * prod(b))
\end{verbatim}

\({{a}\cdot{{\prod}{b}}}\)

\begin{verbatim}
catmath(integrate(sin, 0L, 2L*pi))
\end{verbatim}

\(\int_{{0}}^{{{2}{}\pi}}{{\sin{x}}}\,{d{x}}\)

\begin{verbatim}
catmath(A & B)
\end{verbatim}

\({{A}\land{B}}\)

\begin{verbatim}
catmath(A | B)
\end{verbatim}

\({{A}\lor{B}}\)

\begin{verbatim}
catmath(!A)
\end{verbatim}

\({\lnot{A}}\)

\begin{verbatim}
catmath(xor(A, B))
\end{verbatim}

\({{A}\veebar{B}}\)

\begin{verbatim}
catmath(exp(2L*pi*i))
\end{verbatim}

\({\exp\left({{{2}{}\pi}\cdot{i}}\right)}\)

\begin{verbatim}
catmath(expm1(2L*pi*i))
\end{verbatim}

\({{\exp\left({{{2}{}\pi}\cdot{i}}\right)}-{1}}\)

\begin{verbatim}
catmath(log(x))
\end{verbatim}

\({\log{x}}\)

\begin{verbatim}
catmath(log10(x))
\end{verbatim}

\({{\log_{10}}{x}}\)

\begin{verbatim}
catmath(log2(x))
\end{verbatim}

\({{\log_{2}}{x}}\)

\begin{verbatim}
catmath(logb(x, e))
\end{verbatim}

\({{\log_{e}}{x}}\)

\begin{verbatim}
catmath(log1p(x))
\end{verbatim}

\({{1}+{\log{x}}}\)

\begin{verbatim}
catmath(ceiling(pi))
\end{verbatim}

\({\lceil\pi\rceil}\)

\begin{verbatim}
catmath(floor(pi))
\end{verbatim}

\({\lfloor\pi\rfloor}\)

\begin{verbatim}
catmath(sin(pi/2L))
\end{verbatim}

\({\sin\left({\pi/{2}}\right)}\)

\begin{verbatim}
catmath(cos(pi/2L))
\end{verbatim}

\({\cos\left({\pi/{2}}\right)}\)

\begin{verbatim}
catmath(tan(pi/2L))
\end{verbatim}

\({\tan\left({\pi/{2}}\right)}\)

\hypertarget{grouping}{%
\subsection{Grouping}\label{grouping}}

At first glance, grouping expressions may seem trivial enough. However, when you want to deal with, for instance, exponentiation or multiplication, things can get more complicated. Let us consider a trivial example. If we want to calculate and evaluate the expression \texttt{a\ *\ (b\ +\ c)}, we find that R~automatically gives precedence to the addition even if we do not specify it in the text, as in the following.

\begin{verbatim}
first <- quote(a)
second <- quote(b + c)
mathout(call("*", first, second))
\end{verbatim}

\({{a}\cdot\left({{b}+{c}}\right)}\)

\begin{verbatim}
# precedence
quote(a * (b + c))[[3]][[1]]
\end{verbatim}

\texttt{(}

With exponentiation the situation can be less trivial. If we consider \texttt{x} to the power of \texttt{y\ +\ z}, following R~notation we should add a parenthesis around the two terms. Otherwise R~would consider only \texttt{y} as power of \texttt{x}.

\begin{verbatim}
catmath(x^y + z)
\end{verbatim}

\({{{{x}}^{{y}}}+{z}}\)

\begin{verbatim}
# mathout(call("+", call("^", as.symbol("x"), as.symbol("y")), as.symbol("z")))

catmath(x^(y + z))
\end{verbatim}

\({{{x}}^{\left({{y}+{z}}\right)}}\)

Unfortunately, R~automatically put \texttt{y\ +\ z} in parentheses even if, from a mathematical point of view, the brackets are not needed. To solve this issue two solutions are possible, the first one is cumbersome since it requires many specifications. The second, that can be considered a shortcut, uses a syntax similar to LateX and puts the power in curly brackets. In both cases the output produced is consistent with the mathematical notation.

\begin{verbatim}
# without parenthesis
mathout(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
\end{verbatim}

\({{{x}}^{{{y}+{z}}}}\)

\begin{verbatim}
# shortcut
catmath(x^{y + z})
\end{verbatim}

\({{{x}}^{{{y}+{z}}}}\)

\hypertarget{absolute-value}{%
\subsection{Absolute value}\label{absolute-value}}

Is it also possible to calculate and represent the absolute value, as:

\begin{verbatim}
catmath(abs(x))
\end{verbatim}

\(\left|{{x}}\right|\)

\hypertarget{matrix-and-vectors}{%
\subsection{Matrix and Vectors}\label{matrix-and-vectors}}

Since also Matrices and Vectors are important elements in R, we decided to implement them and render them graphically in the following ways.

\begin{verbatim}
vec <- 1:3
mathout(canonical(vec))
\end{verbatim}

\(\left({{1}{\,{2}{\,{3}}}}\right)\)

\begin{verbatim}
mathout(vec)
\end{verbatim}

\(\left({{1}{\,{2}{\,{3}}}}\right)\)

\begin{verbatim}
mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
mathout(canonical(mat))
\end{verbatim}

\(\left(\begin{array}{cc}{2} & {4}\\ {3} & {5}\\ \end{array}\right)\)

\hypertarget{abbreviations}{%
\subsection{Abbreviations}\label{abbreviations}}

\pkg{mathml} allows also for the use of abbreviations in formulas. Let us consider here the t-statistic for independent samples with equal variance assumed.
The procedure consists of three steps. The first one is to define the hooks to nicely print the variables. Second, we write the t-statistic specifying in the formula how to calculate the pooled variance with the function \texttt{denote(abbr,\ expr,\ info)}. Finally, since \pkg{mathml} also allows computation, we can directly compute the value of the t statistic with \texttt{eval()}.

\begin{verbatim}
hook(quote(x_1), quote(mean(x["1"])))
hook(quote(x_2), quote(mean(x["2"])))
hook(quote(n_1), quote(n["1"]))
hook(quote(n_2), quote(n["2"]))
hook(quote(s_1), quote(s["1"]))
hook(quote(s_2), quote(s["2"]))
hook(quote(s_p), quote(s["pool"]))

t <- quote(dfrac(x_1 - x_2, sqrt(denote(s_p^2L, frac((n_1 - 1L)*s_1^2L + (n_2 - 1L)*s_2^2L, n_1+n_2-2L), "the pooled variance.") * (dfrac(1L, n_1)+dfrac(1L,n_2)))))

mathout(t)
\end{verbatim}

\(\displaystyle{\frac{{\overline{{{x}_\mathrm{1}}}-\overline{{{x}_\mathrm{2}}}}}{\sqrt{{{{s}_\mathrm{pool}^{2}}\cdot\left({\displaystyle{\frac{{1}}{{{n}_\mathrm{1}}}}+\displaystyle{\frac{{1}}{{{n}_\mathrm{2}}}}}\right)}}}}\), with \({{{s}_\mathrm{pool}^{2}}=\frac{{{\left({{{n}_\mathrm{1}}-{1}}\right)\cdot{{s}_\mathrm{1}^{2}}}+{\left({{{n}_\mathrm{2}}-{1}}\right)\cdot{{s}_\mathrm{2}^{2}}}}}{{{{{n}_\mathrm{1}}+{{n}_\mathrm{2}}}-{2}}}}\) denoting the pooled variance.

\begin{verbatim}
# An example

x_1 = 20.5
x_2 = 35.7
n_1 = 150
n_2 = 160
s_1 = 3.4
s_2 = 4.2

eval(t)
\end{verbatim}

{[}1{]} -34.88421

\hypertarget{a-case-study}{%
\section{A case study}\label{a-case-study}}

For a more clear representation of what the \pkg{mathml} package is supposed to do, we provide here a case study, based on the work by Schwarz (1994), in which the features of \pkg{mathml} are implemented.

The reader may recognize that Schwarz' (1994) study can be traced back to the strand of decision-making models, where decision-making is assumed to be a process of
noisy accumulation of information over time (e.g., Ratcliff et al. 2016). The aim of
Schwarz' (1994) study is to present a new explanation of redundancy gains when
observers make speeded responses to stimuli of different sources, and the same
information is presented on two channels. Schwarz (1994) describes a model that assumes
the superposition of channel-specific diffusion processes that eventually reach
an absorbing barrier to elicit the response. For a detailed description the
reader may refer to the original article.

\hypertarget{the-model}{%
\subsubsection{The model}\label{the-model}}

Schwarz' (1994) model refers to two stimuli A and B, presented either
alone or in combination (redundant stimuli), with the redundant stimuli being
presented either simultaneously or with onset asynchrony \(\tau\). The channel
activation is described as a two-dimensional Wiener process with
drifts \(\mu_\mathrm A\), \(\mu_\mathrm B\),
variances \(\sigma^2_\mathrm A\), \(\sigma^2_\mathrm B\),
correlation \(\rho_\mathrm{AB}\) and initial
conditions \(X_i(t=0)=0, i=\mathrm{A, B}\). The response is elicited when the
process reaches the absorbing barrier \(c > 0\) for the first time. In combined
stimuli (``redundant'' stimuli), the overall diffusion process
is \(X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)\) , which is again a
Wiener process with drift \(\mu_\mathrm A + \mu_\mathrm B\) and
variance \(\sigma^2_\mathrm A + \sigma^2_\mathrm B  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B\).

In single-target trials, the first passages at \(c\) are expected at
\(E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}\), whereas in synchronous
redundant-target trials, we have
\(E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}\).

For asynchronous stimuli, Schwarz (1994) derived the expected
first-passage time \(E[D_{\mathrm A(\tau)\mathrm B}]\) as a function of the
stimulus onset asyncrony \(\tau > 0\) (Eq.~10),

\begin{verbatim}
f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
}

mathout(f)
\end{verbatim}

\({{\lambda\left({\tau{\,{c}{\,{\mu_\mathrm{A}}{\,{\sigma_\mathrm{A}}{\,{\mu_\mathrm{B}}{\,{\sigma_\mathrm{B}}}}}}}}\right)}={\displaystyle{\frac{{c}}{{\mu_\mathrm{A}}}}+{\left({\displaystyle{\frac{{1}}{{\mu_\mathrm{A}}}}-\displaystyle{\frac{{1}}{{{\mu_\mathrm{A}}+{\mu_\mathrm{B}}}}}}\right)\cdot\left[{{\left({{{\mu_\mathrm{A}}{}\tau}-{c}}\right)\cdot{\Phi\left(\displaystyle{\frac{{{c}-{{\mu_\mathrm{A}}{}\tau}}}{\sqrt{{{\sigma_\mathrm{A}^{2}}{}\tau}}}}\right)}}-{{\left({{{\mu_\mathrm{A}}{}\tau}+{c}}\right)\cdot{\exp\left(\displaystyle{\frac{{{{2}{}{\mu_\mathrm{A}}}\cdot{c}}}{{\sigma_\mathrm{A}^{2}}}}\right)}}\cdot{\Phi\left(\displaystyle{\frac{{{-{c}}-{{\mu_\mathrm{A}}{}\tau}}}{\sqrt{{{\sigma_\mathrm{A}^{2}}{}\tau}}}}\right)}}}\right]}}}\)

The observable response time is assumed to be the sum of \(D\), the time
employed to reach the threshold for the decision, and \(M\), denoting the of the
other processes.

\(E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M\)

\hypertarget{methods}{%
\subsubsection{Methods}\label{methods}}

Schwarz (1994) applied the model to Miller's (1986) data from
a redundant signals task with 13 onset
asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
where 0 refers to synchronous AB, \(\infty\) refers to the single-target
presentation, and negative \(\tau\) denote those conditions in which B is
presented before A. Each condition was replicated 400 times. The 13 mean RTs and
standard deviations are given in Table 2 below.

Then, from Schwarz' (1994) Equations {[}2{]} {[}3{]} and {[}10{]}, we calculated the expected first time passage for the respective \(\tau\) and the parameters \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)

\({{\lambda\left({\tau{\,{c}{\,{\mu_\mathrm{A}}{\,{\sigma_\mathrm{A}}{\,{\mu_\mathrm{B}}{\,{\sigma_\mathrm{B}}}}}}}}\right)}={{rbind}\left({\displaystyle{\frac{{c}}{{\mu_\mathrm{A}}}},\ \mathrm{if}\ {\tau=\infty}{\,\displaystyle{\frac{{c}}{{\mu_\mathrm{B}}}},\ \mathrm{if}\ {\tau={-\infty}}{\,{{f}\left({\tau{,{c}{,{\mu_\mathrm{A}}{,{\sigma_\mathrm{A}}{,{\mu_\mathrm{B}}{,{\sigma_\mathrm{B}}}}}}}}\right)},\ \mathrm{if}\ {\tau>{0}}{\,\displaystyle{\frac{{c}}{{{\mu_\mathrm{A}}+{\mu_\mathrm{B}}}}},\ \mathrm{if}\ {\tau={0}}{\,{{f}\left({{-\tau}{,{c}{,{\mu_\mathrm{B}}{,{\sigma_\mathrm{B}}{,{\mu_\mathrm{A}}{,{\sigma_\mathrm{A}}}}}}}}\right)},\ \mathrm{if}\ {\tau<{0}}}}}}}\right)}}\)

For instance, we can see that with \(\tau = -167\) the expected first passage time is:

\begin{verbatim}
#> [1] 70.31943
\end{verbatim}

To which we have to add \(\mu_M = 161\) (which represents the amount of time that the movement of providing the answer takes) and we obtain:

\begin{verbatim}
#> [1] 231.3194
\end{verbatim}

Similarly, we obtain the first-time passage for every \(\tau\):

\({{\lambda\left({\tau{\,{c}{\,{\mu_\mathrm{A}}{\,{\sigma_\mathrm{A}}{\,{\mu_\mathrm{B}}{\,{\sigma_\mathrm{B}}{\,{\mu_\mathrm{M}}}}}}}}}\right)}={{{G}\left({\tau{,{c}{,{\mu_\mathrm{A}}{,{\sigma_\mathrm{A}}{,{\mu_\mathrm{B}}{,{\sigma_\mathrm{B}}}}}}}}\right)}+{\mu_\mathrm{M}}}}\)

\begin{verbatim}
#>  [1] 236.7576 231.3194 230.0571 228.4017 226.0523 222.2336 215.0541 238.5977
#>  [9] 262.3817 282.9833 299.7299 312.9791 349.6792
\end{verbatim}

\hypertarget{model-fitting-and-parameters-estimation}{%
\subsubsection{Model fitting and Parameters Estimation}\label{model-fitting-and-parameters-estimation}}

Then we proceeded with the model fitting, using the data provided by Miller (1986) study in Schwarz (1994). First, we calculated the predicted means RT as \(E[RT]=E[T]+\mu_M\).

We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed (Ratcliff and McKoon 2008) with at the first passage of \(c\) \(X(t) \sim N(\mu t , \sigma^2 t)\). Where \(T\) is expressed as \(T= \arg \min_t X(t)\geq c\) with \(T \sim IG (c, \mu, \sigma^2)\) with mean \(E(T)=\displaystyle \frac{c}{\mu}\) and variance \(Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}\). Furthermore, for the Central Limit Theorem \(\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})\)

For these reasons, in order to obtain the Goodness of Fit measure we proceeded with a z-standardisation.

\({{\lambda\left({\overline{{X}}{\,{s}{\,{N}{\,\tau{\,{c}{\,{\mu_\mathrm{A}}{\,{\sigma_\mathrm{A}}{\,{\mu_\mathrm{B}}{\,{\sigma_\mathrm{B}}{\,{\mu_\mathrm{M}}}}}}}}}}}}\right)}={{rbind}\left({{{E}={{h}\left({\tau{,{c}{,{\mu_\mathrm{A}}{,{\sigma_\mathrm{A}}{,{\mu_\mathrm{B}}{,{\sigma_\mathrm{B}}{,{\mu_\mathrm{M}}}}}}}}}\right)}}{\,\displaystyle{\frac{{\overline{{X}}-{E}}}{{{s}/\sqrt{{N}}}}}}}\right)}}\)

\begin{verbatim}
#>  [1] -2.05627706  0.92433511 -0.02852697 -0.70084870  1.21730061 -0.88115730
#>  [7]  1.38996139 -0.42696106  0.47564623 -3.98888722 -1.08120914  1.77700915
#> [13] -0.36505332
\end{verbatim}

Then, we calculated the Goodness of Fit as follows:

\(X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)\).

\({{\lambda\left({{par}{\,\tau{\,\overline{{X}}{\,{s}{\,{N}}}}}}\right)}={{rbind}\left({{{{z}_\tau}={{zf}\left({\overline{{X}}{,{s}{,{N}{,\tau{,{{c}={100}}{,{{\mu_\mathrm{A}}={{par}_\mathrm{mu_A}}}{,{{\sigma_\mathrm{A}}={{par}_\mathrm{sigma_A}}}{,{{\mu_\mathrm{B}}={{par}_\mathrm{mu_B}}}{,{{\sigma_\mathrm{B}}={{par}_\mathrm{sigma_B}}}{,{{\mu_\mathrm{M}}={{par}_\mathrm{mu_M}}}}}}}}}}}}}\right)}}{\,{{\sum}{{z}_\tau^{2}}}}}\right)}}\)

{[}1{]} 30.54472
{[}1{]} 0.0001692742

\({{\lambda\left({\tau{\,\overline{{X}}{\,{s}{\,{N}}}}}\right)}={{rbind}\left({{{start}={{c}\left({{{\mu_\mathrm{A}}={0.30}}{,{{\sigma_\mathrm{A}}={2.00}}{,{{\mu_\mathrm{B}}={1.00}}{,{{\sigma_\mathrm{B}}={5.00}}{,{{\mu_\mathrm{M}}={200.00}}}}}}}\right)}}{\,{{optim}\left({{start}{,{{fn}={gof}}{,{\tau=\tau}{,{\overline{{X}}=\overline{{X}}}{,{{s}={s}}{,{{N}={N}}{,{{hessian}={true}}{,{{control}={{list}\left({{fnscale}={1.00}}\right)}}}}}}}}}}\right)}}}\right)}}\)

\hypertarget{results}{%
\subsubsection{Results}\label{results}}

The calculated Goodness of Fit value is \(X^2= 30.54\), \(p= 0.0002\). Additionally, we proceeded with the parameters' estimation with the Least Squares Method.

The values of the estimated parameters reported in the table below are consistent with those reported by Schwarz (1994).

\begin{longtable}[]{@{}ccc@{}}
\caption{Table 1: Estimated Model's Parameters}\tabularnewline
\toprule()
Parameters & Estimate & 95\% Confidence Interval \\
\midrule()
\endfirsthead
\toprule()
Parameters & Estimate & 95\% Confidence Interval \\
\midrule()
\endhead
\(\mu_ \mathrm A\) & 0.53 & 0.51, 0.56 \\
\(\sigma_ \mathrm A\) & 4.4 & 3.8, 5.1 \\
\(\mu_ \mathrm B\) & 1.32 & 1.2, 1.4 \\
\(\sigma_ \mathrm B\) & 18.3 & 11.6, 24.9 \\
\(\mu_ \mathrm M\) & 161 & 157, 165 \\
\bottomrule()
\end{longtable}

Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i. \(\tau= 100\)) Miller's (1986) observed values are quite similar to the one predicted by Schwarz' (1994) model. Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz (1994) sufficiently fit the data.

\begin{longtable}[]{@{}cccc@{}}
\caption{Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's (1986) study in Schwarz (1994)}\tabularnewline
\toprule()
SOA & Predicted Mean RT & Observed Mean RT & Observed SD \\
\midrule()
\endfirsthead
\toprule()
SOA & Predicted Mean RT & Observed Mean RT & Observed SD \\
\midrule()
\endhead
\(-\infty\) & 237 & 231 & 56 \\
\(-167\) & 231 & 234 & 58 \\
\(-133\) & 230 & 230 & 40 \\
\(-100\) & 228 & 227 & 40 \\
\(-67\) & 226 & 228 & 32 \\
\(-33\) & 222 & 221 & 28 \\
\(0\) & 215 & 217 & 28 \\
\(33\) & 239 & 238 & 28 \\
\(67\) & 262 & 263 & 26 \\
\(100\) & 283 & 277 & 30 \\
\(133\) & 300 & 298 & 32 \\
\(167\) & 313 & 316 & 34 \\
\(\infty\) & 350 & 348 & 92 \\
\bottomrule()
\end{longtable}

\emph{Note}. Parameters: \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)

\hypertarget{mistakes}{%
\subsection{Mistakes}\label{mistakes}}

The case study presented above displays several possible ways of how \pkg{mathml} can be a useful tool in data representation and calculation. Furthermore, the careful reader might have also noticed that \pkg{mathml} allows also for a more smoothly detection of possible mistakes in writing.

An interesting use case of the package is in this sense the readability of self-written
functions in R code. R code is notoriously hard to read, as are most programming
languages (Green 1977). The poor legibility of the language is one of the main
sources of mistakes. For illustration, we consider Equation 10 in
Schwarz (1994), namely the same equation reported above. The first version has a wrong parenthesis, which is barely visible in the R code.

\begin{verbatim}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
}

attributes(f)$name = quote(f)
mathout(f)
\end{verbatim}

\({{{f}\left(\tau\right)}={\displaystyle{\frac{{c}}{{\mu_\mathrm{A}}}}+\left\{{\displaystyle{\frac{{1}}{{\mu_\mathrm{A}}}}-{\displaystyle{\frac{{1}}{{{\mu_\mathrm{A}}+{\mu_\mathrm{B}}}}}\cdot\left[{{\left({{{\mu_\mathrm{A}}{}\tau}-{c}}\right)\cdot{\Phi\left(\displaystyle{\frac{{{c}-{{\mu_\mathrm{A}}{}\tau}}}{\sqrt{{{\sigma_\mathrm{A}^{2}}{}\tau}}}}\right)}}-{{\left({{{\mu_\mathrm{A}}{}\tau}+{c}}\right)\cdot{\exp\left(\displaystyle{\frac{{{{2}{}{\mu_\mathrm{A}}}\cdot\tau}}{{\sigma_\mathrm{A}^{2}}}}\right)}}\cdot{\Phi\left(\displaystyle{\frac{{{-{c}}-{{\mu_\mathrm{A}}{}\tau}}}{\sqrt{{{\sigma_\mathrm{A}^{2}}{}\tau}}}}\right)}}}\right]}}\right\}}}\)

But, if the function is rendered in MathML, the wrong parenthesis appears as a curly
brace, and we immediately see the mistake. The correct version is shown below
for comparison.

\begin{verbatim}
f <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
}

attributes(f)$name = quote(f)
mathout(f)
\end{verbatim}

\({{{f}\left(\tau\right)}={\displaystyle{\frac{{c}}{{\mu_\mathrm{A}}}}+{\left({\displaystyle{\frac{{1}}{{\mu_\mathrm{A}}}}-\displaystyle{\frac{{1}}{{{\mu_\mathrm{A}}+{\mu_\mathrm{B}}}}}}\right)\cdot\left[{{\left({{{\mu_\mathrm{A}}{}\tau}-{c}}\right)\cdot{\Phi\left(\displaystyle{\frac{{{c}-{{\mu_\mathrm{A}}{}\tau}}}{\sqrt{{{\sigma_\mathrm{A}^{2}}{}\tau}}}}\right)}}-{{\left({{{\mu_\mathrm{A}}{}\tau}+{c}}\right)\cdot{\exp\left(\displaystyle{\frac{{{{2}{}{\mu_\mathrm{A}}}\cdot\tau}}{{\sigma_\mathrm{A}^{2}}}}\right)}}\cdot{\Phi\left(\displaystyle{\frac{{{-{c}}-{{\mu_\mathrm{A}}{}\tau}}}{\sqrt{{{\sigma_\mathrm{A}^{2}}{}\tau}}}}\right)}}}\right]}}}\)

This example may seem trivial. However, as the reader well knows, a missed parenthesis is often the cause of untraceable errors and contradictory results. Thus, rendering the expression in MathMl improves the readability of the text, allowing better identification of the sources of possible errors.

\hypertarget{errors}{%
\subsection{Errors}\label{errors}}

An additional feature that this package offers is related to its application in educational scenarios. The functions here reported are intended to identify and highlight possible errors that students make, as well as provide the correct answer for the task.

In this first example the function highlights that the student forgot to add the left-hand side of the expression and it does that by striking the missing part. In this case the letter A and the operator \texttt{+}.
Similarly the function \texttt{omit\_right} does the same for the omissions in the right-hand side of the expression. Finally the function \texttt{omit()} can be used, for instance, to highlight the omission in the elements of a list.

\begin{verbatim}
t <- quote(omit_left(a + b))
mathout(t)
\end{verbatim}

\({{{cancel}\left({{a}{\,+}}\right)}{\,{b}}}\)

\begin{verbatim}
t <- quote(omit_right(a + b))
mathout(t)
\end{verbatim}

\({{a}{\,{{cancel}\left({+{\,{b}}}\right)}}}\)

\begin{verbatim}
t <- list(quote(a), quote(b), quote(omit(c)))
mathout(t)
\end{verbatim}

\({{a}{\,{b}{\,{{cancel}\left({c}\right)}}}}\)

The functions \texttt{add\_left()}, \texttt{add\_right()}, \texttt{add()} reported below follow the same way of reasoning, but they account for the mistakes made in adding incorrect elements in the expression. They highlight the redundant element placing it in a box.

\begin{verbatim}
t <- quote(add_left(a + b))
mathout(t)
\end{verbatim}

\({{{box}\left({{a}{\,+}}\right)}{\,{b}}}\)

\begin{verbatim}
t <- quote(add_right(a + b))
mathout(t)
\end{verbatim}

\({{a}{\,{{box}\left({+{\,{b}}}\right)}}}\)

\begin{verbatim}
t <- list(quote(a), quote(b), quote(add(c)), quote(d))
mathout(t)
\end{verbatim}

\({{a}{\,{b}{\,{{box}\left({c}\right)}{\,{d}}}}}\)

Finally the function \texttt{instead(\ ,\ )} place a curly bracket under the incorrect element and provides its correct version under the curly bracket. In order to provide the correction with this function one must fill it with its wrong and correct terms in this order: \texttt{instead(wrong,\ correct)}. An example of its use its provided below.

\begin{verbatim}
t <- quote(instead(a, b) + c)
mathout(t)
\end{verbatim}

\({{{underbrace}\left({{a}{,{\mathrm{instead of}{\,{b}}}}}\right)}+{c}}\)

It is also possible to use flags for context-dependent rendering. The highlight flag is the default one. It shows where the error is and what is its solution. The ignore flag allows the program to ignore the fact that there is an error and print the expression as if it were correct. Finally, the third flag, fix, tells you what is the term that you are supposed to fix without giving the solution. Below the code with its comment is provided.

\begin{verbatim}
# It highlights what has been omitted
t <- quote(omit_left(a + b))
mathout(t, flags=quote(error(highlight))) # default
\end{verbatim}

\({{{cancel}\left({{a}{\,+}}\right)}{\,{b}}}\)

\begin{verbatim}
# It prints the correct version
mathout(t, flags=quote(error(ignore)))
\end{verbatim}

\({{a}+{b}}\)

\begin{verbatim}
# It prints the correct version
mathout(t, flags=quote(error(asis)))
\end{verbatim}

\({b}\)

\begin{verbatim}
# It highlights the correct version by framing it
mathout(t, flags=quote(error(fix)))
\end{verbatim}

\({{{box}\left({{a}{\,+}}\right)}{\,{b}}}\)

\begin{verbatim}
# It frames the redundant term
t <- quote(add_right(a + b))
mathout(t, flags=quote(error(highlight))) # default
\end{verbatim}

\({{a}{\,{{box}\left({+{\,{b}}}\right)}}}\)

\begin{verbatim}
# It prints the correct expression without the redundant term
mathout(t, flags=quote(error(ignore)))
\end{verbatim}

\({a}\)

\begin{verbatim}
# It strikes the redundant term 
mathout(t, flags=quote(error(fix)))
\end{verbatim}

\({{a}{\,{{cancel}\left({+{\,{b}}}\right)}}}\)

\begin{verbatim}
# It highlights that a is wrong and b is correct
t <- quote(instead(a, b) + c)
mathout(t, flags=quote(error(highlight))) # default
\end{verbatim}

\({{{underbrace}\left({{a}{,{\mathrm{instead of}{\,{b}}}}}\right)}+{c}}\)

\begin{verbatim}
# It provides the correct solution
mathout(t, flags=quote(error(ignore)))
\end{verbatim}

\({{b}+{c}}\)

\begin{verbatim}
# It provides the correct solution framing the correct term
mathout(t, flags=quote(error(fix)))
\end{verbatim}

\({{{box}\left({b}\right)}+{c}}\)

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

This package expands the possibilities of R~by allowing calculations and graphical rendition at the same time. This makes it easier for the user to notice writing errors and correct them smoothly. Building on current features of R~and existing packages for displaying formulas in R~(e.g.~Murrell and Ihaka (2000), Allaire et al. (2018)), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility. To do so, it uses the Prolog's grammar rules and translates the R~expressions to MathML. Then, two possible solutions for typesetting are available to export R~Markdown documents in different formats or ShinyApp webpages: MathML and MathJax.

The article and Appendix show examples of how the package can be used and what operations are currently possible. Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.

In its current version \pkg{mathml} has some limitations. For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation. However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations. Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications.

\pkg{mathml} is available for R~version (\ldots) and later, and can be easily installed using the usual \texttt{install.packages("mathml")}. The source code of the package is found at (\ldots) including installation instructions for Unix, Windows and macOS.

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-allaire2018}{}}%
Allaire, JJ, Rich Iannone, Alison Presmanes Hill, and Yihui Xie. 2018. {``Distill for r Markdown.''} \url{https://rstudio.github.io/distill}.

\leavevmode\vadjust pre{\hypertarget{ref-Chang2022}{}}%
Chang, Winston, Joe Cheng, JJ Allaire, Carson Sievert, Barret Schloerke, Yihui Xie, Jeff Allen, Jonathan McPherson, Alan Dipert, and Barbara Borges. 2022. \emph{Shiny: Web Application Framework for r}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\vadjust pre{\hypertarget{ref-green1977}{}}%
Green, T. R. G. 1977. {``Conditional Program Statements and Their Comprehensibility to Professional Programmers.''} \emph{Journal of Occupational Psychology} 50: 93--109.

\leavevmode\vadjust pre{\hypertarget{ref-miller1986}{}}%
Miller, Jeff. 1986. {``Timecourse of Coactivation in Bimodal Divided Attention.''} \emph{Perception \& {P}sychophysics} 40: 331--43.

\leavevmode\vadjust pre{\hypertarget{ref-murrell2000}{}}%
Murrell, Paul, and Ross Ihaka. 2000. {``An Approach to Providing Mathematical Annotation in Plots.''} \emph{Journal of Computational and Graphical Statistics} 9: 582--99. \url{https://www.jstor.con/stable/1390947}.

\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
R Core Team. 2021. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-ratcliff2008}{}}%
Ratcliff, Roger, and Gail McKoon. 2008. {``The Diffusion Decision Model: Theory and Data for Two-Choice Decision Task.''} \emph{Neural Computation} 20: 873--922.

\leavevmode\vadjust pre{\hypertarget{ref-ratcliff2016}{}}%
Ratcliff, Roger, Philip L. Smith, Scott D. Brown, and Gail McKoon. 2016. {``Diffusion Decision Model: Current Issues and History.''} \emph{Trends in Cognitive Sciences} 20: 260--81.

\leavevmode\vadjust pre{\hypertarget{ref-Sarkar2022}{}}%
Sarkar, Deepayan, and Kurt Hornik. 2022. \emph{Enhancements to {HTML} {D}ocumentation}. \url{https://blog.r-project.org/2022/04/08/enhancements-to-html-documentation/index.html}.

\leavevmode\vadjust pre{\hypertarget{ref-schwarz1994}{}}%
Schwarz, Wolf. 1994. {``Diffusion, Superposition, and the Redundant-Targets Effect.''} \emph{Journal of Mathematical Psychology} 38: 504--20.

\leavevmode\vadjust pre{\hypertarget{ref-Xie2020}{}}%
Xie, Y., C. Dervieux, and E. Riederer. 2020. \emph{R Markdown Cookbook}. Cambridge: Chapman; Hall/CRC.

\end{CSLReferences}

\bibliography{RJreferences.bib}

\address{%
Matthias Gondan\\
University of Innsbruck\\%
Department of Psychology\\ Innsbruck, Austria\\
%
\url{https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0001-9974-0057}{0000-0001-9974-0057}}\\%
\href{mailto:Matthias.Gondan-Rochon@uibk.ac.at}{\nolinkurl{Matthias.Gondan-Rochon@uibk.ac.at}}%
}

\address{%
Irene Alfarone\\
University of Innsbruck\\%
Department of Psychology\\ Innsbruck, Austria\\
%
\url{https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-8409-8900}{0000-0002-8409-8900}}\\%
\href{mailto:Irene.Alfarone@uibk.ac.at}{\nolinkurl{Irene.Alfarone@uibk.ac.at}}%
}

### 2022-12-09 Draft3.Rmd
intro

@@
 
+library(mathml)
 library(plotly)
@@
 of rules for translating R\ objects to mathematical expressions. Section\ 3
-illustrates some important features of the \pkg{mathml} package, potential
-issues and workarounds using examples from the day-to-day perspective of the
-user. Sections\ 4 and\ 5 present two case studies for the use of the package in
-a scientific report and a ShinyApp web page. Section\ 6 concludes with a
+illustrates the main features of the \pkg{mathml} package, potential issues and
+workarounds using examples from the day-to-day perspective of the user.
+Sections\ 4 and\ 5 present two case studies for the use of the package in a
+scientific report and a ShinyApp web page. Section\ 6 concludes with a
 discussion and ideas for further development. A complete list of the R elements
@@
 
-## Prolog: the \pkg{rolog} package
-
-The automatic translation of R expressions to mathematical output is achieved through a Prolog interpreter.
-Prolog is a classical logic programming language with many applications in expert systems, computer linguistics and traditional, that is, symbolic artificial intelligence.
-The main strength of Prolog is its concise representation of facts and rules for the representation of knowledge and grammar, as well as its efficient built-in search engine for closed world domains.
-R, as it is well-known, is a statistical programming language for data analysis and statistical modeling which is widely used in academia and industry.
-Besides the core library, a lot of packages have been developed for all kinds of statistical problems, including statistics-based artificial intelligence tools such as neural networks for machine learning and deep learning.
-Whereas Prolog is weak in statistical computation, but strong in symbolic manipulation, the converse may be said for the R language.
-The \pkg{rolog} package then bridges this gap by providing an interface to a SWI-Prolog distribution in R.
-The communication between the two systems is mainly in the form of queries from R to Prolog, but two predicates allow Prolog to ring back and evaluate terms in R.
-Here is reported a first trivial example. At the Prolog end, there is a handler for `pbinom/3` that translates the term into a pretty MathML syntax like P_bi(X <= k; N, pi).
-
-```{r, echo=TRUE, results='asis'}
-library(rolog)
+The translation of R expressions to mathematical output is achieved through a
+Prolog interpreter. Prolog is a classical logic programming language with many
+applications in expert systems, computer linguistics and symbolic artificial
+intelligence. The main strength of Prolog is its concise representation of facts
+and rules for the representation of knowledge and grammar, as well as its
+efficient built-in search engine for closed world domains. As it is well-known,
+R\ is a statistical programming language for data analysis and statistical
+modeling which is widely used in academia and industry. Besides the core
+library, a lot of packages have been developed for all kinds of statistical
+problems, including statistics-based artificial intelligence tools such as
+neural networks for machine learning and deep learning. Whereas Prolog is weak
+in statistical computation, but strong in symbolic manipulation, the converse
+may be said for the R language. The \pkg{rolog} package bridges this gap by
+providing an interface to a SWI-Prolog distribution in R. The communication
+between the two systems is mainly in the form of queries from R\ to Prolog, but
+two predicates allow Prolog to ring back and evaluate terms in R.
+
+For a first illustration of the package, we consider the binomial probability.
+
+```{r}
+term <- quote(pbinom(k, N, p))
+term
+```
+
+The term is quoted to avoid its immediate evaluation which would raise an error
+anyway since the variables `k`, `N`, `p` have not yet been defined. Experienced
+readers will remember that the quoted expression is a short form for
+
+```r
+term <- call("pbinom", as.name("k"), as.name("N"), as.name("p"))
+```
+
+As is seen from the output above, the variable `term` is not assigned the result
+of the calculation, but an R\ "call" that can eventually be evaluated
+[see, e.g., @Wickham2019, for more details on "non-standard evaluation"].
+
+```{r}
+k <- 10
+N <- 22
+p <- 0.4
+eval(term)
+```
+
+The R package \pkg{mathml} can now be used to render the call in
+MathML, that is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
+below (some of the curly braces are not really needed in this simple example,
+but are necessary in edge cases).
+
+```{r}
 library(mathml)
-term = quote(pbinom(k, N, p))
-
-# Pretty print
+mathjax(term)
+```
+
+We can include the output in a RMarkdown document by specifying `results='asis'`
+in the R code chunk, as has been done below. The R function mathout is just a
+wrapper that invokes `mathml(...)` for HTML output and `mathjax(...)` for
+LaTeX output.
+
+```{r, results='asis'}
 mathout(term)
@@
 
-Beside the graphical rendition of the term, an important feature is that the package allows also for its calculation, for example:
-
-```{r, results = 'asis', echo = TRUE}
-k = 10
-N = 22
-p = 0.4
-eval(term)
-```
-
-Here we present a second less trivial example where Prolog needs to find out the name of
+At the Prolog end, there is a predicate `math/2` that translates the
+call `pbinom(K, N, Pi)` into a "function" `fn/2` with the name `P_Bi`, one
+argument `X =< K`, and the two parameters `N` and `Pi`.
+
+```prolog
+math(pbinom(K, N, Pi), M)
+ => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
+```
+
+Thus, the predicate `math/2` could be considered a "macro" that translates a
+mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
+element, namely `fn(..., ...)`.
+
+The low-level predicates `ml/3` and `jax/3` are used to convert mathematical
+elements to MathML or LaTeX, respectively.
+
+```prolog
+ml(Flags, fn(Name, (Args ; Params)), M)
+ => ml(Flags, Name, N),
+    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Params)])), X),
+    M = mrow([N, mo(&(af)), X]).
+```
+
+The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
+with `N` representing the function's name and `X` its arguments and parameters,
+enclosed in parentheses. The flags can be used for context-sensitive
+translation (see, e.g., the errors below).
+
+# Package \pkg{mathml} in practice
+
+In the second example, Here we present a second less trivial example where Prolog needs to find out the name of
 the integration variable for `sin`.

### 2022-12-09 inst/pl/mathml.pl

@@ -2248,9 +2248,8 @@ math(Flags, dbinom(K, N, Pi), New, X)
  => New = Flags,
     X = fn(subscript('P', "Bi"), (['X' = K] ; [N, Pi])).
 
-math(Flags, pbinom(K, N, Pi), New, X)
- => New = Flags,
-    X = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
+math(pbinom(K, N, Pi), M)
+ => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
 
 math(Flags, upbinom(K, N, Pi), New, X)
  => New = Flags,
@@ -2302,7 +2301,7 @@ math(pnorm(Z), M)
 %
 ml(Flags, fn(Name, (Args ; Params)), M)
  => ml(Flags, Name, F),
-    ml(Flags, paren(list(op(';'), [list(op(','), Args), list(op(','), Params)])), X),
+    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Params)])), X),
     M = mrow([F, mo(&(af)), X]).
 
 jax(Flags, fn(Name, (Args ; Params)), M),

### 2022-12-10 Draft3.Rmd
custom operators

@@
 
+\pkg{mathml} is an R\ package for pretty mathematical representation of
+R\ functions and objects in data analysis, scientific reports and interactive
+web content. The features currently supported are listed below, roughly following
+the order proposed by [@murrell2000].
+
+## Basic calculations
+
+Basic arithmetic operators and parentheses are translated as they are, as
+illustrated below.
+
+```{r, results="asis"}
+term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + {n*o}^{p + q})
+mathout(term)
+```
+
+For multiplications involving only numbers and symbols, the multiplication sign
+is omitted. This heuristic does not always produce the desired result;
+therefore, \pkg{mathml} defines the extra functions `dot/2`, `nodot/2`,
+and `times/2` that produce the respective multiplication signs.
+
+```{r, results="asis"}
+term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
+mathout(term)
+```
+
+Note that for the standard operators with known precedence, \pkg{mathml} is able
+to detect if parentheses are needed; for example, a parenthesis is automatically
+put around `g + h` in the `nodot`-example and around `n*o` above. However, we
+also note an unecessary parenthesis around `l + m` above. This parenthesis
+results is a consequence of `quote(a^(b + c))` actually producing a nested
+R\ call of the form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:
+
+```{r}
+term <- quote(a^(b + c))
+paste(term)
+```
+
+This feature is unfortunate because the extra parenthesis around `b + c` is not
+needed and yields to the extra parenthesis. The preferred result is obtained by
+using the functional representation `quote('^'(k, l + m))` of the power, or, as
+a workaround, curly braces (see last summand above).
+
+## Custom operators
+
+While the parentheses typically follow the precedence in standard infix
+operators, undesirable results may be obtained in custom operators.
+
+```{r, results='asis'}
+term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
+mathout(term)
+```
+
+As can be seen, \pkg{mathml} reminds the user that it is not possible to change
+the precedence of custom operators in R, and that expressions with such
+operators are evaluated strictly from left to right. As in the
+previous example, the solution is to either use the functional form of the
+operator or a curly brace to enforce the correct operator precedence.
+
+```{r, results='asis'}
+term <- quote(`%+-%`(mean(X), 1.96 * s / sqrt(N)))
+term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})
+mathout(term)
+```
+
+_The following can be inserted in a table_
+
+```{r, results='asis', echo=TRUE}
+
+catmath(x*y)
+
+catmath(x %*% y)
+
+catmath(x %.% y)
+
+catmath(X %x% Y)
+
+catmath(x/y)
+
+catmath(x %/% y)
+```
+
 In the second example, Here we present a second less trivial example where Prolog needs to find out the name of
@@
 # Package capabilities
-
-\pkg{mathml} is a package for pretty mathematical representation of R functions or algorithms in data analysis, scientific reports and interactive web content.
-The features currently supported are listed below, following the order proposed by [@murrell2000].
 
@@
 The hooks reported here will be the one used in Section _ref_ to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance
-
-## Arithmetic operators
-The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division. For instance:
-
-```{r, results="asis", echo=TRUE}
-catmath(x + y - z)
-```
-
-In addition, with \pkg{mathml} it is also possible to combine `minus` and `plus` and get a new operator.
-However, in R\ the operator \(\pm\) is not present and this leads to some interesting questions.
-If we let R\ present the expression, this is what we get:
-
-```{r, results='asis', echo=TRUE}
-# Note wrong precedence of custom operator
-catmath(mean(m) %+-% 1.96 * s/sqrt(N))
-```
-
-As can be seen, the new custom operator takes precedence over multiplication, as R\ automatically puts `mean(m) %+-% 1.96` in parentheses.
-Unfortunately, this is not consistent with the rules of mathematics.
-As the reader well knows, it is not possible in R\ to change the precedence of custom operators.
-This leaves two possible solutions.
-The first involves a mathematically unnecessary parenthesis around `(1.96 * s/sqrt(N))`.
-The second is more elegant and mathematically consistent: here `1.96 * s/sqrt(N)` is put in curly brackets and mathematical formality is respected.
-In both solutions R\ is able to produce correct output once the order of operations is made explicit.
-
-```{r, results='asis', echo=TRUE}
-# This is what we actually want
-catmath(mean(m) %+-% (1.96 * s/sqrt(N)))
-
-# Without parentheses
-catmath(mean(m) %+-% {1.96 * s/sqrt(N)})
-```
-
-_The following can be inserted in a table_
-
-```{r, results='asis', echo=TRUE}
-
-catmath(x*y)
-
-catmath(x %*% y)
-
-catmath(x %.% y)
-
-catmath(X %x% Y)
-
-catmath(x/y)
-
-catmath(x %/% y)
-```
 

### 2022-12-10 inst/pl/mathml.pl

@@ -10,20 +10,20 @@
 %
 % R interface: Translate R expression to MathML string
 %
-r2mathml(R, S) :-
-    r2mathml([], R, S).
+r2mathml(R, S)
+ => r2mathml([], R, S).
 
-r2mathml(Flags, R, S) :-
-    mathml(Flags, R, M),
+r2mathml(Flags, R, S)
+ => mathml(Flags, R, M),
     html(M, H, []),
     maplist(atom_string, H, S).
 
 % Same for MathJax/LaTeX
-r2mathjax(R, S) :-
-    r2mathjax([], R, S).
+r2mathjax(R, S)
+ => r2mathjax([], R, S).
 
-r2mathjax(Flags, R, S) :-
-    mathjax(Flags, R, S).
+r2mathjax(Flags, R, S)
+ => mathjax(Flags, R, S).
 
 %
 % Translate R expression to HTML term
@@ -245,10 +245,28 @@ math(Flags, superscript(R, Idx), New, M)
 
 % Render
 ml(Flags, R, M),
+    select(superscript(Pwr), Flags, New),
+    prec(Flags, R, Prec),
+    current_op(P, xfy, ^),
+    Prec >= P
+ => ml(New, paren(R), X),
+    ml(New, Pwr, Y),
+    M = msup([X, Y]).
+
+ml(Flags, R, M),
     select(superscript(Pwr), Flags, New)
  => ml(New, R, X),
     ml(New, Pwr, Y),
     M = msup([X, Y]).
+
+jax(Flags, R, M),
+    prec(Flags, R, Prec),
+    current_op(P, xfy, ^),
+    Prec >= P,
+    select(superscript(Pwr), Flags, New)
+ => jax(New, paren(R), X),
+    jax(New, Pwr, Y),
+    format(string(M), "{{~w}^{~w}}", [X, Y]).
 
 jax(Flags, R, M),
     select(superscript(Pwr), Flags, New)
@@ -1501,6 +1519,10 @@ math(Flags, ~(A, B), New, X)
     current_op(Prec, xfx, =),
     X = yfy(Prec, 'Tilde', A, B).
 
+math(Flags, dot(A, B), New, X)
+ => New = Flags,
+    X = cdot(A, B).
+
 math(Flags, cdot(A, B), New, X)
  => New = Flags,
     current_op(Prec, yfx, *),
@@ -1802,6 +1824,12 @@ jax(Flags, and([A | T]), W)
[…]

### 2022-12-10 Draft3.Rmd
add table

@@
 
-mathout <- function(...)
+mathout <- function(..., flags=NULL)
 {
   if(knitr::is_html_output())
-    return(mathml(...))
+    return(mathml(..., flags=c(flags, list(cat=TRUE))))
 
   if(knitr::is_latex_output())
-    return(mathjax(...))
+    return(mathjax(..., flags=c(flags, list(cat=TRUE))))
 
   warning("no output specified")  
-  mathjax(...)
+  mathjax(..., flags=c(flags, list(cat=TRUE)))
 }
@@
 term <- quote(`%+-%`(mean(X), 1.96 * s / sqrt(N)))
-term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})
+term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
 mathout(term)
@@
 
-_The following can be inserted in a table_
-
-```{r, results='asis', echo=TRUE}
-
-catmath(x*y)
-
-catmath(x %*% y)
-
-catmath(x %.% y)
-
-catmath(X %x% Y)
-
-catmath(x/y)
-
-catmath(x %/% y)
+A few more custom operators are shown in Table\ 1.
+
+```{r, echo=FALSE}
+op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
+  'integer division'=quote(x %/% y))
+m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))
+
+t <- cbind(Operator=as.character(op), Output=m)
+knitr::kable(t, caption = "Custom operators in \\pkg{mathml}", 
+             label="tab:custom")
 ```

### 2022-12-10 R/mathml.R

@@ -27,10 +27,14 @@
 #'
 mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
 {
-  flags = c(attributes(term), flags)
-  t = rolog::once(call("r2mathml", flags, term, expression(X)),
+  flags <- c(attributes(term), flags, list(cat=FALSE))
+  t <- rolog::once(call("r2mathml", flags, term, expression(X)),
     options=list(preproc=mathml_preproc))
-  cat(paste(t$X, collapse=""))
+  r <- paste(t$X, collapse="")
+  if(flags$cat)
+    return(cat(r))
+
+  return(r)
 }
 
 # Prolog representation of not equal etc. (left: R, right: Prolog)
@@ -98,10 +102,14 @@ mathml_preproc <- function(query=quote(2 != 2))
 #'
 mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
 {
-  flags = c(attributes(term), flags)
-  t = rolog::once(call("r2mathjax", flags, term, expression(X)),
+  flags <- c(attributes(term), flags, list(cat=FALSE))
+  t <- rolog::once(call("r2mathjax", flags, term, expression(X)),
     options=list(preproc=mathml_preproc))
-  cat(paste(t$X, collapse=""))
+  r <- paste(t$X, collapse="")
+  if(flags$cat)
+    return(cat(r))
+
+  return(r)
 }
 
 #' Calligraphic font

### 2022-12-11 Draft3.Rmd
table

@@
     toc: no
+    tables: yes
 bibliography: RJreferences.bib
@@
 library(ggplot2)
+library(kableExtra)
 
@@
 as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
-like \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
+like $P_\mathrm{Bi}(X \le k; N, p)$ require explicit Latex commands, that
 is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
@@
 
-A few more custom operators are shown in Table\ 1.
-
-```{r, echo=FALSE}
+A few more custom operators are shown in Table \@ref(tab:custom-operators).
+
+```{r custom-operators, echo=FALSE}
 op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
-  'integer division'=quote(x %/% y))
-m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))
-
+  integerdivision=quote(x %/% y))
+m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
 t <- cbind(Operator=as.character(op), Output=m)
-knitr::kable(t, caption = "Custom operators in \\pkg{mathml}", 
-             label="tab:custom")
+knitr::kable(t, caption = "Custom operators in mathml")
 ```
@@
 alone or in combination (redundant stimuli), with the redundant stimuli being
-presented either simultaneously or with onset asynchrony \(\tau\). The channel
+presented either simultaneously or with onset asynchrony $\tau$. The channel
 activation is described as a two-dimensional Wiener process with
-drifts \(\mu_\mathrm A\), \(\mu_\mathrm B\),
-variances \(\sigma^2_\mathrm A\), \(\sigma^2_\mathrm B\),
-correlation \(\rho_\mathrm{AB}\) and initial
-conditions \(X_i(t=0)=0, i=\mathrm{A, B}\). The response is elicited when the
-process reaches the absorbing barrier \(c > 0\) for the first time. In combined
+drifts $\mu_\mathrm A$, $\mu_\mathrm B$,
+variances $\sigma^2_\mathrm A$, $\sigma^2_\mathrm B$,
+correlation $\rho_\mathrm{AB}$ and initial
+conditions $X_i(t=0)=0, i=\mathrm{A, B}$. The response is elicited when the
+process reaches the absorbing barrier $c > 0$ for the first time. In combined
 stimuli ("redundant" stimuli), the overall diffusion process
-is \(X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)\) , which is again a
-Wiener process with drift \(\mu_\mathrm A + \mu_\mathrm B\) and
-variance \(\sigma^2_\mathrm A + \sigma^2_\mathrm B
-  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B\).
-
-In single-target trials, the first passages at \(c\) are expected at
-\(E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}\), whereas in synchronous
+is $X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)$ , which is again a
+Wiener process with drift $\mu_\mathrm A + \mu_\mathrm B$ and
+variance $\sigma^2_\mathrm A + \sigma^2_\mathrm B
+  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B$.
+
+In single-target trials, the first passages at $c$ are expected at
+$E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}$, whereas in synchronous
 redundant-target trials, we have
-\(E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}\).
+$E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}$.
 
 For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
-first-passage time \(E[D_{\mathrm A(\tau)\mathrm B}]\) as a function of the
-stimulus onset asyncrony \(\tau > 0\) (Eq.\ 10),
+first-passage time $E[D_{\mathrm A(\tau)\mathrm B}]$ as a function of the
+stimulus onset asyncrony $\tau > 0$ (Eq.\ 10),
 
@@
 
-The observable response time is assumed to be the sum of \(D\), the time
-employed to reach the threshold for the decision, and \(M\), denoting the of the
+The observable response time is assumed to be the sum of $D$, the time
+employed to reach the threshold for the decision, and $M$, denoting the of the
 other processes.
 
-\(E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M\)
+$E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M$
 
@@
 a redundant signals task with 13 onset
-asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
-where 0 refers to synchronous AB, \(\infty\) refers to the single-target
-presentation, and negative \(\tau\) denote those conditions in which B is
+asynchronies $0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty$ ms,
+where 0 refers to synchronous AB, $\infty$ refers to the single-target
+presentation, and negative $\tau$ denote those conditions in which B is
 presented before A. Each condition was replicated 400 times. The 13 mean RTs and
@@
 
-Then, from Schwarz' [-@schwarz1994] Equations [2] [3] and [10], we calculated the expected first time passage for the respective \(\tau\) and the parameters \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)
+Then, from Schwarz' [-@schwarz1994] Equations [2] [3] and [10], we calculated the expected first time passage for the respective $\tau$ and the parameters $\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161$
 
@@
 
-For instance, we can see that with \(\tau = -167\) the expected first passage time is:
+For instance, we can see that with $\tau = -167$ the expected first passage time is:
 
@@
 
-To which we have to add \(\mu_M = 161\) (which represents the amount of time that the movement of providing the answer takes) and we obtain:
+To which we have to add $\mu_M = 161$ (which represents the amount of time that the movement of providing the answer takes) and we obtain:
 
@@
 
-Similarly, we obtain the first-time passage for every \(\tau\):
+Similarly, we obtain the first-time passage for every $\tau$:
 
@@
 
-Then we proceeded with the model fitting, using the data provided by Miller [-@miller1986] study in Schwarz [-@schwarz1994]. First, we calculated the predicted means RT as \(E[RT]=E[T]+\mu_M\).
-
-We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed [@ratcliff2008] with at the first passage of \(c\) \(X(t) \sim N(\mu t , \sigma^2 t)\).
-Where \(T\) is expressed as \(T= \arg \min_t X(t)\geq c\) with \(T \sim IG (c, \mu, \sigma^2)\) with mean \(E(T)=\displaystyle \frac{c}{\mu}\) and variance \(Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}\).
-Furthermore, for the Central Limit Theorem \(\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})\)
+Then we proceeded with the model fitting, using the data provided by Miller [-@miller1986] study in Schwarz [-@schwarz1994]. First, we calculated the predicted means RT as $E[RT]=E[T]+\mu_M$.
+
+We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed [@ratcliff2008] with at the first passage of $c$ $X(t) \sim N(\mu t , \sigma^2 t)$.
+Where $T$ is expressed as $T= \arg \min_t X(t)\geq c$ with $T \sim IG (c, \mu, \sigma^2)$ with mean $E(T)=\displaystyle \frac{c}{\mu}$ and variance $Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}$.
+Furthermore, for the Central Limit Theorem $\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})$
 
@@
 
-\(X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)\).
+$X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)$.
 
@@
 
-The calculated Goodness of Fit value is \(X^2= 30.54\), \(p= 0.0002\). Additionally, we proceeded with the parameters' estimation with the Least Squares Method.
+The calculated Goodness of Fit value is $X^2= 30.54$, $p= 0.0002$. Additionally, we proceeded with the parameters' estimation with the Least Squares Method.
 
@@
 
-: Table 1: Estimated Model's Parameters
-
-| Parameters | Estimate | 95% Confidence Interval
-| :---: | :---: | :---:|
-| \( \mu_ \mathrm A \)| 0.53 | 0.51, 0.56
-| \( \sigma_ \mathrm A \)| 4.4 | 3.8, 5.1
-| \( \mu_ \mathrm B \)| 1.32 | 1.2, 1.4
-| \( \sigma_ \mathrm B \)| 18.3 | 11.6, 24.9
-| \( \mu_ \mathrm M \)| 161 | 157, 165
+```{r}
+# Table 1: Estimated Model's Parameters
+#
+# | Parameters | Estimate | 95% Confidence Interval
+# | :---: | :---: | :---:|
+# | $ \mu_ \mathrm A $| 0.53 | 0.51, 0.56
+# | $ \sigma_ \mathrm A $| 4.4 | 3.8, 5.1
+# | $ \mu_ \mathrm B $| 1.32 | 1.2, 1.4
+# | $ \sigma_ \mathrm B $| 18.3 | 11.6, 24.9
+# | $ \mu_ \mathrm M $| 161 | 157, 165
+```
 
 Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i.
-\(\tau= 100\)) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model.
+$\tau= 100$) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model.
 Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.
 
-: Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's [-@miller1986] study in Schwarz [-@schwarz1994]
-
-| SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
-|:--------:|:-----------------:|:----------------:|:-----------:|
-\(-\infty\)| 237 | 231 | 56 |
-\(-167\)   | 231 | 234 | 58 |
-\(-133\)   | 230 | 230 | 40 |
-\(-100\)   | 228 | 227 | 40 |
-\(-67\)    | 226 | 228 | 32 |
-\(-33\)    | 222 | 221 | 28 |
-\(0\)      | 215 | 217 | 28 |
-\(33\)     | 239 | 238 | 28 |
-\(67\)     | 262 | 263 | 26 |
-\(100\)    | 283 | 277 | 30 |
-\(133\)    | 300 | 298 | 32 |
-\(167\)    | 313 | 316 | 34 |
-\(\infty\) | 350 | 348 | 92 |
-
- _Note_.  Parameters: \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)
+```{r}
+# : Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's [-@miller1986] study in Schwarz [-@schwarz1994]
+#
+# | SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
+# |:--------:|:-----------------:|:----------------:|:-----------:|
+# $-\infty$| 237 | 231 | 56 |
+# $-167$   | 231 | 234 | 58 |
+# $-133$   | 230 | 230 | 40 |
+# $-100$   | 228 | 227 | 40 |
+# $-67$    | 226 | 228 | 32 |
+# $-33$    | 222 | 221 | 28 |
+# $0$      | 215 | 217 | 28 |
+# $33$     | 239 | 238 | 28 |
+# $67$     | 262 | 263 | 26 |
+# $100$    | 283 | 277 | 30 |
+# $133$    | 300 | 298 | 32 |
+# $167$    | 313 | 316 | 34 |
+# $\infty$ | 350 | 348 | 92 |
+#
+# _Note_.  Parameters: $\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161$
+```
 

### 2022-12-11 Draft3.Rmd
integrate

@@
 
-In the second example, Here we present a second less trivial example where Prolog needs to find out the name of
-the integration variable for `sin`.
-For that purpose, \pkg{rolog} provides a predicate `r_eval/2` that calls R from Prolog.
-Here, the predicate is used for the  R\ function `formalArgs(args(sin))`, which returns the name of the function argument of `sin`, that is, `x`.
-
-```{r,  results = 'asis', echo = TRUE}
-term = quote(integrate(sin, 0L, 2L*pi))
+## Ringing back to R
+
+R's `integrate` function takes a number of arguments, the most important ones
+being the function to integrate, and the lower and the upper bound of the
+integration. For mathematical typesetting in the form
+of $\int f(x)\, dx$, \pkg{mathml} needs to find out the name of the integration
+variable. For that purpose, the underlying Prolog bridge provides a
+predicate `r_eval/2` that calls R from Prolog. In the example below, this
+predicate is used for a call of `formalArgs(args(sin))`, which returns the names
+of the function arguments of `sin`, namely, `x`.
+
+```{r, results='asis'}
+term <- quote(integrate(sin, 0L, 2L*pi))
 mathout(term)
-```
-
-```{r, echo = TRUE}
-eval(term)
 ```
@@
 accepts only these three arguments in that particular order, and without names,
-that is, `integrate(sin, lower=0L, upper=2L * pi)`, would not print the desired
+that is, `integrate(sin, lower=0L, upper=2L*pi)`, would not print the desired
 result.
 
-Therefore, `pl/mathml.pl` includes two handlers that accept terms with named
-arguments, `integrate(f=Fn, lower=Lower, upper=Upper)`, as well as terms of the
-form `$(integrate(Fn, Lower, Upper), value)` that are needed for the evaluation
-below.
+To "canonicalize" function calls with named arguments and arguments in unusual
+order, \pkg{mathml} provides an auxilliary function `canonical(f, drop)` that
+reorders the argument list of calls to known functions and, if `drop=TRUE`,
+drops the function names.
+
+```{r}
+term <- quote(integrate(lower=0L, upper=2L*pi, sin))
+canonical(term)
+```
 

### 2022-12-11 Draft3.Rmd
hooks

@@
 
-Note that, on the Prolog end, the handler for `integrate/3` is rather rigid; it
-accepts only these three arguments in that particular order, and without names,
-that is, `integrate(sin, lower=0L, upper=2L*pi)`, would not print the desired
-result.
+## Names and order of arguments
+
+One of R's great features is the possibility to refer to function arguments by
+by their names, not only by their position in the list of arguments.
+Unfortunately, the prolog-powered handler for `integrate/3` is rather rigid; it
+accepts only three arguments in that particular order and without names, that
+is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.
 
@@
 
-# Package capabilities
-
-## Hooks
-Before start listing all the possible operations with \pkg{mathml},  we mention here that \pkg{mathml} also allows the user to create hooks for specific functions. As an example:
-
-```{r, echo=TRUE}
-# Improve display
-hook(quote(mu_A), quote(mu["A"]))
-hook(quote(mu_B), quote(mu["B"]))
-hook(quote(sigma_A), quote(sigma["A"]))
-hook(quote(sigma_B), quote(sigma["B"]))
-hook(quote(mu_M), quote(mu["M"]))
-hook(quote(zn), quote(z[tau]))
-hook(quote(M), quote(overline(X)))
-```
-
-The hooks reported here will be the one used in Section _ref_ to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance
+## Customizing the package
+
+In typical R\ functions, variable names are typically longer than just single
+letters, which may yield unsatisfactory results in the mathematical output.
+
+```{r, results='asis'}
+term <- quote(pbinom(successes, Ntotal, prob))
+mathout(term)
+```
+
+To improve the situation, \pkg{mathml} provides a simple hook into the prolog
+predicate `math/2` described in the previous section.
+
+```{r, results='asis'}
+hook(quote(successes), quote(k))
+hook(quote(Ntotal), quote(N))
+hook(quote(prob), quote(pi))
+mathout(term)
+```
+
+More complex hooks are possible. In the example below, `dnorm(X, M, S)` is
+translated to the form `varphi(X; M, S)`.
+
+```{r, results='asis'}
+hook(call("dnorm", expression(X), expression(M), expression(S)), 
+     call("fn", quote(varphi),
+       call(";", list(expression(X)), list(expression(M), expression(S)))))
+term <- quote(dnorm(x, mu, sigma^2L))
+mathout(term)
+```
+
+R's expressions are a bit misused to represent variables in the underlying
+prolog engine.
 

### 2022-12-11 Draft3.Rmd
table

@@
     toc: no
-    tables: yes
 bibliography: RJreferences.bib
+preamble: \usepackage{longtable}
 ---
@@
 op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
-  integerdivision=quote(x %/% y))
+  'integer division'=quote(x %/% y))
 m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
-t <- cbind(Operator=as.character(op), Output=m)
-knitr::kable(t, caption = "Custom operators in mathml")
+
+op <- as.character(op)
+if(knitr::is_latex_output())
+  op <- sapply(op, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op <- sapply(op, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op, Output=m)
+knitr::kable(t, caption="Custom operators in mathml", row.names=FALSE, escape=FALSE)
 ```
@@
 More complex hooks are possible. In the example below, `dnorm(X, M, S)` is
-translated to the form `varphi(X; M, S)`.
+translated to the form `phi(X; M, S)` (instead of the default `varphi`).
 
@@
 hook(call("dnorm", expression(X), expression(M), expression(S)), 
-     call("fn", quote(varphi),
+     call("fn", quote(phi),
        call(";", list(expression(X)), list(expression(M), expression(S)))))

### 2022-12-12 Draft3.Rmd
errors

@@
 It is also possible to use flags for context-dependent rendering.
-The highlight flag is the default one.
+The `highlight` flag is the default one.
 It shows where the error is and what is its solution.
-The ignore flag allows the program to ignore the fact that there is an error and print the expression as if it were correct.
-Finally, the third flag, fix, tells you what is the term that you are supposed to fix without giving the solution.
-Below the code with its comment is provided.
-
-```{r, results="asis"}
-# It highlights what has been omitted
-t <- quote(omit_left(a + b))
+The `ignore` flag allows the program to ignore the fact that there is an error and print the expression as if it were correct.
+The third flag, `fix`, tells you what is the term that you are supposed to fix without giving the solution.
+Finally, the last flag `asis` returns the incorrect expression as the student wrote it, without providing any solutions or corrections.
+Below is provided the code with its comment and an example of use.
+
+Let us consider for example a student that instead of writing `b + c` wrote `a+c`.
+
+With the `asis` flag we report the incorrect expression provided by the student:
+
+```{r, results="asis"}
+# It prints the incorrect version
+t <- quote(instead(a, b) + c)
+mathout(t, flags=quote(error(asis)))
+```
+
+Then, we may want underline the incorrect term and provide the correct solution.
+
+```{r, results="asis"}
+# It underlines what has been incorrect
 mathout(t, flags=quote(error(highlight))) # default
 ```
+
+Alternatively, we can put the correct solution in a box to let the student focus on its correction.
+
+```{r, results="asis"}
+# It highlights what has been corrected
+mathout(t, flags=quote(error(fix))) 
+```
+
+Finally, we can simply provide the correct solution to the problem with the flag `ignore`.
 
@@
 mathout(t, flags=quote(error(ignore)))
-```
-
-```{r, results="asis"}
-# It prints the incorrect version
-mathout(t, flags=quote(error(asis)))
-```
-
-```{r, results="asis"}
-# It highlights the correct version by framing it
-mathout(t, flags=quote(error(fix)))
-```
-
-```{r, results="asis"}
-# It frames the redundant term
-t <- quote(add_right(a + b))
-mathout(t, flags=quote(error(highlight))) # default
-```
-
-```{r, results="asis"}
-# It prints the correct expression without the redundant term
-mathout(t, flags=quote(error(ignore)))
-```
-
-```{r, results="asis"}
-# It strikes the redundant term 
-mathout(t, flags=quote(error(fix)))
-```
-
-```{r, results="asis"}
-# It highlights that a is wrong and b is correct
-t <- quote(instead(a, b) + c)
-mathout(t, flags=quote(error(highlight))) # default
-```
-
-```{r, results="asis"}
-# It provides the correct solution
-mathout(t, flags=quote(error(ignore)))
-```
-
-```{r, results="asis"}
-# It provides the correct solution framing the correct term
-mathout(t, flags=quote(error(fix)))
 ```

### 2022-12-12 inst/pl/mathml.pl
errors

@@ -2143,6 +2143,12 @@ math(Flags, omit_right(Expr), New, M),
     option(error(ignore), Flags, highlight)
  => Flags = New,
     M = Expr.
+    
+math(Flags, omit_right(Expr), New, M),
+    option(error(asis), Flags, highlight),
+    Expr =.. [_Op, L, _R]
+ => Flags = New,
+    M = L.
 
 math(Flags, omit_right(Expr), New, M),
     option(error(fix), Flags, highlight),
@@ -2176,6 +2182,11 @@ math(Flags, add_left(Expr), New, M),
     Expr =.. [_Op, _L, R]
  => Flags = New,
     M = R.
+    
+math(Flags, add_left(Expr), New, M),
+    option(error(asis), Flags, highlight)
+ => Flags = New,
+    M = Expr.
 
 math(Flags, add_left(Expr), New, M),
     option(error(fix), Flags, highlight),
@@ -2231,7 +2242,12 @@ math(Flags, add(_Expr), New, M),
     option(error(ignore), Flags, highlight)
  => Flags = New,
     M = "". % suppress at the next level, in the list
-
+    
+math(Flags, add(Expr), New, M),
+    option(error(asis), Flags, highlight)
+ => Flags = New,
+    M = Expr.    
+  
 math(Flags, add(Expr), New, M),
     option(error(fix), Flags, highlight)
  => Flags = New,
@@ -2246,6 +2262,11 @@ math(Flags, instead(_Wrong, Correct), New, M),
     option(error(ignore), Flags, highlight)
  => Flags = New,
     M = Correct.
+    
+math(Flags, instead(Wrong, _Correct), New, M),
+    option(error(asis), Flags, highlight)
+ => Flags = New,
+    M = Wrong.
 
 math(Flags, instead(_Wrong, Correct), New, M),
     option(error(fix), Flags, highlight)

### 2022-12-12 Draft3.Rmd
errors
small changes

@@
 signs.
-
-*todo*: mention that the _same_ R expression is used for mathematical
-calculations and printing.
+Moreover, in addition to graphical rendering, the same R expression provides its calculation or evaluation.
+This feature of the package saves the user's time and allows faster detection of errors.
 
@@
 
+```{r omit, echo=FALSE}
+op <- list(left=quote(omit_left(a + b)), right=quote(omit_right(a + b)), list=quote(omit(c)))
+m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
+
+op <- as.character(op)
+if(knitr::is_latex_output())
+  op <- sapply(op, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op <- sapply(op, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op, Output=m)
+knitr::kable(t, caption="Omit", row.names=FALSE, escape=FALSE)
+```
+
 The functions `add_left()`, `add_right()`, `add()` reported below follow the same way of reasoning, but they account for the mistakes made in adding incorrect elements in the expression.
@@
 mathout(t)
+```
+
+```{r add, echo=FALSE}
+op <- list(left=quote(add_left(a + b)), right=quote(add_right(a + b)), list=quote(add(c)))
+m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
+
+op <- as.character(op)
+if(knitr::is_latex_output())
+  op <- sapply(op, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op <- sapply(op, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op, Output=m)
+knitr::kable(t, caption="Add", row.names=FALSE, escape=FALSE)
 ```

### 2022-12-12 inst/pl/mathml.pl

@@ -2227,6 +2227,11 @@ math(Flags, omit(Expr), New, M),
     option(error(ignore), Flags, highlight)
  => Flags = New,
     M = Expr.
+    
+math(Flags, omit(_Expr), New, M),
+    option(error(asis), Flags, highlight)
+ => Flags = New,
+    M = "". % suppress at the next level, in the list
 
 math(Flags, omit(Expr), New, M),
     option(error(fix), Flags, highlight)

### 2022-12-12 Draft3.Rmd
clean up

@@
 LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
-rules, for example, to represent specific R elements by custom mathematical
-signs.
+rules, for example, to represent specific R elements by custom mathematical signs.
 Moreover, in addition to graphical rendering, the same R expression provides its calculation or evaluation.
@@
 ```{r add, echo=FALSE}
-op <- list(left=quote(add_left(a + b)), right=quote(add_right(a + b)), list=quote(add(c)))
+op <- list(left=quote(add_left(a + b)), right=quote(add_right(a + b)))
 m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))

### 2022-12-12 Draft3.Rmd
few additions

@@
 
+## Basic elements
+
+`mathml` contains several basic elements used in everyday mathematical expressions. For example it is possible to print the function for the linear regression as:
+
+```{r, results='asis'}
+
+term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*x)
+mathout(term)
+
+# or with greek letters as:
+
+term <- quote(y == beta["0"] + beta["1"]*x + ... + varepsilon["i"])
+mathout(term)
+```
+
+```{r, results='asis'}
+
+f <- function(x, a)
+{
+    if(a == 0) return(+Inf);
+    if(a == 0) return(-Inf);
+    if(a > 1L ) return(frac(1L,2L))
+}
+mathout(f)
+
+```
+
 ## Basic calculations

### 2022-12-13 Draft3.Rmd
small changes

--- 2022-12-12 Draft3.Rmd
@@
 
+## Basic elements
+214
+
+215
+
+`mathml` contains several basic elements used in everyday mathematical expressions. For example it is possible to print the function for the linear regression as:
+216
+
+217
+
+```{r, results='asis'}
+term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*x)
+mathout(term)
+
+# or with greek letters as:
+term <- quote(y == beta["0"] + beta["1"]*x + ... + varepsilon["i"])
+
+mathout(term)
+```
+
+```{r, results='asis'}
+f <- function(x, a)
+{
+    if(a == 0) return(+Inf);
+    if(a == 0) return(-Inf);
+    if(a > 1L ) return(frac(1L,2L))
+}
+
+mathout(f)
+
+```
+
 ## Basic calculations
@@
 if(knitr::is_html_output())
+
 op <- sapply(op, FUN=knitr:::escape_html)
-
 t <- cbind(Operator=op, Output=m)
@@
 op <- as.character(op)
+
 if(knitr::is_latex_output())
@@
 if(knitr::is_html_output())
+
 op <- sapply(op, FUN=knitr:::escape_html)
-
 t <- cbind(Operator=op, Output=m)
+
 knitr::kable(t, caption="Add", row.names=FALSE, escape=FALSE)
@@
 
-Then, we may want underline the incorrect term and provide the correct solution.
-
-```{r, results="asis"}
-# It underlines what has been incorrect
+Then, we may want to underline the incorrect term and provide the correct solution.
+
+```{r, results="asis"}
+# It highlights what has been mistakenly inserted
 mathout(t, flags=quote(error(highlight))) # default
@@
 
-Finally, we can simply provide the correct solution to the problem with the flag `ignore`.
+Finally, we can simply provide the correct solution to the problem wit the flag `ignore`.
 

### 2022-12-13 Draft3.Rmd
small changes

@@
 ## Basic elements
-214
-
-215
-
-`mathml` contains several basic elements used in everyday mathematical expressions. For example it is possible to print the function for the linear regression as:
-216
-
-217
+
+`mathml` contains several basic elements used in everyday mathematical expressions. 
+For example it is possible to print the function for the linear regression as:
 
@@
 
+_I also wanted to add a condition like 1 < a < 2, but apparently it's not possible_ 
+It is also possible to represent a system of equations with different relations.
+Note that you may also assign a name to the function.
+The default name is in fact the greek letter lambda, but with `attributes()$name` the name costumization is performed smoothly.
+
 ```{r, results='asis'}
@@
 {
-    if(a == 0) return(+Inf);
-    if(a == 0) return(-Inf);
-    if(a > 1L ) return(frac(1L,2L))
+    if(a == 0L) return( 0L );
+    if(a <= 1L) return(-Inf);
+    if(a > 1L) return(+Inf)
 }
 
+mathout(f)
+
+attributes(f)$name = quote(f)
 mathout(f)

### 2022-12-13 Draft3.Rmd
latex/cancel

@@
 bibliography: RJreferences.bib
-preamble: \usepackage{longtable}
+preamble:
+- \usepackage{longtable}
+- \usepackage{cancel}
 ---
@@
 op <- list(left=quote(omit_left(a + b)), right=quote(omit_right(a + b)), list=quote(omit(c)))
-m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
+m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))
 
@@
 if(knitr::is_html_output())
-
   op <- sapply(op, FUN=knitr:::escape_html)
+
 t <- cbind(Operator=op, Output=m)

### 2022-12-13 Draft3.Rmd
latex/box

@@
 op <- list(left=quote(add_left(a + b)), right=quote(add_right(a + b)))
-m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
+m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))
 
 op <- as.character(op)
-
 if(knitr::is_latex_output())
@@
 if(knitr::is_html_output())
-
   op <- sapply(op, FUN=knitr:::escape_html)

### 2022-12-14 Draft3.Rmd
table 1 table 2

@@
 
-`mathml` contains several basic elements used in everyday mathematical expressions. 
-For example it is possible to print the function for the linear regression as:
+\pkg{mathml} contains several basic elements used in everyday mathematical expressions, such as superscripts, subscripts, hat, relation and set signs, accents and arrows.
+Below we provide some examples of their application with commented code.
+The other basic elements will be displayed in Table \@ref(tab:basic-elements) and Table \@ref(tab:arrows).
+
+Let us start with the function for the linear regression. As the reader may notice, \pkg{mathml} displays Greek letters, accents and superscripts.
 
 ```{r, results='asis'}
-term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*x)
+term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*x["1"])
 mathout(term)
@@
 # or with greek letters as:
-term <- quote(y == beta["0"] + beta["1"]*x + ... + varepsilon["i"])
+term <- quote(y == beta["0"] + beta["1"]*x["1"] + beta["2"]*x["2"] + ... + beta["i"]*x["i"] + varepsilon["i"])
 
@@
 
-_I also wanted to add a condition like 1 < a < 2, but apparently it's not possible_ 
-It is also possible to represent a system of equations with different relations.
-Note that you may also assign a name to the function.
-The default name is in fact the greek letter lambda, but with `attributes()$name` the name costumization is performed smoothly.
+The second example may be particularly of use when we deal with multiple conditions (see \@ref(case-study)). Here, for instance, we present a trivial one with different relation signs.
 
@@
 mathout(f)
-
+```
+
+Note that you may also assign a name to the function. The default name is in fact the greek letter lambda, but with `attributes()$name` the name costumization is performed smoothly.
+
+```{r, results='asis'}
 attributes(f)$name = quote(f)
 mathout(f)
-
+```
+
+```{r basic-elements, echo=FALSE}
+op <- list(subsup=quote(x["a"]^b), lchoose=quote(lchoose(n, k)), sqrt=quote(sqrt(x)), not=quote(!x), different=quote(x != y), mean=quote(mean(xy)), absolute=quote(abs(x)), tilde=quote(X ~ Y), approx=quote(x %~~% y), equiv=quote(x %==% y), congr=quote(x %=~% y), prop=quote(x %prop% y), 'in'=quote(a %in% A), intersect=quote(intersect(A, B)), union=quote(union(A, B)), intersect=quote(is.null(intersect(A, B))))
+m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
+
+op <- as.character(op)
+if(knitr::is_latex_output())
+  op <- sapply(op, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op <- sapply(op, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op, Output=m)
+knitr::kable(t, caption="Basic elements in mathml", row.names=FALSE, escape=FALSE)
+```
+
+```{r arrows, echo=FALSE}
+op <- list(twosidearrow=quote(x%<->%b), rightarrow=quote(x%->%y), leftarrow=quote(x%<-%y), uparrow=quote(x%up%y), down=quote(x%down%y), iff=quote(x%<=>%y), rArr=quote(x %=>% y), lArr=quote(x %<=% y), dblup=quote(x %dblup% y), dbldown=quote(x %dbldown% y))
+m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
+
+op <- as.character(op)
+if(knitr::is_latex_output())
+  op <- sapply(op, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op <- sapply(op, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op, Output=m)
+knitr::kable(t, caption="Arrows in mathml", row.names=FALSE, escape=FALSE)
 ```
@@
 
-## Subscripts and Superscripts
-Secondly, \pkg{mathml} allows also for the display of subscripts, superscripts and their combination.
-
-```{r, results="asis", echo=TRUE}
-catmath(a["x"])
-
-catmath(a^x)
-
-catmath(x["b"]^"a")
-```
-
-## Fractions
-As in [@murrell2000] the `frac` and `over` functions aim to represent fractions. `choose` and `lchoose`, as in R\ \pkg{base}, display the binomial coefficient.
-
-```{r, results="asis", echo=TRUE}
-catmath(frac(x, y))
-
-catmath(over(x, y))
-
-catmath(choose(n, k))
-
-catmath(lchoose(n, k))
-```
-
-## Radicals
-
-```{r, results="asis", echo=TRUE}
-catmath(sqrt(x))
-```
-
-## Relations
-
-Among the possible relational operators, following [@murrell2000], we included:
-
-```{r, results="asis", echo=TRUE}
-catmath(x==y)
-
-catmath(x < y)
-
-catmath(x <= y)
-
-catmath(x >= y)
-
-catmath(!x)
-
-catmath(x != y)
-
-catmath(X ~ Y)
-
-catmath(x %~~% y)
-
-catmath(x %==% y)
-
-catmath(x %=~% y)
-
-catmath(x %prop% y)
-```
-
-## Set relations
-
-Here is provided a series of useful set relations.
-
-_Still to be implemented_
-
-```{r, results="asis", echo=TRUE}
-catmath(a %in% A)
-
-catmath(intersect(A, B))
-
-catmath(union(A, B))
-
-catmath(is.null(intersect(A, B)))
-```
-
-## Arrows
-
-Following once again [@murrell2000], we provide here a list of useful arrows
-
-```{r, results="asis", echo=TRUE}
-# x %<->% y
-catmath(x %<->% y)
-
-# x %->% y
-catmath(x %->% y)
-
-# x %<-% y
-catmath(x %<-% y)
-
-# x %up% y:
-catmath(x %up% y)
-
-# x %down% y:
-catmath(x %down% y)
-
-# x %<=>% y: solved it using iff
-catmath(x %<=>% y)
-
-# x %=>% y: solved using rArr
-catmath(x %=>% y)
-
-# x %<=% y: solved using lArr
-catmath(x %<=% y)
-
-# x %dblup% y:
-catmath(x %dblup% y)
-
-# x %dbldown% y:
-catmath(x %dbldown% y)
-```
-
-## Accents
-
-```{r, results="asis", echo=TRUE}
-catmath(hat(x))
-
-catmath(hat(xy))
-
-catmath(mean(x))
-
-catmath(mean(xy))
-```
-
-## Symbolic names
-
-```{r, results="asis", echo=TRUE}
-catmath(A-Omega)
-
-catmath(alpha-omega)
-
-catmath(pi-Pi)
-
-catmath(Inf)
-```
-
 ## Operators
@@
 catmath(x^{y + z})
-```
-
-## Absolute value
-
-Is it also possible to calculate and represent the absolute value, as:
-
-```{r, results='asis', echo=TRUE}
-catmath(abs(x))
 ```
@@
 
-# A case study
+# A case study {#case-study}
 

### 2022-12-14 inst/pl/mathml.pl
sum with index

@@ -124,25 +124,26 @@ mathjax(R) :-
 % Summation sign from to (subscript and superscript)
 math(Flags, Sum, New, M),
     compound(Sum),
-    compound_name_arguments(Sum, sum, Arg),
+    compound_name_arguments(Sum, sum, [Arg]),
     select(subscript(From), Flags, New0),
     select(superscript(To), New0, New1)
  => New = New1,
-    M = fn(subsupscript(sum, From, To), Arg).
+    M = fn(subsupscript(sum, From, To), [Arg]).
 
 % Summation sign from (only subscript)
 math(Flags, Sum, New, M),
     compound(Sum),
-    compound_name_arguments(Sum, sum, Arg),
+    compound_name_arguments(Sum, sum, [Arg]),
     select(subscript(Idx), Flags, New0)
  => New = New0,
-    M = fn(subscript(sum, Idx), Arg).
+    M = fn(subscript(sum, Idx), [Arg]).
 
 math(Flags, Sum, New, M),
     compound(Sum),
-    compound_name_arguments(Sum, sum, Arg)
+    compound_name_arguments(Sum, sum, [Arg]),
+    type(Flags, Arg, subscript(Idx))
  => New = Flags,
-    M = fn(sum, Arg).
+    M = fn(subscript(sum, Idx), [Arg]).
 
 mathml :-
     mathml(sum('['(x, i))).
@@ -207,6 +208,11 @@ jax(Flags, R, M),
     jax(New, Pwr, Z),
     format(string(M), "{~w_~w^~w}", [X, Y, Z]).
 
+type(Flags, _R, Type),
+    member(subscript(Idx), Flags),
+    member(superscript(Pwr), Flags)
+ => Type = subsupscript(Idx, Pwr).
+
 mathml :-
     mathml(subsupscript(x, i, 2)).
 
@@ -232,6 +238,10 @@ jax(Flags, R, M),
  => jax(New, R, X),
     jax(New, Idx, Y),
     format(string(M), "{~w_~w}", [X, Y]).
+
+type(Flags, _R, Type),
+    member(subscript(Idx), Flags)
+ => Type = subscript(Idx).
 
 mathml :-
     mathml(subscript(x, i)).
@@ -273,6 +283,10 @@ jax(Flags, R, M),
  => jax(New, R, X),
     jax(New, Pwr, Y),
     format(string(M), "{{~w}^{~w}}", [X, Y]).
+
+type(Flags, _R, Type),
+    member(superscript(Pwr), Flags)
[…]

### 2022-12-14 Draft3.Rmd
sum with index
new table

@@
 ```{r, results='asis'}
-term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*x["1"])
+term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*X["1"])
 mathout(term)
@@
 # or with greek letters as:
-term <- quote(y == beta["0"] + beta["1"]*x["1"] + beta["2"]*x["2"] + ... + beta["i"]*x["i"] + varepsilon["i"])
+term <- quote(y== beta["0"] + beta["1"]*x["1"] + beta["2"]*x["2"] + ... + beta["i"]*x["i"] + varepsilon["i"])
 
@@
 
+op2 <- list(twosidearrow=quote(x%<->%b), rightarrow=quote(x%->%y), leftarrow=quote(x%<-%y), uparrow=quote(x%up%y), down=quote(x%down%y), iff=quote(x%<=>%y), rArr=quote(x %=>% y), lArr=quote(x %<=% y), dblup=quote(x %dblup% y), dbldown=quote(x %dbldown% y))
+m2 <- lapply(op2, FUN=mathjax, flags=list(cat=FALSE))
+
+op2 <- as.character(op2)
+if(knitr::is_latex_output())
+  op2 <- sapply(op2, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op2 <- sapply(op2, FUN=knitr:::escape_html)
+
 op <- as.character(op)
@@
 
-t <- cbind(Operator=op, Output=m)
+t <- cbind(Operator=op, Output=m, Operator=op2, Output=m2)
 knitr::kable(t, caption="Basic elements in mathml", row.names=FALSE, escape=FALSE)
-```
-
-```{r arrows, echo=FALSE}
-op <- list(twosidearrow=quote(x%<->%b), rightarrow=quote(x%->%y), leftarrow=quote(x%<-%y), uparrow=quote(x%up%y), down=quote(x%down%y), iff=quote(x%<=>%y), rArr=quote(x %=>% y), lArr=quote(x %<=% y), dblup=quote(x %dblup% y), dbldown=quote(x %dbldown% y))
-m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
-
-op <- as.character(op)
-if(knitr::is_latex_output())
-  op <- sapply(op, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op <- sapply(op, FUN=knitr:::escape_html)
-
-t <- cbind(Operator=op, Output=m)
-knitr::kable(t, caption="Arrows in mathml", row.names=FALSE, escape=FALSE)
 ```
@@
 # fixme: braces not needed when right to *
-catmath(a * sum(b))
+catmath(a * sum(b[i]))
 
 # Multiplication (capital Pi) is missing
-catmath(a * prod(b))
+catmath(a * prod(b[i]))
 

### 2022-12-15 Draft3.Rmd
small changes Table 2

@@
 op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
-  'integer division'=quote(x %/% y))
+  'integer division'=quote(x %/% y), and=quote(A & B), vel=quote(A | B), xor=quote(xor(A, B)), not=quote(!A))
 m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))

### 2022-12-17 Draft3.Rmd
function bodies

@@
 As is seen from the output above, the variable `term` is not assigned the result
-of the calculation, but an R\ "call" that can eventually be evaluated
-[see, e.g., @Wickham2019, for more details on "non-standard evaluation"].
+of the calculation, but an R\ "call"
+[see, e.g., @Wickham2019, for more details on "non-standard evaluation"]. This
+call can eventually be evaluated with `eval(...)`,
 
@@
 We can include the output in a RMarkdown document by specifying `results='asis'`
-in the R code chunk, as has been done below. The R function mathout is just a
-wrapper that invokes `mathml(...)` for HTML output and `mathjax(...)` for
-LaTeX output.
+in the R code chunk, as is done in the next example. The
+R\ function `mathout(...)` is just a wrapper that invokes `mathml(...)` for HTML
+output and `mathjax(...)` for LaTeX output.
 
@@
 mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
-element, namely `fn(..., ...)`.
-
-The low-level predicates `ml/3` and `jax/3` are used to convert mathematical
-elements to MathML or LaTeX, respectively.
+element, namely `fn(..., ...)`. The low-level predicates `ml/3` and `jax/3`
+convert mathematical elements to MathML or LaTeX, respectively.
 
@@
 R\ functions and objects in data analysis, scientific reports and interactive
-web content. The features currently supported are listed below, roughly following
-the order proposed by [@murrell2000].
+web content. The currently supported features are listed below, roughly
+following the order proposed by [@murrell2000].
 
@@
 
-\pkg{mathml} contains several basic elements used in everyday mathematical expressions, such as superscripts, subscripts, hat, relation and set signs, accents and arrows.
-Below we provide some examples of their application with commented code.
-The other basic elements will be displayed in Table \@ref(tab:basic-elements) and Table \@ref(tab:arrows).
+\pkg{mathml} contains several basic elements used in everyday mathematical
+expressions, such as superscripts, subscripts, hat, relation and set signs,
+accents and arrows. Below we provide some examples of their application with
+commented code. The other basic elements will be displayed in
+Table \@ref(tab:basic-elements) and Table \@ref(tab:arrows).
 

### 2022-12-17 R/mathml.R

@@ -69,7 +69,7 @@ mathml_preproc <- function(query=quote(2 != 2))
 
     index <- which(args[[1]] == names(mathml_operators))
     if(length(index) == 1)
-      args[[1]] <- mathml_operators[index]
+      args[[1]] <- as.name(mathml_operators[index])
 
     args[-1] <- lapply(args[-1], FUN=mathml_preproc)
     return(as.call(args))

### 2022-12-17 inst/pl/mathml.pl

@@ -787,12 +787,19 @@ math(Curly, M),
     compound_name_arguments(Curly, '{', Args)
  => M = body(Args).
 
-math(body([R]), M)
- => M = R.
+ml(Flags, body([R]), M)
+ => ml(Flags, R, M).
 
-math(body(Body), M)
- => findall([A], member(A, Body), List),
-    M = rbind(List).
+ml(Flags, body(Body), M)
+ => maplist(ml(Flags), Body, R),
+    M = mrow([mo('{'), mtable(columnalign(left), R)]).
+
+jax(Flags, body([R]), M)
+ => jax(Flags, R, M).
+
+jax(Flags, body(Body), M)
+ => maplist(jax(Flags), Body, Rows),
+    format(string(M), "\\left\\{\\begin{array}{l}~w\\end{array}\\right.", [Rows]).
 
 % Vectors
 math(Hash, M),

### 2022-12-18 Draft3.Rmd
support for multiple indices

@@
 LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
-rules, for example, to represent specific R elements by custom mathematical signs.
-Moreover, in addition to graphical rendering, the same R expression provides its calculation or evaluation.
-This feature of the package saves the user's time and allows faster detection of errors.
+rules, for example, to represent specific R elements by custom mathematical
+signs.
+
+The main feature of the package that the same R expressions and equations are
+used for both mathematical typesetting and calculations. This saves time and
+reduces mistakes, as well be illustrated below.
 
@@
 of the calculation, but an R\ "call"
-[see, e.g., @Wickham2019, for more details on "non-standard evaluation"]. This
-call can eventually be evaluated with `eval(...)`,
+[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
+can eventually be evaluated with `eval(...)`,
 
@@
 
-The R package \pkg{mathml} can now be used to render the call in
-MathML, that is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
+The R package \pkg{mathml} can now be used to render the call in MathML, that
+is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
 below (some of the curly braces are not really needed in this simple example,
@@
 \pkg{mathml} contains several basic elements used in everyday mathematical
-expressions, such as superscripts, subscripts, hat, relation and set signs,
-accents and arrows. Below we provide some examples of their application with
-commented code. The other basic elements will be displayed in
-Table \@ref(tab:basic-elements) and Table \@ref(tab:arrows).
-
-Let us start with the function for the linear regression. As the reader may notice, \pkg{mathml} displays Greek letters, accents and superscripts.
+expressions, such as numbers, Latin and Greek letters, multi-letter identifiers,
+accents, subscripts, and superscripts.
 
 ```{r, results='asis'}
-term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*X["1"])
+term <- quote(1 + 2L + "a" + a + abc + phi + Phi + varphi + hat(b)[i, j]^2L
+              + NaN + NA + TRUE + FALSE + Inf + -Inf)
 mathout(term)
-
-# or with greek letters as:
-term <- quote(y== beta["0"] + beta["1"]*x["1"] + beta["2"]*x["2"] + ... + beta["i"]*x["i"] + varepsilon["i"])
-
+```
+
+todo: parentheses around -infty, see also $(no)^{p+q}$ below.
+
+## Operators and parentheses
+
+Basic arithmetic operators and parentheses are translated as they are, as
+illustrated below.
+
+```{r, results="asis"}
+term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + {n*o}^{p + q})
 mathout(term)
 ```
+
+For multiplications involving only numbers and symbols, the multiplication sign
+is omitted. This heuristic does not always produce the desired result;
+therefore, \pkg{mathml} defines the extra functions `dot/2`, `nodot/2`,
+and `times/2` that produce the respective multiplication signs.
+
+```{r, results="asis"}
+term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
+mathout(term)
+```
+
+Note that for the standard operators with known precedence, \pkg{mathml} is able
+to detect if parentheses are needed; for example, a parenthesis is automatically
+put around `g + h` in the `nodot`-example and around `n*o` above. However, we
+also note an unecessary parenthesis around `l + m` above. This parenthesis
+results is a consequence of `quote(a^(b + c))` actually producing a nested
+R\ call of the form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:
+
+```{r}
+term <- quote(a^(b + c))
+paste(term)
+```
+
+This feature is unfortunate because the extra parenthesis around `b + c` is not
+needed and yields to the extra parenthesis. The preferred result is obtained by
+using the functional representation `quote('^'(k, l + m))` of the power, or, as
+a workaround, curly braces (see last summand above).
+
+## Custom operators
+
+While the parentheses typically follow the precedence in standard infix
+operators, undesirable results may be obtained in custom operators.
+
+```{r, results='asis'}
+term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
+mathout(term)
+```
+
+As can be seen, \pkg{mathml} reminds the user that it is not possible to change
+the precedence of custom operators in R, and that expressions with such
+operators are evaluated strictly from left to right. As in the
+previous example, the solution is to either use the functional form of the
+operator or a curly brace to enforce the correct operator precedence.
+
+```{r, results='asis'}
+term <- quote(`%+-%`(mean(X), 1.96 * s / sqrt(N)))
+term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
+mathout(term)
+```
+
+More operators are shown in Table \@ref(tab:custom-operators).
+
+```{r custom-operators, echo=FALSE}
+op <- list(
+  cross=quote(x %*% y),
+  dot=quote(x %.% y),
+  kronecker=quote(X %x% Y),
+  'integer division'=quote(x %/% y),
+  and=quote(A & B),
+  vel=quote(A | B),
+  xor=quote(xor(A, B)),
+  not=quote(!A),
+  equal=quote(X == Y),
+  assign=quote(X <- Y),
+  different=quote(x != y),
+  tilde=quote(X ~ Y),
+  approx=quote(x %~~% y),
+  equiv=quote(x %==% y),
+  congr=quote(x %=~% y),
+  prop=quote(x %prop% y),
+  'in'=quote(a %in% A),
+  intersect=quote(intersect(A, B)),
+  union=quote(union(A, B)),
+  intersect=quote(is.null(intersect(A, B)))) # todo: check if duplicate
+
+m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
+
+op <- as.character(op)
+if(knitr::is_latex_output())
+  op <- sapply(op, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op <- sapply(op, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op, Output=m)
+knitr::kable(t, caption="Custom operators in mathml", row.names=FALSE, escape=FALSE)
+```
+
+## R functions
 
@@
 
+A few more elements are displayed in
+Tables \@ref(tab:basic-elements) and \@ref(tab:arrows).
+
 ```{r basic-elements, echo=FALSE}
-op <- list(subsup=quote(x["a"]^b), lchoose=quote(lchoose(n, k)), sqrt=quote(sqrt(x)), not=quote(!x), different=quote(x != y), mean=quote(mean(xy)), absolute=quote(abs(x)), tilde=quote(X ~ Y), approx=quote(x %~~% y), equiv=quote(x %==% y), congr=quote(x %=~% y), prop=quote(x %prop% y), 'in'=quote(a %in% A), intersect=quote(intersect(A, B)), union=quote(union(A, B)), intersect=quote(is.null(intersect(A, B))))
-m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
-
-op2 <- list(twosidearrow=quote(x%<->%b), rightarrow=quote(x%->%y), leftarrow=quote(x%<-%y), uparrow=quote(x%up%y), down=quote(x%down%y), iff=quote(x%<=>%y), rArr=quote(x %=>% y), lArr=quote(x %<=% y), dblup=quote(x %dblup% y), dbldown=quote(x %dbldown% y))
-m2 <- lapply(op2, FUN=mathjax, flags=list(cat=FALSE))
+op1 <- list(
+  lchoose=quote(lchoose(n, k)),
+  sqrt=quote(sqrt(x)),
+  mean=quote(mean(xy)),
+  absolute=quote(abs(x))) # todo: check if duplicate
+m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))
+
+op1 <- as.character(op1)
+if(knitr::is_latex_output())
+  op1 <- sapply(op1, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op1 <- sapply(op1, FUN=knitr:::escape_html)
+
+op2 <- list(
+  twosidearrow=quote(x%<->%b),
+  rightarrow=quote(x%->%y),
+  leftarrow=quote(x%<-%y),
+  uparrow=quote(x%up%y),
+  down=quote(x%down%y),
+  iff=quote(x%<=>%y),
+  rArr=quote(x %=>% y),
+  lArr=quote(x %<=% y),
+  dblup=quote(x %dblup% y),
+  dbldown=quote(x %dbldown% y))
+m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))
 
@@
 
-op <- as.character(op)
-if(knitr::is_latex_output())
-  op <- sapply(op, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op <- sapply(op, FUN=knitr:::escape_html)
-
-t <- cbind(Operator=op, Output=m, Operator=op2, Output=m2)
+t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2)
 knitr::kable(t, caption="Basic elements in mathml", row.names=FALSE, escape=FALSE)
-```
-
-## Basic calculations
-
-Basic arithmetic operators and parentheses are translated as they are, as
-illustrated below.
-
-```{r, results="asis"}
-term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + {n*o}^{p + q})
-mathout(term)
-```
-
-For multiplications involving only numbers and symbols, the multiplication sign
-is omitted. This heuristic does not always produce the desired result;
-therefore, \pkg{mathml} defines the extra functions `dot/2`, `nodot/2`,
-and `times/2` that produce the respective multiplication signs.
-
-```{r, results="asis"}
-term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
-mathout(term)
-```
-
-Note that for the standard operators with known precedence, \pkg{mathml} is able
-to detect if parentheses are needed; for example, a parenthesis is automatically
-put around `g + h` in the `nodot`-example and around `n*o` above. However, we
-also note an unecessary parenthesis around `l + m` above. This parenthesis
-results is a consequence of `quote(a^(b + c))` actually producing a nested
-R\ call of the form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:
-
-```{r}
-term <- quote(a^(b + c))
-paste(term)
-```
-
-This feature is unfortunate because the extra parenthesis around `b + c` is not
-needed and yields to the extra parenthesis. The preferred result is obtained by
-using the functional representation `quote('^'(k, l + m))` of the power, or, as
-a workaround, curly braces (see last summand above).
-
-## Custom operators
-
-While the parentheses typically follow the precedence in standard infix
-operators, undesirable results may be obtained in custom operators.
-
-```{r, results='asis'}
-term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
-mathout(term)
-```
-
-As can be seen, \pkg{mathml} reminds the user that it is not possible to change
-the precedence of custom operators in R, and that expressions with such
-operators are evaluated strictly from left to right. As in the
-previous example, the solution is to either use the functional form of the
-operator or a curly brace to enforce the correct operator precedence.
-
-```{r, results='asis'}
-term <- quote(`%+-%`(mean(X), 1.96 * s / sqrt(N)))
-term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
-mathout(term)
-```
-
-A few more custom operators are shown in Table \@ref(tab:custom-operators).
-
-```{r custom-operators, echo=FALSE}
-op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
-  'integer division'=quote(x %/% y), and=quote(A & B), vel=quote(A | B), xor=quote(xor(A, B)), not=quote(!A))
-m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
-
-op <- as.character(op)
-if(knitr::is_latex_output())
-  op <- sapply(op, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op <- sapply(op, FUN=knitr:::escape_html)
-
-t <- cbind(Operator=op, Output=m)
-knitr::kable(t, caption="Custom operators in mathml", row.names=FALSE, escape=FALSE)
 ```

### 2022-12-18 inst/pl/mathml.pl

@@ -219,8 +219,10 @@ mathml :-
 %
 % Subscript like s_D
 %
-math('['(R, Idx), M)
- => M = subscript(R, Idx).
+math(Sub, M),
+    compound(Sub),
+    compound_name_arguments(Sub, '[', [R | Idx])
+ => M = subscript(R, list("", Idx)).
 
 math(Flags, subscript(R, Idx), New, M)
  => New = [subscript(Idx) | Flags],

### 2022-12-19 Draft3.Rmd
R\ : added backslash after R

@@
 
-The R extension of the markdown language [@Xie2020] enables reproducible
+The R\ extension of the markdown language [@Xie2020] enables reproducible
 statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
-Since recently [@R, version 4.2], R's manual pages include support for
+Since recently [@R, version 4.2], R\'s manual pages include support for
 mathematical expressions [@Sarkar2022], which already is a big improvement.
-However, rules for the translation of R's built-in language elements to their
-mathematical representation are still lacking. So far, R expressions such
+However, rules for the translation of R\'s built-in language elements to their
+mathematical representation are still lacking. So far, R\ expressions such
 as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
@@
 
-The present R package defines a set of rules for an _automatic_ translation of
-R expressions to mathematical output in RMarkdown documents [@Xie2020] and
+The present R\ package defines a set of rules for an _automatic_ translation of
+R\ expressions to mathematical output in RMarkdown documents [@Xie2020] and
 ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
@@
 LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
-rules, for example, to represent specific R elements by custom mathematical
+rules, for example, to represent specific R\ elements by custom mathematical
 signs.
 
-The main feature of the package that the same R expressions and equations are
+The main feature of the package that the same R\ expressions and equations are
 used for both mathematical typesetting and calculations. This saves time and
@@
 The paper is organized as follows. Section\ 2 presents the background of the
-package, including Prolog and the R package \pkg{rolog} and the two main classes
+package, including Prolog and the R\ package \pkg{rolog} and the two main classes
 of rules for translating R\ objects to mathematical expressions. Section\ 3
@@
 scientific report and a ShinyApp web page. Section\ 6 concludes with a
-discussion and ideas for further development. A complete list of the R elements
+discussion and ideas for further development. A complete list of the R\ elements
 covered by the package is found in the Appendix.
@@
 
-The translation of R expressions to mathematical output is achieved through a
+The translation of R\ expressions to mathematical output is achieved through a
 Prolog interpreter. Prolog is a classical logic programming language with many
@@
 in statistical computation, but strong in symbolic manipulation, the converse
-may be said for the R language. The \pkg{rolog} package bridges this gap by
+may be said for the R\ language. The \pkg{rolog} package bridges this gap by
 providing an interface to a SWI-Prolog distribution in R. The communication
@@
 
-The R package \pkg{mathml} can now be used to render the call in MathML, that
+The R\ package \pkg{mathml} can now be used to render the call in MathML, that
 is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
@@
 We can include the output in a RMarkdown document by specifying `results='asis'`
-in the R code chunk, as is done in the next example. The
+in the R\code chunk, as is done in the next example. The
 R\ function `mathout(...)` is just a wrapper that invokes `mathml(...)` for HTML
@@
 As can be seen, \pkg{mathml} reminds the user that it is not possible to change
-the precedence of custom operators in R, and that expressions with such
+the precedence of custom operators in R\, and that expressions with such
 operators are evaluated strictly from left to right. As in the
@@
 
-R's `integrate` function takes a number of arguments, the most important ones
+R\'s `integrate` function takes a number of arguments, the most important ones
 being the function to integrate, and the lower and the upper bound of the
@@
 variable. For that purpose, the underlying Prolog bridge provides a
-predicate `r_eval/2` that calls R from Prolog. In the example below, this
+predicate `r_eval/2` that calls R\ from Prolog. In the example below, this
 predicate is used for a call of `formalArgs(args(sin))`, which returns the names
@@
 
-The extra R function `canonical()` applies `match.call()` to non-primitive R
+The extra R\ function `canonical()` applies `match.call()` to non-primitive R\
 calls, basically cleaning up the arguments and bringing them into the correct
@@
 An interesting use case of the package is in this sense the readability of self-written
-functions in R code. R code is notoriously hard to read, as are most programming
+functions in R\ code. R\ code is notoriously hard to read, as are most programming
 languages [@green1977]. The poor legibility of the language is one of the main

### 2022-12-21 Draft3.Rmd
parentheses around inf + (-inf)

@@
 ```{r, results='asis'}
-term <- quote(1 + 2L + "a" + a + abc + phi + Phi + varphi + hat(b)[i, j]^2L
-              + NaN + NA + TRUE + FALSE + Inf + -Inf)
+term <- quote(1 + 2L + "a" + a + abc + phi + Phi + varphi + hat(b)[i, j]^2L)
 mathout(term)
-```
-
-todo: parentheses around -infty, see also $(no)^{p+q}$ below.
+
+term <- quote(NaN + NA + TRUE + FALSE + Inf + -Inf)
+mathout(term)
+```
 
@@
 ```
+
+todo: parentheses around -infty, see also $(no)^{p+q}$ below.
 

### 2022-12-21 Draft3.Rmd
operators and functions

@@
 ```{r, results="asis"}
-term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + {n*o}^{p + q})
+term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
 mathout(term)
 ```
-
-todo: parentheses around -infty, see also $(no)^{p+q}$ below.
 
@@
 to detect if parentheses are needed; for example, a parenthesis is automatically
-put around `g + h` in the `nodot`-example and around `n*o` above. However, we
-also note an unecessary parenthesis around `l + m` above. This parenthesis
-results is a consequence of `quote(a^(b + c))` actually producing a nested
-R\ call of the form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:
+put around `g + h` in the `nodot`-example. However, we also note an unecessary
+parenthesis around `l + m` above. This parenthesis is a consequence
+of `quote(a^(b + c))` actually producing a nested R\ call of the
+form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:
 
@@
 needed and yields to the extra parenthesis. The preferred result is obtained by
-using the functional representation `quote('^'(k, l + m))` of the power, or, as
-a workaround, curly braces (see last summand above).
+using the functional representation `quote('^'(k + l, m + n))` of the power, or,
+as a workaround, curly braces (see last summand above).
 
@@
 
-As can be seen, \pkg{mathml} reminds the user that it is not possible to change
-the precedence of custom operators in R\, and that expressions with such
-operators are evaluated strictly from left to right. As in the
-previous example, the solution is to either use the functional form of the
-operator or a curly brace to enforce the correct operator precedence.
+This is a reminder that it is not possible to change the precedence of custom
+operators in R\, and that expressions with such operators are evaluated strictly
+from left to right. As in the previous example, the solution is to either use
+the functional form of the operator or a curly brace to enforce the correct
+operator precedence.
 
@@
 
-More operators are shown in Table \@ref(tab:custom-operators).
+More operators are shown in Table \@ref(tab:custom-operators), following the
+suggestions by Murrell and Ihaka [-@murrell2000].
 
 ```{r custom-operators, echo=FALSE}
-op <- list(
+op1 <- list(
   cross=quote(x %*% y),
@@
   equal=quote(X == Y),
-  assign=quote(X <- Y),
+  assign=quote(X <- Y))
+
+m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))
+
+op1 <- as.character(op1)
+if(knitr::is_latex_output())
+  op1 <- sapply(op1, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op1 <- sapply(op1, FUN=knitr:::escape_html)
+
+op2 <- list(
   different=quote(x != y),
@@
   union=quote(union(A, B)),
-  intersect=quote(is.null(intersect(A, B)))) # todo: check if duplicate
-
-m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
-
-op <- as.character(op)
+  intersect=quote(is.null(A)))
+
+m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))
+
+op2 <- as.character(op2)
 if(knitr::is_latex_output())
-  op <- sapply(op, FUN=knitr:::escape_latex)
+  op2 <- sapply(op2, FUN=knitr:::escape_latex)
 if(knitr::is_html_output())
-  op <- sapply(op, FUN=knitr:::escape_html)
-
-t <- cbind(Operator=op, Output=m)
-knitr::kable(t, caption="Custom operators in mathml", row.names=FALSE, escape=FALSE)
+  op2 <- sapply(op2, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2)
+knitr::kable(t, caption="Custom operators in mathml", 
+             row.names=FALSE, escape=FALSE)
 ```
@@
 
-The second example may be particularly of use when we deal with multiple conditions (see \@ref(case-study)). Here, for instance, we present a trivial one with different relation signs.
-
-```{r, results='asis'}
-f <- function(x, a)
-{
-    if(a == 0L) return( 0L );
-    if(a <= 1L) return(-Inf);
-    if(a > 1L) return(+Inf)
-}
-
-mathout(f)
-```
-
-Note that you may also assign a name to the function. The default name is in fact the greek letter lambda, but with `attributes()$name` the name costumization is performed smoothly.
-
-```{r, results='asis'}
-attributes(f)$name = quote(f)
-mathout(f)
-```
-
-A few more elements are displayed in
-Tables \@ref(tab:basic-elements) and \@ref(tab:arrows).
+All functions from R::base are supported, a subset is shown in 
+Tables \@ref(tab:basic-elements).
 
@@
 op1 <- list(
+  sin=quote(sin(x)),
+  cosh=quote(cosh(x)),
+  tanpi=quote(tanpi(alpha)),
+  log=quote(log(p)),
+  log1p=quote(log1p(x)),
+  exp=quote(exp(x)),
+  expm1=quote(expm1(x)),
   lchoose=quote(lchoose(n, k)),
@@
   mean=quote(mean(xy)),
-  absolute=quote(abs(x))) # todo: check if duplicate
+  absolute=quote(abs(x)))
 m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))
@@
 knitr::kable(t, caption="Basic elements in mathml", row.names=FALSE, escape=FALSE)
+```
+
+For self-written functions, matters are a bit more complicated because the
+underlying R package `rolog` [@rolog] cannot easily access the function name,
+hence a default is chosen.
+
+```{r, results='asis'}
+f <- function(x)
+{
+  if(x == 0) return(0L);
+  if(x <= 0) return(-1L);
+  if(x > 0) return(1L)
+}
+
+mathout(f)
+```
+
+Custom names can be given via attributes.
+
+```{r, results='asis'}
+attributes(f)$name = quote(sgn)
+mathout(f)
 ```

### 2022-12-21 inst/pl/mathml.pl

@@ -11,7 +11,7 @@
 % R interface: Translate R expression to MathML string
 %
 r2mathml(R, S)
- => r2mathml([], R, S).
+=> r2mathml([], R, S).
 
 r2mathml(Flags, R, S)
  => mathml(Flags, R, M),

### 2022-12-21 Draft3.Rmd
aestetics

@@
 
-t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2)
+op3 <- list(
+  twosidearrow=quote(x%<->%b),
+  rightarrow=quote(x%->%y),
+  leftarrow=quote(x%<-%y),
+  uparrow=quote(x%up%y),
+  down=quote(x%down%y),
+  iff=quote(x%<=>%y),
+  rArr=quote(x %=>% y),
+  lArr=quote(x %<=% y),
+  dblup=quote(x %dblup% y),
+  dbldown=quote(x %dbldown% y))
+m3 <- lapply(op3, FUN=mathout, flags=list(cat=FALSE))
+
+op3 <- as.character(op3)
+if(knitr::is_latex_output())
+  op3 <- sapply(op3, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op3 <- sapply(op3, FUN=knitr:::escape_html)
+
+t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2, Operator=op3, Arrow=m3)
 knitr::kable(t, caption="Custom operators in mathml", 
@@
 
-op2 <- list(
-  twosidearrow=quote(x%<->%b),
-  rightarrow=quote(x%->%y),
-  leftarrow=quote(x%<-%y),
-  uparrow=quote(x%up%y),
-  down=quote(x%down%y),
-  iff=quote(x%<=>%y),
-  rArr=quote(x %=>% y),
-  lArr=quote(x %<=% y),
-  dblup=quote(x %dblup% y),
-  dbldown=quote(x %dbldown% y))
-m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))
-
-op2 <- as.character(op2)
-if(knitr::is_latex_output())
-  op2 <- sapply(op2, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op2 <- sapply(op2, FUN=knitr:::escape_html)
-
-t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2)
-knitr::kable(t, caption="Basic elements in mathml", row.names=FALSE, escape=FALSE)
+t <- cbind(Function=op1, Output=m1)
+knitr::kable(t, caption="Arithmetic functions", row.names=FALSE, escape=FALSE)
 ```
@@
 {
-  if(x == 0) return(0L);
-  if(x <= 0) return(-1L);
-  if(x > 0) return(1L)
+  if(x == 0L) return(0L);
+  if(x <= 0L) return(-1L);
+  if(x > 0L) return(1L)
 }

### 2022-12-21 Draft3.Rmd
fix problems with parentheses

@@
 
-term <- quote(NaN + NA + TRUE + FALSE + Inf + -Inf)
+term <- quote(NaN + NA + TRUE + FALSE + Inf + (-Inf))
 mathout(term)
@@
 
-Note that for the standard operators with known precedence, \pkg{mathml} is able
-to detect if parentheses are needed; for example, a parenthesis is automatically
-put around `g + h` in the `nodot`-example. However, we also note an unecessary
-parenthesis around `l + m` above. This parenthesis is a consequence
+For the standard operators with known precedence, \pkg{mathml} is generally able
+to detect if parentheses are needed; for example, parentheses are automatically
+put around `g + h` in the `nodot`-example. However, we note unecessary
+parentheses around `l + m` above. This is parenthesis is a consequence
 of `quote(a^(b + c))` actually producing a nested R\ call of the
@@
 underlying R package `rolog` [@rolog] cannot easily access the function name,
-hence a default is chosen.
+hence a default is chosen. The function name can be given as an attribute.
 
@@
   if(x == 0L) return(0L);
-  if(x <= 0L) return(-1L);
+  if(x < 0L) return(-1L);
   if(x > 0L) return(1L)
@@
 
-mathout(f)
-```
-
-Custom names can be given via attributes.
-
-```{r, results='asis'}
 attributes(f)$name = quote(sgn)
@@
 One of R's great features is the possibility to refer to function arguments by
-by their names, not only by their position in the list of arguments.
-Unfortunately, the prolog-powered handler for `integrate/3` is rather rigid; it
-accepts only three arguments in that particular order and without names, that
+their names, not only by their position in a list of arguments. Unfortunately, 
+the Prolog-powered handler for `integrate/3` is rather rigid; it accepts only
+three arguments in that particular order and without names, that
 is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.
@@
 order, \pkg{mathml} provides an auxilliary function `canonical(f, drop)` that
-reorders the argument list of calls to known functions and, if `drop=TRUE`,
-drops the function names.
+reorders the argument list of calls to known functions and, if `drop=TRUE`, also
+removes the names of the arguments.
 
@@
 
+We can then easily feed mixtures of partially named and positional arguments in
+unusual order into the renderer.
+
 ```{r, echo=TRUE, results='asis'}
-# A custom function
 g <- function(u)
@@
 
-# Mixture of (partially) named and positional arguments in unusual order
-term = call("integrate", lower=-Inf, upper=Inf, g)
+# 
+term <- call("integrate", lower=-Inf, upper=Inf, g)
 attributes(term)$name = quote(g)
@@
 
-# Even more complex
+# 2
 term <- call("*", 2L, call("$", call("integrate", low=-Inf, up=Inf, g), "value"))

### 2022-12-21 inst/pl/mathml.pl

@@ -989,8 +989,8 @@ math(Flags, integrate(Fn, Lower, Upper), New, M),
 math(Flags, integrate(Fn, Lower, Upper), New, M),
     r_eval('is.null'('$'('.GlobalEnv', Fn)), false)
  => Flags = New,
-%    r_eval('['(formalArgs(args('$'('.GlobalEnv', Fn))), 1), Arg1),
-    atom_string(DX, "Arg1"),
+    r_eval('['(formalArgs(args('$'('.GlobalEnv', Fn))), 1), Arg1),
+    atom_string(DX, Arg1),
     M = integrate(fn(Fn, [DX]), Lower, Upper, DX).
 
 math(Flags, integrate(Fn, Lower, Upper), New, M)
@@ -1748,7 +1748,7 @@ math(_Flags, left(_, A), M)
  => M = A.
 
 math(right(Prec, A), M)
- => P is Prec - 1,
+ => P is Prec, % - 1,
     M = left(P, A).
 
 denoting(Flags, left(_, A), D)

### 2022-12-29 Draft3.Rmd
update christmas

@@
 abstract: >
-  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
-  thereby enabling pretty mathematical representation of R\ functions or
-  algorithms in data analysis, scientific reports and interactive web content.
-  R\ code text descriptions and mathematical content already coexist,
-  side-by-side, in the RMarkdown document rendering language. The present
-  package allows to use the same R\ objects for both data analysis and
-  typesetting in documents or web content. By tightening the link between the
-  statistical analysis and its verbal or symbolic representation, this new
-  feature is a futher step towards reproducible science. User-defined hooks
-  allow to extend the package by mapping new variable names and functions to
-  new MathML and LaTeX entities. We provide a few working examples that
-  illustrate the use of the package in a little scientific report and in an
-  interactive web page.
+  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
+  thereby allowing for a pretty mathematical representation of R\ objects and
+  functions in data analyses, scientific reports and interactive web content. In
+  the RMarkdown document rendering language, R\ code and mathematical content
+  already exist side-by-side. The present package allows to use the same
+  R\ objects for both data analysis and typesetting in documents or web content.
+  This tightens the link between the statistical analysis and its verbal
+  description or symbolic representation, which is another step towards
+  reproducible science. User-defined hooks allow to extend the package by
+  mapping specific variables or functions to new MathML and LaTeX entities. A
+  few working examples illustrate the use of the package in a scientific report
+  and in on an interactive web page.
 draft: true
@@
 mathematical expressions [@Sarkar2022], which already is a big improvement.
-However, rules for the translation of R\'s built-in language elements to their
+However, rules for the mapping of built-in language elements to their
 mathematical representation are still lacking. So far, R\ expressions such
 as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
-like $P_\mathrm{Bi}(X \le k; N, p)$ require explicit Latex commands, that
+such as \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
 is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
-these commands are tedious to type in and hard to read in their source code.
-
-The present R\ package defines a set of rules for an _automatic_ translation of
+these commands are tedious to type in and their source code is hard to read.
+
+The present R\ package defines a set of rules for the automatic translation of
 R\ expressions to mathematical output in RMarkdown documents [@Xie2020] and
@@
 
-The main feature of the package that the same R\ expressions and equations are
-used for both mathematical typesetting and calculations. This saves time and
-reduces mistakes, as well be illustrated below.
-
-The paper is organized as follows. Section\ 2 presents the background of the
-package, including Prolog and the R\ package \pkg{rolog} and the two main classes
-of rules for translating R\ objects to mathematical expressions. Section\ 3
-illustrates the main features of the \pkg{mathml} package, potential issues and
-workarounds using examples from the day-to-day perspective of the user.
-Sections\ 4 and\ 5 present two case studies for the use of the package in a
-scientific report and a ShinyApp web page. Section\ 6 concludes with a
-discussion and ideas for further development. A complete list of the R\ elements
-covered by the package is found in the Appendix.
+The main feature of the package is that the same R\ expressions and equations
+can be used for both mathematical typesetting and calculations. This saves time
+and reduces mistakes, as will be illustrated below.
+
+The paper is organized as follows. Section\ 2 presents the technical background
+of the package, including the two main classes of Prolog rules for translating
+R\ objects to mathematical expressions. Section\ 3 illustrates the main features
+of the \pkg{mathml} package, potential issues and workarounds using examples
+from the day-to-day perspective of a user. Sections\ 4 and\ 5 present two case
+studies for the use of the package in a scientific report and a ShinyApp web
+page. Section\ 6 concludes with a discussion and ideas for further development.
+A complete list of the R\ elements covered by the package is found in the
+Appendix.
 
@@
 in statistical computation, but strong in symbolic manipulation, the converse
-may be said for the R\ language. The \pkg{rolog} package bridges this gap by
-providing an interface to a SWI-Prolog distribution in R. The communication
-between the two systems is mainly in the form of queries from R\ to Prolog, but
-two predicates allow Prolog to ring back and evaluate terms in R.
-
-For a first illustration of the package, we consider the binomial probability.
+may be said for the R\ language. The \pkg{rolog} package [@rolog] bridges this
+gap by providing an interface to a SWI-Prolog distribution in R. The
+communication between the two systems is mainly in the form of queries from
+R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms in
+R.
+
+For a first illustration of the \pkg{mathml} package, we consider the binomial
+probability.
 
@@
 
-The term is quoted to avoid its immediate evaluation which would raise an error
-anyway since the variables `k`, `N`, `p` have not yet been defined. Experienced
-readers will remember that the quoted expression is a short form for
+The term is quoted to avoid its immediate evaluation (which would raise an error
+anyway since the variables `k`, `N`, `p` have not yet been defined). Experienced
+readers will remember that the quoted expression above is a short form for
 
@@
 [see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
-can eventually be evaluated with `eval(...)`,
+can eventually be evaluated with `eval()`,
 
@@
 The R\ package \pkg{mathml} can now be used to render the call in MathML, that
-is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
-below (some of the curly braces are not really needed in this simple example,
-but are necessary in edge cases).
+is the dialect for mathematical elements on HTML webpages or in MathJax/LaTeX,
+as shown below (some of the curly braces are not really needed in this simple
+example, but are necessary in edge cases).
 
@@
 We can include the output in a RMarkdown document by specifying `results='asis'`
-in the R\code chunk, as is done in the next example. The
-R\ function `mathout(...)` is just a wrapper that invokes `mathml(...)` for HTML
-output and `mathjax(...)` for LaTeX output.
+in the R\ code chunk, as is done in the next example. The
+R\ function `mathout()` is a wrapper that invokes `mathml()` for HTML output
+and `mathjax()` for LaTeX output.
 
@@
 
-At the Prolog end, there is a predicate `math/2` that translates the
-call `pbinom(K, N, Pi)` into a "function" `fn/2` with the name `P_Bi`, one
-argument `X =< K`, and the two parameters `N` and `Pi`.
+At the Prolog end, a predicate `math/2` translates the call `pbinom(K, N, Pi)`
+into a "function" `fn/2` with the name `P_Bi`, one argument `X =< K`, and the
+two parameters `N` and `Pi`.
 
@@
 mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
-element, namely `fn(..., ...)`. The low-level predicates `ml/3` and `jax/3`
-convert mathematical elements to MathML or LaTeX, respectively.
+element, namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is
+used to convert these basic elements to MathML.
 
 ```prolog
-ml(Flags, fn(Name, (Args ; Params)), M)
+ml(Flags, fn(Name, (Args ; Pars)), M)
  => ml(Flags, Name, N),
-    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Params)])), X),
+    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X),
     M = mrow([N, mo(&(af)), X]).
@@
 The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
-with `N` representing the function's name and `X` its arguments and parameters,
-enclosed in parentheses. The flags can be used for context-sensitive
+with `N` representing the name of the function and `X` its arguments and
+parameters, enclosed in parentheses. A corresponding rule `jax/3` does the same
+for MathJax/LaTeX. A list of flags can be used for context-sensitive
 translation (see, e.g., the errors below).
@@
 
-\pkg{mathml} contains several basic elements used in everyday mathematical
-expressions, such as numbers, Latin and Greek letters, multi-letter identifiers,
-accents, subscripts, and superscripts.
+\pkg{mathml} handles the basic elements of everyday mathematical expressions,
+such as numbers, Latin and Greek letters, multi-letter identifiers, accents,
+subscripts, and superscripts.
 
 ```{r, results='asis'}
-term <- quote(1 + 2L + "a" + a + abc + phi + Phi + varphi + hat(b)[i, j]^2L)
+term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + hat(b)[i, j]^2L)
 mathout(term)
@@
 
+An expression such as `1 + -2` may be considered unsatisfactory from an
+aesthetical perspective. It is correct R\ syntax, though, and is reproduced
+accordingly, without the parentheses. Parentheses around negated numbers or
+symbols can be added as shown for `-Inf`.
+
+Note that an R function `hat()` does not exist in base R, it is provided by the
+package for convenience and points to the identity function.
+
 ## Operators and parentheses
 
-Basic arithmetic operators and parentheses are translated as they are, as
-illustrated below.
+Arithmetic operators and parentheses are translated as they are, as illustrated
+below.
 
@@
 mathout(term)
+
+term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
+mathout(term)
 ```
@@
 is omitted. This heuristic does not always produce the desired result;
-therefore, \pkg{mathml} defines the extra functions `dot/2`, `nodot/2`,
-and `times/2` that produce the respective multiplication signs.
-
-```{r, results="asis"}
-term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
-mathout(term)
-```
-
-For the standard operators with known precedence, \pkg{mathml} is generally able
-to detect if parentheses are needed; for example, parentheses are automatically
-put around `g + h` in the `nodot`-example. However, we note unecessary
-parentheses around `l + m` above. This is parenthesis is a consequence
+therefore, \pkg{mathml} defines alternative multiplication
+functions `dot()`, `nodot()`, and `times()` that produce the respective
+multiplication signs.
+
+For standard operators with known precedence, \pkg{mathml} is generally able to
+detect if parentheses are needed; for example, parentheses are automatically
+placed around `g + h` in the `nodot`-example. However, we note unecessary
+parentheses around `l + m` above. Thes parentheses are a consequence
 of `quote(a^(b + c))` actually producing a nested R\ call of the
@@
 
-This feature is unfortunate because the extra parenthesis around `b + c` is not
-needed and yields to the extra parenthesis. The preferred result is obtained by
-using the functional representation `quote('^'(k + l, m + n))` of the power, or,
-as a workaround, curly braces (see last summand above).
+For the present purpose, this feature is unfortunate because extra parentheses
+around `b + c` are not needed. The preferred result is obtained by using the
+functional form `quote('^'(k, l + m))` of the power, or curly braces as a
+workaround (see `p + q` above).
 
@@
 
-While the parentheses typically follow the precedence in standard infix
-operators, undesirable results may be obtained in custom operators.
+Whereas in standard infix operators, the parentheses typically follow the rules
+for precedence, undesirable results may be obtained in custom operators.
 
@@
 mathout(term)
-```
-
-This is a reminder that it is not possible to change the precedence of custom
-operators in R\, and that expressions with such operators are evaluated strictly
-from left to right. As in the previous example, the solution is to either use
-the functional form of the operator or a curly brace to enforce the correct
-operator precedence.
-
-```{r, results='asis'}
-term <- quote(`%+-%`(mean(X), 1.96 * s / sqrt(N)))
+term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N)))
 term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
@@
 
-More operators are shown in Table \@ref(tab:custom-operators), following the
-suggestions by Murrell and Ihaka [-@murrell2000].
+The example is a reminder that it is not possible to define the precedence of
+custom operators in R, and that expressions with such operators are evaluated
+strictly from left to right. As in the previous example, the solution is to
+either use the functional form of the operator or a curly brace to enforce the
+correct operator precedence.
+
+More operators are shown in Table \@ref(tab:custom-operators), including the
+suggestions by Murrell and Ihaka [-@murrell2000] for graphical annotations and
+arrows in R\ figures.
 
@@
 op1 <- list(
-  cross=quote(x %*% y),
-  dot=quote(x %.% y),
-  kronecker=quote(X %x% Y),
-  'integer division'=quote(x %/% y),
+  cross=quote(A %*% B),
+  dot=quote(A %.% B),
+  kronecker=quote(A %x% B),
+  'integer division'=quote(A %/% B),
   and=quote(A & B),
@@
   not=quote(!A),
-  equal=quote(X == Y),
-  assign=quote(X <- Y))
+  equal=quote(A == B),
+  assign=quote(A <- B))
 
@@
 op2 <- list(
-  different=quote(x != y),
-  tilde=quote(X ~ Y),
-  approx=quote(x %~~% y),
-  equiv=quote(x %==% y),
-  congr=quote(x %=~% y),
-  prop=quote(x %prop% y),
-  'in'=quote(a %in% A),
+  different=quote(A != B),
+  tilde=quote(A ~ B),
+  approx=quote(A %~~% B),
+  equiv=quote(A %==% B),
+  congr=quote(A %=~% B),
+  prop=quote(A %prop% B),
+  'in'=quote(A %in% B),
   intersect=quote(intersect(A, B)),
@@
 op3 <- list(
-  twosidearrow=quote(x%<->%b),
-  rightarrow=quote(x%->%y),
-  leftarrow=quote(x%<-%y),
-  uparrow=quote(x%up%y),
-  down=quote(x%down%y),
-  iff=quote(x%<=>%y),
-  rArr=quote(x %=>% y),
-  lArr=quote(x %<=% y),
-  dblup=quote(x %dblup% y),
-  dbldown=quote(x %dbldown% y))
+  twosidearrow=quote(A %<->% B),
+  rightarrow=quote(A %->% B),
+  leftarrow=quote(A %<-% B),
+  uparrow=quote(A %up% B),
+  down=quote(A %down% B),
+  iff=quote(A %<=>% B),
+  rArr=quote(A %=>% B),
+  lArr=quote(A %<=% B),
+  dblup=quote(A %dblup% B),
+  dbldown=quote(A %dbldown% B))
+
 m3 <- lapply(op3, FUN=mathout, flags=list(cat=FALSE))
@@
 
-t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2, Operator=op3, Arrow=m3)
+t <- cbind(Operator=op1, Output=m1,
+  Operator=op2, Output=m2,
+  Operator=op3, Arrow=m3)
+
 knitr::kable(t, caption="Custom operators in mathml",
@@
 
-## R functions
-
-All functions from R::base are supported, a subset is shown in 
-Tables \@ref(tab:basic-elements).
-
-```{r basic-elements, echo=FALSE}
+## Builtin functions
+
+There is support for most functions from \pkg{base}, with adequate use and omission
+of parentheses.
+
+```{r, results='asis'}
+term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
+mathout(term)
+term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + ceiling(x) + round(X))
+mathout(term)
+```
+
+A few more examples are shown in Table \@ref(tab:base-stats), including
+functions from \pkg{stats}.
+
+```{r base-stats, echo=FALSE}
 op1 <- list(
@@
   sqrt=quote(sqrt(x)),
-  mean=quote(mean(xy)),
+  mean=quote(mean(X)["A"]),
   absolute=quote(abs(x)))
@@
 
-t <- cbind(Function=op1, Output=m1)
-knitr::kable(t, caption="Arithmetic functions", row.names=FALSE, escape=FALSE)
-```
+op2 <- list(
+  sin=quote(sin(x)),
+  cosh=quote(cosh(x)),
+  tanpi=quote(tanpi(alpha)),
+  log=quote(log(p)),
+  log1p=quote(log1p(x)),
+  exp=quote(exp(x)),
+  expm1=quote(expm1(x)),
+  lchoose=quote(lchoose(n, k)),
+  sqrt=quote(sqrt(x)),
+  mean=quote(mean(X)["A"]),
+  absolute=quote(abs(x)))
+m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))
+
+op2 <- as.character(op2)
+if(knitr::is_latex_output())
+  op2 <- sapply(op2, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op2 <- sapply(op2, FUN=knitr:::escape_html)
+
+t <- cbind(Function=op1, Output=m1, Function=op2, Output=m2)
+knitr::kable(t, caption="R functions from base and stats",
+  row.names=FALSE, escape=FALSE)
+```
+
+## Custom functions
 
@@
 being the function to integrate, and the lower and the upper bound of the
-integration. For mathematical typesetting in the form
-of $\int f(x)\, dx$, \pkg{mathml} needs to find out the name of the integration
-variable. For that purpose, the underlying Prolog bridge provides a
-predicate `r_eval/2` that calls R\ from Prolog. In the example below, this
-predicate is used for a call of `formalArgs(args(sin))`, which returns the names
-of the function arguments of `sin`, namely, `x`.
+integration.
 
@@
 mathout(term)
+```
+
+In the above example, `sin` is an R symbol, not an R function. For mathematical
+typesetting in the form of $\int f(x)\, dx$, \pkg{mathml} needs to find out the
+name of the integration variable. For that purpose, the underlying Prolog bridge
+provides a predicate `r_eval/2` that calls R\ from Prolog. In the example above,
+this predicate is used for a call of `formalArgs(args(sin))`, which returns the
+names of the arguments of `sin`, namely, `x`.
+
+If integrand points to an R function as in `call("integrate", sin, 0, 2L*pi)`,
+mathml would be unable to find out the function name:
+
+```{r}
+g <- function(x) {sin(x)}
+print(g)
 ```
@@
 
-We can then easily feed mixtures of partially named and positional arguments in
-unusual order into the renderer.
-
-```{r, echo=TRUE, results='asis'}
+This function can be used to feed mixtures of partially named and positional
+arguments into the renderer.
+
+```{r, results='asis'}
 g <- function(u)
@@
 term <- call("integrate", lower=-Inf, upper=Inf, g)
-attributes(term)$name = quote(g)
+attributes(term)$name <- quote(g)
 mathout(canonical(term))

### 2022-12-29 R/mathml.R

@@ -29,7 +29,7 @@ mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
 {
   flags <- c(attributes(term), flags, list(cat=FALSE))
   t <- rolog::once(call("r2mathml", flags, term, expression(X)),
-    options=list(preproc=mathml_preproc))
+    options=list(preproc=list(rolog::preproc, mathml_preproc)))
   r <- paste(t$X, collapse="")
   if(flags$cat)
     return(cat(r))
@@ -39,8 +39,6 @@ mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
 
 # Prolog representation of not equal etc. (left: R, right: Prolog)
 mathml_operators = c(
-  "!=" = "\\=",
-  "<=" = "=<",
   "%.%" = "cdot",
   "%/%" = "div",
   "%+-%" = "pm",
@@ -66,17 +64,16 @@ mathml_preproc <- function(query=quote(2 != 2))
   if(is.call(query))
   {
     args <- as.list(query)
-
-    index <- which(args[[1]] == names(mathml_operators))
-    if(length(index) == 1)
-      args[[1]] <- as.name(mathml_operators[index])
+    index <- (args[[1]] == names(mathml_operators))
+    if(any(index))
+      args[[1]] <- as.symbol(mathml_operators[index])
 
     args[-1] <- lapply(args[-1], FUN=mathml_preproc)
     return(as.call(args))
   }
 
   if(is.list(query))
-    return(lapply(query, FUN=mathml_preproc))
+    query[] <- lapply(query, FUN=mathml_preproc)
 
   if(is.function(query))
     body(query) <- mathml_preproc(body(query))
@@ -107,12 +104,35 @@ mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
 {
   flags <- c(attributes(term), flags, list(cat=FALSE))
   t <- rolog::once(call("r2mathjax", flags, term, expression(X)),
-    options=list(preproc=mathml_preproc))
+    options=list(preproc=list(rolog::preproc, mathml_preproc)))
   r <- paste(t$X, collapse="")
   if(flags$cat)
     return(cat(r))
 
   return(r)
+}
+
+#' Add a name attribute to an element (most often, an R function)
+#'
+#' @param x
+#' an R object, e.g., an R function
+#'
[…]

### 2022-12-29 Draft3.Rmd
sin^2(x)

@@
 mathout(term)
-term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + ceiling(x) + round(X))
+term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + ceiling(x))
 mathout(term)

### 2022-12-29 Draft3.Rmd
qbinom

@@
 op2 <- list(
-  sin=quote(sin(x)),
-  cosh=quote(cosh(x)),
-  tanpi=quote(tanpi(alpha)),
-  log=quote(log(p)),
-  log1p=quote(log1p(x)),
-  exp=quote(exp(x)),
-  expm1=quote(expm1(x)),
-  lchoose=quote(lchoose(n, k)),
-  sqrt=quote(sqrt(x)),
-  mean=quote(mean(X)["A"]),
-  absolute=quote(abs(x)))
+  dbinom=quote(dbinom(k, N, pi)),
+  pbinom=quote(pbinom(k, N, pi)),
+  qbinom=quote(qbinom(p, N, pi)),
+  dnorm=quote(dnorm(x, mu, sigma)),
+  pnorm=quote(pnorm(x, mu, sigma)),
+  qnorm=quote(qnorm(p, mu, sigma)),
+  dchisq=quote(dchisq(x, df)),
+  pchisq=quote(pchisq(x, df)),
+  qchisq=quote(qchisq(p, df)),
+  dt=quote(dt(x, df)),
+  pt=quote(pt(x, df)),
+  qt=quote(qt(p, df)))
 m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

### 2022-12-29 inst/pl/mathml.pl

@@ -2376,55 +2376,18 @@ math(Flags, buggy(Flags, _, B), New, X)
     X = B.
 
 %
-% Binomial distribution
+% Probability distributions
 %
-% Density, distribution etc.
-math(Flags, dbinom(K, N, Pi), New, X)
- => New = Flags,
-    X = fn(subscript('P', "Bi"), (['X' = K] ; [N, Pi])).
+math(dbinom(K, N, Pi), M)
+ => M = fn(subscript('P', "Bi"), (['X' = K] ; [N, Pi])).
 
 math(pbinom(K, N, Pi), M)
  => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
 
-math(Flags, upbinom(K, N, Pi), New, X)
- => New = Flags,
-    X = fn(subscript('P', "Bi"), (['X' >= K] ; [N, Pi])).
+math(qbinom(Alpha, N, Pi), M)
+ => M = fn(subscript("arg min", k),
+          [fn(subscript('P', "Bi"), (['X' =< k] ; [N, Pi])) > Alpha]).
 
-math(Flags, cbinom(Alpha, N, Pi, Tail, Dist), New, X)
- => New = Flags,
-    X = fn(Tail, [fn(subscript('P', "Bi"), ([Dist] ; [N, Pi])) =< Alpha]).
-
-math(Flags, tail("upper"), New, X)
- => New = Flags,
-    X = subscript("argmin", k).
-
-math(Flags, tail("lower"), New, X)
- => New = Flags,
-    X = subscript("argmax", k).
-
-math(Flags, tail("upperdens"), New, X)
- => New = Flags,
-    X = subscript("argmin", k > 'N' * pi).
-
-math(Flags, tail("lowerdens"), New, X)
- => New = Flags,
-    X = subscript("argmax", k < 'N' * pi).
-
-math(Flags, dist("upper"), New, X)
- => New = Flags,
-    X = ('X' >= k).
-
-math(Flags, dist("lower"), New, X)
- => New = Flags,
-    X = ('X' =< k).
-
-math(Flags, dist("density"), New, X)
- => New = Flags,
-    X = ('X' = k).
-
-%
-% Normal distribution
-%
 math(pnorm(X, Mu, Sigma2), M)
  => M = fn('Phi', ([X] ; [Mu, Sigma2])).
 

### 2022-12-29 Draft3.Rmd
dnorm, pnorm, qnorm

--- 2022-12-29 Draft3.Rmd
@@
 op2 <- list(
-  sin=quote(sin(x)),
-  cosh=quote(cosh(x)),
-  tanpi=quote(tanpi(alpha)),
-  log=quote(log(p)),
-  log1p=quote(log1p(x)),
-  exp=quote(exp(x)),
-  expm1=quote(expm1(x)),
-  lchoose=quote(lchoose(n, k)),
-  sqrt=quote(sqrt(x)),
-  mean=quote(mean(X)["A"]),
-  absolute=quote(abs(x)))
+  dbinom=quote(dbinom(k, N, pi)),
+  pbinom=quote(pbinom(k, N, pi)),
+  qbinom=quote(qbinom(p, N, pi)),
+  dnorm=quote(dnorm(x)),
+  dnorm=quote(dnorm(x, mu, sigma)),
+  pnorm=quote(pnorm(x, mu, sigma)),
+  qnorm=quote(qnorm(alpha/2L, mu, sigma)),
+  pchisq=quote(pchisq(x, df)),
+  qchisq=quote(qchisq(p, df)),
+  dt=quote(dt(x, df)),
+  pt=quote(pt(x, df)),
+  qt=quote(qt(p, df)))
 m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

### 2022-12-29 inst/pl/mathml.pl

@@ -2388,11 +2388,23 @@ math(qbinom(Alpha, N, Pi), M)
  => M = fn(subscript("arg min", k),
           [fn(subscript('P', "Bi"), (['X' =< k] ; [N, Pi])) > Alpha]).
 
-math(pnorm(X, Mu, Sigma2), M)
- => M = fn('Phi', ([X] ; [Mu, Sigma2])).
+math(dnorm(Z), M)
+ => M = fn(phi, [Z]).
+
+math(dnorm(X, Mu, Sigma2), M)
+ => M = fn(phi, ([X] ; [Mu, Sigma2])).
 
 math(pnorm(Z), M)
  => M = fn('Phi', [Z]).
+
+math(pnorm(X, Mu, Sigma2), M)
+ => M = fn('Phi', ([X] ; [Mu, Sigma2])).
+
+math(qnorm(Alpha), M)
+ => M = fn('Phi' ^ -1, [Alpha]).
+
+math(qnorm(Alpha, Mu, Sigma2), M)
+ => M = fn('Phi' ^ -1, ([Alpha] ; [Mu, Sigma2])).
 
 %
 % Functions like f(x) and f(x; a, b)

### 2022-12-29 Draft3.Rmd
t dist

--- 2022-12-29 Draft3.Rmd
@@
   sqrt=quote(sqrt(x)),
-  mean=quote(mean(X)["A"]),
+  mean=quote(mean(X)),
   absolute=quote(abs(x)))
@@
 op2 <- list(
-  sin=quote(sin(x)),
-  cosh=quote(cosh(x)),
-  tanpi=quote(tanpi(alpha)),
-  log=quote(log(p)),
-  log1p=quote(log1p(x)),
-  exp=quote(exp(x)),
-  expm1=quote(expm1(x)),
-  lchoose=quote(lchoose(n, k)),
-  sqrt=quote(sqrt(x)),
-  mean=quote(mean(X)["A"]),
-  absolute=quote(abs(x)))
+  dbinom=quote(dbinom(k, N, pi)),
+  pbinom=quote(pbinom(k, N, pi)),
+  qbinom=quote(qbinom(p, N, pi)),
+  dnorm=quote(dnorm(x)),
+  dnorm=quote(dnorm(x, mu, sigma)),
+  pnorm=quote(pnorm(x)),
+  pnorm=quote(pnorm(x, mu, sigma)),
+  qnorm=quote(qnorm(alpha/2L)),
+  pchisq=quote(1-pchisq(x, df)),
+  qchisq=quote(qchisq(1L-alpha, df)),
+  pt=quote(pt(t, N-1L)),
+  qt=quote(qt(alpha/2L, N-1L)))
 m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

### 2022-12-29 inst/pl/mathml.pl

@@ -141,7 +141,7 @@ math(Flags, Sum, New, M),
 math(Flags, Sum, New, M),
     compound(Sum),
     compound_name_arguments(Sum, sum, [Arg]),
-    type(Flags, Arg, subscript(Idx))
+    type(Flags, Arg, subscript(_, Idx))
  => New = Flags,
     M = fn(subscript(sum, Idx), [Arg]).
 
@@ -208,10 +208,10 @@ jax(Flags, R, M),
     jax(New, Pwr, Z),
     format(string(M), "{~w_~w^~w}", [X, Y, Z]).
 
-type(Flags, _R, Type),
+type(Flags, R, Type),
     member(subscript(Idx), Flags),
     member(superscript(Pwr), Flags)
- => Type = subsupscript(Idx, Pwr).
+ => Type = subsupscript(R, Idx, Pwr).
 
 mathml :-
     mathml(subsupscript(x, i, 2)).
@@ -241,9 +241,9 @@ jax(Flags, R, M),
     jax(New, Idx, Y),
     format(string(M), "{~w_~w}", [X, Y]).
 
-type(Flags, _R, Type),
+type(Flags, R, Type),
     member(subscript(Idx), Flags)
- => Type = subscript(Idx).
+ => Type = subscript(R, Idx).
 
 mathml :-
     mathml(subscript(x, i)).
@@ -286,9 +286,9 @@ jax(Flags, R, M),
     jax(New, Pwr, Y),
     format(string(M), "{{~w}^{~w}}", [X, Y]).
 
-type(Flags, _R, Type),
+type(Flags, R, Type),
     member(superscript(Pwr), Flags)
- => Type = superscript(Pwr).
+ => Type = superscript(R, Pwr).
 
 mathml :-
     mathml(superscript(x, 2)).
@@ -2406,6 +2406,18 @@ math(qnorm(Alpha), M)
 math(qnorm(Alpha, Mu, Sigma2), M)
  => M = fn('Phi' ^ -1, ([Alpha] ; [Mu, Sigma2])).
 
+math(pchisq(X, Df), M)
+ => M = fn(subscript('F', fn(chi^2, [list(space, [Df, "df"])])), [X]).
+
+math(qchisq(Alpha, Df), M)
+ => M = fn(subscript('F' ^ -1, fn(chi^2, [list(space, [Df, "df"])])), [Alpha]).
+
+math(pt(T, Df), M)
[…]

### 2022-12-29 Draft3.Rmd
custom functions

@@
   qnorm=quote(qnorm(alpha/2L)),
-  pchisq=quote(1-pchisq(x, df)),
-  qchisq=quote(qchisq(1L-alpha, df)),
+  pchisq=quote(1L-pchisq(x, 1L)),
+  qchisq=quote(qchisq(1L-alpha, 1L)),
   pt=quote(pt(t, N-1L)),
@@
 
-For self-written functions, matters are a bit more complicated because the
-underlying R package `rolog` [@rolog] cannot easily access the function name,
-hence a default is chosen. The function name can be given as an attribute.
+For self-written functions, matters are a bit more complicated. The name of a
+self-written function is not transparent in R, since only the function body is
+represented, as illustrated by the redefinition of `sign` below. The function is
+often a nested R "call" of the form `'{'(L)`, with `L` being a list of
+commands. \pkg{mathml} provides very limited support for typical control
+structures.
 
 ```{r, results='asis'}
-f <- function(x)
+sgn <- function(x)
 {
-  if(x == 0L) return(0L);
-  if(x < 0L) return(-1L);
+  if(x == 0L) return(0L)
+  if(x < 0L) return(-1L)
   if(x > 0L) return(1L)
@@
 
-attributes(f)$name = quote(sgn)
-mathout(f)
-```
+term <- quote(s <- sgn(x))
+mathout(term)
+
+hook(quote(s), quote(sg(x)))
+term <- call("<-", quote(s), sgn)
+mathout(term)
+```
+
+The hook (see below) is used to display \(\mathrm{sg}(x) = \{\ldots\) instead
+of just \(s = \{\).
 

### 2022-12-30 R/mathml.R
mathml/mathjax: optional argument for the environment of r_eval/1,2

@@ -25,11 +25,13 @@
 #' @examples
 #' mathml(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
 #'
-mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
+mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL,
+  env=globalenv())
 {
   flags <- c(attributes(term), flags, list(cat=FALSE))
   t <- rolog::once(call("r2mathml", flags, term, expression(X)),
-    options=list(preproc=list(rolog::preproc, mathml_preproc)))
+    options=list(preproc=list(rolog::preproc, mathml_preproc)),
+    env=env)
   r <- paste(t$X, collapse="")
   if(flags$cat)
     return(cat(r))
@@ -100,11 +102,13 @@ mathml_preproc <- function(query=quote(2 != 2))
 #' @examples
 #' mathjax(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
 #'
-mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL)
+mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL,
+  env=globalenv())
 {
   flags <- c(attributes(term), flags, list(cat=FALSE))
   t <- rolog::once(call("r2mathjax", flags, term, expression(X)),
-    options=list(preproc=list(rolog::preproc, mathml_preproc)))
+    options=list(preproc=list(rolog::preproc, mathml_preproc)),
+    env=env)
   r <- paste(t$X, collapse="")
   if(flags$cat)
     return(cat(r))

### 2023-01-02 inst/pl/mathml.pl
poc multiline equation

@@ -186,7 +186,7 @@ jax(Flags, R, M),
  => jax(New, R, X),
     jax(New, Idx, Y),
     jax(New, Pwr, Z),
-    format(string(M), "{~w_~w^~w}", [X, Y, Z]).
+    format(string(M), "{~w}_{~w}^{~w}", [X, Y, Z]).
 
 type(Flags, R, Type),
     member(subscript(Idx), Flags),
@@ -219,7 +219,7 @@ jax(Flags, R, M),
     select(subscript(Idx), Flags, New)
  => jax(New, R, X),
     jax(New, Idx, Y),
-    format(string(M), "{~w_~w}", [X, Y]).
+    format(string(M), "{~w}_{~w}", [X, Y]).
 
 type(Flags, R, Type),
     member(subscript(Idx), Flags)
@@ -258,13 +258,13 @@ jax(Flags, R, M),
     select(superscript(Pwr), Flags, New)
  => jax(New, paren(R), X),
     jax(New, Pwr, Y),
-    format(string(M), "{{~w}^{~w}}", [X, Y]).
+    format(string(M), "{~w}^{~w}", [X, Y]).
 
 jax(Flags, R, M),
     select(superscript(Pwr), Flags, New)
  => jax(New, R, X),
     jax(New, Pwr, Y),
-    format(string(M), "{{~w}^{~w}}", [X, Y]).
+    format(string(M), "{~w}^{~w}", [X, Y]).
 
 type(Flags, R, Type),
     member(superscript(Pwr), Flags)
@@ -317,7 +317,7 @@ ml(_Flags, greek(R), M)
  => M = mi(&(R)).
 
 jax(_Flags, greek(R), M)
- => format(string(M), "{\\~w}", [R]).
+ => format(string(M), "\\~w", [R]).
 
 type(_Flags, greek(_), T)
  => T = atomic.
@@ -338,7 +338,7 @@ ml(_Flags, boolean(R), M)
  => M = mi(R).
 
 jax(_Flags, boolean(R), M)
- => format(string(M), "{~w}", [R]).
+ => format(string(M), "~w", [R]).
 
 type(_Flags, boolean(_), T)
  => T = atomic.
@@ -483,7 +483,7 @@ jax(Flags, ident(R), M),
  => format(string(M), "\\mathcal{~w}", [R]).
 
 jax(_Flags, ident(R), M)
[…]

### 2023-01-05 R/mathml.R
modulo

@@ -43,6 +43,7 @@ mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL,
 mathml_operators = c(
   "%.%" = "cdot",
   "%/%" = "div",
+  "%%"="rem",
   "%+-%" = "pm",
   "%*%" = "times",
   "%~~%" = "approx",
@@ -557,6 +558,7 @@ e1*e2
 #'
 '%dbldown%' <- function(x, y)
   x=y
+
 
 #' denote
 #' This is a function that allows the user to insert abbreviations in the formula,

### 2023-01-05 inst/pl/mathml3.pl

@@ -1986,6 +1986,11 @@ math(over(Num, Den), M)
 math(div(Num, Den), M)
  => M = floor(Num / Den).
 
+% Modulo
+math(rem(Num, Den), M)
+ => M = ceiling(Num / Den).
+
+
 %
 % Large font ("displaystyle")
 %

### 2023-01-05 inst/pl/mathml3.pl
case study comma separated list

@@ -2418,6 +2418,12 @@ prec(fn(Name, _Args), Prec, Flags)
 type(fn(_Name, _Args), Type, _Flags)
  => Type = function.
 
+% Comma-separated list
+math(R, M),
+    compound(R),
+    compound_name_arguments(R, ',', Args)
+ => M = list(',', Args).
+
 % Default compounds
 %
 % Can't use the macros here because of left recursion

### 2023-01-10 rj/mathml.Rmd
remove unnecessary packages

--- 2022-12-29 Draft3.Rmd
@@
 - \usepackage{cancel}
+- \usepackage{amsmath}
+Vignette: >
+  %\VignetteIndexEntry{mathml}
+  %\VignetteEngine{knitr::rmarkdown}
+  %\VignetteEncoding{UTF-8}
 ---
@@
 library(mathml)
-library(plotly)
-library(ggplot2)
-library(kableExtra)
-
-mathout <- function(..., flags=NULL)
+
+mathout <- function(..., flags=NULL, env=parent.frame())
 {
   if(knitr::is_html_output())
-    return(mathml(..., flags=c(flags, list(cat=TRUE))))
+    return(mathml(..., flags=c(flags, list(cat=TRUE)), env=env))
 
   if(knitr::is_latex_output())
-    return(mathjax(..., flags=c(flags, list(cat=TRUE))))
-
-  warning("no output specified")  
-  mathjax(..., flags=c(flags, list(cat=TRUE)))
+    return(mathjax(..., flags=c(flags, list(cat=TRUE)), env=env))
+
+  warning("mathout: no knitr output specified")  
+  mathjax(..., flags=c(flags, list(cat=TRUE)), env=env)
 }
@@
 as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
-such as \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
-is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
-these commands are tedious to type in and their source code is hard to read.
+such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit Latex commands, that
+is, `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for minimalistic use
+cases, these commands are tedious to type in and their source code is hard to
+read.
 
@@
 from the day-to-day perspective of a user. Sections\ 4 and\ 5 present two case
-studies for the use of the package in a scientific report and a ShinyApp web
-page. Section\ 6 concludes with a discussion and ideas for further development.
-A complete list of the R\ elements covered by the package is found in the
-Appendix.
+studies for the use of the package in a scientific report and 
+a **ShinyApp web page**. Section\ 6 concludes with a discussion and ideas for
+further development. A complete list of the R\ elements covered by the package
+is found in the Appendix.
+
+*shinyapp missing so far
 
@@
 
-term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
+term <- quote(dot(a, b) + frac(1L, nodot(c, d + e)) + dfrac(1L, times(g, h)))
 mathout(term)
@@
 therefore, \pkg{mathml} defines alternative multiplication
-functions `dot()`, `nodot()`, and `times()` that produce the respective
-multiplication signs.
+functions `dot()`, `nodot()`, and `times()` that calculate the product and
+produce the respective multiplication signs. Similarly, `frac()` and `dfrac()`
+can be used for small and large fractions.
 
@@
   'integer division'=quote(A %/% B),
+  modulo=quote(A %% B),
   and=quote(A & B),
@@
   union=quote(union(A, B)),
+  crossprod=quote(crossprod(A, B)),
   intersect=quote(is.null(A)))
@@
   tanpi=quote(tanpi(alpha)),
+  asinh=quote(asinh(alpha)),
   log=quote(log(p)),
   log1p=quote(log1p(x)),
+  logb=quote(logb(x, e)),
   exp=quote(exp(x)),
   expm1=quote(expm1(x)),
+  choose=quote(choose(n,k)),
   lchoose=quote(lchoose(n, k)),
+  factorial=quote(factorial(n)), 
+  lfactorial=quote(lfactorial(n)),
   sqrt=quote(sqrt(x)),
@@
   qbinom=quote(qbinom(p, N, pi)),
-  dnorm=quote(dnorm(x)),
+  dpois=quote(dpois(k, lambda)),
+  ppois=quote(ppois(k, lambda)),
+  qpois=quote(qpois(p, lambda)),
+  dexp=quote(dexp(x, lambda)),
+  pexp=quote(pexp(x, lambda)),
+  qexp=quote(qexp(p, lambda)),
   dnorm=quote(dnorm(x, mu, sigma)),
-  pnorm=quote(pnorm(x)),
   pnorm=quote(pnorm(x, mu, sigma)),
@@
 
-term <- quote(s <- sgn(x))
-mathout(term)
-
-hook(quote(s), quote(sg(x)))
-term <- call("<-", quote(s), sgn)
+term <- quote(mysign <- sgn(x))
+mathout(term)
+
+hook(quote(mysign), quote(sg(x)))
+term <- call("<-", quote(mysign), sgn)
 mathout(term)
@@
 of just \(s = \{\).
+
+## Indices and powers
+
+Indices in square brackets are rendered as subscripts, powers are rendered as
+superscript. Moreover, \pkg{mathml} defines the
+functions `subscript()`, `superscript()` and `subsuperscript()` that simply
+return their first argument. The other arguments serve as decorations, for
+example, for summation and product signs.
+
+```{r, results='asis'}
+term <- quote(S[Y]^2L <- frac(1L, N) * sum(Y[i] - mean(Y))^2L)
+mathout(term)
+
+term <- quote(log(subsupscript(prod(L[i]), i==1L, N)) <- 
+          subsupscript(sum(log(L[i])), i==1L, N))
+mathout(term)
+```
 
@@
 mathout(term)
-```
-
-In the above example, `sin` is an R symbol, not an R function. For mathematical
-typesetting in the form of $\int f(x)\, dx$, \pkg{mathml} needs to find out the
-name of the integration variable. For that purpose, the underlying Prolog bridge
-provides a predicate `r_eval/2` that calls R\ from Prolog. In the example above,
-this predicate is used for a call of `formalArgs(args(sin))`, which returns the
-names of the arguments of `sin`, namely, `x`.
-
-If integrand points to an R function as in `call("integrate", sin, 0, 2L*pi)`,
-mathml would be unable to find out the function name:
-
-```{r}
-g <- function(x) {sin(x)}
-print(g)
-```
+eval(term)
+```
+
+For mathematical typesetting in the form
+of \(\int f(x)\, dx\), \pkg{mathml} needs to find out the name of the
+integration variable. For that purpose, the underlying Prolog bridge provides a
+predicate `r_eval/2` that calls R\ from Prolog. In the example above, this
+predicate evaluates `formalArgs(args(sin))`, which returns the names of the
+arguments of `sin`, namely, `x`.
+
+Note that in the example above, the quoted term is an abbreviation
+for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
+function. R's `integrate()` can handle both. However, as already mentioned in
+the previous subsection, \pkg{mathml} would be unable to find out the function
+name, it needs the symbol.
 
@@
 One of R's great features is the possibility to refer to function arguments by
-their names, not only by their position in a list of arguments. Unfortunately, 
-the Prolog-powered handler for `integrate/3` is rather rigid; it accepts only
-three arguments in that particular order and without names, that
-is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.
+their names, not only by their position in the list of arguments. At the other
+end, most Prolog-powered handlers for the R\ calls are rather rigid, for
+example, `integrate/3` accepts exactly three arguments in a particular order and
+without names, that is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print
+the desired result.
 
 To "canonicalize" function calls with named arguments and arguments in unusual
-order, \pkg{mathml} provides an auxilliary function `canonical(f, drop)` that
-reorders the argument list of calls to known functions and, if `drop=TRUE`, also
-removes the names of the arguments.
+order, \pkg{mathml} provides an auxilliary R\ function `canonical(f, drop)` that
+reorders the argument list of calls to known R\ functions and, 
+if `drop=TRUE` (which is the default), also removes the names of the arguments.
 
@@
 
+```{r, results='asis'}
+mathout(canonical(term))
+```
+
 This function can be used to feed mixtures of partially named and positional
-arguments into the renderer.
-
-```{r, results='asis'}
-g <- function(u)
-{
-  sin(u)
-}
-
-# 
-term <- call("integrate", lower=-Inf, upper=Inf, g)
-attributes(term)$name <- quote(g)
-mathout(canonical(term))
-
-# 2
-term <- call("*", 2L, call("$", call("integrate", low=-Inf, up=Inf, g), "value"))
-attributes(term)$name = quote(g)
-mathout(canonical(term))
-
-# It is a bit of a mystery that R knows the result of this integral.
-eval(term)
-```
-
-The extra R\ function `canonical()` applies `match.call()` to non-primitive R\
-calls, basically cleaning up the arguments and bringing them into the correct
-order.
+arguments into the renderer. For details, see the R\ function `match.call()`.
+
+## Matrices and Vectors
+
+Of course, \pkg{mathml} also supports matrices and vectors.
+
+```{r, results='asis'}
+v <- 1:3
+mathout(call("t", v))
+
+A <- matrix(data=11:16, nrow=2, ncol=3)
+B <- matrix(data=21:26, nrow=2, ncol=3)
+term <- call("+", A, B)
+mathout(term)
+```
+
+Note that the seemingly more convient `term <- quote(A + B)` yields \(A + B\) in
+the output---instead of the desired matrix representation.
+
+## Abbreviations
+
+We consider the \(t\)-statistic for independent samples with equal variance. To
+avoid clutter in the equation, the pooled variance \(s^2_{\mathrm{pool}}\) is
+abbreviated, and a comment is given with the expression
+for \(s^2_{\mathrm{pool}}\). For this purpose, \pkg{mathml} provides a
+function `denote(abbr, expr, info)`, with `expr` actually being
+evaluated, `abbr` being rendered, plus a comment of the
+form "with `expr` denoting `info`". The hooks are described in the next section.
+
+```{r, results='asis'}
+hook(quote(m_A), quote(mean(X)["A"])); hook(quote(s2_A), quote(s["A"]^2L));
+hook(quote(n_A), quote(n["A"]))
+hook(quote(m_B), quote(mean(X)["B"])); hook(quote(s2_B), quote(s["B"]^2L))
+hook(quote(n_B), quote(n["B"])); hook(quote(s2_p), quote(s["pool"]^2L))
+
+term <- quote(t <- dfrac(m_A - m_B, 
+    sqrt(denote(s2_p, frac((n_A - 1L)*s2_A + (n_B - 1L)*s2_B, n_A + n_B - 2L),
+                "the pooled variance.") * (frac(1L, n_A) + frac(1L, n_B)))))
+mathout(term)
+```
+
+The term is evaluated below. `print()` is needed because the return value of the
+assignment in the term `t <- dfrac(...)` is invisible.
+
+```{r}
+m_A <- 1.5; s2_A <- 2.4^2; n_A <- 27; m_B <- 3.9; s2_B <- 2.8^2; n_B <- 20
+print(eval(term))
+```
 
@@
 mathout(term)
-```
-
-To improve the situation, \pkg{mathml} provides a simple hook into the prolog
-predicate `math/2` described in the previous section.
-
-```{r, results='asis'}
+
 hook(quote(successes), quote(k))
@@
 
-More complex hooks are possible. In the example below, `dnorm(X, M, S)` is
-translated to the form `phi(X; M, S)` (instead of the default `varphi`).
-
-```{r, results='asis'}
-hook(call("dnorm", expression(X), expression(M), expression(S)), 
-     call("fn", quote(phi),
-       call(";", list(expression(X)), list(expression(M), expression(S)))))
-term <- quote(dnorm(x, mu, sigma^2L))
-mathout(term)
-```
-
-R's expressions are a bit misused to represent variables in the underlying
-prolog engine.
-
-## Operators
-
-A list of possible operators is available here.
-
-```{r, results='asis', echo=TRUE}
-catmath(sum(x[i], i==1L, n))
-
-# fixme: braces not needed when right to *
-catmath(a * sum(b[i]))
-
-# Multiplication (capital Pi) is missing
-catmath(a * prod(b[i]))
-
-catmath(integrate(sin, 0L, 2L*pi))
-
-catmath(A & B)
-
-catmath(A | B)
-
-catmath(!A)
-
-catmath(xor(A, B))
-
-catmath(exp(2L*pi*i))
-
-catmath(expm1(2L*pi*i))
-
-catmath(log(x))
-
-catmath(log10(x))
-
-catmath(log2(x))
-
-catmath(logb(x, e))
-
-catmath(log1p(x))
-
-catmath(ceiling(pi))
-
-catmath(floor(pi))
-
-catmath(sin(pi/2L))
-
-catmath(cos(pi/2L))
-
-catmath(tan(pi/2L))
-```
-
-## Grouping
-
-At first glance, grouping expressions may seem trivial enough.
-However, when you want to deal with, for instance, exponentiation or multiplication, things can get more complicated.
-Let us consider a trivial example.
-If we want to calculate and evaluate the expression `a * (b + c)`, we find that R\ automatically gives precedence to the addition even if we do not specify it in the text, as in the following.
-
-```{r, results='asis', echo=TRUE}
-first <- quote(a)
-second <- quote(b + c)
-mathout(call("*", first, second))
-
-# precedence
-quote(a * (b + c))[[3]][[1]]
-```
-
-With exponentiation the situation can be less trivial.
-If we consider `x` to the power of `y + z`, following R\ notation we should add a parenthesis around the two terms.
-Otherwise R\ would consider only `y` as power of `x`.
-
-```{r, results='asis', echo=TRUE}
-catmath(x^y + z)
-
-# mathout(call("+", call("^", as.symbol("x"), as.symbol("y")), as.symbol("z")))
-
-catmath(x^(y + z))
-```
-
-Unfortunately, R\ automatically put `y + z` in parentheses even if, from a mathematical point of view, the brackets are not needed.
-To solve this issue two solutions are possible, the first one is cumbersome since it requires many specifications.
-The second, that can be considered a shortcut, uses a syntax similar to LateX and puts the power in curly brackets.
-In both cases the output produced is consistent with the mathematical notation.
-
-```{r, results='asis', echo=TRUE}
-# without parenthesis
-mathout(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
-
-# shortcut
-catmath(x^{y + z})
-```
-
-## Matrix and Vectors
-
-Since also Matrices and Vectors are important elements in R\, we decided to implement them and render them graphically in the following ways.
-
-```{r, results='asis', echo=TRUE}
-vec <- 1:3
-mathout(canonical(vec))
-mathout(vec)
-
-mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
-mathout(canonical(mat))
-```
-
-## Abbreviations
-
-\pkg{mathml} allows also for the use of abbreviations in formulas. Let us consider here the t-statistic for independent samples with equal variance assumed. 
-The procedure consists of three steps.
-The first one is to define the hooks to nicely print the variables.
-Second, we write the t-statistic specifying in the formula how to calculate the pooled variance with the function  `denote(abbr, expr, info)`.
-Finally, since \pkg{mathml} also allows computation, we can directly compute the value of the t statistic with `eval()`.
-
-```{r, results='asis', echo=TRUE}
-
-hook(quote(x_1), quote(mean(x["1"])))
-hook(quote(x_2), quote(mean(x["2"])))
-hook(quote(n_1), quote(n["1"]))
-hook(quote(n_2), quote(n["2"]))
-hook(quote(s_1), quote(s["1"]))
-hook(quote(s_2), quote(s["2"]))
-hook(quote(s_p), quote(s["pool"]))
-
-t <- quote(dfrac(x_1 - x_2, sqrt(denote(s_p^2L, frac((n_1 - 1L)*s_1^2L + (n_2 - 1L)*s_2^2L, n_1+n_2-2L), "the pooled variance.") * (dfrac(1L, n_1)+dfrac(1L,n_2)))))
-
-mathout(t)
-
-# An example
-
-x_1 = 20.5
-x_2 = 35.7
-n_1 = 150
-n_2 = 160
-s_1 = 3.4
-s_2 = 4.2
-
-eval(t)
-```
+To improve the situation, \pkg{mathml} provides a simple hook that can be used
+to replace elements (e.g., verbose variable names) of the code by concise
+mathematical symbols, as illustrated in the example. Care should be taken to
+avoid recursive hooks such as `hook(quote(s), quote(s["A"]))` that endlessly
+replace the \(s\) from \(s_{\mathrm{A}}\) as
+in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).
+
+more stuff: math/2, ml/3, jax/3
 
@@
 
-For a more clear representation of what the \pkg{mathml} package is supposed to do, we provide here a case study, based on the work by Schwarz [-@schwarz1994], in which the features of \pkg{mathml} are implemented.
-
-The reader may recognize that Schwarz' [-@schwarz1994] study can be traced back to the strand of decision-making models, where decision-making is assumed to be a process of
-noisy accumulation of information over time [e.g., @ratcliff2016]. The aim of
-Schwarz' [-@schwarz1994] study is to present a new explanation of redundancy gains when
-observers make speeded responses to stimuli of different sources, and the same
-information is presented on two channels. Schwarz [-@schwarz1994] describes a model that assumes
-the superposition of channel-specific diffusion processes that eventually reach
-an absorbing barrier to elicit the response. For a detailed description the
-reader may refer to the original article.
-
-### The model
+This case study describes a model by Schwarz [-@schwarz1994] from mathematical
+Psychology using the features of package \pkg{mathml}. Schwarz presents a new
+explanation of redundancy gains that occur when observers respond to stimuli of
+different sources, and the same information is presented on two or more
+channels. In Schwarz' model, decision-making builds on a process of noisy
+accumulation of information over time [e.g., @ratcliff2016]. In redundant
+stimuli, the model assumes a superposition of channel-specific diffusion
+processes that eventually reach an absorbing barrier to elicit the response. For
+a detailed description the reader may refer to the original article.
+
+## The model
+
+```{r, echo=FALSE}
+hook(quote(mu_A), quote(mu["A"]))
+hook(quote(mu_B), quote(mu["B"]))
+hook(quote(sigma_A), quote(sigma["A"]))
+hook(quote(sigma_B), quote(sigma["B"]))
+hook(quote(mu_M), quote(mu["M"]))
+hook(quote(zn), quote(z[tau]))
+hook(quote(M), quote(overline(X)))
+```
 
 Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
-alone or in combination (redundant stimuli), with the redundant stimuli being
-presented either simultaneously or with onset asynchrony $\tau$. The channel
-activation is described as a two-dimensional Wiener process with
-drifts $\mu_\mathrm A$, $\mu_\mathrm B$,
-variances $\sigma^2_\mathrm A$, $\sigma^2_\mathrm B$,
-correlation $\rho_\mathrm{AB}$ and initial
-conditions $X_i(t=0)=0, i=\mathrm{A, B}$. The response is elicited when the
-process reaches the absorbing barrier $c > 0$ for the first time. In combined
-stimuli ("redundant" stimuli), the overall diffusion process
-is $X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)$ , which is again a
-Wiener process with drift $\mu_\mathrm A + \mu_\mathrm B$ and
-variance $\sigma^2_\mathrm A + \sigma^2_\mathrm B
-  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B$.
-
-In single-target trials, the first passages at $c$ are expected at
-$E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}$, whereas in synchronous
-redundant-target trials, we have
-$E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}$.
-
-For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
-first-passage time $E[D_{\mathrm A(\tau)\mathrm B}]$ as a function of the
-stimulus onset asyncrony $\tau > 0$ (Eq.\ 10),
-
-```{r, results="asis", echo=TRUE}
-f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+alone or in combination (AB, redundant stimuli), with the redundant stimuli
+being presented either simultaneously or with onset asynchrony \(\tau\). The
+channel activation is described as a two-dimensional Wiener process with
+drifts \(\mu_i\), variances \(\sigma^2_i\), and initial
+conditions \(X_i(t = 0) = 0, i = \mathrm{A, B}\). The channels may be correlated
+with \(\rho_{\mathrm{AB}}\), but we assume \(\rho_{\mathrm{AB}} = 0\) for
+simplicity.
+
+A response is elicited when the process reaches an absorbing barrier \(c > 0\)
+for the first time. In single-target trials, the first passages at \(c\) are
+expected at
+
+```{r, results='asis'}
+hook(call("ED_single", expression(c), expression(mu)), 
+     call("E", call(";", quote(D), call(",", expression(c), expression(mu)))))
+
+ED_single <- function(c, mu)
+  dfrac(c, mu)
+
+mathout(call("=", quote(ED_single(c, mu)), ED_single))
+```
+
+In redundant stimuli, the activation of the channel-specific diffusion processes
+adds up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\), hence
+the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process with
+drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
+variance \(\sigma^2_{\mathrm A} + \sigma^2_{\mathrm B}\). For the expected
+first-passage time, we have
+
+```{r, results='asis'}
+mathout(call("=", quote(E(D["AB"])), quote(ED_single(c, mu_A + mu_B))))
+```
+
+For asynchronous stimuli, Schwarz [-@schwarz1994; Eq. 10] derived the expected
+first-passage time as a function of the stimulus onset asyncrony \(\tau\),
+
+```{r, results="asis"}
+hook(call("ED_async", expression(tau), expression(c), expression(mu_A), expression(sigma_A), expression(mu_B)), 
+     call("E", call(";", call("[", quote(D), expression(tau)), 
+                    call(",", expression(c), expression(mu_A), expression(sigma_A), expression(mu_B)))))
+
+ED_async <- function(tau, c, mu_A, sigma_A, mu_B)
 { dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
@@
 
-mathout(f)
-```
-
-The observable response time is assumed to be the sum of $D$, the time
-employed to reach the threshold for the decision, and $M$, denoting the of the
-other processes.
-
-$E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M$
-
-### Methods
-
-Schwarz [-@schwarz1994] applied the model to Miller’s [-@miller1986] data from
-a redundant signals task with 13 onset
-asynchronies $0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty$ ms,
-where 0 refers to synchronous AB, $\infty$ refers to the single-target
-presentation, and negative $\tau$ denote those conditions in which B is
-presented before A. Each condition was replicated 400 times. The 13 mean RTs and
-standard deviations are given in Table 2 below.
-
-Then, from Schwarz' [-@schwarz1994] Equations [2] [3] and [10], we calculated the expected first time passage for the respective $\tau$ and the parameters $\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161$
-
-````{r, echo=FALSE}
-mu_A = 0.53
-mu_B = 1.32
-sigma_A = 4.4
-sigma_B = 18.3
-tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
-c = 100
-mu_M = 161
-````
-
-```{r, results="asis", echo=FALSE}
-# Conditions
-
-g <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
+mathout(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
+```
+
+For negative onset asynchrony (i.e., B before A), the parameters need to be
+switched.
+
+```{r, results='asis'}
+hook(call("ED", expression(tau), expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B)), 
+     call("E", call(";", call("[", quote(D), expression(tau)), 
+                    call(",", expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B)))))
+
+ED <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
 {
-    if( tau == Inf) return(dfrac(c, mu_A));
-    if( tau == -Inf) return(dfrac(c, mu_B));
-    if( tau > 0L)  return(f(tau, c, mu_A, sigma_A, mu_B, sigma_B));
-    if( tau == 0L) return(dfrac(c, mu_A + mu_B));
-    if( tau < 0L) return(f(-tau, c,  mu_B, sigma_B, mu_A, sigma_A))
+  if(tau == Inf) return(ED_single(c, mu_A)) ;
+  if(tau == -Inf) return(ED_single(c, mu_B)) ;
+  if(tau == 0L) return(ED_single(c, mu_A + mu_B)) ;
+  if(tau > 0L) return(ED_async(tau, c, mu_A, sigma_A, mu_B)) ;
+  if(tau < 0L) return(ED_async(abs(tau), c, mu_B, sigma_B, mu_A))
 }
 
-mathout(g)
-```
-
-For instance, we can see that with $\tau = -167$ the expected first passage time is:
-
-```{r, echo=FALSE}
-mu_A = 0.53
-mu_B = 1.32
-sigma_A = 4.4
-sigma_B = 18.3
-tau = -167
-c = 100
-mu_M = 161
-
-g(tau, c, mu_A, sigma_A, mu_B, sigma_B)
-```
-
-To which we have to add $\mu_M = 161$ (which represents the amount of time that the movement of providing the answer takes) and we obtain:
-
-```{r, echo=FALSE}
-g(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
-```
-
-Similarly, we obtain the first-time passage for every $\tau$:
-
-```{r, results="asis", echo=FALSE}
-
-G <- Vectorize(g, vectorize.args = 'tau')
-
-h <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+mathout(call("=", quote(ED(tau, c, mu_A, sigma_A, mu_B, sigma_B)), ED))
+```
+
+The observable response time is assumed to be the sum of \(D\), the time
+employed to reach the threshold for the decision, and a residual \(M\) denoting
+other processes such a motor preparation and execution. Correspondingly, the
+expected response time amounts to
+
+```{r, results='asis'}
+ET <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+  ED(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
+
+mathout(call("=", quote(E(T[tau])), ET))
+```
+
+## Methods
+
+Schwarz [-@schwarz1994] applied the model to data from a redundant signals task
+with 13 onset
+asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
+where \(0\) refers to the synchronous condition, and \(\pm\infty\) to the
+single-target presentations. Each condition was replicated 400 times. The
+observed mean response times and their standard deviations are given in
+Table \@ref(tab:miller-data) below.
+
+```{r miller-data, echo=FALSE}
+tau  <- list(-Inf, -167L, -133L, -100L, -67L, -33L, 0L, 33L, 67L, 100L, 133L, 167L, Inf)
+m    <- c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
+s    <- c( 56,  58,  40,  40,  32,  28,  28,  28,  26,  30,  32,  34,  92)
+n    <- c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400)
+
+ttau <- lapply(tau, FUN=mathout, flags=list(cat=FALSE))
+tm   <- lapply(as.integer(m), FUN=mathout, flags=list(cat=FALSE))
+ts   <- lapply(as.integer(s), FUN=mathout, flags=list(cat=FALSE))
+tn   <- lapply(as.integer(n), FUN=mathout, flags=list(cat=FALSE))
+
+t <- cbind('$\\tau$'=ttau, '$m$'=tm, '$s$'=ts, '$n$'=tn)
+knitr::kable(t, caption="Miller data", row.names=FALSE, escape=FALSE)
+```
+
+Assuming that the model is correct, the observable mean reaction times follow an
+approximate Normal distribution around the model prediction \(E(T)\). We can,
+therefore, use a standard goodness-of-fit measure by \(z\)-standardisation.
+
+```{r, results='asis'}
+z <- function(m, s, n, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
+  dfrac(m - denote(E, ET(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M),
+              "the expected mean response time"),
+    s / sqrt(n))
+
+mathout(call("=", quote(z[tau]), z))
+```
+
+The overall goodness-of-fit is the sum of the squared \(z\)-statistics for each
+onset asynchrony. Assuming again that the architecture of the model is correct,
+but the parameters are adjusted to the data, it follows
+a \(\chi^2(8~\mathrm{df})\)-distribution. The degrees of freedom result from
+the 13 predicted mean response times, minus 5 free parameters.
+
+$X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\ df)$.
+
+```{r, results='asis'}
+gof <- function(par, tau, m, s, n)
 {
-  G(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
+  zn <- Vectorize(z, vectorize.args='tau')
+  zi <- zn(m, s, n, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])
+  sum(zi^2L)
 }
 
-mathout(h)
-```
-
-```{r, echo=FALSE}
-
-mu_M = 161
-mu_A = 0.53
-mu_B = 1.32
-sigma_A = 4.4
-sigma_B = 18.3
-tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
-c = 100
-
-h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
-```
-
-### Model fitting and Parameters Estimation
-
-Then we proceeded with the model fitting, using the data provided by Miller [-@miller1986] study in Schwarz [-@schwarz1994]. First, we calculated the predicted means RT as $E[RT]=E[T]+\mu_M$.
-
-We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed [@ratcliff2008] with at the first passage of $c$ $X(t) \sim N(\mu t , \sigma^2 t)$.
-Where $T$ is expressed as $T= \arg \min_t X(t)\geq c$ with $T \sim IG (c, \mu, \sigma^2)$ with mean $E(T)=\displaystyle \frac{c}{\mu}$ and variance $Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}$.
-Furthermore, for the Central Limit Theorem $\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})$
-
-For these reasons, in order to obtain the Goodness of Fit measure we proceeded with a z-standardisation.
-
-```{r, results="asis", echo=FALSE}
-zf <-  function(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
-{
-  E <- h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
-  dfrac(M-E, s/sqrt(N))
-}
-
-mathout(zf)
-```
-
-```{r, echo=FALSE}
-# The vector X contains the observed values, the vector s their observed standard deviation and N = 400 are the number of trials.
-
-M = c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
-s = c(56, 58, 40, 40, 32, 28, 28, 28, 26, 30, 32, 34, 92)
-N = c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400)
-
-zn <- zf(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
-zn
-```
-
-Then, we calculated the Goodness of Fit as follows:
-
-$X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)$.
-
-```{r, results="asis", echo=FALSE}
-par <- c(mu_A = 0.53, sigma_A = 4.4, mu_B = 1.32, sigma_B = 18.3, mu_M = 161)
-
-# Goodness of Fit
-
-gof <- function(par, tau, M, s, N)
-{
-  zn <- zf(M, s, N, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])
-  sum(zn^2L)
-}
-
 mathout(gof)
-```
-
-```{r, results="asis", echo=FALSE}
-n <- gof(par, tau, M, s, N)
-n
-pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)
 ```
@@
   optim(start, fn = gof,
-    tau = tau, M = M, s = s, N = N, hessian = TRUE, control = list(fnscale = 1))
+    tau = tau, m = M, s = s, n = N, hessian = TRUE, control = list(fnscale = 1))
 }
@@
 # Parameters' estimates
-fit <-  lsq(tau, M, s, N)
+fit <-  lsq(tau, m, s, n)
 
@@
 
-```{r}
-# Table 1: Estimated Model's Parameters
-#
-# | Parameters | Estimate | 95% Confidence Interval
-# | :---: | :---: | :---:|
-# | $ \mu_ \mathrm A $| 0.53 | 0.51, 0.56
-# | $ \sigma_ \mathrm A $| 4.4 | 3.8, 5.1
-# | $ \mu_ \mathrm B $| 1.32 | 1.2, 1.4
-# | $ \sigma_ \mathrm B $| 18.3 | 11.6, 24.9
-# | $ \mu_ \mathrm M $| 161 | 157, 165
-```
-
 Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i.
@@
 Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.
-
-```{r}
-# : Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's [-@miller1986] study in Schwarz [-@schwarz1994]
-#
-# | SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
-# |:--------:|:-----------------:|:----------------:|:-----------:|
-# $-\infty$| 237 | 231 | 56 |
-# $-167$   | 231 | 234 | 58 |
-# $-133$   | 230 | 230 | 40 |
-# $-100$   | 228 | 227 | 40 |
-# $-67$    | 226 | 228 | 32 |
-# $-33$    | 222 | 221 | 28 |
-# $0$      | 215 | 217 | 28 |
-# $33$     | 239 | 238 | 28 |
-# $67$     | 262 | 263 | 26 |
-# $100$    | 283 | 277 | 30 |
-# $133$    | 300 | 298 | 32 |
-# $167$    | 313 | 316 | 34 |
-# $\infty$ | 350 | 348 | 92 |
-#
-# _Note_.  Parameters: $\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161$
-```
-
-## Mistakes
-
-The case study presented above displays several possible ways of how \pkg{mathml} can be a useful tool in data representation and calculation.
-Furthermore, the careful reader might have also noticed that \pkg{mathml} allows also for a more smoothly detection of possible mistakes in writing.
-
-An interesting use case of the package is in this sense the readability of self-written
-functions in R\ code. R\ code is notoriously hard to read, as are most programming
-languages [@green1977]. The poor legibility of the language is one of the main
-sources of mistakes. For illustration, we consider Equation 10 in
-Schwarz [-@schwarz1994], namely the same equation reported above. The first version has a wrong parenthesis, which is barely visible in the R code.
-
-````{r, results="asis", echo=TRUE}
-f <- function(tau)
-{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
-    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
-      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
-        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
-}
-
-attributes(f)$name = quote(f)
-mathout(f)
-````
-
-But, if the function is rendered in MathML, the wrong parenthesis appears as a curly
-brace, and we immediately see the mistake. The correct version is shown below
-for comparison.
-
-````{r, results="asis", echo=TRUE}
-f <- function(tau)
-{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
-    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
-      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
-        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
-}
-
-attributes(f)$name = quote(f)
-mathout(f)
-````
-
-This example may seem trivial.
-However, as the reader well knows, a missed parenthesis is often the cause of untraceable errors and contradictory results.
-Thus, rendering the expression in MathMl improves the readability of the text, allowing better identification of the sources of possible errors.
 
@@
 
+Also: no linebreaks in long equations. Mostly because it is not properly
+supported by MathML renderers, and also in Latex, package "breqn" (insert
+reference) is mostly a proof of concept. No support for `\begin{align*}` with
+manual linebreaks etc.
+
 \pkg{mathml} is available for R\ version (…) and later, and can be easily installed using the usual `install.packages("mathml")`.
@@
 
+## Mistakes
+
+The case study presented above shows that \pkg{mathml} can help to improve
+data analyses and statistical reports from an aesthetical perspective, as well
+as regarding reproducibility of research. Furthermore, the package may also
+allow for a better detection of possible mistakes in R programs.
+
+Similar to most programming languages [@green1977], R\ code is notoriously hard
+to read, and the poor legibility of the language is one of the main sources of
+mistakes. For illustration, we consider again Equation 10 in
+Schwarz [-@schwarz1994].
+
+````{r, results="asis"}
+f1 <- function(tau)
+{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
+    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
+      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
+        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
+}
+
+mathout(f1)
+````
+
+The first version has a wrong parenthesis, which is barely visible in the code,
+whereas the wrong curly brace in the mathematical representation is immediately
+obvious. The correct version is shown below for comparison.
+
+````{r, results="asis"}
+f2 <- function(tau)
+{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
+    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
+      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
+        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
+}
+
+mathout(f2)
+````
+
+As the reader may know from own experience, missed parentheses are frequent
+causes of wrong results and errors that are hard to locate in programming code.
+This particular example shows that mathematical rendering can help to
+substantially reduce the amount of careless errors in programming.
+
 # Acknowledgments
+
+## Appendix
+
+Please check if the examples below
+
+* are already presented somewhere
+* can be added to one of the tables
+* or if they should go into a separate appendix
+
+_I removed the examples since they were already present._

### 2023-01-10 rj/mathml.Rmd
case study gof

@@
   dblup=quote(A %dblup% B),
-  dbldown=quote(A %dbldown% B))
+  dbldown=quote(A %dbldown% B),
+  " "="")
 
@@
 but the parameters are adjusted to the data, it follows
-a \(\chi^2(8~\mathrm{df})\)-distribution. The degrees of freedom result from
-the 13 predicted mean response times, minus 5 free parameters.
-
-$X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\ df)$.
-
-```{r, results='asis'}
+a \(\chi^2(8~\mathrm{df})\)-distribution.
+
+```{r, results='asis'}
+zv <- Vectorize(z, vectorize.args='tau')
+
+hook(call("zv", expression(m), expression(s), expression(n), expression(tau), 
+        expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B), expression(mu_M)), quote(z[tau]))
+
 gof <- function(par, tau, m, s, n)
-{
-  zn <- Vectorize(z, vectorize.args='tau')
-  zi <- zn(m, s, n, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])
-  sum(zi^2L)
-}
-
-mathout(gof)
-```
+  sum(zv(m, s, n, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], 
+         mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])^2L)
+
+mathout(call("=", quote(X["8 df"]^2L), gof))
+```
+
+The degrees of freedom result from the 13 predicted mean response times,
+minus 5 free parameters (\(c\) is a redundant scaling parameter).
 

### 2023-01-10 inst/pl/mathml.pl

@@ -793,10 +793,13 @@ math(function(na, Body, _), M),
 math(function(na, Body, _), M)
  => M = body([[Body]]).
 
-math(Curly, M),
+math(Curly, M, Flags),
     compound(Curly),
     compound_name_arguments(Curly, '{', Args)
- => M = body(Args).
+ => exclude(invisible_(Flags), Args, Args1),
+    M = body(Args1).
+
+invisible_(_Flags, invisible(_)).
 
 ml(body([R]), M, Flags)
  => ml(R, M, Flags).
@@ -811,6 +814,10 @@ jax(body([R]), M, Flags)
 jax(body(Body), M, Flags)
  => maplist(jax_(Flags), Body, Rows),
     format(string(M), "\\left\\{\\begin{array}{l}~w\\end{array}\\right.", [Rows]).
+
+% Hide
+math(invisible(_), M, _Flags)
+ => M = ''.
 
 % Vectors
 math(Hash, M),

### 2023-01-10 rj/mathml.Rmd
improve table with custom operators

@@
 op1 <- list(
-  cross=quote(A %*% B),
-  dot=quote(A %.% B),
-  kronecker=quote(A %x% B),
-  'integer division'=quote(A %/% B),
-  modulo=quote(A %% B),
-  and=quote(A & B),
-  vel=quote(A | B),
-  xor=quote(xor(A, B)),
-  not=quote(!A),
-  equal=quote(A == B),
-  assign=quote(A <- B))
+  "A\\ %*%\\ B"=quote(A %*% B),
+  "A\\ %.%\\ B"=quote(A %.% B),
+  "A\\ %x%\\ B"=quote(A %x% B),
+  "A\\ %/%\\ B"=quote(A %/% B),
+  "A\\ %%\\ B"=quote(A %% B),
+  "A\\ &\\ B"=quote(A & B),
+  "A\\ |\\ B"=quote(A | B),
+  "xor(A,\\ B)"=quote(xor(A, B)),
+  "!A"=quote(!A),
+  "A\\ ==\\ B"=quote(A == B),
+  "A\\ <-\\ B"=quote(A <- B))
 
@@
 
-op1 <- as.character(op1)
+op1 <- names(op1)
 if(knitr::is_latex_output())
@@
 op2 <- list(
-  different=quote(A != B),
-  tilde=quote(A ~ B),
-  approx=quote(A %~~% B),
-  equiv=quote(A %==% B),
-  congr=quote(A %=~% B),
-  prop=quote(A %prop% B),
-  'in'=quote(A %in% B),
-  intersect=quote(intersect(A, B)),
-  union=quote(union(A, B)),
-  crossprod=quote(crossprod(A, B)),
-  intersect=quote(is.null(A)))
+  "A\\ !=\\ B"=quote(A != B),
+  "A\\ ~ B"=quote(A ~ B),
+  "A\\ %~~%\\ B"=quote(A %~~% B),
+  "A\\ %==%\\ B"=quote(A %==% B),
+  "A\\ %=~%\\ B"=quote(A %=~% B),
+  "A\\ %prop%\\ B"=quote(A %prop% B),
+  "A\\ %in%\\ B"=quote(A %in% B),
+  "intersect(A,\\ B)"=quote(intersect(A, B)),
+  "union(A,\\ B)"=quote(union(A, B)),
+  "crossprod(A,\\ B)"=quote(crossprod(A, B)),
+  "is.null(A)"=quote(is.null(A)))
 
@@
 
-op2 <- as.character(op2)
+op2 <- names(op2)
 if(knitr::is_latex_output())
@@
 op3 <- list(
-  twosidearrow=quote(A %<->% B),
-  rightarrow=quote(A %->% B),
-  leftarrow=quote(A %<-% B),
-  uparrow=quote(A %up% B),
-  down=quote(A %down% B),
-  iff=quote(A %<=>% B),
-  rArr=quote(A %=>% B),
-  lArr=quote(A %<=% B),
-  dblup=quote(A %dblup% B),
-  dbldown=quote(A %dbldown% B),
+  "A\\ %<->%\\ B"=quote(A %<->% B),
+  "A\\ %->%\\ B"=quote(A %->% B),
+  "A\\ %<-%\\ B"=quote(A %<-% B),
+  "A\\ %up%\\ B"=quote(A %up% B),
+  "A\\ %down%\\ B"=quote(A %down% B),
+  "A\\ %<=>%\\ B"=quote(A %<=>% B),
+  "A\\ %=>%\\ B"=quote(A %=>% B),
+  "A\\ %<=%\\ B"=quote(A %<=% B),
+  "A\\ %dblup%\\ B"=quote(A %dblup% B),
+  "A\\ %dbldown%\\ B"=quote(A %dbldown% B),
   " "="")
@@
 
-op3 <- as.character(op3)
+op3 <- names(names(op3))
 if(knitr::is_latex_output())
@@
 ```{r, results='asis'}
-zv <- Vectorize(z, vectorize.args='tau')
+zv <- Vectorize(z, vectorize.args = c('tau', 'm', 's', 'n'))
 
@@
 The degrees of freedom result from the 13 predicted mean response times, 
-minus 5 free parameters (\(c\) is a redundant scaling parameter).
+minus 5 free parameters
+\(\theta = \langle\mu_{\mathrm A}, \sigma_{\mathrm A}, \mu_{\mathrm B}, \sigma_{\mathrm B}, \mu_{\mathrm M}\rangle\)
+(\(c\) is a redundant scaling parameter).
 
 ```{r, results="asis", echo=FALSE}
-lsq = function(tau, M, s, N)
+lsq = function(tau, m, s, n)
 {
-  start = c(mu_A = 0.3, sigma_A = 2, mu_B = 1 , sigma_B = 5, mu_M = 200)
-  optim(start, fn = gof,
-    tau = tau, m = M, s = s, n = N, hessian = TRUE, control = list(fnscale = 1))
+  invisible(theta <- c(mu_A=0.53, sigma_A=4.3, mu_B=1.34, sigma_B=11.7, mu_M=160))
+  optim(theta, gof, tau=tau, m=m, s=s, n=n, hessian=TRUE)
 }
 
-mathout(lsq)
-
-# Parameters' estimates
+mathout(call("=", quote(hat(theta)), lsq))
+```
+
+The best fitting parameter values and their confidence intervals are given in
+Table \@ref(tab:params) below.
+
+```{r params, echo=FALSE}
 fit <- lsq(tau, m, s, n)
-
-est = fit$par
+theta <- fit$par
 se = sqrt(diag(solve(fit$hessian)))
-cilo = qnorm(0.025, est, se)
-ciup = qnorm(0.975, est, se)
-
-# Lower bound
-#cilo
-
-# Upper bound
-#ciup
-
-```
-
-### Results
-
-The calculated Goodness of Fit value is $X^2= 30.54$, $p= 0.0002$. Additionally, we proceeded with the parameters' estimation with the Least Squares Method.
-
-The values of the estimated parameters reported in the table below are consistent with those reported by Schwarz [-@schwarz1994].
-
-Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i.
-$\tau= 100$) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model.
-Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.
+cilo = qnorm(0.025, theta, se)
+ciup = qnorm(0.975, theta, se)
+
+ntheta <- lapply(lapply(FUN=as.symbol, names(theta)), FUN=mathout, flags=list(cat=FALSE))
+ttheta <- lapply(theta, FUN=mathout, flags=list(cat=FALSE))
+tcilo <- lapply(cilo, FUN=mathout, flags=list(cat=FALSE))
+tciup <- lapply(ciup, FUN=mathout, flags=list(cat=FALSE))
+
+t <- cbind(Parameter=ntheta, Estimate=ttheta, Lo=tcilo, Up=tciup)
+knitr::kable(t, caption="Model fit", row.names=FALSE, escape=FALSE)
+```
+
+The statistic goodness-of-fit value is \(X^2(8) = 30.54\), \(p = .0002\), 
+indicating some lack of fit. Given the large trial numbers in the original
+study, some lack of fit is expected. For more detail, especially on fitting the
+observed standard deviations, the reader is referred to the original
+paper [@schwarz1994].
 

### 2023-01-10 rj/mathml.Rmd
improve Table 3

@@
 op1 <- list(
-  sin=quote(sin(x)),
-  cosh=quote(cosh(x)),
-  tanpi=quote(tanpi(alpha)),
-  asinh=quote(asinh(alpha)),
-  log=quote(log(p)),
-  log1p=quote(log1p(x)),
-  logb=quote(logb(x, e)),
-  exp=quote(exp(x)),
-  expm1=quote(expm1(x)),
-  choose=quote(choose(n,k)),
-  lchoose=quote(lchoose(n, k)),
-  factorial=quote(factorial(n)), 
-  lfactorial=quote(lfactorial(n)),
-  sqrt=quote(sqrt(x)),
-  mean=quote(mean(X)),
-  absolute=quote(abs(x)))
+  "sin(x)"=quote(sin(x)),
+  "cosh(x)"=quote(cosh(x)),
+  "tanpi(alpha)"=quote(tanpi(alpha)),
+  "asinh(alpha)"=quote(asinh(alpha)),
+  "log(p)"=quote(log(p)),
+  "log1p(x)"=quote(log1p(x)),
+  "logb(x,\\ e)"=quote(logb(x, e)),
+  "exp(x)"=quote(exp(x)),
+  "expm1(x)"=quote(expm1(x)),
+  "choose(n,\\ k)"=quote(choose(n, k)),
+  "lchoose(n,\\ k)"=quote(lchoose(n, k)),
+  "factorial(n)"=quote(factorial(n)), 
+  "lfactorial(n)"=quote(lfactorial(n)),
+  "sqrt(x)"=quote(sqrt(x)),
+  "mean(X)"=quote(mean(X)),
+  "abs(x)"=quote(abs(x)))
 m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))
 
-op1 <- as.character(op1)
+op1 <- names(op1)
 if(knitr::is_latex_output())
@@
 op2 <- list(
-  dbinom=quote(dbinom(k, N, pi)),
-  pbinom=quote(pbinom(k, N, pi)),
-  qbinom=quote(qbinom(p, N, pi)),
-  dpois=quote(dpois(k, lambda)),
-  ppois=quote(ppois(k, lambda)),
-  qpois=quote(qpois(p, lambda)),
-  dexp=quote(dexp(x, lambda)),
-  pexp=quote(pexp(x, lambda)),
-  qexp=quote(qexp(p, lambda)),
-  dnorm=quote(dnorm(x, mu, sigma)),
-  pnorm=quote(pnorm(x, mu, sigma)),
-  qnorm=quote(qnorm(alpha/2L)),
-  pchisq=quote(1L-pchisq(x, 1L)),
-  qchisq=quote(qchisq(1L-alpha, 1L)),
-  pt=quote(pt(t, N-1L)),
-  qt=quote(qt(alpha/2L, N-1L)))
+  "dbinom(k,\\ N,\\ pi)"=quote(dbinom(k, N, pi)),
+  "pbinom(k,\\ N,\\ pi)"=quote(pbinom(k, N, pi)),
+  "qbinom(p,\\ N,\\ pi)"=quote(qbinom(p, N, pi)),
+  "dpois(k,\\ lambda)"=quote(dpois(k, lambda)),
+  "ppois(k,\\ lambda)"=quote(ppois(k, lambda)),
+  "qpois(p,\\ lambda)"=quote(qpois(p, lambda)),
+  "dexp(x,\\ lambda)"=quote(dexp(x, lambda)),
+  "pexp(x,\\ lambda)"=quote(pexp(x, lambda)),
+  "qexp(p,\\ lambda)"=quote(qexp(p, lambda)),
+  "dnorm(x,\\ mu,\\ sigma)"=quote(dnorm(x, mu, sigma)),
+  "pnorm(x,\\ mu,\\ sigma)"=quote(pnorm(x, mu, sigma)),
+  "qnorm(alpha/2L)"=quote(qnorm(alpha/2L)),
+  "1L\\ -\\ pchisq(x,\\ 1L)"=quote(1L - pchisq(x, 1L)),
+  "qchisq(1L\\ -\\ alpha,\\ 1L)"=quote(qchisq(1L-alpha, 1L)),
+  "pt(t,\\ N\\ -\\ 1L)"=quote(pt(t, N-1L)),
+  "qt(alpha/2L,\\ N\\ -\\ 1L)"=quote(qt(alpha/2L, N-1L)))
 m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))
 
-op2 <- as.character(op2)
+op2 <- names(op2)
 if(knitr::is_latex_output())

### 2023-01-11 inst/pl/mathml.pl
diamond

@@ -331,6 +331,25 @@ mathml :-
     mathml(alpha).
 
 %
+% Symbols
+%
+math(R, M),
+    atom(R),
+    memberchk(R, [diamond])
+ => M = symbol(R).
+
+ml(symbol(R), M, _Flags)
+ => M = mi(&(R)).
+
+jax(symbol(R), M, _Flags)
+ => format(string(M), "\\~w", [R]).
+
+type(symbol(_), T, _Flags)
+ => T = atomic.
+
+
+
+%
 % Booleans
 %
 math(true, M)

### 2023-01-12 rj/mathml.Rmd
changes

@@
 Then, two possible solutions for typesetting are available to export R\ Markdown documents in different formats or ShinyApp webpages: MathML and MathJax.
+A further, not-yet-mentioned, use of \pkg{mathml} is related to presentations  made using R Markdown.
+By adopting \pkg{mathml} researchers or teachers can use the \R markdown to both conduct analysis and show results, smoothing the process and uniforming the graphical output.
 
@@
 
-In its current version \pkg{mathml} has some limitations.
+In its current version, however, \pkg{mathml} has some limitations.
 For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation.
@@
 _I removed the examples since they were already present._
+
+```{r, results='asis', echo=TRUE}
+mathout(quote(T^diamond))
+mathout(quote(sinpi(alpha)))
+```

### 2023-01-13 rj/mathml.Rmd
fix broken table column

@@
 
-op3 <- names(names(op3))
+op3 <- names(op3)
 if(knitr::is_latex_output())

### 2023-01-13 rj/mathml.Rmd
context-dependent rendering of mistakes

@@
 
+## Context-dependent rendering
+
+Consider an educational scenario in which we want to highlight a certain
+element of a term, for example, that a student has forgotten to subtract the
+null hypothesis in a \(t\)-ratio:
+
+```{r, results='asis'}
+t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
+mathout(t, flags=list(error='highlight'))
+```
+
+The R function `omit_right(a + b)` uses non-standard evaluation
+techniques [e.g., @Wickham2019] to return only the left part an operation,
+and cancels the right part. This may not always be desired, for example, when
+illustrating how to fix the mistake.
+
+For this purpose, the functions `mathml()` or `mathjax()` have an optional
+argument `flags` which is a list with named elements. In this example, we use
+this argument to tell \pkg{mathml} how to render such erroneous
+expressions (asis, highlight, fix, or ignore), see Table \@ref(tab:mistakes).
+
+```{r mistakes, echo=FALSE}
+op1 <- list(
+  "omit_left(a\\ +\\ b)"=quote(omit_left(a + b)),
+  "omit_right(a\\ +\\ b)"=quote(omit_right(a + b)),
+  "list(quote(a),\\ quote(omit(b)))"=list(quote(a), quote(omit(b))),
+  "add_left(a\\ +\\ b)"=quote(add_left(a + b)),
+  "add_right(a\\ +\\ b)"=quote(add_right(a + b)),
+  "list(quote(a),\\ quote(add(b)))"=list(quote(a), quote(add(b))),
+  "instead(a,\\ b)\\ +\\ c"=quote(instead(a, b) + c))
+
+asis <- lapply(op1, FUN=mathout, flags=list(cat=FALSE, error="asis"))
+high <- lapply(op1, FUN=mathout, flags=list(cat=FALSE, error="highlight"))
+fix  <- lapply(op1, FUN=mathout, flags=list(cat=FALSE, error="fix"))
+igno <- lapply(op1, FUN=mathout, flags=list(cat=FALSE, error="ignore"))
+
+op1 <- names(op1)
+if(knitr::is_latex_output())
+  op1 <- sapply(op1, FUN=knitr:::escape_latex)
+if(knitr::is_html_output())
+  op1 <- sapply(op1, FUN=knitr:::escape_html)
+
+t <- cbind(Operation=op1, 
+  "error asis"=asis, highlight=high, fix=fix, ignore=igno)
+knitr::kable(t, caption="Highlighting elements of a term",
+  row.names=FALSE, escape=FALSE)
+```
+
 ## Customizing the package

### 2023-01-13 R/mathml.R

@@ -32,7 +32,7 @@
 mathml <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL,
   env=globalenv())
 {
-  flags <- c(attributes(term), flags, list(cat=FALSE))
+  flags <- c(flags, list(cat=FALSE))
   t <- rolog::once(call("r2mathml", term, expression(X), flags),
     options=list(preproc=list(rolog::preproc, mathml_preproc)),
     env=env)
@@ -112,7 +112,7 @@ mathml_preproc <- function(query=quote(5 %% 2))
 mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL,
   env=globalenv())
 {
-  flags <- c(attributes(term), flags, list(cat=FALSE))
+  flags <- c(flags, list(cat=FALSE))
   t <- rolog::once(call("r2mathjax", term, expression(X), flags),
     options=list(preproc=list(rolog::preproc, mathml_preproc)),
     env=env)

### 2023-01-13 rj/mathml.Rmd
context-dependent rendering of mistakes

@@
 mathout(t, flags=list(error='highlight'))
+mathout(t, flags=list(error='fix'))
 ```
@@
 this argument to tell \pkg{mathml} how to render such erroneous
-expressions (asis, highlight, fix, or ignore), see Table \@ref(tab:mistakes).
+expressions using the flag `error` which can be asis, highlight, fix, or ignore.
+For more examples, see Table \@ref(tab:mistakes).
 
@@
 t <- cbind(Operation=op1, 
-  "error asis"=asis, highlight=high, fix=fix, ignore=igno)
+  "error\\ =\\ asis"=asis, highlight=high, fix=fix, ignore=igno)
 knitr::kable(t, caption="Highlighting elements of a term",

### 2023-01-13 inst/pl/mathml.pl

@@ -2163,140 +2163,129 @@ type(underbrace(A, _), Type, Flags)
 %
 % Mistakes
 %
+option_(NameOption, Flags) :-
+    option(NameOption, Flags).
+
+option_(NameOption, Flags) :-
+    compound_name_arguments(NameOption, Name, [Option]),
+    atom_string(Option, String),
+    member(Name-String, Flags).
+
 math(omit_left(Expr), M, Flags),
-    option(error(ignore), Flags, highlight)
+    option_(error(ignore), Flags)
  => M = Expr.
 
 math(omit_left(Expr), M, Flags),
-    option(error(asis), Flags, highlight),
+    option_(error(asis), Flags),
     Expr =.. [_Op, _L, R]
  => M = R.
 
 math(omit_left(Expr), M, Flags),
-    option(error(fix), Flags, highlight),
+    option_(error(fix), Flags),
     Expr =.. [Op, L, R]
  => M = list(space, [box(list(space, [L, op(Op)])), R]).
 
-math(omit_left(Expr), M, Flags),
-    option(error(highlight), Flags, highlight),
+math(omit_left(Expr), M, _Flags), % default
     Expr =.. [Op, L, R]
  => M = list(space, [cancel(list(space, [L, op(Op)])), R]).
 
 math(omit_right(Expr), M, Flags),
-    option(error(ignore), Flags, highlight)
+    option_(error(ignore), Flags)
  => M = Expr.
 
 math(omit_right(Expr), M, Flags),
-    option(error(asis), Flags, highlight),
+    option_(error(asis), Flags),
     Expr =.. [_Op, L, _R]
  => M = L.
 
 math(omit_right(Expr), M, Flags),
-    option(error(fix), Flags, highlight),
+    option_(error(fix), Flags),
     Expr =.. [Op, L, R]
  => M = list(space, [L, box(list(space, [op(Op), R]))]).
 
-math(omit_right(Expr), M, Flags),
-    option(error(highlight), Flags, highlight),
+math(omit_right(Expr), M, _Flags),
     Expr =.. [Op, L, R]
  => M = list(space, [L, cancel(list(space, [op(Op), R]))]).
-
-math(omit(Expr), M, Flags),
-    option(error(ignore), Flags, highlight)
- => M = Expr.
-
-math(omit(Expr), M, Flags),
-    option(error(fix), Flags, highlight)
- => M = box(Expr).
-
[…]

### 2023-01-13 rj/mathml.Rmd
add ref

@@
 Since recently [@R, version 4.2], R\'s manual pages include support for
-mathematical expressions [@Sarkar2022], which already is a big improvement.
-However, rules for the mapping of built-in language elements to their
-mathematical representation are still lacking. So far, R\ expressions such
+mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a big
+improvement. However, rules for the mapping of built-in language elements to
+their mathematical representation are still lacking. So far, R\ expressions such
 as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae

### 2023-01-14 rj/mathml.Rmd
simplified hooks

@@
 
-hook(quote(mysign), quote(sg(x)))
+hook(mysign, sg(x))
 term <- call("<-", quote(mysign), sgn)
@@
 ```{r, results='asis'}
-hook(quote(m_A), quote(mean(X)["A"])); hook(quote(s2_A), quote(s["A"]^2L));
-hook(quote(n_A), quote(n["A"]))
-hook(quote(m_B), quote(mean(X)["B"])); hook(quote(s2_B), quote(s["B"]^2L))
-hook(quote(n_B), quote(n["B"])); hook(quote(s2_p), quote(s["pool"]^2L))
+hook(m_A, mean(X)["A"]) ; hook(s2_A, s["A"]^2L) ;
+hook(n_A, n["A"])
+hook(m_B, mean(X)["B"]) ; hook(s2_B, s["B"]^2L)
+hook(n_B, n["B"]) ; hook(s2_p, s["pool"]^2L)
 
@@
 
-hook(quote(successes), quote(k))
-hook(quote(Ntotal), quote(N))
-hook(quote(prob), quote(pi))
+hook(successes, k)
+hook(Ntotal, N)
+hook(prob, pi)
 mathout(term)
@@
 mathematical symbols, as illustrated in the example. Care should be taken to
-avoid recursive hooks such as `hook(quote(s), quote(s["A"]))` that endlessly
-replace the \(s\) from \(s_{\mathrm{A}}\) as
+avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
+the \(s\) from \(s_{\mathrm{A}}\) as
 in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).
@@
 
-```{r, echo=FALSE}
-hook(quote(mu_A), quote(mu["A"]))
-hook(quote(mu_B), quote(mu["B"]))
-hook(quote(sigma_A), quote(sigma["A"]))
-hook(quote(sigma_B), quote(sigma["B"]))
-hook(quote(mu_M), quote(mu["M"]))
-hook(quote(zn), quote(z[tau]))
-hook(quote(M), quote(overline(X)))
-```
-
 Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
@@
 ```{r, results='asis'}
-hook(call("ED_single", expression(c), expression(mu)), 
-     call("E", call(";", quote(D), call(",", expression(c), expression(mu)))))
-
 ED_single <- function(c, mu)
@@
 
+# display as E(D; mu), c is a scaling parameter
+hook(ED_single(.C, .Mu), E(`;`(D, .Mu)))
 mathout(call("=", quote(ED_single(c, mu)), ED_single))
@@
 ```{r, results='asis'}
+hook(mu_A, mu["A"])
+hook(mu_B, mu["B"])
+hook(sigma_A, sigma["A"])
+hook(sigma_B, sigma["B"])
+hook(mu_M, mu["M"])
+hook(M, overline(X))
+
 mathout(call("=", quote(E(D["AB"])), quote(ED_single(c, mu_A + mu_B))))
@@
 ```{r, results="asis"}
-hook(call("ED_async", expression(tau), expression(c), expression(mu_A), expression(sigma_A), expression(mu_B)), 
-     call("E", call(";", call("[", quote(D), expression(tau)), 
-                    call(",", expression(c), expression(mu_A), expression(sigma_A), expression(mu_B)))))
-
 ED_async <- function(tau, c, mu_A, sigma_A, mu_B)
@@
 
+hook(ED_async(.Tau, .C, .MA, .SA, .MB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB))))
 mathout(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
@@
 ```{r, results='asis'}
-hook(call("ED", expression(tau), expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B)), 
-     call("E", call(";", call("[", quote(D), expression(tau)), 
-                    call(",", expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B)))))
-
 ED <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
@@
 
+hook(ED(.Tau, .C, .MA, .SA, .MB, .SB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB, .SB))))
 mathout(call("=", quote(ED(tau, c, mu_A, sigma_A, mu_B, sigma_B)), ED))
@@
 
-hook(call("zv", expression(m), expression(s), expression(n), expression(tau), 
-        expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B), expression(mu_M)), quote(z[tau]))
-
 gof <- function(par, tau, m, s, n)
@@
 
+hook(zv(.M, .S, .N, .Tau, .C, .MA, .SA, .MB, .MM), z(.Tau))
 mathout(call("=", quote(X["8 df"]^2L), gof))

### 2023-01-14 R/mathml.R

@@ -285,17 +285,38 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' @param display
 #' an R call or symbol/number. This is shown instead of _term_.
 #'
+#' @param quote (default is TRUE)
+#' indicates that _term_ and _display_ should be quoted.
+#'
+#' @param as.rolog (default is TRUE)
+#' indicates that simplified quasi-quotoation is to be used.
+#'
 #' @return
 #' TRUE on success
 #'
 #' @md
 #'
 #' @examples
-#' hook(term=quote(t0), display=quote(subscript(t, 0)))
+#' hook(t0, subscript(t, 0))
 #' mathml(quote(t0))
 #'
-hook <- function(term=quote(t0), display=quote(subscript(t, 0)))
+#' hook(term=quote(t0), display=quote(subscript(t, 0)), quote=FALSE)
+#' mathml(quote(t0))
+#'
+hook <- function(term, display, quote=TRUE, as.rolog=TRUE)
 {
+  if(quote)
+  {
+    term <- substitute(term)
+    display <- substitute(display)
+  }
+
+  if(as.rolog)
+  {
+    term <- rolog::as.rolog(term)
+    display <- rolog::as.rolog(display)
+  }
+
   r <- rolog::once(call("assert", call("math_hook", term, display)))
   if(isFALSE(r))
     return(FALSE)

### 2023-01-14 rj/mathml.Rmd
amendment to previous

@@
 ```{r, results='asis'}
-zv <- Vectorize(z, vectorize.args = c('tau', 'm', 's', 'n'))
+zv <- Vectorize(z, vectorize.args = c('m', 's', 'n', 'tau'))
 
@@
 
-hook(zv(.M, .S, .N, .Tau, .C, .MA, .SA, .MB, .MM), z(.Tau))
+hook(zv(.M, .S, .N, .Tau, .C, .MA, .SA, .MB, .SB, .MM), z[.Tau])
 mathout(call("=", quote(X["8 df"]^2L), gof))
@@
 paper [@schwarz1994].
-
-## Errors
-
-An additional feature that this package offers is related to its application in educational scenarios.
-The functions here reported are intended to identify and highlight possible errors that students make, as well as provide the correct answer for the task.
-
-In this first example the function highlights that the student forgot to add the left-hand side of the expression and it does that by striking the missing part.
-In this case the letter A and the operator `+`.
-Similarly the function `omit_right` does the same for the omissions in the right-hand side of the expression.
-Finally the function `omit()` can be used, for instance, to highlight the omission in the elements of a list.
-
-```{r, results="asis"}
-t <- quote(omit_left(a + b))
-mathout(t)
-```
-
-```{r, results="asis"}
-t <- quote(omit_right(a + b))
-mathout(t)
-```
-
-```{r, results="asis"}
-t <- list(quote(a), quote(b), quote(omit(c)))
-mathout(t)
-```
-
-```{r omit, echo=FALSE}
-op <- list(left=quote(omit_left(a + b)), right=quote(omit_right(a + b)), list=quote(omit(c)))
-m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))
-
-op <- as.character(op)
-if(knitr::is_latex_output())
-  op <- sapply(op, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op <- sapply(op, FUN=knitr:::escape_html)
-
-t <- cbind(Operator=op, Output=m)
-knitr::kable(t, caption="Omit", row.names=FALSE, escape=FALSE)
-```
-
-The functions `add_left()`, `add_right()`, `add()` reported below follow the same way of reasoning, but they account for the mistakes made in adding incorrect elements in the expression.
-They highlight the redundant element placing it in a box.
-
-```{r, results="asis"}
-t <- quote(add_left(a + b))
-mathout(t)
-```
-
-```{r, results="asis"}
-t <- quote(add_right(a + b))
-mathout(t)
-```
-
-```{r, results="asis"}
-t <- list(quote(a), quote(b), quote(add(c)), quote(d))
-mathout(t)
-```
-
-```{r add, echo=FALSE}
-op <- list(left=quote(add_left(a + b)), right=quote(add_right(a + b)))
-m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))
-
-op <- as.character(op)
-if(knitr::is_latex_output())
-  op <- sapply(op, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op <- sapply(op, FUN=knitr:::escape_html)
-t <- cbind(Operator=op, Output=m)
-
-knitr::kable(t, caption="Add", row.names=FALSE, escape=FALSE)
-```
-
-Finally the function `instead( , )` place a curly bracket under the incorrect element  and provides its correct version under the curly bracket.
-In order to provide the correction with this function one must fill it with its wrong and correct terms in this order: `instead(wrong, correct)`.
-An example of its use its provided below.
-
-```{r, results="asis"}
-t <- quote(instead(a, b) + c)
-mathout(t)
-```
-
-It is also possible to use flags for context-dependent rendering.
-The `highlight` flag is the default one.
-It shows where the error is and what is its solution.
-The `ignore` flag allows the program to ignore the fact that there is an error and print the expression as if it were correct.
-The third flag, `fix`, tells you what is the term that you are supposed to fix without giving the solution.
-Finally, the last flag `asis` returns the incorrect expression as the student wrote it, without providing any solutions or corrections.
-Below is provided the code with its comment and an example of use.
-
-Let us consider for example a student that instead of writing `b + c` wrote `a+c`.
-
-With the `asis` flag we report the incorrect expression provided by the student:
-
-```{r, results="asis"}
-# It prints the incorrect version
-t <- quote(instead(a, b) + c)
-mathout(t, flags=quote(error(asis)))
-```
-
-Then, we may want to underline the incorrect term and provide the correct solution.
-
-```{r, results="asis"}
-# It highlights what has been mistakenly inserted
-mathout(t, flags=quote(error(highlight))) # default
-```
-
-Alternatively, we can put the correct solution in a box to let the student focus on its correction.
-
-```{r, results="asis"}
-# It highlights what has been corrected
-mathout(t, flags=quote(error(fix)))
-```
-
-Finally, we can simply provide the correct solution to the problem wit the flag `ignore`.
-
-```{r, results="asis"}
-# It prints the correct version
-mathout(t, flags=quote(error(ignore)))
-```
 

### 2023-01-14 rj/mathml.Rmd
finished the vignette

@@
   mapping specific variables or functions to new MathML and LaTeX entities. A
-  few working examples illustrate the use of the package in a scientific report
-  and in on an interactive web page.
+  few working examples illustrate the use of the package in a scientific report.
 draft: true
@@
 
-The paper is organized as follows. Section\ 2 presents the technical background
-of the package, including the two main classes of Prolog rules for translating
-R\ objects to mathematical expressions. Section\ 3 illustrates the main features
-of the \pkg{mathml} package, potential issues and workarounds using examples
-from the day-to-day perspective of a user. Sections\ 4 and\ 5 present two case
-studies for the use of the package in a scientific report and 
-a **ShinyApp web page**. Section\ 6 concludes with a discussion and ideas for
-further development. A complete list of the R\ elements covered by the package
-is found in the Appendix.
-
-*shinyapp missing so far
+The vignette is organized as follows. Section\ 2 presents the technical
+background of the package, including the two main classes of Prolog rules for
+translating R\ objects to mathematical expressions. Section\ 3 illustrates the
+main features of the \pkg{mathml} package, potential issues and workarounds
+using examples from the day-to-day perspective of a user. Section\ 4 present a
+case study for the use of the package in a scientific report. Section\ 5
+concludes with a discussion and ideas for further development.
 
@@
 hook(successes, k)
-hook(Ntotal, N)
+hook(quote(Ntotal), quote(N), quote=FALSE)
 hook(prob, pi)
@@
 to replace elements (e.g., verbose variable names) of the code by concise
-mathematical symbols, as illustrated in the example. Care should be taken to
+mathematical symbols, as illustrated in the example. To simplify notation,
+the `quote` flag of `hook()` defaults to TRUE, and `hook()` uses non-standard
+evaluation to unpack its arguments. If quote is FALSE, as shown above, the user
+has to provide the quoted expressions. Care should be taken to
 avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
@@
 
-more stuff: math/2, ml/3, jax/3
+Further customizations require the assertion of custom Prolog rules `math/2`,
+`ml/3`, `jax/3`, as shown in the introduction.
 
@@
 
+hook(ET(.Tau, .C, .MA, .SA, .MB, .SB, .MM), 
+     E(`;`(T[.Tau], `,`(.MA, .SA, .MB, .SB, .MM))))
 mathout(call("=", quote(E(T[tau])), ET))
@@
 ```{r, results="asis", echo=FALSE}
-lsq = function(tau, m, s, n)
+lsq <- function(tau, m, s, n)
 {
@@
 theta <- fit$par
-se = sqrt(diag(solve(fit$hessian)))
-cilo = qnorm(0.025, theta, se)
-ciup = qnorm(0.975, theta, se)
+se <- sqrt(diag(solve(fit$hessian)))
+cilo <- qnorm(0.025, theta, se)
+ciup <- qnorm(0.975, theta, se)
 
@@
 
-This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time.
-This makes it easier for the user to notice writing errors and correct them smoothly.
-Building on current features of R\ and existing packages for displaying formulas in R\ (e.g.
-Murrell and Ihaka [-@murrell2000], Allaire et al.
-[-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility.
-To do so, it uses the Prolog's grammar rules and translates the R\ expressions to MathML.
-Then, two possible solutions for typesetting are available to export R\ Markdown documents in different formats or ShinyApp webpages: MathML and MathJax.
-A further, not-yet-mentioned, use of \pkg{mathml} is related to presentations  made using R Markdown.
-By adopting \pkg{mathml} researchers or teachers can use the \R markdown to both conduct analysis and show results, smoothing the process and uniforming the graphical output.
-
-The article and Appendix show examples of how the package can be used and what operations are currently possible.
-Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.
-
-In its current version, however, \pkg{mathml} has some limitations.
-For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation.
-However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations.
-Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications.
-
-Also: no linebreaks in long equations. Mostly because it is not properly
-supported by MathML renderers, and also in Latex, package "breqn" (insert
-reference) is mostly a proof of concept. No support for `\begin{align*}` with
-manual linebreaks etc.
-
-\pkg{mathml} is available for R\ version (…) and later, and can be easily installed using the usual `install.packages("mathml")`.
-The source code of the package is found at (…) including installation instructions for Unix, Windows and macOS.
-
-## Mistakes
+This package expands the possibilities of R\ by allowing calculations and
+graphical rendering of the same terms. Building on current features of R\ and
+existing packages for displaying formulas in
+R\ [@murrell2000;@allaire2018]), the \pkg{mathml} package bridges the gap
+between computational needs, presentation of results, and their reproducibility.
+Two possible modes for typesetting are available to export R\ Markdown documents
+in different formats or ShinyApp webpages: MathML and MathJax. A further, not-yet-mentioned, use of \pkg{mathml} is related to presentations made using
+RMarkdown. By adopting \pkg{mathml} researchers or teachers can use the
+RMarkdown to both conduct analysis and show results, smoothing the process and
+uniforming the graphical output.
 
@@
 
-# Acknowledgments
+In its current version \pkg{mathml} has some limitations. For example, it is
+currently not possible to use the functions of \pkg{mathml} for writing
+intratextual formulas; here, the user has to adopt the usual LateX notation.
+However, we do not consider this to be a serious limitation, since the most
+productive use of \pkg{mathml} still comes from more complicated calculations.
 
-## Appendix
+Also: In long equations, we did not yet find a convenient way to insert
+linebreaks. This is mostly because it is not properly supported by MathML and
+LaTeX renderers. In its current stage, the LaTeX package \pkg{breqn} [@breqn]
+is mostly a proof of concept.
+
+\pkg{mathml} is available for R\ version (…) and later, and can be easily
+installed using the usual `install.packages("mathml")`. The source code of the
+package is found at https://github.com/mgondan/mathml.
+
+# Appendix
 
@@
 ```
+
+# References

### 2023-01-16 rj/mathml.Rmd
amend
Update authors

@@
 #
-# paste(x, y, z)
-# Spacing
-# Comma separated lists
-# Ellipsis
-# Radicals
-# Subsets
-# Greek letters in MathJax
-# Typeface
-# Stylechanges
-# Phantoms
+# paste(x, y, z)--> z, y, z (Covered by quote ())
+
+# Spacing: operator for separating elements in expression. E.g.,
+## x ~ y --> x  y ;
+## x ~ ~ y --> x    y
+
+# Ellipsis: ldots or cdots. E.g., 
+## x[1] + ... + x[n]
+
+# Radicals sqrt(x, y) --> y√x
+
+# Sets: some symbols in set relations are missing:
+## supset, supseteq, notsubset, subset, subseteq, notin
+
+# Phantom: command that produces a space in place of an argument (partially covered by omit() (?)). E.g.:
+## x + phantom(0) + y --> x +    + y
+## x + over(1, vphantom(0)) --> x + 1/
+
+# Typeface: plain, bold, italic, bolditalic
+
+# Stylechanges: textstyle(x), scriptstyle(x) scriptscript(x)
+
+# linear models: DepVar ~ X0 + X1
+
+# hat, overline, cancel, box, invisible, phantom
 
@@
 ```
-
-# References

### 2023-01-16 R/mathml.R
Update man page for option env

@@ -8,6 +8,8 @@
 
 #' MathML output
 #'
+#' @md
+#' 
 #' @param term
 #' an R call or symbol/number. This function translates _term_ into a
 #' MathML string.
@@ -16,15 +18,18 @@
 #' list of flags that control the translation
 #'
 #' @param env (default globalenv())
-#' The R environment in which r_eval is being executed (see vignette for
-#' details, "Ringing back to R").
-#'
+#' The R environment in which r_eval is being executed.
+#' 
 #' @return
 #' A string with the MathML representation or _term_.
 #'
-#' @md
-#'
 #' @seealso [mathjax()]
+#'
+#' @details In some functions, the Prolog code may ring back R, for example, to
+#' find the names of function arguments. For example (see vignette), when
+#' rendering the call `integrate(g, lower=0L, upper=Inf)` as Int_0^Inf g(x) dx,
+#' Prolog needs to know that the function g is a function of x. The Prolog rule
+#' then searches for the formalArgs of g in the environment _env_.
 #'
 #' @examples
 #' mathml(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
@@ -103,6 +108,12 @@ mathml_preproc <- function(query=quote(5 %% 2))
 #' A string with the MathJax representation or _term_.
 #'
 #' @md
+#'
+#' @details In some functions, the Prolog code may ring back R, for example, to
+#' find the names of function arguments. For example (see vignette), when
+#' rendering the call `integrate(g, lower=0L, upper=Inf)` as Int_0^Inf g(x) dx,
+#' Prolog needs to know that the function g is a function of x. The Prolog rule
+#' then searches for the formalArgs of g in the environment _env_.
 #'
 #' @seealso [mathml()]
 #'

### 2023-01-21 inst/pl/mathml.pl
Section 3

@@ -2040,7 +2040,7 @@ paren(frac(_, _), P, _Flags)
 
 prec(frac(_, _), P, _Flags)
  => current(P0, yfx, /),
-    P is P0. % was - 1
+    P is P0 + 1. % was - 1
 
 type(frac(_, _), Type, _Flags)
   => Type = fraction.

### 2023-01-21 rj/mathml.Rmd
preamble

@@
 ---
-title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
-date: "2022-12-09"
+title: "mathml: Translate R expressions to MathML and LaTeX"
+date: "2023-01-21"
 abstract: >
@@
 author:  
-  # see ?rjournal_article for more information
   - name: Matthias Gondan
-    affiliation: University of Innsbruck
+    affiliation: Universität Innsbruck
     address:
@@
     orcid: 0000-0002-8409-8900
-    affiliation: University of Innsbruck
+    affiliation: Universität Innsbruck
     address:
@@
     toc: no
-bibliography: RJreferences.bib
+bibliography: bibliography.bib
 preamble:
-- \usepackage{longtable}
 - \usepackage{cancel}
@@
 Vignette: >
-  %\VignetteIndexEntry{mathml}
+  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
   %\VignetteEngine{knitr::rmarkdown}

### 2023-01-21 rj/mathml.Rmd
intro
pdf

@@
   mapping specific variables or functions to new MathML and LaTeX entities. A
-  few working examples illustrate the use of the package in a scientific report.
+  number of examples and a case study illustrate the use of the package in a
+  scientific report.
 draft: true
@@
 # Radicals sqrt(x, y) --> y√x
+
+# Chains: a < b < c
 
@@
 
-The vignette is organized as follows. Section\ 2 presents the technical
-background of the package, including the two main classes of Prolog rules for
-translating R\ objects to mathematical expressions. Section\ 3 illustrates the
-main features of the \pkg{mathml} package, potential issues and workarounds
-using examples from the day-to-day perspective of a user. Section\ 4 present a
-case study for the use of the package in a scientific report. Section\ 5
-concludes with a discussion and ideas for further development.
+The paper is organized as follows. Section\ 2 presents the technical background
+of the package, including the two main classes of Prolog rules for translating
+R\ objects to mathematical expressions. Section\ 3 illustrates the main features
+of the \pkg{mathml} package, potential issues and workarounds using examples
+from the day-to-day perspective of a user. Section\ 4 present a case study for
+the use of the package in a scientific report. Section\ 5 concludes with a
+discussion and ideas for further development.
 
@@
 The translation of R\ expressions to mathematical output is achieved through a
-Prolog interpreter. Prolog is a classical logic programming language with many
-applications in expert systems, computer linguistics and symbolic artificial
-intelligence. The main strength of Prolog is its concise representation of facts
-and rules for the representation of knowledge and grammar, as well as its
-efficient built-in search engine for closed world domains. As it is well-known,
-R\ is a statistical programming language for data analysis and statistical
-modeling which is widely used in academia and industry. Besides the core
-library, a lot of packages have been developed for all kinds of statistical
-problems, including statistics-based artificial intelligence tools such as
-neural networks for machine learning and deep learning. Whereas Prolog is weak
-in statistical computation, but strong in symbolic manipulation, the converse
-may be said for the R\ language. The \pkg{rolog} package [@rolog] bridges this
-gap by providing an interface to a SWI-Prolog distribution in R. The
-communication between the two systems is mainly in the form of queries from
-R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms in
-R.
+Prolog interpreter from R\ package \pkg{rolog} [@rolog]. Prolog is a classical logic
+programming language with many applications in expert systems, computer
+linguistics and symbolic artificial intelligence. The main strength of Prolog is
+its concise representation of facts and rules for knowledge and grammar, as well
+as its efficient built-in search engine for closed world domains. As it is
+well-known, R\ is a statistical programming language for data analysis and
+statistical modeling. Whereas Prolog is weak in statistical computation, but
+strong in symbolic manipulation, the converse may be said for the R\ language.
+The \pkg{rolog} package bridges this gap by providing an interface to a SWI-Prolog
+distribution [@swipl] in R. The communication between the two systems is mainly
+in the form of queries from R\ to Prolog, but two predicates allow Prolog to
+ring back and evaluate terms in R.
 
@@
 As is seen from the output above, the variable `term` is not assigned the result
-of the calculation, but an R\ "call"
-[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
+of the calculation, but an R\ call
+[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. Such a call
 can eventually be evaluated with `eval()`,
@@
 in the R\ code chunk, as is done in the next example. The
-R\ function `mathout()` is a wrapper that invokes `mathml()` for HTML output
-and `mathjax()` for LaTeX output.
+R\ function `mathout()` has been defined in this vignette; it invokes `mathml()`
+for HTML output and `mathjax()` for LaTeX output.
 
@@
 for MathJax/LaTeX. A list of flags can be used for context-sensitive
-translation (see, e.g., the errors below).
+translation (see, e.g., the section on errors below).
 
@@
 accordingly, without the parentheses. Parentheses around negated numbers or
-symbols can be added as shown for `-Inf`.
+symbols can be added as shown above for `+ (-Inf)`.
 
-Note that an R function `hat()` does not exist in base R, it is provided by the
-package for convenience and points to the identity function.
+Note that an R\ function `hat()` does not exist in base R, it is provided by the
+package for convenience and points to the identity function. Quite a few such
+convenience functions have been defined
+in \pkg{mathml} (see, e.g., next section, and the manual for details).
 
@@
 is omitted. This heuristic does not always produce the desired result;
-therefore, \pkg{mathml} defines alternative multiplication
-functions `dot()`, `nodot()`, and `times()` that calculate the product and
-produce the respective multiplication signs. Similarly, `frac()` and `dfrac()`
-can be used for small and large fractions.
+therefore, \pkg{mathml} defines alternative R\ functions `dot()`, `nodot()`,
+and `times()`. These functions calculate a product and produce the respective
+multiplication signs. Similarly, `frac()` and `dfrac()` can be used for small
+and large fractions.
 
@@
 mathout(term)
-term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N)))
+term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N))) # functional form of '%+-%'
 term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
@@
 custom operators in R, and that expressions with such operators are evaluated
-strictly from left to right. As in the previous example, the solution is to
-either use the functional form of the operator or a curly brace to enforce the
-correct operator precedence.
+strictly from left to right. Again, the problem can be worked around by the
+functional form of the operator, or a curly brace to hide the parenthesis but
+enforce the correct operator precedence.
 
@@
 
-There is support for most functions from \pkg{base}, with adequate use and omission
-of parentheses.
+There is support for most functions from package _base_, with adequate use and
+omission of parentheses.
 
@@
 mathout(term)
-term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + ceiling(x))
+term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
 mathout(term)
@@
   "tanpi(alpha)"=quote(tanpi(alpha)),
-  "asinh(alpha)"=quote(asinh(alpha)),
+  "asinh(x)"=quote(asinh(x)),
   "log(p)"=quote(log(p)),
@@
 
-For self-written functions, matters are a bit more complicated. The name of a
-self-written function is not transparent in R, since only the function body is
-represented, as illustrated by the redefinition of `sign` below. The function is
-often a nested R "call" of the form `'{'(L)`, with `L` being a list of
-commands. \pkg{mathml} provides very limited support for typical control
-structures.
+For self-written functions, matters are a bit more complicated. For a function
+such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
+only the function body is represented, as illustrated by the redefinition
+of `sign` below.
 
@@
 
-term <- quote(mysign <- sgn(x))
-mathout(term)
-
-hook(mysign, sg(x))
-term <- call("<-", quote(mysign), sgn)
-mathout(term)
+mathout(sgn)
+mathout(call("<-", quote(sgn(x)), sgn))
 ```
 
-The hook (see below) is used to display \(\mathrm{sg}(x) = \{\ldots\) instead
-of just \(s = \{\).
+As shown in the example, we can still display functions in the
+form `head(x) = body` if we embed the object to be shown into a
+call `"<-"(head, body)`.
+
+The function body is generally a nested R\ call of the form `'{'(L)`, with `L`
+being a list of commands (the semicolon, not necessary in R, is translated to a
+newline). As illustrated in the example, \pkg{mathml} provides limited support
+for control structures such as `if`.
 
@@
 integration variable. For that purpose, the underlying Prolog bridge provides a
-predicate `r_eval/2` that calls R\ from Prolog. In the example above, this
+predicate `r_eval/3` that calls R\ from Prolog. In the example above, this
 predicate evaluates `formalArgs(args(sin))`, which returns the names of the
@@
 for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
-function. R's `integrate()` can handle both. However, as already mentioned in
-the previous subsection, \pkg{mathml} would be unable to find out the function
-name, it needs the symbol.
+function. While the R\ function `integrate()` can handle both symbols and
+functions, \pkg{mathml} needs the symbol because it is unable to determine the
+function name of custom fuctions.
 
@@
 their names, not only by their position in the list of arguments. At the other
-end, most Prolog-powered handlers for the R\ calls are rather rigid, for
-example, `integrate/3` accepts exactly three arguments in a particular order and
-without names, that is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print
-the desired result.
+end, Prolog does not have such a feature. Therefore, the Prolog handlers for
+R\ calls are rather rigid, for example, `integrate/3` accepts exactly three
+arguments in a particular order and without names, that
+is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.
 
 To "canonicalize" function calls with named arguments and arguments in unusual
-order, \pkg{mathml} provides an auxilliary R\ function `canonical(f, drop)` that
+order, \pkg{mathml} provides an auxiliary R\ function `canonical(f, drop)` that
 reorders the argument list of calls to known R\ functions and, 
@@
 Note that the seemingly more convient `term <- quote(A + B)` yields \(A + B\) in
-the output---instead of the desired matrix representation.
+the output---instead of the desired matrix representation. This behavior is
+expected because quotation of R calls also quote the components of the
+call (here, _A_ and _B_).
+
+## Short mathematical names for R symbols
+
+In typical R\ functions, variable names are typically longer than just single
+letters, which may yield unsatisfactory results in the mathematical output.
+
+```{r, results='asis'}
+term <- quote(pbinom(successes, Ntotal, prob))
+mathout(term)
+
+hook(successes, k)
+hook(quote(Ntotal), quote(N), quote=FALSE)
+hook(prob, pi)
+mathout(term)
+```
+
+To improve the situation, \pkg{mathml} provides a simple hook that can be used
+to replace elements (e.g., verbose variable names) of the code by concise
+mathematical symbols, as illustrated in the example. To simplify notation,
+the `quote` flag of `hook()` defaults to TRUE, and `hook()` uses non-standard
+evaluation to unpack its arguments. If quote is FALSE, as shown above, the user
+has to provide the quoted expressions. Care should be taken to
+avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
+the \(s\) from \(s_{\mathrm{A}}\) as
+in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).
+
+The hooks can also be used for more complex elements such as R\ calls, with
+dotted symbols representing Prolog variables.
+
+```{r, results='asis'}
+hook(pbinom(.K, .N, .P), subsupscript(sum(dbinom(i, .N, .P)), i=0L, .K))
+mathout(term)
+```
 
@@
 evaluated, `abbr` being rendered, plus a comment of the
-form "with `expr` denoting `info`". The hooks are described in the next section.
+form "with `expr` denoting `info`".
 
@@
 
-The term is evaluated below. `print()` is needed because the return value of the
-assignment in the term `t <- dfrac(...)` is invisible.
+The term is evaluated below. `print()` is needed because the return value of an
+assignment of the form `t <- dfrac(...)` is not visible in R.
 
@@
 t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
-mathout(t, flags=list(error='highlight'))
-mathout(t, flags=list(error='fix'))
+mathout(t, flags=list(error="highlight"))
+mathout(t, flags=list(error="fix"))
 ```
@@
 This case study describes a model by Schwarz [-@schwarz1994] from mathematical
-Psychology using the features of package \pkg{mathml}. Schwarz presents a new
+psychology using the features of package \pkg{mathml}. Schwarz presents a new
 explanation of redundancy gains that occur when observers respond to stimuli of
@@
 
-## The model
-
 Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
@@
 drifts \(\mu_i\), variances \(\sigma^2_i\), and initial
-conditions \(X_i(t = 0) = 0, i = \mathrm{A, B}\). The channels may be correlated
-with \(\rho_{\mathrm{AB}}\), but we assume \(\rho_{\mathrm{AB}} = 0\) for
-simplicity.
+conditions \(X_i(t = 0) = 0, i = \mathrm{A, B}\). The buildup of 
+channel-specific activation may be correlated with \(\rho_{\mathrm{AB}}\), but
+we assume \(\rho_{\mathrm{AB}} = 0\) for simplicity.
 
 A response is elicited when the process reaches an absorbing barrier \(c > 0\)
-for the first time. In single-target trials, the first passages at \(c\) are
+for the first time. In single-target trials, the first passages of \(c\) are
 expected at
@@
 
-In redundant stimuli, the activation of the channel-specific diffusion processes
-adds up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\), hence
-the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process with
-drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
+(One would typically use chunk option `echo=FALSE` to suppress the R code.)
+In redundant stimuli, the activations from the channel-specific diffusion
+processes add up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\),
+hence the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process
+with drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
 variance \(\sigma^2_{\mathrm A} + \sigma^2_{\mathrm B}\). For the expected
@@
     ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
-      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
+      - (mu_A*tau + c) * exp(dfrac(2L*c*mu_A, sigma_A^2L))
         * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
@@
 
-For negative onset asynchrony (i.e., B before A), the parameters need to be
+For negative onset asynchrony (i.e., B before A), the parameters are simply
 switched.
@@
 {
-  if(tau == Inf) return(ED_single(c, mu_A)) ;
-  if(tau == -Inf) return(ED_single(c, mu_B)) ;
-  if(tau == 0L) return(ED_single(c, mu_A + mu_B)) ;
-  if(tau > 0L) return(ED_async(tau, c, mu_A, sigma_A, mu_B)) ;
+  if(tau == Inf) return(ED_single(c, mu_A))
+  if(tau == -Inf) return(ED_single(c, mu_B))
+  if(tau == 0L) return(ED_single(c, mu_A + mu_B))
+  if(tau > 0L) return(ED_async(tau, c, mu_A, sigma_A, mu_B))
   if(tau < 0L) return(ED_async(abs(tau), c, mu_B, sigma_B, mu_A))
@@
 
-## Methods
-
 Schwarz [-@schwarz1994] applied the model to data from a redundant signals task
@@
 ```{r miller-data, echo=FALSE}
-tau  <- list(-Inf, -167L, -133L, -100L, -67L, -33L, 0L, 33L, 67L, 100L, 133L, 167L, Inf)
+tau  <- list(-Inf,-167L,-133L,-100L,-67L,-33L,0L,33L,67L,100L,133L,167L,Inf)
 m    <- c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
@@
 
-t <- cbind('$\\tau$'=ttau, '$m$'=tm, '$s$'=ts, '$n$'=tn)
+t <- cbind('\\(\\tau\\)'=ttau, '\\(m\\)'=tm, '\\(s\\)'=ts, '\\(n\\)'=tn)
 knitr::kable(t, caption="Miller data", row.names=FALSE, escape=FALSE)
@@
 Assuming that the model is correct, the observable mean reaction times follow an
-approximate Normal distribution around the model prediction \(E(T)\). We can,
-therefore, use a standard goodness-of-fit measure by \(z\)-standardisation.
+approximate Normal distribution around the model prediction \(E(T_\tau)\) for
+each condition. We can, therefore, use a standard goodness-of-fit measure
+by \(z\)-standardisation.
 
@@
 z <- function(m, s, n, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
-  dfrac(m - denote(E, ET(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M),
+  dfrac(m - denote(mu[tau], ET(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M),
               "the expected mean response time"),
@@
 but the parameters are adjusted to the data, it follows
-a \(\chi^2(8~\mathrm{df})\)-distribution.
+a \(\chi^2(8\ \mathrm{df})\)-distribution.
 
@@
 zv <- Vectorize(z, vectorize.args = c('m', 's', 'n', 'tau'))
+hook(zv(.M, .S, .N, .Tau, .C, .MA, .SA, .MB, .SB, .MM), z[.Tau])
 
@@
 
-hook(zv(.M, .S, .N, .Tau, .C, .MA, .SA, .MB, .SB, .MM), z[.Tau])
 mathout(call("=", quote(X["8 df"]^2L), gof))
@@
 
-The degrees of freedom result from the 13 predicted mean response times, 
-minus 5 free parameters
-\(\theta = \langle\mu_{\mathrm A}, \sigma_{\mathrm A}, \mu_{\mathrm B}, \sigma_{\mathrm B}, \mu_{\mathrm M}\rangle\)
-(\(c\) is a redundant scaling parameter).
+with the degrees of freedom given by the difference between the number of
+observations (`r length(tau)`) and the number of free model
+parameters \(\theta = \langle\mu_{\mathrm A}, \sigma_{\mathrm A},
+\mu_{\mathrm B}, \sigma_{\mathrm B}, \mu_{\mathrm M}\rangle\); the barrier \(c\)
+is only a scaling parameter.
 
@@
 {
-  invisible(theta <- c(mu_A=0.53, sigma_A=4.3, mu_B=1.34, sigma_B=11.7, mu_M=160))
+  invisible(theta <- c(mu_A=0.5, sigma_A=5, mu_B=0.5, sigma_B=5, mu_M=150))
   optim(theta, gof, tau=tau, m=m, s=s, n=n, hessian=TRUE)
@@
 se <- sqrt(diag(solve(fit$hessian)))
-cilo <- qnorm(0.025, theta, se)
-ciup <- qnorm(0.975, theta, se)
+ci <- mapply(FUN=qnorm, theta, se, MoreArgs=list(p=c(0.025, 0.975)),
+    SIMPLIFY=FALSE)
 
-ntheta <- lapply(lapply(FUN=as.symbol, names(theta)), FUN=mathout, flags=list(cat=FALSE))
+ntheta <- lapply(lapply(FUN=as.symbol, names(theta)), FUN=mathout,
+    flags=list(cat=FALSE))
 ttheta <- lapply(theta, FUN=mathout, flags=list(cat=FALSE))
-tcilo <- lapply(cilo, FUN=mathout, flags=list(cat=FALSE))
-tciup <- lapply(ciup, FUN=mathout, flags=list(cat=FALSE))
+tci <- lapply(ci, FUN=mathout, flags=list(cat=FALSE))
 
-t <- cbind(Parameter=ntheta, Estimate=ttheta, Lo=tcilo, Up=tciup)
+t <- cbind(Parameter=ntheta, Estimate=ttheta, CI=tci)
 knitr::kable(t, caption="Model fit", row.names=FALSE, escape=FALSE)
@@
 
-The statistic goodness-of-fit value is \(X^2(8) = 30.54\), \(p = .0002\), 
-indicating some lack of fit. Given the large trial numbers in the original
-study, some lack of fit is expected. For more detail, especially on fitting the
-observed standard deviations, the reader is referred to the original
-paper [@schwarz1994].
+The goodness-of-fit statistic indicates some lack of
+fit, \(X^2(8\ \mathrm{df}) = `r sprintf("%.2f", fit$value)`, 
+  p = `r sprintf("%.4f", 1-pchisq(fit$value, df=8))`\). Given the large trial
+numbers in the original study, this is not an unexpected result. For more
+detail, especially on fitting the observed standard deviations, the reader is
+referred to the original paper [@schwarz1994].
 
@@
 
-This package expands the possibilities of R\ by allowing calculations and
-graphical rendering of the same terms. Building on current features of R\ and
-existing packages for displaying formulas in
-R\ [@murrell2000;@allaire2018]), the \pkg{mathml} package bridges the gap
-between computational needs, presentation of results, and their reproducibility.
-Two possible modes for typesetting are available to export R\ Markdown documents
-in different formats or ShinyApp webpages: MathML and MathJax. A further, not-yet-mentioned, use of \pkg{mathml} is related to presentations made using
-RMarkdown. By adopting \pkg{mathml} researchers or teachers can use the
-RMarkdown to both conduct analysis and show results, smoothing the process and
-uniforming the graphical output.
+This package allows R\ to render its terms in pretty mathematical equations. It
+extends the current features of R\ and existing packages for displaying
+mathematical formulas in R\ [@murrell2000;@allaire2018]), bit most
+importantly, \pkg{mathml} bridges the gap between computational needs, presentation
+of results, and their reproducibility. The package supports both MathML and
+Latex/MathJax for use in R\ Markdown documents, presentations and ShinyApp
+webpages.
 
-The case study presented above shows that \pkg{mathml} can help to improve
-data analyses and statistical reports from an aesthetical perspective, as well
-as regarding reproducibility of research. Furthermore, the package may also
-allow for a better detection of possible mistakes in R programs.
+Researchers or teachers can already use RMarkdown to conduct analyses
+and show results, and \pkg{mathml} smoothes this process and allows for integrated
+calculations and output. As shown in the case study of the previous
+section, \pkg{mathml} can help to improve data analyses and statistical reports from
+an aesthetical perspective, as well as regarding reproducibility of research.
 
-Similar to most programming languages [@green1977], R\ code is notoriously hard
-to read, and the poor legibility of the language is one of the main sources of
-mistakes. For illustration, we consider again Equation 10 in
-Schwarz [-@schwarz1994].
+Furthermore, the package may also allow for a better detection of possible
+mistakes in R\ programs. Similar to most programming languages [@green1977],
+R\ code is notoriously hard to read, and the poor legibility of the language is
+one of the main sources of mistakes. For illustration, we consider again
+Equation\ 10 in Schwarz [-@schwarz1994].
 
@@
 f1 <- function(tau)
-{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
-    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
-      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
-        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
+{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B) * 
+    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
+      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*tau, sigma_A^2L))
+        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau)))))
 }
@@
 The first version has a wrong parenthesis, which is barely visible in the code,
-whereas the wrong curly brace in the mathematical representation is immediately
-obvious. The correct version is shown below for comparison.
+whereas in the mathematical representation, the wrong curly brace is immediately
+obvious (the correct version is shown below for comparison).
 
@@
 f2 <- function(tau)
-{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
-    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
-      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
-        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
+{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) * 
+    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
+      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*tau, sigma_A^2L))
+        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
 }
@@
 currently not possible to use the functions of \pkg{mathml} for writing
-intratextual formulas; here, the user has to adopt the usual LateX notation.
-However, we do not consider this to be a serious limitation, since the most
-productive use of \pkg{mathml} still comes from more complicated calculations.
+intratextual formulas, here, the user has to adopt the usual LateX notation.
+Moreover, for long equations, we did not yet find a convenient way to insert
+linebreaks. This is mostly due to lacking support by MathML and LaTeX renderers.
+For example, in its current stage, the LaTeX package \pkg{breqn} [@breqn] is mostly
+a proof of concept.
 
-Also: In long equations, we did not yet find a convenient way to insert
-linebreaks. This is mostly because it is not properly supported by MathML and
-LaTeX renderers. In its current stage, the LaTeX package \pkg{breqn} [@breqn]
-is mostly a proof of concept.
-
-\pkg{mathml} is available for R\ version (…) and later, and can be easily
-installed using the usual `install.packages("mathml")`. The source code of the
-package is found at https://github.com/mgondan/mathml.
-
-# Appendix
-
-Please check if the examples below
-
-* are already presented somewhere
-* can be added to one of the tables
-* or if they should go into a separate appendix
-
-_I removed the examples since they were already present._
-
-```{r, results='asis', echo=TRUE}
-mathout(quote(T^diamond))
-mathout(quote(sinpi(alpha)))
-```
+The package \pkg{mathml} is available for R\ version (…) and later, and can be
+easily installed using the usual `install.packages("mathml")`. The source code
+of the package is found at https://github.com/mgondan/mathml.

### 2023-01-22 rj/mathml.Rmd
customizing

@@
 asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
-where \(0\) refers to the synchronous condition, and \(\pm\infty\) to the
+where \(\tau = 0\) refers to the synchronous condition, and \(\pm\infty\) to the
 single-target presentations. Each condition was replicated 400 times. The
@@
 of the package is found at https://github.com/mgondan/mathml.
+
+# Appendix: Customizing the package
+
+Several ways exist for translating new R\ terms to their mathematical
+representation. We have already seen above how to use "hooks" to translate long
+variable names from R to compact mathematical signs, as well as functions such
+as cumulative probabilities \(P(X \le k)\) to different representations
+like \(\sum_{i=0}^k P(X = i)\). Obviously, the hooks require that there already
+exists a rule to translate the target representation into MathML and MathJax.
+
+In this appendix we describe a few more ways to extend the set of translations
+according to a user's needs. As stated in the background section, the Prolog end
+provides two classes of rules for translation, macros `math/2,3,4` mirroring the
+R\ hooks mentioned above, and the low-level predicates `ml/3` and `jax/3` that
+create proper MathML and Latex terms.
+
+## \(n\)-th root
+
+Base R does not provide a function like `cuberoot(x)` or `nthroot(x, n)`, and
+the present package does not support the respective representation. To obtain a
+cube root, a programmer would typically type `x^(1/3)` or better `x^{1/3}` (see
+the practice section why the curly brace is preferred in an exponent), resulting
+in \(x^{1/3}\) which may still not match everyone's taste. Here describe the
+steps needed to represent the \(n\)-th root as \(\sqrt[n]x\).
+
+We assume that `nthroot(x, n)` is available in the current
+namespace [manually defined, or from R package \pkg{pracma}, @pracma], so that
+the names of the arguments and their order are accessible to `canonical()` if
+needed. As we can see below, \pkg{mathml} uses a default
+representation `name(arguments)` for such unknown functions.
+
+```{r, results='asis'}
+nthroot <- function(x, n)
+  x^{1L/n}
+
+term <- canonical(quote(nthroot(n=3L, 2L)))
+mathout(term)
+```
+
+A proper MathML term is obtained by `mlx/3` (the x in mlx indicates that it is
+an extension and is prioritized over the default ml/3 rules). `mlx/3`
+recursively invokes `ml/3` for translating the function arguments _X_ and _N_,
+and then contructs the correct MathML entity `<mroot>...</mroot>`.
+
+```prolog
+mlx(nthroot(X, N), M, Flags) :-
+    ml(X, X1, Flags),
+    ml(N, N1, Flags),
+    M = mroot([X1, N1]).
+```
+
+The explicit unification `M = ...` in the last line serves to avoid clutter in
+the head of `mlx/3`. The Prolog file `nthroot.pl` also includes the respective
+rule for Latex and can be consulted from the package folder via the underlying
+package \pkg{rolog}.
+
+```{r, results='asis'}
+rolog::consult(system.file(file.path("pl", "nthroot.pl"), package="mathml"))
+
+term <- quote(nthroot(a * (b + c), 3L)^2L)
+mathout(term)
+term <- quote(a^(1L/3L) + a^{1L/3L} + a^(1.0/3L))
+mathout(term)
+```
+
+The file `nthroot.pl` includes three more statements `precx/3` and `parenx/3`,
+as well as a `math_hook/2` macro. The first sets the operator precedence of the
+cubic root above the power, thereby putting a parentheses around nthroot
+in \((\sqrt[3]{\ldots})^2\). The second tells the system to increase the counter
+of the parentheses below the root, such that the outer parenthesis becomes a
+square bracket.
+
+The last rule maps powers like `a^(1L/3L)` to `nthroot/3`, as shown in the
+first summand. Of course, \pkg{mathml} is not a proper computer algebra system.
+As is illustrated by the other terms in the sum, such macros are limited to
+purely syntactical matching, and terms like `a^{1L/3L}` with the curly brace
+or `a^(1.0/3L)` with a floating point number in the numerator are not detected.

### 2023-01-22 inst/pl/mathml.pl

@@ -1,4 +1,5 @@
-:- discontiguous math/2, math/3, math/4, current/3, paren/3, prec/3, type/3, denoting/3, ml/3, jax/3, mathml/0.
+:- discontiguous mathml/0, math/2, math/3, math/4, current/3, paren/3, prec/3, type/3, denoting/3, ml/3, jax/3.
+
 :- use_module(library(http/html_write)).
 
 % Here you can define your own macros
@@ -6,6 +7,7 @@
 % Example: assert(math_hook(t0, subscript(t, 0)))
 %
 :- dynamic math_hook/2.
+:- multifile math_hook/2, mlx/3, jaxx/3, precx/3, parenx/3, typex/3.
 
 %
 % R interface: Translate R expression to MathML string
@@ -79,23 +81,43 @@ ml(R, M, Flags),
     macro(R, R1, Flags, Flags1)
  => ml(R1, M, Flags1).
 
+ml(R, M, Flags),
+    mlx(R, R1, Flags)
+ => M = R1.
+
 % Same for MathJax/LaTeX
 jax(R, M, Flags),
     macro(R, R1, Flags, Flags1)
  => jax(R1, M, Flags1).
+
+jax(R, M, Flags),
+    jaxx(R, R1, Flags)
+ => M = R1.
 
 % Same for precedence checks, parentheses and types
 prec(R, Prec, Flags),
     macro(R, R1, Flags, Flags1)
  => prec(R1, Prec, Flags1).
 
+prec(R, P, Flags),
+    precx(R, P1, Flags)
+ => P = P1.
+
 paren(R, Paren, Flags),
     macro(R, R1, Flags, Flags1)
  => paren(R1, Paren, Flags1).
 
+paren(R, P, Flags),
+    parenx(R, P1, Flags)
+ => P = P1.
+
 type(R, Type, Flags),
     macro(R, R1, Flags, Flags1)
  => type(R1, Type, Flags1).
+
+type(R, T, Flags),
+    typex(R, T1, Flags)
+ => T = T1.
 
 %
 % Suppress the names of function arguments from R
@@ -185,29 +207,44 @@ mathml :-
 % function can handle them. This is needed, e.g., if colors are to be
 % skipped.
 %
-math(subsupscript(R, Idx, Pwr), M, Flags, Flags1)
[…]

### 2023-01-22 inst/pl/nthroot.pl

@@ -1,0 +1,23 @@
+% Render in MathML
+mlx(nthroot(X, N), M, Flags) :-
+    ml(X, X1, Flags),
+    ml(N, N1, Flags),
+    M = mroot([X1, N1]).
+
+% Render in MathJax
+jaxx(nthroot(X, N), M, Flags) :-
+    jax(X, X1, Flags),
+    jax(N, N1, Flags),
+    format(string(M), "\\sqrt[~w]{~w}", [N1, X1]).
+
+% Precedence above power to ensure parenthesis around (nthroot)^2
+precx(nthroot(_X, _N), P, _Flags) :-
+    current_op(P0, xfy, ^),
+    P is P0 + 1.
+
+% Continue counting parentheses below root
+parenx(nthroot(X, _N), P, Flags) :-
+    paren(X, P, Flags).
+
+% Show x^(1/n) as nthroot(x, n)
+math_hook(X ^ '('(1/N), nthroot(X, N)).

### 2023-01-23 R/mathml.R
new organization of subscripts

@@ -175,83 +175,55 @@ name <- function(x, name)
 #'
 cal <- identity
 
-#' Subscript. On the R side, this function is a wrapper of identity, but allows
-#' for decorations.
+#' sum over a range. On the R side, this function just returns the first
+#' argument, but allows for decorations.
 #'
-#' @param sub
-#' an R symbol or call, e.g., i
+#' @param x
+#' the object to be summed
 #'
-#' @param fun
-#' an R call or symbol, e.g. sum(x). This is the return value of the function.
+#' @param from
+#' decoration for sum_from^to x[i]
+#'
+#' @param to
+#' decoration for sum_from^to x[i]
 #'
 #' @return
-#' The function over is a wrapper for the identity function, returning _fun_
+#' The function returns sum(x)
 #'
 #' @md
 #'
-#' @seealso [identity()]
+#' @seealso [sum(), prod_over()]
 #'
 #' @examples
-#' mathjax(quote(subscript(sub=i, fun=x)))
+#' mathjax(quote(sum_over(x[i], i=1L, N)))
 #'
-subscript <- function(fun=quote(x), sub=quote(i))
-{
-  return(fun)
-}
+sum_over <- function(x, from, to)
+  sum(x)
 
-#' Superscript. This is a wrapper for the identity function, but decorates the
-#' result with a superscript.
+#' product over a range. On the R side, this function just returns the product
+#' of the first argument, but allows for decorations.
 #'
-#' @param sup
-#' an R symbol, e.g., "*"
+#' @param x
+#' the object to be multiplied
 #'
-#' @param fun
-#' an R call or symbol, e.g. x. This is the return value of the function.
+#' @param from
+#' decoration for prod_from^to x[i]
+#'
+#' @param to
+#' decoration for prod_from^to x[i]
 #'
 #' @return
-#' The function over is a wrapper for the identity function, returning _fun_
+#' The function returns prod(x)
 #'
 #' @md
 #'
-#' @seealso [identity()]
+#' @seealso [prod(), sum_over()]
 #'
 #' @examples
-#' mathjax(quote(superscript(fun=A, sup="*")))
+#' mathjax(quote(prod_over(x[i], i=1L, N)))
 #'
-superscript <- function(fun=quote(A), sup="*")
-{
-  return(fun)
-}
-
[…]

### 2023-01-23 inst/pl/mathml.pl

@@ -126,13 +126,6 @@ math(name(_) = R, M)
  => M = R.
 
 %
-% Check if the current element is to be replaced
-%
-math(R, M, Flags, Flags1),
-    member(replace(R, _), Flags)
- => select(replace(R, M), Flags, Flags1).
-
-%
 % Examples
 %
 mathml(R) :-
@@ -159,175 +152,168 @@ denoting_(Flags, R, Den)
  => denoting(R, Den, Flags).
 
 %
-% Content starts here
+% Summation and product sign
 %
+math(sum_over(Arg, Range), M)
+ => M = fn(subscript(sum, Range), [Arg]).
 
-% Summation sign from to (subscript and superscript)
-math(Sum, M, Flags, New),
-    compound(Sum),
-    compound_name_arguments(Sum, sum, [Arg]),
-    select(subscript(From), Flags, Flags1),
-    select(superscript(To), Flags1, Flags2)
- => New = Flags2,
-    M = fn(subsupscript(sum, From, To), [Arg]).
-
-math(Sum, M, Flags, New),
-    compound(Sum),
-    compound_name_arguments(Sum, sum, [Arg]),
-    select(subscript(Index), Flags, Flags1)
- => New = Flags1,
-    M = fn(subscript(sum, Index), [Arg]).
+math(sum_over(Arg, From, To), M)
+ => M = fn(subsupscript(sum, From, To), [Arg]).
 
 mathml :-
-    mathml(subscript(sum('['(x, i)), i)).
-
-% Same for product sign
-math(Prod, M, Flags, New),
-    compound(Prod),
-    compound_name_arguments(Prod, prod, [Arg]),
-    select(subscript(From), Flags, Flags1),
-    select(superscript(To), Flags1, Flags2)
- => New = Flags2,
-    M = fn(subsupscript(prod, From, To), [Arg]).
-
-math(Prod, M, Flags, New),
-    compound(Prod),
-    compound_name_arguments(Prod, prod, [Arg]),
-    select(subscript(Index), Flags, Flags1)
- => New = Flags1,
-    M = fn(subscript(prod, Index), [Arg]).
+    mathml(sum('['(x, i), i)).
 
 mathml :-
[…]

### 2023-01-28 R/mathml.R
font styles

@@ -733,13 +733,54 @@ add <- function(expr)
 #' in particular adds a curly bracket under the wrong term and it provides the
 #' correct solutions.
 #'
-#' @param inst instead
+#' @param inst
+#' the wrong term
 #'
-#' @param of of
+#' @param of
+#' the correct term
 #'
-#' @return inst , e.g. a + c instead of a + b
+#' @return
+#' inst
+#'
+#' @examples
+#' 1 + instead(2, 3)
+#'
+#' @examples
+#' mathml(term=quote(1 + instead(2, 3)))
 #'
 instead <- function(inst, of)
 {
   return(inst)
 }
+
+#' Identity functions for different font styles
+#'
+#' @name fontstyles
+#'
+#' @param expr
+#' the expression to render
+#'
+#' @return
+#' expr
+#'
+#' @examples
+#' plain(1) + bold(2) + italic(3) + bolditalic(4)
+#'
+#' @examples
+#' mathml(term=quote(plain(1) + bold(2) + italic(3) + bolditalic(4)))
+
+#' @rdname fontstyles
+#' @export
+plain <- identity
+
+#' @rdname fontstyles
+#' @export
+italic <- identity
+
+#' @rdname fontstyles
+#' @export
+bold <- identity
+
+#' @rdname fontstyles
+#' @export
+bolditalic <- identity

### 2023-01-28 inst/pl/mathml.pl

@@ -556,6 +556,22 @@ ml(ident(R), M, Flags),
     member(mathvariant(calligraphy), Flags)
  => M = mi(mathvariant(script), R).
 
+ml(ident(R), M, Flags),
+    member(mathvariant(plain), Flags)
+ => M = mi(mathvariant(normal), R).
+
+ml(ident(R), M, Flags),
+    member(mathvariant(italic), Flags)
+ => M = mi(mathvariant(italic), R).
+
+ml(ident(R), M, Flags),
+    member(mathvariant(bold), Flags)
+ => M = mi(mathvariant(bold), R).
+
+ml(ident(R), M, Flags),
+    member(mathvariant(bolditalic), Flags)
+ => M = mi(mathvariant('bold-italic'), R).
+
 ml(ident(R), M, _Flags)
  => M = mi(R).
 
@@ -568,6 +584,33 @@ jax(ident(R), M, _Flags)
 
 type(ident(_), T, _Flags)
  => T = atomic.
+
+%
+% Font styles
+%
+math(plain(R), M, Flags0, Flags1)
+ => M = R,
+    Flags1 = [mathvariant(plain) | Flags0].
+
+math(bold(R), M, Flags0, Flags1)
+ => M = R,
+    Flags1 = [mathvariant(bold) | Flags0].
+
+math(italic(R), M, Flags0, Flags1)
+ => M = R,
+    Flags1 = [mathvariant(italic) | Flags0].
+
+math(bolditalic(R), M, Flags0, Flags1)
+ => M = R,
+    Flags1 = [mathvariant(bolditalic) | Flags0].
+
+jax(ident(R), M, Flags),
+    member(mathvariant(calligraphy), Flags)
+ => format(string(M), "\\mathcal{~w}", [R]).
+
+jax(ident(R), M, _Flags)
+ => format(string(M), "~w", [R]).
+
 
 %
 % Linear model

### 2023-01-28 rj/mathml.Rmd
new organization of subscripts
decorations

@@
 title: "mathml: Translate R expressions to MathML and LaTeX"
-date: "2023-01-21"
+date: "2023-01-28"
 abstract: >
@@
 ```{r setup, include=FALSE}
-# To do
-#
-# paste(x, y, z)--> z, y, z (Covered by quote ())
-
-# Spacing: operator for separating elements in expression. E.g.,
-## x ~ y --> x  y ;
-## x ~ ~ y --> x    y
-
-# Ellipsis: ldots or cdots. E.g., 
-## x[1] + ... + x[n]
-
-# Radicals sqrt(x, y) --> y√x
-
-# Chains: a < b < c
-
-# Sets: some symbols in set relations are missing:
-## supset, supseteq, notsubset, subset, subseteq, notin
-
-# Phantom: command that produces a space in place of an argument (partially covered by omit() (?)). E.g.:
-## x + phantom(0) + y --> x +    + y
-## x + over(1, vphantom(0)) --> x + 1/
-
-# Typeface: plain, bold, italic, bolditalic
-
-# Stylechanges: textstyle(x), scriptstyle(x) scriptscript(x)
-
 # linear models: DepVar ~ X0 + X1
-
-# hat, overline, cancel, box, invisible, phantom
 
@@
 ```{r, results='asis'}
-term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + hat(b)[i, j]^2L)
+term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + roof(b)[i, j]^2L)
 mathout(term)
@@
 
-Note that an R\ function `hat()` does not exist in base R, it is provided by the
-package for convenience and points to the identity function. Quite a few such
-convenience functions have been defined
-in \pkg{mathml} (see, e.g., next section, and the manual for details).
+To avoid name clashes with package \pkg{stats}, `roof()` is used to put a hat
+on a symbol (see next section for further decorations). Note that an 
+R\ function `roof()` does not exist in base R, it is provided by the package
+for convenience and points to the identity function.
+
+## Decorations
+
+The package offers some support for different fonts as well as accents and
+boxes etc. Internally, these decorations are implemented as identity functions,
+so they can be introduced into R expressions without side-effects.
+
+```{r, results='asis'}
+term <- quote(bold(b[x, 5L]) + bold(b[italic(x)]) + italic(ab) + italic(42L))
+mathout(term)
+
+term <- quote(tilde(a) + mean(X) + box(c) + cancel(d) + phantom(e) + prime(f))
+mathout(term)
+```
+
+Note that the font styles only affect the display of identifiers, whereas
+numbers, character strings etc. are left untouched.
 
@@
 superscript. Moreover, \pkg{mathml} defines the
-functions `subscript()`, `superscript()` and `subsuperscript()` that simply
-return their first argument. The other arguments serve as decorations, for
-example, for summation and product signs.
+functions `sum_over(x, from, to)`, and `prod_over(x, from, to)` that simply
+return their first argument. The other two arguments serve as
+decorations (_to_ is optional), for example, for summation and product signs.
 
@@
 
-term <- quote(log(subsupscript(prod(L[i]), i==1L, N)) <- 
-          subsupscript(sum(log(L[i])), i==1L, N))
+term <- quote(log(prod_over(L[i], i==1L, N)) <- sum_over(log(L[i]), i==1L, N))
 mathout(term)
@@
 ```{r, results='asis'}
-hook(pbinom(.K, .N, .P), subsupscript(sum(dbinom(i, .N, .P)), i=0L, .K))
+hook(pbinom(.K, .N, .P), sum_over(dbinom(i, .N, .P), i=0L, .K))
 mathout(term)
@@
 ```{r, results='asis'}
-term <- quote(pbinom(successes, Ntotal, prob))
-mathout(term)
-
 hook(successes, k)
@@
 hook(prob, pi)
+
+term <- quote(dbinom(successes, Ntotal, prob))
 mathout(term)
@@
 
-For asynchronous stimuli, Schwarz [-@schwarz1994; Eq. 10] derived the expected
+For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
 first-passage time as a function of the stimulus onset asyncrony \(\tau\),

### 2023-01-28 inst/pl/mathml.pl

@@ -552,6 +552,22 @@ math(R, M),
     atom(R)
  => M = ident(R).
 
+math(plain(R), M, Flags0, Flags1)
+ => M = R,
+    Flags1 = [mathvariant(plain) | Flags0].
+
+math(bold(R), M, Flags0, Flags1)
+ => M = R,
+    Flags1 = [mathvariant(bold) | Flags0].
+
+math(italic(R), M, Flags0, Flags1)
+ => M = R,
+    Flags1 = [mathvariant(italic) | Flags0].
+
+math(cal(A), M, Flags, New)
+ => New = [mathvariant(calligraphy) | Flags],
+    M = A.
+
 ml(ident(R), M, Flags),
     member(mathvariant(calligraphy), Flags)
  => M = mi(mathvariant(script), R).
@@ -568,10 +584,6 @@ ml(ident(R), M, Flags),
     member(mathvariant(bold), Flags)
  => M = mi(mathvariant(bold), R).
 
-ml(ident(R), M, Flags),
-    member(mathvariant(bolditalic), Flags)
- => M = mi(mathvariant('bold-italic'), R).
-
 ml(ident(R), M, _Flags)
  => M = mi(R).
 
@@ -579,38 +591,23 @@ jax(ident(R), M, Flags),
     member(mathvariant(calligraphy), Flags)
  => format(string(M), "\\mathcal{~w}", [R]).
 
+jax(ident(R), M, Flags),
+    member(mathvariant(plain), Flags)
+ => format(string(M), "\\mathrm{~w}", [R]).
+
+jax(ident(R), M, Flags),
+    member(mathvariant(italic), Flags)
+ => format(string(M), "\\mathit{~w}", [R]).
+
+jax(ident(R), M, Flags),
+    member(mathvariant(bold), Flags)
+ => format(string(M), "\\mathbf{~w}", [R]).
+
 jax(ident(R), M, _Flags)
  => format(string(M), "~w", [R]).
 
 type(ident(_), T, _Flags)
  => T = atomic.
-
-%
-% Font styles
-%
-math(plain(R), M, Flags0, Flags1)
- => M = R,
-    Flags1 = [mathvariant(plain) | Flags0].
-
-math(bold(R), M, Flags0, Flags1)
- => M = R,
-    Flags1 = [mathvariant(bold) | Flags0].
-
-math(italic(R), M, Flags0, Flags1)
- => M = R,
[…]

### 2023-01-28 rj/mathml.Rmd
linear model

@@
 
+## Linear models
+
+To render the model equation of a linear model such
+as `lm(EOT ~ T0 + Therapy, data=d)` in mathematical form, it is sufficient to
+map the `Formula` in `lm(Formula, Data)` to its respective equation. This can in
+two ways, using either the hooks described above, or a new `math/2` macro at the
+Prolog end.
+
+````r
+hook(lm(.Formula, .Data), .Formula)
+````
+
+The hook is simple, but is a bit limited because only R's tilde-form
+of linear models is shown, and it only works for a call with exactly two
+arguments.
+
+Below is an example how to build a linear equation of the
+form \(Y = b_0 + b_1X_1 + ...\) using the Prolog macros from \pkg{mathml}.
+
+````prolog
+math_hook(LM, M) :-
+    compound(LM),
+    LM =.. [lm, ~(Y, Sum) | _Tail],
+    summands(Sum, Predictors),
+    findall(subscript(b, X) * X, member(X, Predictors), Terms),
+    summands(Model, Terms),
+    M = (Y == subscript(b, 0) + Model + epsilon).
+````
+
+The predicate `summands/2` unpacks an expression `A + B + C` to a
+list `[C, B, A]` and vice-versa (see the file `lm.pl` for details).
+
+````{r, results='asis'}
+rolog::consult(system.file(file.path("pl", "lm.pl"), package="mathml"))
+
+term <- quote(lm(EOT ~ T0 + Therapy, data=d, na.action=na.fail))
+mathout(term)
+````
+
 ## \(n\)-th root

### 2023-01-28 R/mathml.R

@@ -792,10 +792,10 @@ bold <- identity
 #' x
 #'
 #' @examples
-#' roof(1) + mean(2) + box(3) + cancel(4) + phantom(5) + prime(6) + tilde(7)
+#' roof(1) + mean(2) + boxed(3) + cancel(4) + phantom(5) + prime(6) + tilde(7)
 #'
 #' @examples
-#' mathml(quote(roof(b) + mean(X) + box(3) + cancel(4) + phantom(5) + prime(6)))
+#' mathml(quote(roof(b) + mean(X) + boxed(3) + cancel(4) + phantom(5)))
 
 #' @rdname decorations
 #' @export
@@ -803,7 +803,7 @@ roof <- identity
 
 #' @rdname decorations
 #' @export
-box <- identity
+boxed <- identity
 
 #' @rdname decorations
 #' @export

### 2023-01-28 inst/pl/lm.pl

@@ -1,0 +1,21 @@
+% A + B + C => [C, B, A]
+summands(A + B, X) :-
+    var(X),
+    !,
+    summands(A, Rest),
+    X = [B | Rest].
+
+% A + B + C <= [C, B, A]
+summands(S + A, [A, B | Rest]) :-
+    summands(S, [B | Rest]).
+
+% Base case
+summands(A, [A]).
+
+math_hook(LM, M) :-
+    compound(LM),
+    LM =.. [lm, ~(Y, Sum) | _Tail],
+    summands(Sum, Predictors),
+    findall(subscript(b, X) * X, member(X, Predictors), Terms),
+    summands(Model, Terms),
+    M = (Y == subscript(b, 0) + Model + epsilon).

### 2023-01-28 inst/pl/mathml.pl

@@ -1218,6 +1218,9 @@ prec(cancel(A), Prec, Flags)
 type(cancel(A), Type, Flags)
  => type(A, Type, Flags).
 
+math(boxed(A), M)
+ => M = box(A).
+
 ml(box(A), M, Flags)
  => ml(A, X, Flags),
     M = menclose(notation(roundedbox), X).

### 2023-01-29 inst/pl/mathml.pl
checks passed

@@ -937,7 +937,14 @@ math(invisible(_), M, _Flags)
  => M = ''.
 
 % Vectors
-math(Hash, M),
+math(Hash, M, Flags),
+    option_(sep(Sep), Flags),
+    compound(Hash),
+    compound_name_arguments(Hash, Name, Elements),
+    member(Name, ['#', '$$', '%', '!'])
+ => M = paren(list(Sep, Elements)).
+
+math(Hash, M, _Flags),
     compound(Hash),
     compound_name_arguments(Hash, Name, Elements),
     member(Name, ['#', '$$', '%', '!'])
@@ -1697,9 +1704,11 @@ math(A - B, X)
 % Use dot or no dot instead of asterisk
 math(A * B, X, Flags),
     type(A, TypeA, Flags),
-    member(TypeA, [atomic, subscript(_, _), superscript(_, _), subsupscript(_, _, _)]),
+    member(TypeA, [atomic, subscript(_, _), superscript(_, _),
+                   subsupscript(_, _, _)]),
     type(B, TypeB, Flags),
-    member(TypeB, [atomic, subscript(_, _), superscript(_, _), subsupscript(_, _, _)])
+    member(TypeB, [atomic, subscript(_, _), superscript(_, _),
+                   subsupscript(_, _, _)])
  => X = nodot(A, B).
 
 % Todo: This should be checked
@@ -2205,8 +2214,7 @@ paren(frac(_, _), P, _Flags)
  => P = 0.
 
 prec(frac(_, _), P, _Flags)
- => current(P0, yfx, /),
-    P is P0 + 1. % was - 1
+ => current(P, yfx, /). % was P - 1
 
 type(frac(_, _), Type, _Flags)
   => Type = fraction.
@@ -2272,8 +2280,8 @@ option_(NameOption, Flags) :-
 
 option_(NameOption, Flags) :-
     compound_name_arguments(NameOption, Name, [Option]),
-    atom_string(Option, String),
-    member(Name-String, Flags).
+    member(Name-String, Flags),
+    atom_string(Option, String).
 
 math(omit_left(Expr), M, Flags),
     option_(error(ignore), Flags)

### 2023-01-29 rj/mathml.Rmd
checks passed
minor

@@
 ```{r setup, include=FALSE}
-# linear models: DepVar ~ X0 + X1
-
 library(mathml)
@@
 ttheta <- lapply(theta, FUN=mathout, flags=list(cat=FALSE))
-tci <- lapply(ci, FUN=mathout, flags=list(cat=FALSE))
+tci <- lapply(ci, FUN=mathout, flags=list(cat=FALSE, sep=","))
 

### 2023-01-30 rj/mathml.Rmd
acknowledgement
initial checks

@@
 ---
-title: "mathml: Translate R expressions to MathML and LaTeX"
+title: "`mathml`: Translate R Expressions to MathML and LaTeX"
 date: "2023-01-28"
@@
 R\ objects to mathematical expressions. Section\ 3 illustrates the main features
-of the \pkg{mathml} package, potential issues and workarounds using examples
+of the \CRANpkg{mathml} package, potential issues and workarounds using examples
 from the day-to-day perspective of a user. Section\ 4 present a case study for
@@
 The translation of R\ expressions to mathematical output is achieved through a
-Prolog interpreter from R\ package \pkg{rolog} [@rolog]. Prolog is a classical logic
+Prolog interpreter from R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic
 programming language with many applications in expert systems, computer
@@
 
-term <- quote(NaN + NA + TRUE + FALSE + Inf + (-Inf))
+term <- quote(round(3.1415, 3L) + NaN + NA + TRUE + FALSE + Inf + (-Inf))
 mathout(term)
@@
 
-To avoid name clashes with package \pkg{stats}, `roof()` is used to put a hat
+To avoid name clashes with package `stats`, `roof()` is used to put a hat
 on a symbol (see next section for further decorations). Note that an 
@@
 detect if parentheses are needed; for example, parentheses are automatically
-placed around `g + h` in the `nodot`-example. However, we note unecessary
+placed around `g + h` in the `nodot`-example. However, we note unnecessary
 parentheses around `l + m` above. Thes parentheses are a consequence
@@
 A few more examples are shown in Table \@ref(tab:base-stats), including
-functions from \pkg{stats}.
+functions from `base` and `stats`.
 
@@
 functions, \pkg{mathml} needs the symbol because it is unable to determine the
-function name of custom fuctions.
+function name of custom functions.
 
@@
 
-Note that the seemingly more convient `term <- quote(A + B)` yields \(A + B\) in
-the output---instead of the desired matrix representation. This behavior is
+Note that the seemingly more convenient `term <- quote(A + B)` yields \(A + B\)
+in the output---instead of the desired matrix representation. This behavior is
 expected because quotation of R calls also quote the components of the
@@
 
-Further customizations require the assertion of custom Prolog rules `math/2`,
-`ml/3`, `jax/3`, as shown in the introduction.
+Further customization requires the assertion of new Prolog rules `math/2`,
+`ml/3`, `jax/3`, as shown in the Appendix.
 
@@
 For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
-first-passage time as a function of the stimulus onset asyncrony \(\tau\),
+first-passage time as a function of the stimulus onset asynchrony \(\tau\),
 
@@
 each condition. We can, therefore, use a standard goodness-of-fit measure
-by \(z\)-standardisation.
+by \(z\)-standardization.
 
@@
 currently not possible to use the functions of \pkg{mathml} for writing
-intratextual formulas, here, the user has to adopt the usual LateX notation.
+inline formulas, here, the user has to adopt the usual LateX notation.
 Moreover, for long equations, we did not yet find a convenient way to insert
-linebreaks. This is mostly due to lacking support by MathML and LaTeX renderers.
-For example, in its current stage, the LaTeX package \pkg{breqn} [@breqn] is mostly
-a proof of concept.
+line breaks. This is mostly due to lacking support by MathML and LaTeX
+renderers. For example, in its current stage, the LaTeX package `breqn` [@breqn]
+is mostly a proof of concept.
 
@@
 We assume that `nthroot(x, n)` is available in the current
-namespace [manually defined, or from R package \pkg{pracma}, @pracma], so that
+namespace [manually defined, or from R package \CRANpkg{pracma}, @pracma], so that
 the names of the arguments and their order are accessible to `canonical()` if
@@
 recursively invokes `ml/3` for translating the function arguments _X_ and _N_,
-and then contructs the correct MathML entity `<mroot>...</mroot>`.
+and then constructs the correct MathML entity `<mroot>...</mroot>`.
 
@@
 or `a^(1.0/3L)` with a floating point number in the numerator are not detected.
+
+# Acknowledgment
+
+Supported by the Erasmus+ program of the European
+Commission (2019-1-EE01-KA203-051708).

### 2023-01-31 rj/mathml.Rmd
improve manuscript

@@
 as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
-such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit Latex commands, that
-is, `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for minimalistic use
+such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit Latex commands
+like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
 cases, these commands are tedious to type in and their source code is hard to
@@
 can be used for both mathematical typesetting and calculations. This saves time
-and reduces mistakes, as will be illustrated below.
-
-The paper is organized as follows. Section\ 2 presents the technical background
-of the package, including the two main classes of Prolog rules for translating
-R\ objects to mathematical expressions. Section\ 3 illustrates the main features
-of the \CRANpkg{mathml} package, potential issues and workarounds using examples
-from the day-to-day perspective of a user. Section\ 4 present a case study for
-the use of the package in a scientific report. Section\ 5 concludes with a
-discussion and ideas for further development.
+and potentially reduces mistakes, as will be illustrated below.
+
+The paper is organized as follows. We start with a description of the
+technical background of the package, including the two main classes of rules for
+translating R\ objects to mathematical expressions. The next section illustrates
+the main features of the \CRANpkg{mathml} package, potential issues and their
+workarounds using examples from the day-to-day perspective of a user. A case
+study follows with a scientific report written with the help of the package. The
+last section concludes with a discussion and ideas for further development.
 
@@
 
-The translation of R\ expressions to mathematical output is achieved through a
-Prolog interpreter from R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic
-programming language with many applications in expert systems, computer
-linguistics and symbolic artificial intelligence. The main strength of Prolog is
-its concise representation of facts and rules for knowledge and grammar, as well
-as its efficient built-in search engine for closed world domains. As it is
-well-known, R\ is a statistical programming language for data analysis and
-statistical modeling. Whereas Prolog is weak in statistical computation, but
-strong in symbolic manipulation, the converse may be said for the R\ language.
-The \pkg{rolog} package bridges this gap by providing an interface to a SWI-Prolog
-distribution [@swipl] in R. The communication between the two systems is mainly
-in the form of queries from R\ to Prolog, but two predicates allow Prolog to
-ring back and evaluate terms in R.
-
-For a first illustration of the \pkg{mathml} package, we consider the binomial
-probability.
-
-```{r}
+Similar to other high-level programming languages, R is homoiconic, that is,
+R\ commands (i.e., R\ "calls") are, themselves, symbolic data structures that
+can be created, parsed and modified. Because the default response of the
+R\ interpreter is to evaluate a call and return its result, this property is not
+transparent to the general user. There exist, however, a number of built-in
+R\ functions (e.g., `quote()`, `call()` etc.) that allow to create R\ calls
+which can be stored in regular variables and then, for example, be evaluated at
+a later stage or in a specific environment [@Wickham2019]. The present package
+includes a set of rules that translate such calls to a mathematical
+representation in MathML and LaTeX. For a first illustration of the \pkg{mathml}
+package, we consider the binomial probability.
+
+````{r}
 term <- quote(pbinom(k, N, p))
-term
-```
+````
 
@@
 anyway since the variables `k`, `N`, `p` have not yet been defined). Experienced
-readers will remember that the quoted expression above is a short form for
-
-```r
+R\ users will recognize that the expression is a short form for
+
+````{r}
 term <- call("pbinom", as.name("k"), as.name("N"), as.name("p"))
-```
-
-As is seen from the output above, the variable `term` is not assigned the result
-of the calculation, but an R\ call
+term
+````
+
+As is seen from the output, the variable `term` is not assigned the result of
+the calculation, but an R\ call
 [see, e.g., @Wickham2019, for details on "non-standard evaluation"]. Such a call
@@
 
-```{r}
+````{r}
 k <- 10
@@
 eval(term)
-```
-
-The R\ package \pkg{mathml} can now be used to render the call in MathML, that
-is the dialect for mathematical elements on HTML webpages or in MathJax/LaTeX,
-as shown below (some of the curly braces are not really needed in this simple
-example, but are necessary in edge cases).
-
-```{r}
+````
+
+The R\ package \pkg{mathml} can now be used to render the call in MathML or in
+MathJax/LaTeX. MathML is the dialect for mathematical elements on HTML webpages,
+whereas LaTeX is typically used for typesetting printed documents, as shown
+below.
+
+````{r}
 library(mathml)
 mathjax(term)
-```
-
-We can include the output in a RMarkdown document by specifying `results='asis'`
-in the R\ code chunk, as is done in the next example. The
-R\ function `mathout()` has been defined in this vignette; it invokes `mathml()`
-for HTML output and `mathjax()` for LaTeX output.
-
-```{r, results='asis'}
-mathout(term)
-```
+````
+
+Some of the curly braces are not really needed in this simple example, but are
+necessary in edge cases. We can include the output in a RMarkdown document by
+specifying `results='asis'` in the R\ code chunk of RMarkdown, as is done in all
+examples below. The R\ function `mathout()` has been defined in the preamble of
+this article; it wraps a call to `mathml()` for HTML output and `mathjax()` for
+LaTeX output.
+
+````{r, results='asis'}
+mathout(term)
+````
+
+For convenience, the translation of the R expressions is achieved through a
+Prolog interpreter provided by another package 
+R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic programming
+language with many applications in expert systems, computer linguistics and
+symbolic artificial intelligence. The strength of Prolog lies in its concise
+representation of facts and rules for knowledge and grammar, as well as its
+efficient built-in search engine for closed world domains. Whereas Prolog is
+weak in statistical computation, but strong in symbolic manipulation, the
+converse may be said for the R\ language. \pkg{rolog} bridges this gap by
+providing an interface to a SWI-Prolog distribution [@swipl] in R. The
+communication between the two systems is mainly in the form of queries from
+R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms
+in R.
 
@@
 
-```prolog
+````prolog
 math(pbinom(K, N, Pi), M)
  => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
-```
-
-Thus, the predicate `math/2` could be considered a "macro" that translates a
-mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
+````
+
+Thus, the predicate `math/2` could be considered a "macro" that translates one
+mathematical element (here, `pbinom(K, N, Pi)`) to another mathematical
 element, namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is
@@
 
-```prolog
-ml(Flags, fn(Name, (Args ; Pars)), M)
- => ml(Flags, Name, N),
-    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X),
+````prolog
+ml(fn(Name, (Args ; Pars)), M, Flags)
+ => ml(Name, N, Flags),
+    ml(paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X, Flags),
     M = mrow([N, mo(&(af)), X]).
-```
+````
 
@@
 with `N` representing the name of the function and `X` its arguments and
-parameters, enclosed in parentheses. A corresponding rule `jax/3` does the same
+parameters enclosed in parentheses. A corresponding rule `jax/3` does the same
 for MathJax/LaTeX. A list of flags can be used for context-sensitive
@@
 
-\pkg{mathml} is an R\ package for pretty mathematical representation of
-R\ functions and objects in data analysis, scientific reports and interactive
-web content. The currently supported features are listed below, roughly
-following the order proposed by [@murrell2000].
+The currently supported R\ objects are listed below, roughly following the order
+proposed by [@murrell2000].
 
@@
 \pkg{mathml} handles the basic elements of everyday mathematical expressions,
-such as numbers, Latin and Greek letters, multi-letter identifiers, accents,
-subscripts, and superscripts.
+such as integers, floating-point numbers, Latin and Greek letters, multi-letter
+identifiers, accents, subscripts, and superscripts.
 
@@
 
-An expression such as `1 + -2` may be considered unsatisfactory from an
-aesthetical perspective. It is correct R\ syntax, though, and is reproduced
-accordingly, without the parentheses. Parentheses around negated numbers or
-symbols can be added as shown above for `+ (-Inf)`.
+An expression such as `1 + -2` may be considered aesthetically unsatisfactory.
+It is correct R\ syntax, though, and is reproduced accordingly, without the
+parentheses. Parentheses around negative numbers or symbols can be added as
+shown above for `+ (-Inf)`.
 
@@
 boxes etc. Internally, these decorations are implemented as identity functions,
-so they can be introduced into R expressions without side-effects.
+so that they can be introduced into R expressions without side-effects.
 
@@
 placed around `g + h` in the `nodot`-example. However, we note unnecessary
-parentheses around `l + m` above. Thes parentheses are a consequence
+parentheses around `l + m` above. These parentheses are a consequence
 of `quote(a^(b + c))` actually producing a nested R\ call of the
@@
 
-For the present purpose, this feature is unfortunate because extra parentheses
-around `b + c` are not needed. The preferred result is obtained by using the
-functional form `quote('^'(k, l + m))` of the power, or curly braces as a
+For the present purpose, this R\ feature is unfortunate because the extra
+parentheses around `b + c` are not needed. The preferred result is obtained by
+the functional form `quote('^'(k, l + m))` of the power, or curly braces as a
 workaround (see `p + q` above).
@@
 strictly from left to right. Again, the problem can be worked around by the
-functional form of the operator, or a curly brace to hide the parenthesis but
-enforce the correct operator precedence.
+functional form of the operator or a curly brace to hide the parenthesis, and,
+at the same time, enforce the correct operator precedence.
 
@@
 
-There is support for most functions from package _base_, with adequate use and
+There is support for most functions from package `base`, with adequate use and
 omission of parentheses.
@@
 A few more examples are shown in Table \@ref(tab:base-stats), including
-functions from `base` and `stats`.
+functions from `stats`.
 
@@
 such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
-only the function body is represented, as illustrated by the redefinition
-of `sign` below.
+only the function body is represented. We can still display functions in the
+form `head(x) = body` if we embed the object to be shown into a
+call `"<-"(head, body)`.
 
@@
 
-As shown in the example, we can still display functions in the
-form `head(x) = body` if we embed the object to be shown into a
-call `"<-"(head, body)`.
-
 The function body is generally a nested R\ call of the form `'{'(L)`, with `L`
 being a list of commands (the semicolon, not necessary in R, is translated to a
-newline). As illustrated in the example, \pkg{mathml} provides limited support
-for control structures such as `if`.
+newline). The example also illustrates that \pkg{mathml} provides limited
+support for control structures such as `if` that is internally represented
+as `if(condition, action)`.
 
@@
 integration variable. For that purpose, the underlying Prolog bridge provides a
-predicate `r_eval/3` that calls R\ from Prolog. In the example above, this
-predicate evaluates `formalArgs(args(sin))`, which returns the names of the
-arguments of `sin`, namely, `x`.
+predicate `r_eval/2` that calls R\ from Prolog. This predicate is used to
+evaluate `formalArgs(args(sin))` and returns the names of the arguments
+of `sin()`, namely, _x_.
 
@@
 their names, not only by their position in the list of arguments. At the other
-end, Prolog does not have such a feature. Therefore, the Prolog handlers for
-R\ calls are rather rigid, for example, `integrate/3` accepts exactly three
-arguments in a particular order and without names, that
-is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.
+end, the Prolog handlers for R\ calls are rather rigid, for
+example, `integrate/3` accepts exactly three arguments in a particular order and
+without names, that is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print
+the desired result.
 
@@
 mathematical symbols, as illustrated in the example. To simplify notation,
-the `quote` flag of `hook()` defaults to TRUE, and `hook()` uses non-standard
-evaluation to unpack its arguments. If quote is FALSE, as shown above, the user
-has to provide the quoted expressions. Care should be taken to
-avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
-the \(s\) from \(s_{\mathrm{A}}\) as
+`hook()` uses non-standard evaluation of its arguments. If the `quote` flag
+of `hook()` is set to `FALSE`, the user has to provide the quoted expressions.
+Care should be taken to avoid recursive hooks such as `hook(s, s["A"])` that
+endlessly replace the \(s\) from \(s_{\mathrm{A}}\) as
 in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).
@@
 ```
+
+Further customization requires the assertion of new Prolog rules `math/2`,
+`ml/3`, `jax/3`, as shown in the Appendix.
 
@@
 this argument to tell \pkg{mathml} how to render such erroneous
-expressions using the flag `error` which can be asis, highlight, fix, or ignore.
-For more examples, see Table \@ref(tab:mistakes).
+expressions using the flag `error` which can be "asis", "highlight", "fix", 
+or "ignore". For more examples, see Table \@ref(tab:mistakes).
 
@@
 
-## Customizing the package
-
-In typical R\ functions, variable names are typically longer than just single
-letters, which may yield unsatisfactory results in the mathematical output.
-
-```{r, results='asis'}
-hook(successes, k)
-hook(quote(Ntotal), quote(N), quote=FALSE)
-hook(prob, pi)
-
-term <- quote(dbinom(successes, Ntotal, prob))
-mathout(term)
-```
-
-To improve the situation, \pkg{mathml} provides a simple hook that can be used
-to replace elements (e.g., verbose variable names) of the code by concise
-mathematical symbols, as illustrated in the example. To simplify notation,
-the `quote` flag of `hook()` defaults to TRUE, and `hook()` uses non-standard
-evaluation to unpack its arguments. If quote is FALSE, as shown above, the user
-has to provide the quoted expressions. Care should be taken to
-avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
-the \(s\) from \(s_{\mathrm{A}}\) as
-in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).
-
-Further customization requires the assertion of new Prolog rules `math/2`,
-`ml/3`, `jax/3`, as shown in the Appendix.
-
-# A case study {#case-study}
+# A case study
 
@@
 
-(One would typically use chunk option `echo=FALSE` to suppress the R code.)
-In redundant stimuli, the activations from the channel-specific diffusion
-processes add up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\),
+One would typically use chunk option `echo=FALSE` to suppress the R code.
+
+In redundant stimuli, the activation from the channel-specific diffusion
+processes adds up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\),
 hence the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process
@@
 
-# Conclusions
+# Conclusion
 
@@
 mathematical formulas in R\ [@murrell2000;@allaire2018]), bit most
-importantly, \pkg{mathml} bridges the gap between computational needs, presentation
-of results, and their reproducibility. The package supports both MathML and
-Latex/MathJax for use in R\ Markdown documents, presentations and ShinyApp
-webpages.
-
-Researchers or teachers can already use RMarkdown to conduct analyses
-and show results, and \pkg{mathml} smoothes this process and allows for integrated
+importantly, \pkg{mathml} bridges the gap between computational needs,
+presentation of results, and their reproducibility. The package supports both
+MathML and Latex/MathJax for use in R\ Markdown documents, presentations and
+ShinyApp webpages.
+
+Researchers or teachers can already use RMarkdown to conduct analyses and show
+results, and \pkg{mathml} smoothes this process and allows for integrated
 calculations and output. As shown in the case study of the previous
-section, \pkg{mathml} can help to improve data analyses and statistical reports from
-an aesthetical perspective, as well as regarding reproducibility of research.
+section, \pkg{mathml} can help to improve data analyses and statistical reports
+from an aesthetical perspective, as well as regarding reproducibility of
+research.
 
@@
 
-The package \pkg{mathml} is available for R\ version (…) and later, and can be
+The package \pkg{mathml} is available for R\ version 4.2 and later, and can be
 easily installed using the usual `install.packages("mathml")`. The source code

### 2023-04-07 rj/mathml.Rmd
small changes in the formatting

@@
 transparent to the general user. There exist, however, a number of built-in
-R\ functions (e.g., `quote()`, `call()` etc.) that allow to create R\ calls
+R\ functions (e.g., `quote()`, `call()` etc.) that allow the user to create R\ calls
 which can be stored in regular variables and then, for example, be evaluated at
@@
 op1 <- list(
-  "A\\ %*%\\ B"=quote(A %*% B),
-  "A\\ %.%\\ B"=quote(A %.% B),
-  "A\\ %x%\\ B"=quote(A %x% B),
-  "A\\ %/%\\ B"=quote(A %/% B),
-  "A\\ %%\\ B"=quote(A %% B),
-  "A\\ &\\ B"=quote(A & B),
-  "A\\ |\\ B"=quote(A | B),
-  "xor(A,\\ B)"=quote(xor(A, B)),
+  "A%*%B"=quote(A %*% B),
+  "A%.%B"=quote(A %.% B),
+  "A%x%B"=quote(A %x% B),
+  "A%/%B"=quote(A %/% B),
+  "A%%B"=quote(A %% B),
+  "A & B"=quote(A & B),
+  "A | B"=quote(A | B),
+  "xor(A, B)"=quote(xor(A, B)),
   "!A"=quote(!A),
-  "A\\ ==\\ B"=quote(A == B),
-  "A\\ <-\\ B"=quote(A <- B))
+  "A == B"=quote(A == B),
+  "A <- B"=quote(A <- B))
 
@@
 op2 <- list(
-  "A\\ !=\\ B"=quote(A != B),
-  "A\\ ~ B"=quote(A ~ B),
-  "A\\ %~~%\\ B"=quote(A %~~% B),
-  "A\\ %==%\\ B"=quote(A %==% B),
-  "A\\ %=~%\\ B"=quote(A %=~% B),
-  "A\\ %prop%\\ B"=quote(A %prop% B),
-  "A\\ %in%\\ B"=quote(A %in% B),
-  "intersect(A,\\ B)"=quote(intersect(A, B)),
-  "union(A,\\ B)"=quote(union(A, B)),
-  "crossprod(A,\\ B)"=quote(crossprod(A, B)),
+  "A != B"=quote(A != B),
+  "A ~ B"=quote(A ~ B),
+  "A%~~%B"=quote(A %~~% B),
+  "A%==%B"=quote(A %==% B),
+  "A%=~%B"=quote(A %=~% B),
+  "A%prop%B"=quote(A %prop% B),
+  "A%in%B"=quote(A %in% B),
+  "intersect(A,B)"=quote(intersect(A, B)),
+  "union(A,B)"=quote(union(A, B)),
+  "crossprod(A,B)"=quote(crossprod(A, B)),
   "is.null(A)"=quote(is.null(A)))
@@
 op3 <- list(
-  "A\\ %<->%\\ B"=quote(A %<->% B),
-  "A\\ %->%\\ B"=quote(A %->% B),
-  "A\\ %<-%\\ B"=quote(A %<-% B),
-  "A\\ %up%\\ B"=quote(A %up% B),
-  "A\\ %down%\\ B"=quote(A %down% B),
-  "A\\ %<=>%\\ B"=quote(A %<=>% B),
-  "A\\ %=>%\\ B"=quote(A %=>% B),
-  "A\\ %<=%\\ B"=quote(A %<=% B),
-  "A\\ %dblup%\\ B"=quote(A %dblup% B),
-  "A\\ %dbldown%\\ B"=quote(A %dbldown% B),
+  "A%<->%B"=quote(A %<->% B),
+  "A%->%B"=quote(A %->% B),
+  "A%<-%B"=quote(A %<-% B),
+  "A%up%B"=quote(A %up% B),
+  "A%down%B"=quote(A %down% B),
+  "A%<=>%B"=quote(A %<=>% B),
+  "A%=>%B"=quote(A %=>% B),
+  "A%<=%B"=quote(A %<=% B),
+  "A%dblup%B" =quote(A %dblup% B),
+  "A%dbldown%B"=quote(A %dbldown% B),
   " "="")
@@
   "log1p(x)"=quote(log1p(x)),
-  "logb(x,\\ e)"=quote(logb(x, e)),
+  "logb(x,e)"=quote(logb(x, e)),
   "exp(x)"=quote(exp(x)),
   "expm1(x)"=quote(expm1(x)),
-  "choose(n,\\ k)"=quote(choose(n, k)),
-  "lchoose(n,\\ k)"=quote(lchoose(n, k)),
+  "choose(n,k)"=quote(choose(n, k)),
+  "lchoose(n,k)"=quote(lchoose(n, k)),
   "factorial(n)"=quote(factorial(n)), 
@@
 op2 <- list(
-  "dbinom(k,\\ N,\\ pi)"=quote(dbinom(k, N, pi)),
-  "pbinom(k,\\ N,\\ pi)"=quote(pbinom(k, N, pi)),
-  "qbinom(p,\\ N,\\ pi)"=quote(qbinom(p, N, pi)),
-  "dpois(k,\\ lambda)"=quote(dpois(k, lambda)),
-  "ppois(k,\\ lambda)"=quote(ppois(k, lambda)),
-  "qpois(p,\\ lambda)"=quote(qpois(p, lambda)),
-  "dexp(x,\\ lambda)"=quote(dexp(x, lambda)),
-  "pexp(x,\\ lambda)"=quote(pexp(x, lambda)),
-  "qexp(p,\\ lambda)"=quote(qexp(p, lambda)),
-  "dnorm(x,\\ mu,\\ sigma)"=quote(dnorm(x, mu, sigma)),
-  "pnorm(x,\\ mu,\\ sigma)"=quote(pnorm(x, mu, sigma)),
+  "dbinom(k,N,pi)"=quote(dbinom(k, N, pi)),
+  "pbinom(k,N,pi)"=quote(pbinom(k, N, pi)),
+  "qbinom(p,N,pi)"=quote(qbinom(p, N, pi)),
+  "dpois(k,lambda)"=quote(dpois(k, lambda)),
+  "ppois(k,lambda)"=quote(ppois(k, lambda)),
+  "qpois(p,lambda)"=quote(qpois(p, lambda)),
+  "dexp(x,lambda)"=quote(dexp(x, lambda)),
+  "pexp(x,lambda)"=quote(pexp(x, lambda)),
+  "qexp(p,lambda)"=quote(qexp(p, lambda)),
+  "dnorm(x,mu,sigma)"=quote(dnorm(x, mu, sigma)),
+  "pnorm(x,mu,sigma)"=quote(pnorm(x, mu, sigma)),
   "qnorm(alpha/2L)"=quote(qnorm(alpha/2L)),
-  "1L\\ -\\ pchisq(x,\\ 1L)"=quote(1L - pchisq(x, 1L)),
-  "qchisq(1L\\ -\\ alpha,\\ 1L)"=quote(qchisq(1L-alpha, 1L)),
-  "pt(t,\\ N\\ -\\ 1L)"=quote(pt(t, N-1L)),
-  "qt(alpha/2L,\\ N\\ -\\ 1L)"=quote(qt(alpha/2L, N-1L)))
+  "1L-pchisq(x,1L)"=quote(1L - pchisq(x, 1L)),
+  "qchisq(1L-alpha,1L)"=quote(qchisq(1L-alpha, 1L)),
+  "pt(t,N-1L)"=quote(pt(t, N-1L)),
+  "qt(alpha/2L,N-1L)"=quote(qt(alpha/2L, N-1L)))
 m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))
@@
 This case study describes a model by Schwarz [-@schwarz1994] from mathematical
-psychology using the features of package \pkg{mathml}. Schwarz presents a new
+psychology using the features of package \pkg{mathml}. Schwarz [-@schwarz1994] presents a new
 explanation of redundancy gains that occur when observers respond to stimuli of
 different sources, and the same information is presented on two or more
-channels. In Schwarz' model, decision-making builds on a process of noisy
+channels. In Schwarz' [-@schwarz1994] model, decision-making builds on a process of noisy
 accumulation of information over time [e.g., @ratcliff2016]. In redundant
@@
 The best fitting parameter values and their confidence intervals are given in
-Table \@ref(tab:params) below.
+Table \@ref(tab:params).
 

### 2023-04-07 inst/pl/mathml.pl

@@ -1174,6 +1174,10 @@ ml(tilde(A), M, Flags)
  => ml(A, X, Flags),
     M = mover(accent(true), [X, mo(&(tilde))]).
 
+jax(tilde(A), M, Flags)
+ => jax(A, X, Flags),
+    format(string(M), "\\tilde{~w}", [X]).
+
 paren(tilde(A), Paren, Flags)
  => paren(A, Paren, Flags).
 

### 2023-04-07 rj/mathml.Rmd
\ removed

@@
 op1 <- list(
-  "omit_left(a\\ +\\ b)"=quote(omit_left(a + b)),
-  "omit_right(a\\ +\\ b)"=quote(omit_right(a + b)),
-  "list(quote(a),\\ quote(omit(b)))"=list(quote(a), quote(omit(b))),
-  "add_left(a\\ +\\ b)"=quote(add_left(a + b)),
-  "add_right(a\\ +\\ b)"=quote(add_right(a + b)),
-  "list(quote(a),\\ quote(add(b)))"=list(quote(a), quote(add(b))),
-  "instead(a,\\ b)\\ +\\ c"=quote(instead(a, b) + c))
+  "omit_left(a+b)"=quote(omit_left(a + b)),
+  "omit_right(a+b)"=quote(omit_right(a + b)),
+  "list(quote(a),quote(omit(b)))"=list(quote(a), quote(omit(b))),
+  "add_left(a+b)"=quote(add_left(a + b)),
+  "add_right(a+b)"=quote(add_right(a + b)),
+  "list(quote(a),quote(add(b)))"=list(quote(a), quote(add(b))),
+  "instead(a,b)+c"=quote(instead(a, b) + c))
 

### 2023-04-08 rj/mathml.Rmd
Update mathml.Rmd

@@
 translating R\ objects to mathematical expressions. The next section illustrates
-the main features of the \CRANpkg{mathml} package, potential issues and their
+the main features of the \pkg{mathml} package, potential issues and their
 workarounds using examples from the day-to-day perspective of a user. A case
@@
 Prolog interpreter provided by another package 
-R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic programming
+R\ package \pkg{rolog} [@rolog]. Prolog is a classical logic programming
 language with many applications in expert systems, computer linguistics and
@@
 We assume that `nthroot(x, n)` is available in the current
-namespace [manually defined, or from R package \CRANpkg{pracma}, @pracma], so that
+namespace [manually defined, or from R package \pkg{pracma}, @pracma], so that
 the names of the arguments and their order are accessible to `canonical()` if

### 2023-04-08 rj/mathml.Rmd
Update mathml.Rmd

@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-01-28"
+date: "2023-04-08"
 abstract: >

### 2023-04-20 R/mathml.R
Prepare version 0.7

@@ -3,14 +3,6 @@
   if(!requireNamespace("rolog", quietly=TRUE))
     stop("Could not attach library rolog.")
 
-  options(rolog.boolvec="!!")
-  options(rolog.boolmat="!!!")
-  options(rolog.intvec="%%")
-  options(rolog.intmat="%%%")
-  options(rolog.charvec="$$")
-  options(rolog.charmat="$$$")
-  options(rolog.realvec="##")
-  options(rolog.realmat="###")
   rolog::consult(system.file("pl/mathml.pl", package=pkgname))
 }
 

### 2023-04-21 rj/mathml.Rmd
manuscript draft

--- 2023-04-07 rj/mathml.Rmd
@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-01-28"
+date: "2023-04-21"
 abstract: >
@@
 ```{r custom-operators, echo=FALSE}
+escape <- function(s)
+{
+  if(knitr::is_latex_output())
+  {
+    s = lapply(s, FUN=gsub, pattern="\\\\ ", replacement=" ")
+    return(sapply(s, FUN=knitr:::escape_latex))
+  }
+
+  if(knitr::is_html_output())
+    return(sapply(s, FUN=knitr:::escape_html))
+
+  warning("escape: no knitr output specified")
+  sapply(s, FUN=knitr:::escape_html)
+}
+
 op1 <- list(
-  "A%*%B"=quote(A %*% B),
-  "A%.%B"=quote(A %.% B),
-  "A%x%B"=quote(A %x% B),
-  "A%/%B"=quote(A %/% B),
-  "A%%B"=quote(A %% B),
-  "A & B"=quote(A & B),
-  "A | B"=quote(A | B),
-  "xor(A, B)"=quote(xor(A, B)),
+  "A\\ %*%\\ B"=quote(A %*% B),
+  "A\\ %.%\\ B"=quote(A %.% B),
+  "A\\ %x%\\ B"=quote(A %x% B),
+  "A\\ %/%\\ B"=quote(A %/% B),
+  "A\\ %%\\ B"=quote(A %% B),
+  "A\\ & B"=quote(A & B),
+  "A\\ | B"=quote(A | B),
+  "xor(A,\\ B)"=quote(xor(A, B)),
   "!A"=quote(!A),
-  "A == B"=quote(A == B),
-  "A <- B"=quote(A <- B))
+  "A\\ ==\\ B"=quote(A == B),
+  "A\\ <-\\ B"=quote(A <- B))
 
@@
 op1 <- names(op1)
-if(knitr::is_latex_output())
-  op1 <- sapply(op1, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op1 <- sapply(op1, FUN=knitr:::escape_html)
+op1 <- escape(op1)
 
 op2 <- list(
-  "A != B"=quote(A != B),
-  "A ~ B"=quote(A ~ B),
-  "A%~~%B"=quote(A %~~% B),
-  "A%==%B"=quote(A %==% B),
-  "A%=~%B"=quote(A %=~% B),
-  "A%prop%B"=quote(A %prop% B),
-  "A%in%B"=quote(A %in% B),
-  "intersect(A,B)"=quote(intersect(A, B)),
-  "union(A,B)"=quote(union(A, B)),
-  "crossprod(A,B)"=quote(crossprod(A, B)),
+  "A\\ !=\\ B"=quote(A != B),
+  "A\\ ~\\ B"=quote(A ~ B),
+  "A\\ %~~%\\ B"=quote(A %~~% B),
+  "A\\ %==%\\ B"=quote(A %==% B),
+  "A\\ %=~%\\ B"=quote(A %=~% B),
+  "A\\ %prop%\\ B"=quote(A %prop% B),
+  "A\\ %in%\\ B"=quote(A %in% B),
+  "intersect(A,\\ B)"=quote(intersect(A, B)),
+  "union(A,\\ B)"=quote(union(A, B)),
+  "crossprod(A,\\ B)"=quote(crossprod(A, B)),
   "is.null(A)"=quote(is.null(A)))
@@
 op2 <- names(op2)
-if(knitr::is_latex_output())
-  op2 <- sapply(op2, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op2 <- sapply(op2, FUN=knitr:::escape_html)
+op2 <- escape(op2)
 
 op3 <- list(
-  "A%<->%B"=quote(A %<->% B),
-  "A%->%B"=quote(A %->% B),
-  "A%<-%B"=quote(A %<-% B),
-  "A%up%B"=quote(A %up% B),
-  "A%down%B"=quote(A %down% B),
-  "A%<=>%B"=quote(A %<=>% B),
-  "A%=>%B"=quote(A %=>% B),
-  "A%<=%B"=quote(A %<=% B),
-  "A%dblup%B" =quote(A %dblup% B),
-  "A%dbldown%B"=quote(A %dbldown% B),
+  "A\\ %<->%\\ B"=quote(A %<->% B),
+  "A\\ %->%\\ B"=quote(A %->% B),
+  "A\\ %<-%\\ B"=quote(A %<-% B),
+  "A\\ %up%\\ B"=quote(A %up% B),
+  "A\\ %down%\\ B"=quote(A %down% B),
+  "A\\ %<=>%\\ B"=quote(A %<=>% B),
+  "A\\ %=>%\\ B"=quote(A %=>% B),
+  "A\\ %<=%\\ B"=quote(A %<=% B),
+  "A\\ %dblup%\\ B" =quote(A %dblup% B),
+  "A\\ %dbldown%\\ B"=quote(A %dbldown% B),
   " "="")
@@
 op3 <- names(op3)
-if(knitr::is_latex_output())
-  op3 <- sapply(op3, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op3 <- sapply(op3, FUN=knitr:::escape_html)
+op3 <- escape(op3)
 
@@
   "log1p(x)"=quote(log1p(x)),
-  "logb(x,e)"=quote(logb(x, e)),
+  "logb(x,\\ e)"=quote(logb(x, e)),
   "exp(x)"=quote(exp(x)),
   "expm1(x)"=quote(expm1(x)),
-  "choose(n,k)"=quote(choose(n, k)),
-  "lchoose(n,k)"=quote(lchoose(n, k)),
+  "choose(n,\\ k)"=quote(choose(n, k)),
+  "lchoose(n,\\ k)"=quote(lchoose(n, k)),
   "factorial(n)"=quote(factorial(n)), 
@@
 op1 <- names(op1)
-if(knitr::is_latex_output())
-  op1 <- sapply(op1, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op1 <- sapply(op1, FUN=knitr:::escape_html)
+op1 <- escape(op1)
 
 op2 <- list(
-  "dbinom(k,N,pi)"=quote(dbinom(k, N, pi)),
-  "pbinom(k,N,pi)"=quote(pbinom(k, N, pi)),
-  "qbinom(p,N,pi)"=quote(qbinom(p, N, pi)),
-  "dpois(k,lambda)"=quote(dpois(k, lambda)),
-  "ppois(k,lambda)"=quote(ppois(k, lambda)),
-  "qpois(p,lambda)"=quote(qpois(p, lambda)),
-  "dexp(x,lambda)"=quote(dexp(x, lambda)),
-  "pexp(x,lambda)"=quote(pexp(x, lambda)),
-  "qexp(p,lambda)"=quote(qexp(p, lambda)),
-  "dnorm(x,mu,sigma)"=quote(dnorm(x, mu, sigma)),
-  "pnorm(x,mu,sigma)"=quote(pnorm(x, mu, sigma)),
+  "dbinom(k,\\ N,\\ pi)"=quote(dbinom(k, N, pi)),
+  "pbinom(k,\\ N,\\ pi)"=quote(pbinom(k, N, pi)),
+  "qbinom(p,\\ N,\\ pi)"=quote(qbinom(p, N, pi)),
+  "dpois(k,\\ lambda)"=quote(dpois(k, lambda)),
+  "ppois(k,\\ lambda)"=quote(ppois(k, lambda)),
+  "qpois(p,\\ lambda)"=quote(qpois(p, lambda)),
+  "dexp(x,\\ lambda)"=quote(dexp(x, lambda)),
+  "pexp(x,\\ lambda)"=quote(pexp(x, lambda)),
+  "qexp(p,\\ lambda)"=quote(qexp(p, lambda)),
+  "dnorm(x,\\ mu,\\ sigma)"=quote(dnorm(x, mu, sigma)),
+  "pnorm(x,\\ mu,\\ sigma)"=quote(pnorm(x, mu, sigma)),
   "qnorm(alpha/2L)"=quote(qnorm(alpha/2L)),
-  "1L-pchisq(x,1L)"=quote(1L - pchisq(x, 1L)),
-  "qchisq(1L-alpha,1L)"=quote(qchisq(1L-alpha, 1L)),
-  "pt(t,N-1L)"=quote(pt(t, N-1L)),
-  "qt(alpha/2L,N-1L)"=quote(qt(alpha/2L, N-1L)))
+  "1L\\ -\\ pchisq(x,\\ 1L)"=quote(1L - pchisq(x, 1L)),
+  "qchisq(1L\\ -\\ alpha,\\ 1L)"=quote(qchisq(1L-alpha, 1L)),
+  "pt(t,\\ N\\ -\\ 1L)"=quote(pt(t, N-1L)),
+  "qt(alpha/2L,\\ N\\ -\\ 1L)"=quote(qt(alpha/2L, N-1L)))
 m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))
@@
 op2 <- names(op2)
-if(knitr::is_latex_output())
-  op2 <- sapply(op2, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op2 <- sapply(op2, FUN=knitr:::escape_html)
+op2 <- escape(op2)
 
@@
 op1 <- list(
-  "omit_left(a+b)"=quote(omit_left(a + b)),
-  "omit_right(a+b)"=quote(omit_right(a + b)),
-  "list(quote(a),quote(omit(b)))"=list(quote(a), quote(omit(b))),
-  "add_left(a+b)"=quote(add_left(a + b)),
-  "add_right(a+b)"=quote(add_right(a + b)),
-  "list(quote(a),quote(add(b)))"=list(quote(a), quote(add(b))),
-  "instead(a,b)+c"=quote(instead(a, b) + c))
+  "omit_left(a\\ +\\ b)"=quote(omit_left(a + b)),
+  "omit_right(a\\ +\\ b)"=quote(omit_right(a + b)),
+  "list(quote(a),\\ quote(omit(b)))"=list(quote(a), quote(omit(b))),
+  "add_left(a\\ +\\ b)"=quote(add_left(a + b)),
+  "add_right(a\\ +\\ b)"=quote(add_right(a + b)),
+  "list(quote(a),\\ quote(add(b)))"=list(quote(a), quote(add(b))),
+  "instead(a,\\ b)\\ +\\ c"=quote(instead(a, b) + c))
 
@@
 op1 <- names(op1)
-if(knitr::is_latex_output())
-  op1 <- sapply(op1, FUN=knitr:::escape_latex)
-if(knitr::is_html_output())
-  op1 <- sapply(op1, FUN=knitr:::escape_html)
+op1 <- escape(op1)
 

### 2023-04-21 inst/pl/mathml.pl

@@ -2249,6 +2249,11 @@ ml(underbrace(A, U), M, Flags)
     M = munder([munder(accentunder(true),
                   [X, mo(stretchy(true), &('UnderBrace'))]), Y]).
 
+jax(underbrace(A, U), M, Flags)
+ => jax(A, X, Flags),
+    jax(U, Y, Flags),
+    format(string(M), "\\underbrace{~w}_{~w}", [X, Y]).
+
 paren(underbrace(A, _), Paren, Flags)
  => paren(A, Paren, Flags).
 
@@ -2385,7 +2390,7 @@ math(instead(_Wrong, Correct), M, Flags),
  => M = box(Correct).
 
 math(instead(Wrong, Correct), M, _Flags)
- => M = underbrace(Wrong, list(space, ["instead of", Correct])).
+ => M = underbrace(Wrong, list(space, ["instead", "of", Correct])).
 
 % Find minimum
 math(Optim, M),

### 2023-04-21 tests/testthat/test-mathml.R

@@ -856,8 +856,7 @@ t <- quote(instead(a, b) + c)
 test_that("instead",
           {
             q <- mathml(t)
-#            expect_equal(q,  "<math><mrow><munder><munder accentunder=\"true\"><mi>a</mi><mo stretchy=\"true\">&UnderBrace;</mo></munder><mrow><mtext>instead</mtext><mspace width=\"thinmathspace\"></mspace><mtext>of</mtext><mspace width=\"thinmathspace\"></mspace><mi>b</mi></mrow></munder><mo>+</mo><mi>c</mi></mrow></math>")
-            expect_equal(q,  "<math><mrow><munder><munder accentunder=\"true\"><mi>a</mi><mo stretchy=\"true\">&UnderBrace;</mo></munder><mrow><mtext>instead of</mtext><mspace width=\"thinmathspace\"></mspace><mi>b</mi></mrow></munder><mo>+</mo><mi>c</mi></mrow></math>")
+            expect_equal(q,  "<math><mrow><munder><munder accentunder=\"true\"><mi>a</mi><mo stretchy=\"true\">&UnderBrace;</mo></munder><mrow><mtext>instead</mtext><mspace width=\"thinmathspace\"></mspace><mtext>of</mtext><mspace width=\"thinmathspace\"></mspace><mi>b</mi></mrow></munder><mo>+</mo><mi>c</mi></mrow></math>")
           })
 
 test_that("Tdiamond",

### 2023-04-24 rj/mathml.Rmd
Update mathml.Rmd

@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-04-21"
+date: "2023-04-24"
 abstract: >

### 2023-04-25 rj/mathml.Rmd
date updated

@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-04-24"
+date: "2023-04-25"
 abstract: >

### 2023-06-16 rj/mathml.Rmd
Reviewer 1 (except new code)

@@
   functions in data analyses, scientific reports and interactive web content. In
-  the RMarkdown document rendering language, R\ code and mathematical content
+  the R\ Markdown document rendering language, R\ code and mathematical content
   already exist side-by-side. The present package allows to use the same
@@
 The R\ extension of the markdown language [@Xie2020] enables reproducible
-statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
+statistical reports with nice typesetting in HTML, Microsoft Word, and LaTeX.
 Since recently [@R, version 4.2], R\'s manual pages include support for
 mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a big
-improvement. However, rules for the mapping of built-in language elements to
-their mathematical representation are still lacking. So far, R\ expressions such
-as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
-such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit Latex commands
+improvement. However, except for special cases such as regression
+models [@equatiomatic] and R's own plotmath annotation, rules for the mapping of
+built-in language elements to their mathematical representation are still
+lacking. So far, R\ expressions such as `pbinom(k, N, p)` are printed as they
+are; pretty mathematical formulae such as \(P_{\mathrm{Bi}}(X \le k; N, p)\)
+require explicit LaTeX commands
 like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
@@
 The present R\ package defines a set of rules for the automatic translation of
-R\ expressions to mathematical output in RMarkdown documents [@Xie2020] and
+R\ expressions to mathematical output in R\ Markdown documents [@Xie2020] and
 ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
@@
 Some of the curly braces are not really needed in this simple example, but are
-necessary in edge cases. We can include the output in a RMarkdown document by
-specifying `results='asis'` in the R\ code chunk of RMarkdown, as is done in all
+necessary in edge cases. We can include the output in a R\ Markdown document by
+specifying `results='asis'` in the R\ code chunk of R\ Markdown, as is done in all
 examples below. The R\ function `mathout()` has been defined in the preamble of
@@
 communication between the two systems is mainly in the form of queries from
-R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms
+R\ to Prolog, but two Prolog functions allow to ring back and evaluate terms
 in R.
 
-At the Prolog end, a predicate `math/2` translates the call `pbinom(K, N, Pi)`
-into a "function" `fn/2` with the name `P_Bi`, one argument `X =< K`, and the
-two parameters `N` and `Pi`.
+The proper term for a Prolog "function" is predicate, and it is typically
+written with name and arity (i.e., number of arguments), separated by a forward
+slash. Thus, at the Prolog end, the predicate `math/2` translates the
+representation of the R call `pbinom(K, N, Pi)` into a more general 
+representation of an R function `fn/2` with the name `P_Bi`, one 
+argument `X =< K`, and the two parameters `N` and `Pi`, as shown below.
 
@@
 
-Thus, the predicate `math/2` could be considered a "macro" that translates one
-mathematical element (here, `pbinom(K, N, Pi)`) to another mathematical
-element, namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is
-used to convert these basic elements to MathML.
+`math/2` operates like macro that translates one mathematical
+element (here, `pbinom(K, N, Pi)`) to another mathematical element,
+namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is used to
+convert these basic elements to MathML.
 
@@
 detect if parentheses are needed; for example, parentheses are automatically
-placed around `g + h` in the `nodot`-example. However, we note unnecessary
+placed around `d + e` in the `nodot`-example. However, we note unnecessary
 parentheses around `l + m` above. These parentheses are a consequence
@@
 extends the current features of R\ and existing packages for displaying
-mathematical formulas in R\ [@murrell2000;@allaire2018]), bit most
+mathematical formulas in R\ [@murrell2000;@allaire2018]), but most
 importantly, \pkg{mathml} bridges the gap between computational needs,
 presentation of results, and their reproducibility. The package supports both
-MathML and Latex/MathJax for use in R\ Markdown documents, presentations and
+MathML and LaTeX/MathJax for use in R\ Markdown documents, presentations and
 ShinyApp webpages.
 
-Researchers or teachers can already use RMarkdown to conduct analyses and show
+Researchers or teachers can already use R\ Markdown to conduct analyses and show
 results, and \pkg{mathml} smoothes this process and allows for integrated
@@
 currently not possible to use the functions of \pkg{mathml} for writing
-inline formulas, here, the user has to adopt the usual LateX notation.
+inline formulas, here, the user has to adopt the usual LaTeX notation. The
+reason is the missing chunk option results="asis" in inline output from
+R\ Markdown, such that the HTML or LaTeX commands are quoted.
 Moreover, for long equations, we did not yet find a convenient way to insert
@@
 renderers. For example, in its current stage, the LaTeX package `breqn` [@breqn]
-is mostly a proof of concept.
+is mostly a proof of concept. Finally, \pkg{mathml} is one-way road, that is,
+translation from LaTeX or HTML back to R is not 
+possible [see @latex2r, for an example].
 
@@
 R\ hooks mentioned above, and the low-level predicates `ml/3` and `jax/3` that
-create proper MathML and Latex terms.
+create proper MathML and LaTeX terms.
 
@@
 To render the model equation of a linear model such
-as `lm(EOT ~ T0 + Therapy, data=d)` in mathematical form, it is sufficient to
-map the `Formula` in `lm(Formula, Data)` to its respective equation. This can in
-two ways, using either the hooks described above, or a new `math/2` macro at the
-Prolog end.
+as `lm(EOT ~ T0 + Therapy, data=d)` in mathematical
+form [see also @equatiomatic], it is sufficient to map the `Formula`
+in `lm(Formula, Data)` to its respective equation. This can be done in two ways,
+using either the hooks described above, or a new `math/2` macro at the Prolog
+end.
 
@@
 the head of `mlx/3`. The Prolog file `nthroot.pl` also includes the respective
-rule for Latex and can be consulted from the package folder via the underlying
+rule for LaTeX and can be consulted from the package folder via the underlying
 package \pkg{rolog}.

### 2023-06-18 rj/mathml.Rmd
Review 1

@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-04-25"
+date: "2023-06-17"
 abstract: >
@@
 library(mathml)
-
-mathout <- function(..., flags=NULL, env=parent.frame())
-{
-  if(knitr::is_html_output())
-    return(mathml(..., flags=c(flags, list(cat=TRUE)), env=env))
-
-  if(knitr::is_latex_output())
-    return(mathjax(..., flags=c(flags, list(cat=TRUE)), env=env))
-
-  warning("mathout: no knitr output specified")  
-  mathjax(..., flags=c(flags, list(cat=TRUE)), env=env)
-}
-
-catmath <- function(term, ...)
-{
-  s <- substitute(term)
-  mathout(canonical(s), ...)
-}
 ```
@@
 library(mathml)
+substr(mathml(term), 1, 70)
 mathjax(term)
@@
 
-Some of the curly braces are not really needed in this simple example, but are
-necessary in edge cases. We can include the output in a R\ Markdown document by
-specifying `results='asis'` in the R\ code chunk of R\ Markdown, as is done in all
-examples below. The R\ function `mathout()` has been defined in the preamble of
-this article; it wraps a call to `mathml()` for HTML output and `mathjax()` for
-LaTeX output.
-
-````{r, results='asis'}
-mathout(term)
-````
+Some of the curly braces are not really needed in the LaTeX output, but are
+necessary in edge cases. The package also includes a function `mathout()` that
+wraps a call to `mathml()` for HTML output and `mathjax()` for LaTeX output.
+Moreover, the function `math(x)` adds the class `"math"` to its argument, such
+that a special knitr printing function is invoked. We can include the output in
+a R\ Markdown document by specifying `results="asis"` in the R\ code chunk of
+R\ Markdown, as is done in all examples below.
+
+````{r, results="asis"}
+math(term)
+````
+
+Similarly, `inline()` produces inline output: `r inline(term)`.
 
@@
 term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + roof(b)[i, j]^2L)
-mathout(term)
+math(term)
 
 term <- quote(round(3.1415, 3L) + NaN + NA + TRUE + FALSE + Inf + (-Inf))
-mathout(term)
+math(term)
 ```
@@
 term <- quote(bold(b[x, 5L]) + bold(b[italic(x)]) + italic(ab) + italic(42L))
-mathout(term)
+math(term)
 
 term <- quote(tilde(a) + mean(X) + box(c) + cancel(d) + phantom(e) + prime(f))
-mathout(term)
+math(term)
 ```
@@
 term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
-mathout(term)
+math(term)
 
 term <- quote(dot(a, b) + frac(1L, nodot(c, d + e)) + dfrac(1L, times(g, h)))
-mathout(term)
+math(term)
 ```
@@
 term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
-mathout(term)
+math(term)
 term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N))) # functional form of '%+-%'
 term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
-mathout(term)
+math(term)
 ```
@@
 term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
-mathout(term)
+math(term)
 term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
-mathout(term)
+math(term)
 ```
@@
 
-mathout(sgn)
-mathout(call("<-", quote(sgn(x)), sgn))
+math(sgn)
+math(call("<-", quote(sgn(x)), sgn))
 ```
@@
 term <- quote(S[Y]^2L <- frac(1L, N) * sum(Y[i] - mean(Y))^2L)
-mathout(term)
+math(term)
 
 term <- quote(log(prod_over(L[i], i==1L, N)) <- sum_over(log(L[i]), i==1L, N))
-mathout(term)
+math(term)
 ```
@@
 term <- quote(integrate(sin, 0L, 2L*pi))
-mathout(term)
+math(term)
 eval(term)
@@
 ```{r, results='asis'}
-mathout(canonical(term))
+math(canonical(term))
 ```
@@
 v <- 1:3
-mathout(call("t", v))
+math(call("t", v))
 
@@
 term <- call("+", A, B)
-mathout(term)
+math(term)
 ```
@@
 term <- quote(pbinom(successes, Ntotal, prob))
-mathout(term)
+math(term)
 
@@
 hook(prob, pi)
-mathout(term)
+math(term)
 ```
@@
 hook(pbinom(.K, .N, .P), sum_over(dbinom(i, .N, .P), i=0L, .K))
-mathout(term)
+math(term)
 ```
@@
                 "the pooled variance.") * (frac(1L, n_A) + frac(1L, n_B)))))
-mathout(term)
+math(term)
 ```
@@
 t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
-mathout(t, flags=list(error="highlight"))
-mathout(t, flags=list(error="fix"))
+math(t, flags=list(error="highlight"))
+math(t, flags=list(error="fix"))
 ```
@@
 
-For this purpose, the functions `mathml()` or `mathjax()` have an optional
-argument `flags` which is a list with named elements. In this example, we use
-this argument to tell \pkg{mathml} how to render such erroneous
-expressions using the flag `error` which can be "asis", "highlight", "fix", 
-or "ignore". For more examples, see Table \@ref(tab:mistakes).
+For this purpose, the functions `mathml()`, `mathjax()`, `mathout()`
+and `math()` have an optional argument `flags` which is a list with named
+elements. In this example, we use this argument to tell \pkg{mathml} how to
+render such erroneous expressions using the flag `error` which can be "asis",
+"highlight", "fix", or "ignore". For more examples,
+see Table \@ref(tab:mistakes).
 
@@
 This case study describes a model by Schwarz [-@schwarz1994] from mathematical
-psychology using the features of package \pkg{mathml}. Schwarz [-@schwarz1994] presents a new
-explanation of redundancy gains that occur when observers respond to stimuli of
-different sources, and the same information is presented on two or more
-channels. In Schwarz' [-@schwarz1994] model, decision-making builds on a process of noisy
-accumulation of information over time [e.g., @ratcliff2016]. In redundant
-stimuli, the model assumes a superposition of channel-specific diffusion
-processes that eventually reach an absorbing barrier to elicit the response. For
-a detailed description the reader may refer to the original article.
+psychology using the features of package \pkg{mathml}. Schwarz [-@schwarz1994]
+presents a new explanation of redundancy gains that occur when observers respond
+to stimuli of different sources, and the same information is presented on two or
+more channels. In Schwarz' [-@schwarz1994] model, decision-making builds on a
+process of noisy accumulation of information over time [e.g., @ratcliff2016]. In
+redundant stimuli, the model assumes a superposition of channel-specific
+diffusion processes that eventually reach an absorbing barrier to elicit the
+response. For a detailed description the reader may refer to the original
+article.
 
@@
 hook(ED_single(.C, .Mu), E(`;`(D, .Mu)))
-mathout(call("=", quote(ED_single(c, mu)), ED_single))
+math(call("=", quote(ED_single(c, mu)), ED_single))
 ```
@@
 
-mathout(call("=", quote(E(D["AB"])), quote(ED_single(c, mu_A + mu_B))))
+math(call("=", quote(E(D["AB"])), quote(ED_single(c, mu_A + mu_B))))
 ```
@@
 hook(ED_async(.Tau, .C, .MA, .SA, .MB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB))))
-mathout(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
+math(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
 ```
@@
 hook(ED(.Tau, .C, .MA, .SA, .MB, .SB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB, .SB))))
-mathout(call("=", quote(ED(tau, c, mu_A, sigma_A, mu_B, sigma_B)), ED))
+math(call("=", quote(ED(tau, c, mu_A, sigma_A, mu_B, sigma_B)), ED))
 ```
@@
      E(`;`(T[.Tau], `,`(.MA, .SA, .MB, .SB, .MM))))
-mathout(call("=", quote(E(T[tau])), ET))
+math(call("=", quote(E(T[tau])), ET))
 ```
@@
 
-mathout(call("=", quote(z[tau]), z))
+math(call("=", quote(z[tau]), z))
 ```
@@
 
-mathout(call("=", quote(X["8 df"]^2L), gof))
+math(call("=", quote(X["8 df"]^2L), gof))
 ```
@@
 
-mathout(call("=", quote(hat(theta)), lsq))
+math(call("=", quote(hat(theta)), lsq))
 ```
@@
 
-mathout(f1)
+math(f1)
 ````
@@
 
-mathout(f2)
+math(f2)
 ````
@@
 
-In its current version \pkg{mathml} has some limitations. For example, it is
-currently not possible to use the functions of \pkg{mathml} for writing
-inline formulas, here, the user has to adopt the usual LaTeX notation. The
-reason is the missing chunk option results="asis" in inline output from
-R\ Markdown, such that the HTML or LaTeX commands are quoted.
-Moreover, for long equations, we did not yet find a convenient way to insert
-line breaks. This is mostly due to lacking support by MathML and LaTeX
-renderers. For example, in its current stage, the LaTeX package `breqn` [@breqn]
-is mostly a proof of concept. Finally, \pkg{mathml} is one-way road, that is,
-translation from LaTeX or HTML back to R is not 
-possible [see @latex2r, for an example].
+One limitation of the package is the lack of a convenient way to insert line
+breaks. This is mostly due to lacking support by MathML and LaTeX renderers. 
+For example, in its current stage, the LaTeX package `breqn` [@breqn] is mostly
+a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, translation
+from LaTeX or HTML back to R is not possible [see @latex2r, for an example].
 
@@
 term <- quote(lm(EOT ~ T0 + Therapy, data=d, na.action=na.fail))
-mathout(term)
+math(term)
 ````
@@
 term <- canonical(quote(nthroot(n=3L, 2L)))
-mathout(term)
+math(term)
 ```
@@
 term <- quote(nthroot(a * (b + c), 3L)^2L)
-mathout(term)
+math(term)
 term <- quote(a^(1L/3L) + a^{1L/3L} + a^(1.0/3L))
-mathout(term)
+math(term)
 ```

### 2023-06-18 R/mathml.R

@@ -1,3 +1,8 @@
+.onLoad <- function(...)
+{
+  registerS3method("knit_print", "math", .knit_print.math)
+}
+
 .onAttach <- function(libname, pkgname)
 {
   if(!requireNamespace("rolog", quietly=TRUE))
@@ -21,7 +26,7 @@
 #' The R environment in which r_eval is being executed.
 #' 
 #' @return
-#' A string with the MathML representation or _term_.
+#' A string with the MathML representation of _term_.
 #'
 #' @seealso [mathjax()]
 #'
@@ -105,7 +110,7 @@ mathml_preproc <- function(query=quote(5 %% 2))
 #' details, "Ringing back to R").
 #'
 #' @return
-#' A string with the MathJax representation or _term_.
+#' A string with the MathJax representation of _term_.
 #'
 #' @md
 #'
@@ -132,6 +137,89 @@ mathjax <- function(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L), flags=NULL,
     return(cat(r))
 
   return(r)
+}
+
+#' MathML or MathJax output, depending on the knitr context
+#'
+#' @param term
+#' an R call or symbol/number. This function translates _term_ into a
+#' LaTeX/MathJax string.
+#'
+#' @param flags (default NULL)
+#' list of flags that control the translation
+#'
+#' @param env (default parent.frame())
+#' The R environment in which r_eval is being executed (see vignette for
+#' details, "Ringing back to R").
+#'
+#' @return
+#' A string with the MathML or MathJax representation of _term_.
+#'
+#' @md
+#'
+#' @details This function checks knitr::is_html_output() 
+#' and knitr::is_html_output() and invokes the respective function mathml() or
+#' mathjax(). Outside of knitr context, MathML is returned, and a warning is
+#' given.
+#'
+#' @seealso [mathml()], [mathjax()], [inline()]
+#'
+#' @examples
+#' mathout(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
+#'
+#' inline(term=quote((a + b)^2L == a^2L + 2L*a*b + b^2L))
+#'
+mathout <- function(term, flags=NULL, env=parent.frame())
+{
+  if(knitr::is_html_output())
+    return(mathml(term, flags=c(flags, list(cat=TRUE)), env=env))
+
+  if(knitr::is_latex_output())
+    return(mathjax(term, flags=c(flags, list(cat=TRUE)), env=env))
+
[…]

### 2023-06-18 rj/mathml.Rmd
Reviewer 3

--- 2023-06-16 rj/mathml.Rmd
@@
 
-The R\ extension of the markdown language [@Xie2020] enables reproducible
-statistical reports with nice typesetting in HTML, Microsoft Word, and LaTeX.
-Since recently [@R, version 4.2], R\'s manual pages include support for
-mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a big
-improvement. However, except for special cases such as regression
+The R\ extension of the markdown language [@Xie2018;rmarkdown] enables
+reproducible statistical reports with nice typesetting in HTML, Microsoft Word,
+and LaTeX. Since recently [@R, version 4.2], R\'s manual pages include support
+for mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a
+big improvement. However, except for special cases such as regression
 models [@equatiomatic] and R's own plotmath annotation, rules for the mapping of
@@
 R\ expressions to mathematical output in R\ Markdown documents [@Xie2020] and
-ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
+Shiny apps [@Chang2022]. The translation is done by an embedded Prolog
 interpreter that maps nested expressions recursively to MathML and
@@
 can be used for both mathematical typesetting and calculations. This saves time
-and potentially reduces mistakes, as will be illustrated below.
+and potentially reduces mistakes, as will be illustrated below. Readers should
+have basic knowledge of knitr and R\ Markdown to be able to follow this
+article [@knitr;@rmarkdown]. To extend and customize the package, some basic
+knowledge of Prolog is needed.
 
@@
 
-Some of the curly braces are not really needed in this simple example, but are
-necessary in edge cases. We can include the output in a R\ Markdown document by
-specifying `results='asis'` in the R\ code chunk of R\ Markdown, as is done in all
-examples below. The R\ function `mathout()` has been defined in the preamble of
-this article; it wraps a call to `mathml()` for HTML output and `mathjax()` for
-LaTeX output.
-
-````{r, results='asis'}
-mathout(term)
-````
-
-For convenience, the translation of the R expressions is achieved through a
-Prolog interpreter provided by another package 
-R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic programming
-language with many applications in expert systems, computer linguistics and
-symbolic artificial intelligence. The strength of Prolog lies in its concise
-representation of facts and rules for knowledge and grammar, as well as its
-efficient built-in search engine for closed world domains. Whereas Prolog is
-weak in statistical computation, but strong in symbolic manipulation, the
-converse may be said for the R\ language. \pkg{rolog} bridges this gap by
-providing an interface to a SWI-Prolog distribution [@swipl] in R. The
-communication between the two systems is mainly in the form of queries from
-R\ to Prolog, but two Prolog functions allow to ring back and evaluate terms
-in R.
-
-The proper term for a Prolog "function" is predicate, and it is typically
-written with name and arity (i.e., number of arguments), separated by a forward
-slash. Thus, at the Prolog end, the predicate `math/2` translates the
-representation of the R call `pbinom(K, N, Pi)` into a more general 
-representation of an R function `fn/2` with the name `P_Bi`, one 
-argument `X =< K`, and the two parameters `N` and `Pi`, as shown below.
-
-````prolog
-math(pbinom(K, N, Pi), M)
- => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
-````
-
-`math/2` operates like macro that translates one mathematical
-element (here, `pbinom(K, N, Pi)`) to another mathematical element,
-namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is used to
-convert these basic elements to MathML.
-
-````prolog
-ml(fn(Name, (Args ; Pars)), M, Flags)
- => ml(Name, N, Flags),
-    ml(paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X, Flags),
-    M = mrow([N, mo(&(af)), X]).
-````
-
-The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
-with `N` representing the name of the function and `X` its arguments and
-parameters enclosed in parentheses. A corresponding rule `jax/3` does the same
-for MathJax/LaTeX. A list of flags can be used for context-sensitive
-translation (see, e.g., the section on errors below).
+Some of the curly braces are not really needed in the LaTeX output, but are
+necessary in edge cases. The package also includes a function `mathout()` that
+wraps a call to `mathml()` for HTML output and `mathjax()` for LaTeX output.
+Moreover, the function `math(x)` adds the class `"math"` to its argument, such
+that a special knitr printing function is invoked. We can include the output in
+a R\ Markdown document by specifying `results="asis"` in the R\ code chunk of
+R\ Markdown, as is done in all examples below.
+
+``````markdown
+`r ''`````{r, results="asis"}
+math(term)
+````
+``````
+
+````{r, results="asis", echo=FALSE}
+math(term)
+````
+
+Similarly, `inline()` produces inline
+output, `` `r "\x60r inline(term)\x60"` `` yields `r inline(term)`.
 
@@
 # Appendix: Customizing the package
+
+## Implementation details
+
+For convenience, the translation of the R expressions is achieved through a
+Prolog interpreter provided by another R\ package \CRANpkg{rolog} [@rolog]. If a
+version of SWI-Prolog [@swipl] is found on the system, \CRANpkg{rolog} connects
+to it. Alternatively, the SWI-Prolog runtime libraries can be conveniently
+accessed by installing the R\ package \CRANpkg{rswipl} [@rswipl]. Prolog is a
+classical logic programming language with many applications in expert systems,
+computer linguistics and symbolic artificial intelligence. The strength of
+Prolog lies in its concise representation of facts and rules for knowledge and
+grammar, as well as its efficient built-in search engine for closed world
+domains. Whereas Prolog is weak in statistical computation, but strong in
+symbolic manipulation, the converse may be said for the R\ language. \pkg{rolog}
+bridges this gap by providing an interface to a SWI-Prolog
+distribution [@swipl] in R. The communication between the two systems is mainly
+in the form of queries from R\ to Prolog, but two Prolog functions allow to ring
+back and evaluate terms in R.
+
+The proper term for a Prolog "function" is predicate, and it is typically
+written with name and arity (i.e., number of arguments), separated by a forward
+slash. Thus, at the Prolog end, the predicate `math/2` translates the
+representation of the R call `pbinom(K, N, Pi)` into a more general 
+representation of an R function `fn/2` with the name `P_Bi`, one 
+argument `X =< K`, and the two parameters `N` and `Pi`, as shown below.
+
+````prolog
+math(pbinom(K, N, Pi), M)
+ => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
+````
+
+`math/2` operates like macro that translates one mathematical
+element (here, `pbinom(K, N, Pi)`) to another mathematical element,
+namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is used to
+convert these basic elements to MathML.
+
+````prolog
+ml(fn(Name, (Args ; Pars)), M, Flags)
+ => ml(Name, N, Flags),
+    ml(paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X, Flags),
+    M = mrow([N, mo(&(af)), X]).
+````
+
+The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
+with `N` representing the name of the function and `X` its arguments and
+parameters enclosed in parentheses. A corresponding rule `jax/3` does the same
+for MathJax/LaTeX. A list of flags can be used for context-sensitive
+translation (see, e.g., the section on errors above).
 

### 2023-06-20 rj/mathml.Rmd
small changes

@@
 
-The R\ extension of the markdown language [@Xie2018;rmarkdown] enables
+The R\ extension of the markdown language [@Xie2018;@rmarkdown] enables
 reproducible statistical reports with nice typesetting in HTML, Microsoft Word,
-and LaTeX. Since recently [@R, version 4.2], R\'s manual pages include support
-for mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a
-big improvement. However, except for special cases such as regression
-models [@equatiomatic] and R's own plotmath annotation, rules for the mapping of
-built-in language elements to their mathematical representation are still
-lacking. So far, R\ expressions such as `pbinom(k, N, p)` are printed as they
-are; pretty mathematical formulae such as \(P_{\mathrm{Bi}}(X \le k; N, p)\)
-require explicit LaTeX commands
+and LaTeX. Moreover, since recently [@R, version 4.2], R\'s manual pages 
+include support for mathematical expressions [@Sarkar2022;@Viechtbauer2022],
+which already is a big improvement. However, except for special cases 
+such as regression models [@equatiomatic] and R's own plotmath annotation, 
+rules for the mapping of built-in language elements to their mathematical 
+representation are still lacking. So far, R\ expressions such 
+as `pbinom(k, N, p)` are printed as they are and pretty mathematical formulae 
+such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit LaTeX commands
 like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
@@
 R\ expressions to mathematical output in R\ Markdown documents [@Xie2020] and
-Shiny apps [@Chang2022]. The translation is done by an embedded Prolog
+Shiny Apps [@Chang2022]. The translation is done by an embedded Prolog
 interpreter that maps nested expressions recursively to MathML and
@@
 have basic knowledge of knitr and R\ Markdown to be able to follow this
-article [@knitr;@rmarkdown]. To extend and customize the package, some basic
-knowledge of Prolog is needed.
+article [@knitr;@rmarkdown], while to extend and customize the package, 
+some basic knowledge of Prolog is needed.
 
@@
 R\ interpreter is to evaluate a call and return its result, this property is not
-transparent to the general user. There exist, however, a number of built-in
+transparent to the general user. There exists, however, a number of built-in
 R\ functions (e.g., `quote()`, `call()` etc.) that allow the user to create R\ calls
-which can be stored in regular variables and then, for example, be evaluated at
+which can be stored in regular variables and then, for example, evaluated at
 a later stage or in a specific environment [@Wickham2019]. The present package
@@
 
-As is seen from the output, the variable `term` is not assigned the result of
-the calculation, but an R\ call
-[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. Such a call
-can eventually be evaluated with `eval()`,
+As it can be seen from the output, to the variable `term` is not assigned the 
+result of the calculation, yet an R\ call [see, e.g., @Wickham2019, for details 
+on "non-standard evaluation"], which can eventually be evaluated with `eval()`.
 
@@
 
-For self-written functions, matters are a bit more complicated. For a function
+For self-written functions, the matter is somewhat more complicated. For a function
 such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
@@
 MathML and LaTeX/MathJax for use in R\ Markdown documents, presentations and
-ShinyApp webpages.
+Shiny App webpages.
 
@@
 For example, in its current stage, the LaTeX package `breqn` [@breqn] is mostly
-a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, translation
-from LaTeX or HTML back to R is not possible [see @latex2r, for an example].
+a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, it is not 
+possible to translate from LaTeX or HTML back to R [see @latex2r, for an example].
 

### 2023-06-20 rj/mathml.Rmd
Small changes in the text (references added)

@@
 result of the calculation, yet an R\ call [see, e.g., @Wickham2019, for details 
-on "non-standard evaluation"], which can eventually be evaluated with `eval()`.
+on "non-standard evaluation"], which can eventually be evaluated with `eval()`,
 

### 2023-06-20 rj/mathml.Rmd
The Sections of the Vignette have been reorganized.

@@
 The currently supported R\ objects are listed below, roughly following the order
-proposed by [@murrell2000].
+proposed by Murrell and Ihaka [-@murrell2000].
 
@@
 extends the current features of R\ and existing packages for displaying
-mathematical formulas in R\ [@murrell2000;@allaire2018]), but most
+mathematical formulas in R\ [@murrell2000], but most
 importantly, \pkg{mathml} bridges the gap between computational needs,
@@
 
-`math/2` operates like macro that translates one mathematical
+`math/2` operates like a macro that translates one mathematical
 element (here, `pbinom(K, N, Pi)`) to another mathematical element,

### 2023-06-20 R/mathml.R
Suppress warning from is.null(x)

@@ -217,9 +217,9 @@ math <- function(term, flags=NULL)
 {
   flags <- attr(x, "flags")
   if(inline)
-    return(structure(inline(x, flags), class="knit_asis"))
+    return(knitr::asis_output(as.character(inline(x, flags=flags))))
 
-  structure(mathout(x, flags), class="knit_asis")
+  structure(knitr::asis_output(as.character(mathout(x, flags=flags))))
 }
 
 #' Add a name attribute to an element (most often, an R function)

### 2023-06-29 rj/mathml.Rmd
math()
small typo

@@
 include support for mathematical expressions [@Sarkar2022;@Viechtbauer2022],
-which already is a big improvement. However, except for special cases 
+which is already a big improvement. However, except for special cases 
 such as regression models [@equatiomatic] and R's own plotmath annotation, 

### 2023-06-29 rj/mathml.Rmd

@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-06-17"
+date: "2023-06-29"
 abstract: >

### 2023-07-06 rj/mathml.Rmd

@@
 Moreover, the function `math(x)` adds the class `"math"` to its argument, such
-that a special knitr printing function is invoked. We can include the output in
-a R\ Markdown document by specifying `results="asis"` in the R\ code chunk of
-R\ Markdown, as is done in all examples below.
-
-``````markdown
-`r ''`````{r, results="asis"}
-math(term)
-````
-``````
-
-````{r, results="asis", echo=FALSE}
+that a special knitr printing function is
+invoked [see the vignette on custom print methods in @knitr].
+
+````{r, echo=FALSE}
 math(term)
@@
 
-```{r, results='asis'}
+```{r}
 term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + roof(b)[i, j]^2L)
-mathout(term)
+math(term)
 
 term <- quote(round(3.1415, 3L) + NaN + NA + TRUE + FALSE + Inf + (-Inf))
-mathout(term)
+math(term)
 ```
@@
 
-```{r, results='asis'}
+```{r}
 term <- quote(bold(b[x, 5L]) + bold(b[italic(x)]) + italic(ab) + italic(42L))
-mathout(term)
+math(term)
 
 term <- quote(tilde(a) + mean(X) + box(c) + cancel(d) + phantom(e) + prime(f))
-mathout(term)
+math(term)
 ```
@@
 
-```{r, results='asis'}
+```{r}
 term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
-mathout(term)
+math(term)
 term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N))) # functional form of '%+-%'
 term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
-mathout(term)
+math(term)
 ```
@@
 
-```{r, results='asis'}
+```{r}
 term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
-mathout(term)
+math(term)
 term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
-mathout(term)
+math(term)
 ```
@@
 
-```{r, results='asis'}
+```{r}
 sgn <- function(x)
@@
 
-```{r, results='asis'}
+```{r}
 term <- quote(S[Y]^2L <- frac(1L, N) * sum(Y[i] - mean(Y))^2L)
-mathout(term)
+math(term)
 
 term <- quote(log(prod_over(L[i], i==1L, N)) <- sum_over(log(L[i]), i==1L, N))
-mathout(term)
+math(term)
 ```
@@
 
-```{r, results='asis'}
+```{r}
 term <- quote(integrate(sin, 0L, 2L*pi))
-mathout(term)
+math(term)
 eval(term)
@@
 
-```{r, results='asis'}
-mathout(canonical(term))
+```{r}
+math(canonical(term))
 ```
@@
 
-```{r, results='asis'}
+```{r}
 v <- 1:3
-mathout(call("t", v))
+math(call("t", v))
 
@@
 
-```{r, results='asis'}
+```{r}
 term <- quote(pbinom(successes, Ntotal, prob))
-mathout(term)
+math(term)
 
@@
 
-```{r, results='asis'}
+```{r}
 hook(pbinom(.K, .N, .P), sum_over(dbinom(i, .N, .P), i=0L, .K))
-mathout(term)
+math(term)
 ```
@@
 
-```{r, results='asis'}
+```{r}
 hook(m_A, mean(X)["A"]) ; hook(s2_A, s["A"]^2L) ;
@@
 
-```{r, results='asis'}
+```{r}
 t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
-mathout(t, flags=list(error="highlight"))
-mathout(t, flags=list(error="fix"))
+math(t, flags=list(error="highlight"))
+math(t, flags=list(error="fix"))
 ```
@@
 
-```{r, results='asis'}
+```{r}
 ED_single <- function(c, mu)
@@
 
-```{r, results='asis'}
+```{r}
 hook(mu_A, mu["A"])
@@
 
-```{r, results='asis'}
+```{r}
 ED <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
@@
 
-```{r, results='asis'}
+```{r}
 ET <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
@@
 
-```{r, results='asis'}
+```{r}
 z <- function(m, s, n, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
@@
 
-```{r, results='asis'}
+```{r}
 zv <- Vectorize(z, vectorize.args = c('m', 's', 'n', 'tau'))
@@
 
-````{r, results='asis'}
+````{r}
 rolog::consult(system.file(file.path("pl", "lm.pl"), package="mathml"))
@@
 
-```{r, results='asis'}
+```{r}
 nthroot <- function(x, n)
@@
 
-```{r, results='asis'}
+```{r}
 rolog::consult(system.file(file.path("pl", "nthroot.pl"), package="mathml"))

### 2023-07-11 rj/mathml.Rmd
results=asis no longer needed

@@
 
-```{r, results="asis"}
+```{r}
 term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
@@
 
-```{r, results="asis"}
+```{r}
 ED_async <- function(tau, c, mu_A, sigma_A, mu_B)
@@
 
-```{r, results="asis", echo=FALSE}
+```{r, echo=FALSE}
 lsq <- function(tau, m, s, n)
@@
 
-````{r, results="asis"}
+````{r}
 f1 <- function(tau)
@@
 
-````{r, results="asis"}
+````{r}
 f2 <- function(tau)

### 2023-07-11 R/mathml.R

@@ -222,7 +222,7 @@ math <- function(term, flags=NULL)
   if(inline)
     return(knitr::asis_output(as.character(inline(x, flags=flags))))
 
-  structure(knitr::asis_output(as.character(mathout(x, flags=flags))))
+  structure(knitr::asis_output(as.character(mathout(x, flags=c(cat=FALSE, flags)))))
 }
 
 #' Add a name attribute to an element (most often, an R function)

### 2023-07-12 R/mathml.R
Support MS Word

@@ -180,6 +180,9 @@ mathout <- function(term, flags=NULL, env=parent.frame())
   if(knitr::is_latex_output())
     return(mathjax(term, flags=c(flags, list(cat=TRUE)), env=env))
 
+  if(knitr::pandoc_to("docx"))
+    return(mathjax(term, flags=c(flags, list(cat=TRUE)), env=env))
+
   warning("knitr output not specified. Use mathml() or mathjax() outside of code chunks.")  
   mathjax(term, flags=c(flags, list(cat=TRUE)), env=env)
 }
@@ -493,7 +496,7 @@ fname <- function(fname, body)
 #'
 '%+-%' <- function(x, y)
 {
-  return(c(x - y, x + y))
+  c(x - y, x + y)
 }
 
 
@@ -534,10 +537,10 @@ fname <- function(fname, body)
 #' second argument
 #'
 #' @return 
-#' x=y , e.g., a = b
+#' x == y
 #'
 '%==%' <- function(x, y)
-  x=y
+  x == y
 
 
 #'Congruence, shown as x =~ y
@@ -549,10 +552,10 @@ fname <- function(fname, body)
 #' second argument
 #'
 #' @return 
-#' x=y , e.g., a cong b
+#' x == y , e.g., a cong b
 #'
 '%=~%' <- function(x, y)
-  x=y
+  x == y
 
 #'Proportional, shown as x prop y
 #'

### 2023-07-15 rj/mathml.Rmd
Support MS Word
Revision 2

@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-06-29"
+date: "2023-07-15"
 abstract: >
@@
 a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, it is not 
-possible to translate from LaTeX or HTML back to R [see @latex2r, for an example].
+possible to translate from LaTeX or HTML back to
+R [see @latex2r, for an example].
 
 The package \pkg{mathml} is available for R\ version 4.2 and later, and can be
-easily installed using the usual `install.packages("mathml")`. The source code
-of the package is found at https://github.com/mgondan/mathml.
+easily installed using the usual `install.packages("mathml")`. At its present
+stage, it supports output in HTML, LaTeX, and Microsoft 
+Word [via pandoc, @pandoc]. The source code of the package is found
+at https://github.com/mgondan/mathml.
 

### 2023-11-20 rj/mathml.Rmd
proofs IA

@@
 title: "`mathml`: Translate R Expressions to MathML and LaTeX"
-date: "2023-07-15"
+date: "2023-11-20"
 abstract: >
@@
   the R\ Markdown document rendering language, R\ code and mathematical content
-  already exist side-by-side. The present package allows to use the same
+  already exist side-by-side. The present package enables use of the same
   R\ objects for both data analysis and typesetting in documents or web content.
@@
   description or symbolic representation, which is another step towards
-  reproducible science. User-defined hooks allow to extend the package by
-  mapping specific variables or functions to new MathML and LaTeX entities. A
-  number of examples and a case study illustrate the use of the package in a
-  scientific report.
+  reproducible science. User-defined hooks enable extension of the package by
+  mapping specific variables or functions to new MathML and LaTeX entities. 
+  The use of the package is displayed throughout the paper by examples and its
+  application is illustrated as a case study in a scientific report.
+
 draft: true
@@
 like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
-cases, these commands are tedious to type in and their source code is hard to
+cases, these commands are tedious to type and their source code is hard to
 read.
@@
 interpreter that maps nested expressions recursively to MathML and
-LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
+LaTeX/MathJax, respectively. User-defined hooks enable extension of the set of
 rules, for example, to represent specific R\ elements by custom mathematical
@@
 
-As it can be seen from the output, to the variable `term` is not assigned the 
-result of the calculation, yet an R\ call [see, e.g., @Wickham2019, for details 
-on "non-standard evaluation"], which can eventually be evaluated with `eval()`,
+As can be seen from the output, to the variable `term` is not assigned the 
+result of the calculation, but instead an R\ call [see, e.g., @Wickham2019, 
+for details on "non-standard evaluation"], which can eventually be evaluated 
+with `eval()`,
 
@@
 mathjax(term)
+````
+
+````{r, echo=FALSE}
+math(term)
 ````
@@
 invoked [see the vignette on custom print methods in @knitr].
-
-````{r, echo=FALSE}
-math(term)
-````
 
@@
 
-Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
+Schwarz's [-@schwarz1994] model refers to two stimuli A and B, presented either
 alone or in combination (AB, redundant stimuli), with the redundant stimuli
@@
 employed to reach the threshold for the decision, and a residual \(M\) denoting
-other processes such a motor preparation and execution. Correspondingly, the
+other processes such as motor preparation and execution. Correspondingly, the
 expected response time amounts to
@@
 t <- cbind('\\(\\tau\\)'=ttau, '\\(m\\)'=tm, '\\(s\\)'=ts, '\\(n\\)'=tn)
-knitr::kable(t, caption="Miller data", row.names=FALSE, escape=FALSE)
+knitr::kable(t, caption="Observed RT Means and SDs of [@miller]'s study in 
+             [@schwarz1994, p. 513]", row.names=FALSE, escape=FALSE)
 ```
@@
 
-As the reader may know from own experience, missed parentheses are frequent
+As the reader may know from their own experience, missed parentheses are frequent
 causes of wrong results and errors that are hard to locate in programming code.
@@
 For example, in its current stage, the LaTeX package `breqn` [@breqn] is mostly
-a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, it is not 
-possible to translate from LaTeX or HTML back to
-R [see @latex2r, for an example].
+a proof of concept. Moreover, \pkg{mathml} only works in one direction,
+that is, it is not possible to translate from LaTeX or HTML back to R 
+[see @latex2r, for an example].
 
@@
 distribution [@swipl] in R. The communication between the two systems is mainly
-in the form of queries from R\ to Prolog, but two Prolog functions allow to ring
-back and evaluate terms in R.
+in the form of queries from R\ to Prolog, but two Prolog functions allow ring
+back and evaluation of terms in R.
 
@@
 
-Below is an example how to build a linear equation of the
+Below is an example of how to build a linear equation of the
 form \(Y = b_0 + b_1X_1 + ...\) using the Prolog macros from \pkg{mathml}.
@@
 the practice section why the curly brace is preferred in an exponent), resulting
-in \(x^{1/3}\) which may still not match everyone's taste. Here describe the
+in \(x^{1/3}\) which may still not match everyone's taste. Here we describe the
 steps needed to represent the \(n\)-th root as \(\sqrt[n]x\).

### 2023-11-20 rj/mathml.Rmd
proofs MG

@@
   mapping specific variables or functions to new MathML and LaTeX entities. 
-  The use of the package is displayed throughout the paper by examples and its
-  application is illustrated as a case study in a scientific report.
-
+  Throughout the paper, examples are given for the functions of the package, and
+  a case study illustrates its use in a scientific report.
 draft: true
@@
 output: 
-  rjtools::rjournal_web_article:
+  rjtools::rjournal_article:
     self_contained: yes
@@
 - \usepackage{amsmath}
-Vignette: >
-  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
 ---
@@
 
-````{r, echo=FALSE}
-math(term)
-````
-
 Some of the curly braces are not really needed in the LaTeX output, but are
@@
 that a special knitr printing function is
-invoked [see the vignette on custom print methods in @knitr].
+invoked [see the vignette on custom print methods in @knitr]. An R\ Markdown
+code chunk with `mathout(term)` thus produces:
+
+````{r, echo=FALSE}
+math(term)
+````
 
@@
 hook(ED_async(.Tau, .C, .MA, .SA, .MB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB))))
-math(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
+math(call("=", quote(E(D[tau])), ED_async))
 ```
@@
 
-Schwarz [-@schwarz1994] applied the model to data from a redundant signals task
-with 13 onset
+Schwarz [-@schwarz1994] applied the model to data from a redundant signals 
+task [@miller] with 13 onset
 asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
@@
 t <- cbind('\\(\\tau\\)'=ttau, '\\(m\\)'=tm, '\\(s\\)'=ts, '\\(n\\)'=tn)
-knitr::kable(t, caption="Observed RT Means and SDs of [@miller]'s study in 
-             [@schwarz1994, p. 513]", row.names=FALSE, escape=FALSE)
+knitr::kable(t, caption="Miller (1986) data", row.names=FALSE, escape=FALSE)
 ```
@@
 breaks. This is mostly due to lacking support by MathML and LaTeX renderers. 
-For example, in its current stage, the LaTeX package `breqn` [@breqn] is mostly
+For example, in its current stage, the LaTeX package breqn [@breqn] is mostly
 a proof of concept. Moreover, \pkg{mathml} only works in one direction,

### 2023-11-22 rj/mathml.Rmd
final proof IA

@@
 to stimuli of different sources, and the same information is presented on two or
-more channels. In Schwarz' [-@schwarz1994] model, decision-making builds on a
+more channels. In Schwarz's [-@schwarz1994] model, decision-making builds on a
 process of noisy accumulation of information over time [e.g., @ratcliff2016]. In
@@
 observed mean response times and their standard deviations are given in
-Table \@ref(tab:miller-data) below.
+Table \@ref(tab:miller-data).
 

### 2023-12-21 rj/mathml.Rmd
IA: draft changes

@@
 ---
-title: "`mathml`: Translate R Expressions to MathML and LaTeX"
+title: "mathml: Translate R Expressions to MathML and LaTeX"
 date: "2023-11-20"
@@
 on a symbol (see next section for further decorations). Note that an 
-R\ function `roof()` does not exist in base R, it is provided by the package
+R\ function `roof()` does not exist in base R\, it is provided by the package
 for convenience and points to the identity function.
@@
 boxes etc. Internally, these decorations are implemented as identity functions,
-so that they can be introduced into R expressions without side-effects.
+so that they can be introduced into R\ expressions without side-effects.
 
@@
 The example is a reminder that it is not possible to define the precedence of
-custom operators in R, and that expressions with such operators are evaluated
+custom operators in R\, and that expressions with such operators are evaluated
 strictly from left to right. Again, the problem can be worked around by the
@@
 ```
-
-## Builtin functions
-
-There is support for most functions from package `base`, with adequate use and
-omission of parentheses.
-
-```{r}
-term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
-math(term)
-term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
-math(term)
-```
-
-A few more examples are shown in Table \@ref(tab:base-stats), including
-functions from `stats`.
 
@@
 
+## Builtin functions
+
+There is support for most functions from package `base`, with adequate use and
+omission of parentheses.
+
+```{r}
+term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
+math(term)
+term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
+math(term)
+```
+
+A few more examples are shown in Table \@ref(tab:base-stats), including
+functions from `stats`.
+
 ## Custom functions
 
-For self-written functions, the matter is somewhat more complicated. For a function
-such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
-only the function body is represented. We can still display functions in the
-form `head(x) = body` if we embed the object to be shown into a
-call `"<-"(head, body)`.
+For self-written functions, the matter is somewhat more complicated. For instance, 
+if we consider a function such as `g <- function(...) ...`, the name _g_ is not 
+transparent to R, because only the function body is represented. 
+We can still display functions in the form `head(x) = body` if we embed the object to be shown into a call `"<-"(head, body)`.
 
@@
 
-Note that in the example above, the quoted term is an abbreviation
+Above, the quoted term is an abbreviation
 for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
@@
 in the output---instead of the desired matrix representation. This behavior is
-expected because quotation of R calls also quote the components of the
+expected because quotation of R\ calls also quote the components of the
 call (here, _A_ and _B_).
@@
 
-The R function `omit_right(a + b)` uses non-standard evaluation
+The R\ function `omit_right(a + b)` uses non-standard evaluation
 techniques [e.g., @Wickham2019] to return only the left part an operation,
@@
 
-One would typically use chunk option `echo=FALSE` to suppress the R code.
+One would typically use chunk option `echo=FALSE` to suppress the R\ code.
 
@@
 a proof of concept. Moreover, \pkg{mathml} only works in one direction,
-that is, it is not possible to translate from LaTeX or HTML back to R 
+that is, it is not possible to translate from LaTeX or HTML back to R\ 
 [see @latex2r, for an example].
@@
 
-For convenience, the translation of the R expressions is achieved through a
+For convenience, the translation of the R\ expressions is achieved through a
 Prolog interpreter provided by another R\ package \CRANpkg{rolog} [@rolog]. If a
@@
 slash. Thus, at the Prolog end, the predicate `math/2` translates the
-representation of the R call `pbinom(K, N, Pi)` into a more general 
-representation of an R function `fn/2` with the name `P_Bi`, one 
+representation of the R\ call `pbinom(K, N, Pi)` into a more general 
+representation of an R\ function `fn/2` with the name `P_Bi`, one 
 argument `X =< K`, and the two parameters `N` and `Pi`, as shown below.
@@
 representation. We have already seen above how to use "hooks" to translate long
-variable names from R to compact mathematical signs, as well as functions such
+variable names from R\ to compact mathematical signs, as well as functions such
 as cumulative probabilities \(P(X \le k)\) to different representations
@@
 
+See Section 3.11 for replacing lengthy R\ labels by short mathematical symbols.
+
 ## \(n\)-th root
 
-Base R does not provide a function like `cuberoot(x)` or `nthroot(x, n)`, and
+Base R\ does not provide a function like `cuberoot(x)` or `nthroot(x, n)`, and
 the present package does not support the respective representation. To obtain a
@@
 We assume that `nthroot(x, n)` is available in the current
-namespace [manually defined, or from R package \CRANpkg{pracma}, @pracma], so that
+namespace [manually defined, or from R\ package \CRANpkg{pracma}, @pracma], so that
 the names of the arguments and their order are accessible to `canonical()` if

### 2024-01-30 R/mathml.R
add unhook

@@ -355,6 +355,9 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 }
 
 #' Hook for custom symbols
+#' 
+#' hook(term, display)
+#' unhook(term)
 #'
 #' @param term
 #' an R call or symbol/number. This is the expression to replace.
@@ -366,7 +369,7 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' indicates that _term_ and _display_ should be quoted.
 #'
 #' @param as.rolog (default is TRUE)
-#' indicates that simplified quasi-quotoation is to be used.
+#' indicates that simplified quasi-quotation is to be used.
 #'
 #' @return
 #' TRUE on success
@@ -376,9 +379,10 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' @examples
 #' hook(t0, subscript(t, 0))
 #' mathml(quote(t0))
-#'
+#' unhook(t0)
 #' hook(term=quote(t0), display=quote(subscript(t, 0)), quote=FALSE)
 #' mathml(quote(t0))
+#' unhook(quote(t0), quote=FALSE)
 #'
 hook <- function(term, display, quote=TRUE, as.rolog=TRUE)
 {
@@ -398,6 +402,23 @@ hook <- function(term, display, quote=TRUE, as.rolog=TRUE)
   if(isFALSE(r))
     return(FALSE)
 
+  invisible(r)
+}
+
+#' @rdname hook
+#' @export
+unhook <- function(term, quote=TRUE, as.rolog=TRUE)
+{
+  if(quote)
+    term <- substitute(term)
+
+  if(as.rolog)
+    term <- rolog::as.rolog(term)
+
+  r <- rolog::once(call("retractall", call("math_hook", term, expression(X))))
+  if(isFALSE(r))
+    return(FALSE)
+  
   invisible(r)
 }
 

### 2024-01-30 R/mathml.R
update hook

@@ -379,10 +379,10 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' @examples
 #' hook(t0, subscript(t, 0))
 #' mathml(quote(t0))
-#' unhook(t0)
-#' hook(term=quote(t0), display=quote(subscript(t, 0)), quote=FALSE)
+#' hook(term=quote(t0), display=quote(superscript(t, 0)), quote=FALSE)
 #' mathml(quote(t0))
-#' unhook(quote(t0), quote=FALSE)
+#' unhook(t0)
+#' mathml(quote(t0))
 #'
 hook <- function(term, display, quote=TRUE, as.rolog=TRUE)
 {
@@ -398,7 +398,7 @@ hook <- function(term, display, quote=TRUE, as.rolog=TRUE)
     display <- rolog::as.rolog(display)
   }
 
-  r <- rolog::once(call("assert", call("math_hook", term, display)))
+  r <- rolog::once(call("asserta", call("math_hook", term, display)))
   if(isFALSE(r))
     return(FALSE)
 

### 2024-03-15 inst/pl/bussproofs.pl
Stub for bussproofs

@@ -1,0 +1,28 @@
+% Render in MathML
+%
+%      <mpadded height="+.5em" width="+.5em" voffset="-.15em" semantics="bspr_prooflabel:right">
+%        <mstyle displaystyle="false" scriptlevel="0">
+%          <mrow data-mjx-texclass="ORD">
+%            <mstyle mathsize="0.7em">
+%              <mrow data-mjx-texclass="ORD">
+%                <mi>R</mi>
+%                <mo accent="false" stretchy="false">&#x2192;</mo>
+%              </mrow>
+%            </mstyle>
+%          </mrow>
+%        </mstyle>
+%      </mpadded>
+%
+
+% Render in MathML
+mlx(rcond(A, B), M, Flags) :-
+    ml(frac(B, A), F1, Flags),
+    ml('%->%'('R', ''), R1, Flags),
+    M = mrow([F1, mstyle([mathsize('0.7em')], R1)]).
+
+% Render in MathJax
+% jaxx(rcond(A, B), M, Flags) :-
+%     jax(A, A1, Flags),
+%     jax(B, B1, Flags),
+%     format(string(M), "\\frac{~w}{~w}", [B1, A1]).
+

### 2024-04-29 inst/pl/bussproofs.pl
Add trivial bussproof example

@@ -1,3 +1,15 @@
+:- module(bussproofs,
+  [ op(500, fy, ~),     % negation
+    op(1000, xfy, &),   % conjunction
+    op(1100, xfy, '|'), % disjunction
+    op(1110, xfy, =>),  % conditional
+    op(1120, xfy, <=>)  % biconditional
+  ]).
+
+:- reexport(library(mathml)).
+
+:- multifile mlx/3.
+
 % Render in MathML
 %
 %      <mpadded height="+.5em" width="+.5em" voffset="-.15em" semantics="bspr_prooflabel:right">
@@ -15,145 +27,144 @@
 %
 % math_hook(rcond(A, B), [frac(B, A), '%->%'('R', xx)]).
 
+mathml:math_hook([], '').
+mathml:math_hook(&(A, B), and(A, B)).
+mathml:math_hook('|'(A, B), or(A, B)).
+mathml:math_hook(=>(A, B), '%->%'(A, B)).
+mathml:math_hook(<=>(A, B), '%<->%'(A, B)).
+
 % Render in MathML
-mlx(proof(Denominator, Numerator), M, Flags) :-
+mathml:mlx(proof(Denominator, Numerator), M, Flags) :-
     X1 =.. ['###1', '##'(Numerator)],
     X =.. ['###2', '##'(X1), '##1'(Denominator)],
-    ml(proof_tree(X), M, Flags).
+    mathml:ml(proof_tree(X), M, Flags).
 
-mlx(proof(Denominator, Numerator1, Numerator2), M, Flags) :-
+mathml:mlx(proof(Denominator, Numerator1, Numerator2), M, Flags) :-
     X1 =.. ['###1', '##'(Numerator1, '', Numerator2)],
     X =.. ['###2', '##'(X1), '##1'(Denominator)],
-    ml(proof_tree(X), M, Flags).
+    mathml:ml(proof_tree(X), M, Flags).
 
-mlx(format_operator(Op), R, Flags) :-
-    ml(Op, R1, Flags),
-    ml(size(R1, '0.7em'), R2, Flags),
-    ml(mstyle_right(R2), R3, Flags),
-    ml(mpadded_right(R3), R, Flags).
+mathml:mlx(format_operator(Op), R, Flags) :-
+    mathml:ml(Op, R1, Flags),
+    mathml:ml(size(R1, '0.7em'), R2, Flags),
[…]

### 2025-03-06 prolog/mathml.pl
percent/1

@@ -1768,7 +1768,8 @@ math(round(A, D), M, Flags0, Flags1)
 ml(pos(A), M, Flags)
  => option_(digits(D), Flags, 2),
     format(atom(Mask), '~~~wf', [D]),
-    format(string(X), Mask, [A]),
+    option_(mult(F), Flags, 1),
+    format(string(X), Mask, [F*A]),
     M = mn(X).
 
 jax(posint(A), M, _Flags)

### 2025-03-06 prolog/pval.pl

@@ -37,7 +37,12 @@ mathml:math_hook(pval(A, P), M, _Flags) :-
     M = (P == pval(A)).
 
 mathml:math_hook(tstat(A), M, Flags, Flags1) :-
-    !,
     M = A,
     Flags1 = [digits(2) | Flags].
+
+mathml:math_hook(percent(A), M, Flags, Flags1) :-
+    option(digits(D), Flags, 2),
+    D1 is D - 2,
+    Flags1 = [digits(D1), mult(100) | Flags],
+    M = list("", [A, '%']).
 

### 2025-05-17 R/mathml.R
hooked to display results of math_hook

@@ -362,6 +362,7 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' 
 #' hook(term, display)
 #' unhook(term)
+#' hooked(term)
 #'
 #' @param term
 #' an R call or symbol/number. This is the expression to replace.
@@ -376,12 +377,14 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' indicates that simplified quasi-quotation is to be used.
 #'
 #' @return
-#' TRUE on success
+#' hook and unhook return TRUE on success. hooked returns the hooked
+#' expression or FALSE on failure.
 #'
 #' @md
 #'
 #' @examples
 #' hook(t0, subscript(t, 0))
+#' hooked(t0)
 #' mathml(quote(t0))
 #' hook(term=quote(t0), display=quote(superscript(t, 0)), quote=FALSE)
 #' mathml(quote(t0))
@@ -424,6 +427,17 @@ unhook <- function(term, quote=TRUE, as.rolog=TRUE)
     return(FALSE)
   
   invisible(r)
+}
+
+#' @rdname hook
+#' @export
+hooked <- function(term)
+{
+  r <- rolog::once(call("math_hooked", term, expression(X)))
+  if(isFALSE(r))
+    return(FALSE)
+  
+  return(r$X)
 }
 
 #' Multiplication

### 2025-05-17 R/mathml.R
hooked to display results of math_hook

@@ -362,6 +362,7 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' 
 #' hook(term, display)
 #' unhook(term)
+#' hooked(term)
 #'
 #' @param term
 #' an R call or symbol/number. This is the expression to replace.
@@ -376,12 +377,14 @@ canonical <- function(term=quote(`%in%`(table=Table, x=X)), drop=TRUE)
 #' indicates that simplified quasi-quotation is to be used.
 #'
 #' @return
-#' TRUE on success
+#' hook and unhook return TRUE on success. hooked returns the hooked
+#' expression or FALSE on failure.
 #'
 #' @md
 #'
 #' @examples
 #' hook(t0, subscript(t, 0))
+#' hooked(quote(t0))
 #' mathml(quote(t0))
 #' hook(term=quote(t0), display=quote(superscript(t, 0)), quote=FALSE)
 #' mathml(quote(t0))
@@ -424,6 +427,17 @@ unhook <- function(term, quote=TRUE, as.rolog=TRUE)
     return(FALSE)
   
   invisible(r)
+}
+
+#' @rdname hook
+#' @export
+hooked <- function(term)
+{
+  r <- rolog::once(call("math_hooked", term, expression(X)))
+  if(isFALSE(r))
+    return(FALSE)
+  
+  return(r$X)
 }
 
 #' Multiplication

### 2025-05-17 inst/pl/mathml.pl

@@ -93,6 +93,24 @@ macro(R, M, Flags, Flags) :-
     math(R, M),                  % math/2 ignoring the flags
     dif(R, M).
 
+% Apply hook to entire expression
+math_hooked1(A, A1) :-
+    math_hook(A, A0),
+    !,
+    A1 = A0.
+
+math_hooked1(A, A).
+    
+math_hooked(A, A1) :-
+    compound(A),
+    !,
+    mapargs(math_hooked, A, A0),
+    math_hooked1(A0, A1).
+
+math_hooked(A, A1) :-
+    !,
+    math_hooked1(A, A1).
+
 % Main MathML translation
 %
 % R: R expression
